### 2020-01-17 README.Rmd
Initial commit

---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
# Please put your title here to include it in the file below.
Title <- "Title of your paper goes here"
```

# Robinson2020

[![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh///master?urlpath=rstudio)

This repository contains the data and code for our paper:

> Authors, (YYYY). _`r Title`_. Name of journal/book <https://doi.org/xxx/xxx>

Our pre-print is online here:

> Authors, (YYYY). _`r Title`_. Name of journal/book, Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>

### How to cite

Please cite this compendium as:

> Authors, (`r format(Sys.Date(), "%Y")`). _Compendium of R code and data for `r Title`_. Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>

### How to download or install

You can download the compendium as a zip from from this URL: </archive/master.zip>

Or you can install this compendium as an R package, Robinson2020, from GitHub with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
remotes::install_github("/")
```

### Licenses

**Text and figures :**  [CC-BY-4.0](http://creativecommons.org/licenses/by/4.0/)

**Code :** See the [DESCRIPTION](DESCRIPTION) file

**Data :** [CC-0](http://creativecommons.org/publicdomain/zero/1.0/) attribution requested in reuse

### Contributions

We welcome contributions from everyone. Before you get started, please see our [contributor guidelines](CONTRIBUTING.md). Please note that this project is released with a [Contributor Code of Conduct](CONDUCT.md). By participating in this project you agree to abide by its terms.

### 2020-01-17 Robinson2020.Rmd
Initial commit

---
title: "Robinson et al. 2020 Supplemental Information: Comparing the Radiocarbon and Tree-ring Records for the Upland US Southwest"
author: "Kyle Bocinsky"
date: "1/12/2020"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# devtools::install_github("village-ecodynamics/villager")
# purrr::walk(c("magrittr",
#               "tidyverse",
#               "googledrive",
#               "sf",
#               "lwgeom",
#               "units",
#               "rcarbon"), 
#             devtools::install_cran)

packages <- c("magrittr",
              "tidyverse",
              "googledrive",
              "sf",
              "lwgeom",
              "units",
              "rcarbon",
              "villager"
)

purrr::walk(packages,
            library, 
            character.only = TRUE)

source("R/get_df.R")

```

## Summary of methods
Here, we directly compare two large collections of radiocarbon and tree-ring dates drawn from the Upland US Southwest --- roughly defined as the 'Four Corners' states of Arizona, Colorado, New Mexico, and Utah, but excluding the Phoenix and Tucson basins.

After harmonizing those datasets to with one another and calibrating Radiocarbon dates are drawn from the [DARCY: DATABASE AND REFERENCE HERE], and are available [HERE]. Tree-ring data are those presented in @Bocinsky2016_SA and archived on the Digital Archaeological Record (https://doi.org/10.6067/XCV82J6D7B).

## Study Area

```{r}
# Define the Bocinsky et al. 2016 study area
UUSW <- sf::st_bbox(c(xmin = -113,
                      xmax = -105,
                      ymin = 32,
                      ymax = 38), 
                    crs = 4326) %>%
  sf::st_as_sfc()

# VEP study areas
vep_study_areas <- 
  list(CMV = villager::get_vep_boundary(version = "vepii_cmv"),
       NRG = villager::get_vep_boundary(version = "vepii_nrg")) %>%
  purrr::map(sf::st_transform,
             crs = 4326) %>%
  purrr::map(tibble::as_tibble) %>%
  dplyr::bind_rows(.id = "Study Area") %>%
  sf::st_as_sf(crs = 4326)

```

## Datasets
These data contain sensitive location information about cultural sites, and are restricted on tDAR. Accordingly, we pre-processed the data using the script below, and provide data as supplemental information without location information represented.
```{r}
# Read in the Bocinsky et al. 2016 tree ring data
# https://doi.org/10.6067/XCV82J6D7B
dendro_data <- 
  readr::read_csv("./data-raw/TreeRing/Bocinsky_et_al_2016_dendrochronology.csv") %>%
  # Convert to spatial object
  sf::st_as_sf(coords = c("LON","LAT"),
               crs = 4326) %>%
  # Crop to SWUS bounding box
  sf::st_intersection(UUSW) %>%
  # Convert dates to BP
  dplyr::mutate(Date_BP = 1950 - Outer_Date_AD,
                Type = ifelse(C_Level %in% c(2,3),
                              "Cutting or Near-cutting",
                              "Non-cutting")) %>%
  # drop C_Level
  dplyr::select(Seq_Number,
                Site_Number,
                Site_Name,
                Lab_Number,
                Type,
                Date_BP) %>%
  dplyr::rename(SiteID = Seq_Number) %T>%
  # Write the output to a new file (geojson)
  sf::write_sf("./data-derived/dendro_data.geojson",
               delete_dsn = TRUE)

# Check R/T
# sf::read_sf("./data-derived/dendro_data.geojson")

```

We draw on the tree-ring date database presented in @Bocinsky2016_SA and archived on the Digital Archaeological Record (https://doi.org/10.6067/XCV82J6D7B) --- `r nrow(readr::read_csv("./data-raw/TreeRing/Bocinsky_et_al_2016_dendrochronology.csv"))` dates from the Upland US Southwest. Here, we use only the subset of those data

## Preparing the radiocarbon data

```{r}
# Define the Bocinsky et al. 2016 study area
SWUS_bbox <- sf::st_bbox(c(xmin = -113,
                           xmax = -105,
                           ymin = 32,
                           ymax = 38), 
                         crs = 4326) %>%
  sf::st_as_sfc()

# Define the Phoenix Basin
# Defined as SW AZ counties, using the National Map small scale data
# "https://prd-tnm.s3.amazonaws.com/StagedProducts/Small-scale/data/Boundaries/countyp010g.gdb_nt00933.tar.gz" %>%
#   download.file(destfile="../data-raw/countyp010g.gdb_nt00933.tar.gz", 
#                 mode='wb')
# untar("../data-raw/countyp010g.gdb_nt00933.tar.gz",
#       exdir="../data-raw/countyp010g")
phx_basin <- "./data-raw/countyp010g/countyp010g.gdb/" %>%
  sf::st_read() %>%
  dplyr::filter(NAME %in% 
                  c("Maricopa",
                    "Pinal",
                    "Pima")) %>%
  sf::st_transform(4326) %>%
  sf::st_union()

# Read in the Bocinsky et al. 2016 tree ring data
radiocarbon_data <- #readr::read_csv("./data-raw/Radiocarbon/14C_Raw.csv") %>%
  readr::read_csv("./data-raw/Radiocarbon/0403_sbox_clean.csv") %>%
  # Convert to spatial object
  sf::st_as_sf(coords = c("Long","Lat"),
               crs = 4326) %>%
  # Crop to SWUS bounding box
  sf::st_intersection(SWUS_bbox) %>%
  # Remove Phoenix Basin dates
  sf::st_difference(phx_basin) %>%
  # Select salient columns
  dplyr::select(labnumber,
                date,
                sd,
                #diff,
                #d13,
                #mat,
                #Source,
                Site,
                SiteID) %>%
  dplyr::rename(Lab_Number = labnumber,
                Date_BP = date,
                SD = sd,
                # Diff = diff,
                # Material = mat,
                Site_Number = Site) %>%
  sf::st_intersection("./data-raw/countyp010g/countyp010g.gdb/" %>%
                        sf::st_read() %>%
                        sf::st_transform(4326) %>%
                        dplyr::mutate(County = paste(NAME, STATE, sep = ", ")) %>%
                        dplyr::select(County)) %T>%
  # Write the output to a new file (geojson)
  sf::write_sf("./data-derived/radiocarbon_data.geojson",
               delete_dsn = TRUE)

# Check R/T
# sf::read_sf("./data-derived/radiocarbon_data.geojson")

```

```{r}
radiocarbon_data <-
  sf::read_sf("./data-derived/radiocarbon_data.geojson") %>%
  dplyr::mutate(Type = "Radiocarbon",
                SiteID = SiteID %>% stringr::str_c("Radiocarbon - ", .)) %>%
  dplyr::mutate(`Study Area` = ifelse(County == "Montezuma, CO",
                                      "CMV", 
                                      NA),
                `Study Area` = ifelse(County %in% c("Rio Arriba, NM",
                                                    "Sandoval, NM",
                                                    "Santa Fe, NM",
                                                    "Taos, NM"),
                                      "NRG", 
                                      `Study Area`)) %>%
  dplyr::select(-County)

dendro_data <- 
  sf::read_sf("./data-derived/dendro_data.geojson") %>%
  dplyr::mutate(Type = Type %>% stringr::str_c("Dendro - ", .),
                SiteID = SiteID %>% stringr::str_c("Dendro - ", .),
                SD = 0) %>%
  dplyr::filter(Type == "Dendro - Cutting or Near-cutting") %>%
  dplyr::mutate(Type = "Tree-ring") %>%
  sf::st_join(vep_study_areas) %>%
  dplyr::select(-Site_Name) %>%
  # dplyr::rename(`Study Area` = `Study.Area`) %>%
  dplyr::select(names(radiocarbon_data))

rbind(radiocarbon_data,
      dendro_data) %>%
  sf::st_drop_geometry() %>%
  group_by(Type,`Study Area`) %>%
  dplyr::summarise(Count = n()) %>%
  tidyr::spread(Type, Count)

```

## Preparing the radiocarbon SPDs

```{r}
radiocarbon_spd <-
  radiocarbon_data %>%
  split(.,.$`Study Area`) %>%
  tibble::tibble(`Study Area` = names(.),
                 Dates = .) %>%
  dplyr::bind_rows(tibble::tibble(`Study Area` = "UUSW",
                                  Dates = list(radiocarbon_data))) %>%
  dplyr::rowwise() %>%
  dplyr::mutate(Calibrated = list(Dates %$%
                                    rcarbon::calibrate(x = Date_BP,
                                                       errors = SD,
                                                       calMatrix = TRUE,
                                                       normalised = TRUE)),
                Bins = list(rcarbon::binPrep(sites = Dates$SiteID,
                                             ages = Calibrated,
                                             h = 100)),
                SPD = list(spd(x = Calibrated,
                               bins = Bins,
                               timeRange = c(1949,0),
                               runm = 21))) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(Density = SPD %>%
                  purrr::map(function(x){
                    x$grid %>%
                      tibble::as_tibble() %$%
                      tibble::tibble(Type = "Radiocarbon Density",
                                     Date_BP = calBP,
                                     Density = PrDens / sum(PrDens, na.rm = T))
                  })
  ) %T>%
  readr::write_rds("./data-derived/radiocarbon_spd.rds",
                   compress = "gz")

radiocarbon_spd <- readr::read_rds("./data-derived/radiocarbon_spd.rds")

```

To prepare the radiocarbon dataset, we used functions provided by the *rcarbon* package for R [@rcarbon]. We calibrated the radiocarbon dates using the methods of @Bronk2008 and implemented in @rcarbon. To control for inconsistent sample size across the archaeological sites represented in the database, we aggregated radiocarbon dates known to be from the same phase of same archaeological site using the `rcarbon::binprep` function with an `h`-value of 100. We averaged probability densities of binned dates, and then constructed summed probability density (SPD) distributions across bins.

## 'Calibrating' tree-ring dates, and preparing tree-ring SPDs

```{r}
calibrate_tree_ring_dates <- function(x){
  out <- list(
    metadata = tibble::tibble(
      DateID = 1:length(x),
      CRA = x,
      Error = 0,
      Details = NA,
      CalCurve = "intcal13",
      ResOffsets = 0,
      ResErrors = 0,
      StartBP = 50000,
      EndBP = 0,
      Normalised = TRUE,
      F14C = FALSE,
      CalEPS = 1e-05
    ),
    grids = NA,
    calmatrix = 
      tibble::tibble(index = 1:length(x),
                     year_bp = x,
                     value = TRUE) %>%
      tidyr::pivot_wider(names_from = index,
                         values_from = value) %>%
      dplyr::bind_rows(tibble::tibble(year_bp = 50000:0)) %>%
      dplyr::filter(!duplicated(year_bp)) %>%
      dplyr::arrange(-year_bp) %>%
      dplyr::filter(year_bp >= 0) %>%
      tibble::column_to_rownames("year_bp") %>%
      as.matrix() %>%
      tidyr::replace_na(0) %>%
      magrittr::set_colnames(NULL)
  )

  attr(out,"class") <- c("CalDates", "list")

  return(out)
}

dendro_spd <-
  dendro_data %>%
  split(.,.$`Study Area`) %>%
  tibble::tibble(`Study Area` = names(.),
                 Dates = .) %>%
  dplyr::bind_rows(tibble::tibble(`Study Area` = "UUSW",
                                  Dates = list(dendro_data))) %>%
  dplyr::rowwise() %>%
  dplyr::mutate(Calibrated = list(Dates %$%
                                    calibrate_tree_ring_dates(x = Date_BP)),
                Bins = list(rcarbon::binPrep(sites = Dates$SiteID,
                                             ages = Dates$Date_BP,
                                             h = 100)),
                SPD = list(spd(x = Calibrated,
                               bins = Bins,
                               timeRange = c(1949,0),
                               runm = 21))) %>%

  dplyr::ungroup() %>%
  dplyr::mutate(Density = SPD %>%
                  purrr::map(function(x){
                    x$grid %>%
                      tibble::as_tibble() %$%
                      tibble::tibble(Type = "Tree-ring Density",
                                     Date_BP = calBP,
                                     Density = PrDens / sum(PrDens, na.rm = T))
                  })
  ) %T>%
  readr::write_rds("./data-derived/dendro_spd.rds",
                   compress = "gz")

dendro_spd <- readr::read_rds("./data-derived/dendro_spd.rds")

```

## Comparing radiocarbon and tree-ring SPDs

```{r}
all_density <- dendro_spd %>%
  dplyr::select(`Study Area`,Density) %>%
  tidyr::unnest(cols = c(Density)) %>%
  bind_rows(radiocarbon_spd %>%
              dplyr::select(`Study Area`,Density) %>%
              tidyr::unnest(cols = c(Density))) %>%
  dplyr::mutate(`Year AD` = 1950 - Date_BP) %>%
  dplyr::select(`Year AD`,
                Date_BP,
                `Study Area`,
                Type, 
                Density) %>%
  dplyr::rename(Value = Density)
```

```{r}
vep_demography <- 
  readxl::read_excel("./data-raw/vepii_demography_reconstructions.xlsx") %>%
  dplyr::rowwise() %>%
  dplyr::mutate(`Year AD` = list((Start+1):End)) %>%
  dplyr::select(`Study Area`,`Year AD`,Population) %>%
  tidyr::unnest(cols = c(`Year AD`)) %>%
  dplyr::mutate(Type = "Population",
                Date_BP = 1950 - `Year AD`) %>%
  dplyr::rename(Value = Population)
```

```{r}
all_density %>%
  dplyr::bind_rows(vep_demography) %>%
  dplyr::mutate(Type = factor(Type, 
                              levels = c("Radiocarbon Density",
                                         "Tree-ring Density",
                                         "Population"),
                              ordered = TRUE)#,
                # `Study Area` = ifelse(`Study Area` == "CMV", "Central Mesa Verde", "Northern Rio Grande")
  ) %>%
  ggplot(aes(x = `Year AD`)) +
  geom_line(aes(y = Value,
                color = `Study Area`),
            size = 1.25) +
  # coord_cartesian(xlim = c(600,1600)) +
  xlab("Year AD") +
  theme_minimal(24) +
  theme(legend.title = element_blank(),
        legend.justification = c(0, 1), 
        legend.position = "bottom",
        axis.title.y = element_blank(),
        axis.text.y = element_blank(),
        panel.grid.major.y = element_blank(),
        panel.grid.minor.y = element_blank(),
        plot.margin=unit(c(0.25,0.5,0.25,0.25),"in")
  ) +
  ggplot2::facet_wrap("Type", 
                      nrow = 3,
                      scales = "free_y")

ggsave("./output/c14_tr_pop.pdf",
       height = 10,
       width = 10)

```

```{r}
four_corners_lcc_proj <- "+proj=lcc +lat_1=37 +lon_0=-109.045225"

# Define the Bocinsky et al. 2016 study area
SWUS_bbox <- sf::st_bbox(c(xmin = -113,
                           xmax = -105,
                           ymin = 32,
                           ymax = 38), 
                         crs = 4326) %>%
  sf::st_as_sfc() %>%
  sf::st_set_crs(NA) %>%
  sf::st_segmentize(dfMaxLength = 0.01) %>%
  sf::st_set_crs(4326) %>%
  sf::st_transform(four_corners_lcc_proj)

four_corners_states <- "./data-raw/statesp010g/statesp010g.gdb/" %>%
  sf::st_read() %>%
  dplyr::filter(NAME %in% 
                  c("Arizona",
                    "Colorado",
                    "New Mexico",
                    "Utah")) %>%
  sf::st_transform(four_corners_lcc_proj)

swus_hillshade_500m <- readr::read_rds("./data-derived/swus_hillshade_500m.rds") %>%
  raster::crop(four_corners_states) %>%
  raster::mask(four_corners_states)

dendro_data <- sf::read_sf("./data-derived/dendro_data.geojson") %>%
  sf::st_transform(four_corners_lcc_proj) %>%
  dplyr::filter(Date_BP >= 550,
                Date_BP <= 1450) %>%
  dplyr::group_by(Site_Number) %>%
  dplyr::summarise(Date_BP = mean(Date_BP, na.rm = T),
                   geometry = mean(geometry)) %>%
  dplyr::mutate(`Date Type` = "Dendro") %>%
  dplyr::select(-Site_Number) %>%
  sf::st_centroid()

radiocarbon_data <- sf::read_sf("./data-derived/radiocarbon_data.geojson") %>%
  sf::st_transform(four_corners_lcc_proj) %>%
  dplyr::filter(Date_BP >= 550,
                Date_BP <= 1450) %>%
  dplyr::group_by(Site_Number) %>%
  dplyr::summarise(Date_BP = mean(Date_BP, na.rm = T)) %>%
  sf::st_jitter(20000) %>%
  dplyr::mutate(`Date Type` = "C14") %>%
  dplyr::select(-Site_Number) %>%
  sf::st_centroid()

dates <- rbind(dendro_data,
               radiocarbon_data)

theme_map <- function(base_size = 6.5, 
                      base_family = ""){
  ggplot2::theme_bw(base_size = base_size, 
                    base_family = base_family) %+replace%
    ggplot2::theme(axis.line = ggplot2::element_blank(),
                   axis.text = ggplot2::element_blank(),
                   axis.ticks = ggplot2::element_blank(),
                   axis.title = ggplot2::element_blank(),
                   panel.background = ggplot2::element_blank(),
                   panel.border = ggplot2::element_blank(),
                   panel.grid = ggplot2::element_blank(),
                   panel.grid.major = ggplot2::element_line(colour = 'transparent'),
                   panel.grid.minor = ggplot2::element_line(colour = 'transparent'),
                   legend.key = element_blank(),
                   legend.position = "top",
                   # legend.justification = c(0,0),
                   # legend.position = "top",
                   # legend.position = c(0.5,1),
                   # legend.background = ggplot2::element_blank(),
                   # legend.key.width = unit(0.15,"in"),
                   # legend.key.height = unit(0.15,"in"),
                   # legend.text = element_text(size = ggplot2::rel(1)),
                   plot.background = ggplot2::element_blank(),
                   plot.margin = ggplot2::margin(t = 0, r = 0, b = 0, l = 0, unit = "npc")
    )
}

ggplot() + 
  geom_sf(data = four_corners_states) +
  ggplot2::geom_raster(data = swus_hillshade_500m %>%
                         get_df(),
                       mapping = aes(x = x,
                                     y = y,
                                     alpha = ID),
                       na.rm = TRUE) +
  scale_alpha(range = c(0.8, 0),
              na.value = 0,
              limits = c(0,255),
              guide = "none") +
  geom_sf(data = four_corners_states,
          fill = NA) +
  geom_sf(data = SWUS_bbox,
          fill = NA) +
  geom_sf(data = dates,
          mapping = aes(shape = `Date Type`,
                        color = Date_BP),
          show.legend = "point") +
  scale_shape(
    name = 'Date Type',
    solid = FALSE,
    guide = guide_legend(position = "top",
                         direction = "horizontal",
                         title.position = "top",
                         label.position = "bottom")
  ) +
  viridis::scale_color_viridis(
    name = 'Year BP',
    limits = c(450,1450),
    breaks = c(450,1450),
    direction = -1,
    guide = guide_colorbar(position = "top",
                           direction = "horizontal",
                           title.position = "top",
                           label.position = "bottom",
                           reverse = TRUE)) +
  theme_map(16)

ggsave('./swus_sample_map.pdf',
       height = 8,
       width = 6)
```

### 2020-01-17 R/Recursive14Ccode.R

@@ -1,0 +1,175 @@
+setwd("~/R/19_entropyregimechange/")
+
+library(plyr)
+library(searchable)
+library(reshape)
+library(sf)
+library(robustbase)
+library(rcarbon)
+#library(tidyverse)
+#remove.packages(tidyverse)
+
+
+
+#### 2.  select relevant datess -------
+Rawdata <- read.csv("14C_Raw.csv")
+Rawdata$state <- Rawdata$state
+
+# Define the Bocinsky et al. 2016 study area
+SWUS_bbox <- sf::st_bbox(c(xmin = -113,
+                           xmax = -105,
+                           ymin = 32,
+                           ymax = 38), 
+                         crs = 4326) %>%
+  sf::st_as_sfc() 
+
+Rawdata <-subset(Rawdata, Rawdata$Country == c("USA")) %>%
+  sf::st_as_sf(coords = c("Long","Lat"),
+               crs = 4326)%>%
+  #Remove SWUS bounding box
+  sf::st_difference(SWUS_bbox)
+
+Rawdata$state <- gsub("UT", "Northern Southwest", Rawdata$state, fixed= TRUE)
+Rawdata$state <- gsub("CO", "Northern Southwest", Rawdata$state, fixed= TRUE)
+Rawdata$state <- gsub("AZ", "Southern Southwest", Rawdata$state, fixed= TRUE)
+Rawdata$state <- gsub("NM", "Southern Southwest", Rawdata$state, fixed= TRUE)
+
+Rawdata$state <- as.factor(Rawdata$state)
+
+write.csv(Rawdata, "Rawdata2.csv")
+
+
+
+
+##### make directory -----
+Directory <- count(RawData, vars = "state")
+names(Directory) <- c("state", "n")
+Directory<-Directory[!(Directory$n<200),]
+write.csv(Directory, "Directory_state.csv", row.names = FALSE)
+###
+
+#### 3. Calibrate each sampling unit recursively --------
+
+#Sometimes r treats values within a dataframe in a way you cannot use. The lines ensure our calibration will work.
+Rawdata$date <- as.numeric(Rawdata$date)
+Rawdata$sd <- as.numeric(Rawdata$sd)
+Rawdata$labnumber <- as.character(Rawdata$labnumber)
+Rawdata$state <- as.character(Rawdata$state)
+
+##Turn each sampling unit into a data.frame and make a list of them. 
+stateList <- list()
+for(i in 1:length(unique(Directory$state))){
[…]

### 2020-01-17 R/get_df.R

@@ -1,0 +1,21 @@
+get_df <- function(x){
+  out <- cbind(raster::xyFromCell(x, seq_len(raster::ncell(x))),
+               tibble::tibble(ID = raster::getValues(x))) %>%
+    tibble::as_tibble()
+  
+  if(is.factor(x)){
+    
+    levels <- levels(x)[[1]] %>%
+      dplyr::mutate_all(.funs = list(ordered)) %>%
+      tibble::as_tibble()
+    
+    fact <- out$ID %>%
+      ordered(levels = levels(levels$ID))
+    
+    out %<>%
+      dplyr::mutate(ID = fact) %>%
+      dplyr::left_join(levels)
+  }
+  
+  return(out)
+}

### 2020-01-17 R/mdt_theme_map.R

@@ -1,0 +1,26 @@
+mdt_theme_map <- function(base_size = 6.5, 
+                          base_family = ""){
+  ggplot2::theme_bw(base_size = base_size, 
+                    base_family = base_family) %+replace%
+    ggplot2::theme(axis.line = ggplot2::element_blank(),
+                   axis.text = ggplot2::element_blank(),
+                   axis.ticks = ggplot2::element_blank(),
+                   axis.title = ggplot2::element_blank(),
+                   panel.background = ggplot2::element_blank(),
+                   panel.border = ggplot2::element_blank(),
+                   panel.grid = ggplot2::element_blank(),
+                   panel.grid.major = ggplot2::element_line(colour = 'transparent'),
+                   panel.grid.minor = ggplot2::element_line(colour = 'transparent'),
+                   legend.key = element_blank(),
+                   legend.position = "top",
+                   # legend.justification = c(0,0),
+                   # legend.position = "top",
+                   # legend.position = c(0.5,1),
+                   # legend.background = ggplot2::element_blank(),
+                   # legend.key.width = unit(0.15,"in"),
+                   # legend.key.height = unit(0.15,"in"),
+                   # legend.text = element_text(size = ggplot2::rel(1)),
+                   plot.background = ggplot2::element_blank(),
+                   plot.margin = ggplot2::margin(t = 0, r = 0, b = 0, l = 0, unit = "npc")
+    )
+}

### 2020-01-17 R/plot_dates_map.R

@@ -1,0 +1,121 @@
+packages <- c("magrittr",
+              "tidyverse",
+              "googledrive",
+              "sf",
+              "lwgeom",
+              "units"
+) 
+# purrr::walk(packages, devtools::install_cran) %>%
+purrr::walk(packages,
+            library, 
+            character.only = TRUE)
+
+source("./R/get_df.R")
+
+# A nice projection for the Four Corners states
+four_corners_lcc_proj <- "+proj=lcc +lat_1=37 +lon_0=-109.045225"
+
+# Define the Bocinsky et al. 2016 study area
+SWUS_bbox <- sf::st_bbox(c(xmin = -113,
+                           xmax = -105,
+                           ymin = 32,
+                           ymax = 38), 
+                         crs = 4326) %>%
+  sf::st_as_sfc() %>%
+  sf::st_set_crs(NA) %>%
+  sf::st_segmentize(dfMaxLength = 0.01) %>%
+  sf::st_set_crs(4326) %>%
+  sf::st_transform(four_corners_lcc_proj)
+
+# Define the Four Corners states
+# "https://prd-tnm.s3.amazonaws.com/StagedProducts/Small-scale/data/Boundaries/statesp010g.gdb_nt00937.tar.gz" %>%
+#   download.file(destfile="../Data/data-raw/statesp010g.gdb_nt00937.tar.gz",
+#                 mode='wb')
+# untar("../Data/data-raw/statesp010g.gdb_nt00937.tar.gz",
+#       exdir="../Data/data-raw/statesp010g")
+four_corners_states <- "./Data/data-raw/statesp010g/statesp010g.gdb/" %>%
+  sf::st_read() %>%
+  dplyr::filter(NAME %in% 
+                  c("Arizona",
+                    "Colorado",
+                    "New Mexico",
+                    "Utah")) %>%
+  sf::st_transform(four_corners_lcc_proj)
+
+swus_hillshade_500m <- readr::read_rds("./Data/data-derived/swus_hillshade_500m.rds") %>%
+  raster::crop(four_corners_states) %>%
+  raster::mask(four_corners_states)
+
+dendro_data <- sf::read_sf("./Data/data-derived/dendro_data.geojson") %>%
+  sf::st_transform(four_corners_lcc_proj) %>%
+  dplyr::filter(Date_BP >= 550,
+                Date_BP <= 1450) %>%
+  dplyr::group_by(Site_Number) %>%
[…]

### 2020-01-17 R/prepare_dendro_dates.R

@@ -1,0 +1,47 @@
+packages <- c("magrittr",
+              "tidyverse",
+              "googledrive",
+              "sf",
+              "lwgeom"
+) 
+# purrr::walk(packages, devtools::install_cran) %>%
+purrr::walk(packages,
+            library, 
+            character.only = TRUE)
+
+# Define the Bocinsky et al. 2016 study area
+SWUS_bbox <- sf::st_bbox(c(xmin = -113,
+                           xmax = -105,
+                           ymin = 32,
+                           ymax = 38), 
+                         crs = 4326) %>%
+  sf::st_as_sfc() 
+  
+
+# Read in the Bocinsky et al. 2016 tree ring data
+dendro_data <- 
+  readr::read_csv("./data-raw/TreeRing/Bocinsky_et_al_2016_dendrochronology.csv") %>%
+  # Convert to spatial object
+  sf::st_as_sf(coords = c("LON","LAT"),
+            crs = 4326) %>%
+  # Crop to SWUS bounding box
+  sf::st_intersection(SWUS_bbox) %>%
+  # Convert dates to BP
+  dplyr::mutate(Date_BP = 1950 - Outer_Date_AD,
+                Type = ifelse(C_Level %in% c(2,3),
+                              "Cutting or Near-cutting",
+                              "Non-cutting")) %>%
+  # drop C_Level
+  dplyr::select(Seq_Number,
+                Site_Number,
+                Site_Name,
+                Lab_Number,
+                Type,
+                Date_BP) %>%
+  dplyr::rename(SiteID = Seq_Number) %T>%
+  # Write the output to a new file (geojson)
+  sf::write_sf("./data-derived/dendro_data.geojson",
+               delete_dsn = TRUE)
+  
+# Check R/T
+# sf::read_sf("./data-derived/dendro_data.geojson")

### 2020-01-17 R/prepare_radiocarbon_dates.R

@@ -1,0 +1,72 @@
+packages <- c("magrittr",
+              "tidyverse",
+              "googledrive",
+              "sf",
+              "lwgeom"
+) 
+# purrr::walk(packages, devtools::install_cran) %>%
+purrr::walk(packages,
+            library, 
+            character.only = TRUE)
+
+# Define the Bocinsky et al. 2016 study area
+SWUS_bbox <- sf::st_bbox(c(xmin = -113,
+                           xmax = -105,
+                           ymin = 32,
+                           ymax = 38), 
+                         crs = 4326) %>%
+  sf::st_as_sfc() 
+
+# Define the Phoenix Basin
+# Defined as SW AZ counties, using the National Map small scale data
+# "https://prd-tnm.s3.amazonaws.com/StagedProducts/Small-scale/data/Boundaries/countyp010g.gdb_nt00933.tar.gz" %>%
+#   download.file(destfile="../data-raw/countyp010g.gdb_nt00933.tar.gz", 
+#                 mode='wb')
+# untar("../data-raw/countyp010g.gdb_nt00933.tar.gz",
+#       exdir="../data-raw/countyp010g")
+phx_basin <- "./data-raw/countyp010g/countyp010g.gdb/" %>%
+  sf::st_read() %>%
+  dplyr::filter(NAME %in% 
+                  c("Maricopa",
+                    "Pinal",
+                    "Pima")) %>%
+  sf::st_transform(4326) %>%
+  sf::st_union()
+
+# Read in the Bocinsky et al. 2016 tree ring data
+radiocarbon_data <- #readr::read_csv("./data-raw/Radiocarbon/14C_Raw.csv") %>%
+  readr::read_csv("./data-raw/Radiocarbon/0403_sbox_clean.csv") %>%
+  # Convert to spatial object
+  sf::st_as_sf(coords = c("Long","Lat"),
+               crs = 4326) %>%
+  # Crop to SWUS bounding box
+  sf::st_intersection(SWUS_bbox) %>%
+  # Remove Phoenix Basin dates
+  sf::st_difference(phx_basin) %>%
+  # Select salient columns
+  dplyr::select(labnumber,
+                date,
+                sd,
+                #diff,
+                #d13,
+                #mat,
+                #Source,
+                Site,
+                SiteID) %>%
+  dplyr::rename(Lab_Number = labnumber,
+                Date_BP = date,
+                SD = sd,
+                # Diff = diff,
+                # Material = mat,
[…]

### 2020-01-17 R/swus_hillshade.R

@@ -1,0 +1,93 @@
+packages <- c("magrittr",
+              "tidyverse",
+              "googledrive",
+              "sf",
+              "lwgeom",
+              "units",
+              "FedData",
+              "rgdal",
+              "raster"
+) 
+# purrr::walk(packages, devtools::install_cran) %>%
+purrr::walk(packages,
+            library, 
+            character.only = TRUE)
+
+# A nice projection for the Four Corners states
+four_corners_lcc_proj <- "+proj=lcc +lat_1=37 +lon_0=-109.045225"
+
+# Define the Four Corners states
+# "https://prd-tnm.s3.amazonaws.com/StagedProducts/Small-scale/data/Boundaries/statesp010g.gdb_nt00937.tar.gz" %>%
+#   download.file(destfile="../Data/data-raw/statesp010g.gdb_nt00937.tar.gz",
+#                 mode='wb')
+# untar("../Data/data-raw/statesp010g.gdb_nt00937.tar.gz",
+#       exdir="../Data/data-raw/statesp010g")
+four_corners_states <- "./data-raw/statesp010g/statesp010g.gdb/" %>%
+  sf::st_read() %>%
+  dplyr::filter(NAME %in% 
+                  c("Arizona",
+                    "Colorado",
+                    "New Mexico",
+                    "Utah")) %>%
+  sf::st_transform(four_corners_lcc_proj) %>%
+  sf::st_union()
+
+swus_rast_500 <- raster::raster(raster::extent(-566500,643000,3815000,5052500),
+                                nrows = 2475,
+                                ncols = 2419,
+                                crs = CRS(four_corners_lcc_proj))
+
+swus_ned <- FedData::get_ned(template = four_corners_states %>%
+                             sf::st_buffer(20000) %>%
+                             as("Spatial"),
+                           label = "swus_4C",
+                           raw.dir = "./data-raw/ned",
+                           extraction.dir = "./data-raw/")
+
+system("gdaldem hillshade ./data-raw/swus_4C_NED_1.tif ./data-derived/swus_4C_hillshade.tif -z 2 -s 111120 -multidirectional -co 'COMPRESS=DEFLATE' -co 'ZLEVEL=9'")
+
+## Generate a hillshade for statewide mapping
[…]

### 2020-01-17 Robinson2020.R

@@ -1,0 +1,258 @@
+packages <- c("magrittr",
+              "tidyverse",
+              "googledrive",
+              "sf",
+              "lwgeom",
+              "units",
+              "rcarbon",
+              "villager"
+)
+# devtools::install_github("benmarwick/rrtools")
+# devtools::install_github("village-ecodynamics/villager")
+
+# purrr::walk(packages, devtools::install_cran)
+purrr::walk(packages,
+            library, 
+            character.only = TRUE)
+
+# VEP study areas
+vep_study_areas <- 
+  list(CMV = villager::get_vep_boundary(version = "vepii_cmv"),
+       NRG = villager::get_vep_boundary(version = "vepii_nrg")) %>%
+  purrr::map(sf::st_transform,
+             crs = 4326) %>%
+  purrr::map(tibble::as_tibble) %>%
+  dplyr::bind_rows(.id = "Study Area") %>%
+  sf::st_as_sf(crs = 4326)
+
+#SPDs for SW region overlapping Bocinsky et al. 2016 dendro cutting date database
+radiocarbon_data <-
+  sf::read_sf("./data-derived/radiocarbon_data.geojson") %>%
+  dplyr::mutate(Type = "Radiocarbon",
+                SiteID = SiteID %>% stringr::str_c("Radiocarbon - ", .)) %>%
+  dplyr::mutate(`Study Area` = ifelse(County == "Montezuma, CO",
+                                      "CMV", 
+                                      NA),
+                `Study Area` = ifelse(County %in% c("Rio Arriba, NM",
+                                                    "Sandoval, NM",
+                                                    "Santa Fe, NM",
+                                                    "Taos, NM"),
+                                      "NRG", 
+                                      `Study Area`)) %>%
+  dplyr::select(-County)
+
+# We can do many of the same things for the tr dates
+dendro_data <- 
+  sf::read_sf("./data-derived/dendro_data.geojson") %>%
+  dplyr::mutate(Type = Type %>% stringr::str_c("Dendro - ", .),
+                SiteID = SiteID %>% stringr::str_c("Dendro - ", .),
+                SD = 0) %>%
+  dplyr::filter(Type == "Dendro - Cutting or Near-cutting") %>%
+  dplyr::mutate(Type = "Tree-ring") %>%
+  sf::st_join(vep_study_areas) %>%
+  dplyr::select(-Site_Name) %>%
+  # dplyr::rename(`Study Area` = `Study.Area`) %>%
[…]

### 2020-01-18 analysis/paper/Robinson2020.Rmd
initial commit

--- 2020-01-17 Robinson2020.Rmd
@@
 knitr::opts_chunk$set(echo = TRUE)
-
-# devtools::install_github("village-ecodynamics/villager")
-# purrr::walk(c("magrittr",
-#               "tidyverse",
-#               "googledrive",
-#               "sf",
-#               "lwgeom",
-#               "units",
-#               "rcarbon"), 
-#             devtools::install_cran)
-
-packages <- c("magrittr",
-              "tidyverse",
-              "googledrive",
-              "sf",
-              "lwgeom",
-              "units",
-              "rcarbon",
-              "villager"
-)
-
-purrr::walk(packages,
-            library, 
-            character.only = TRUE)
-
-source("R/get_df.R")
-
+library(Robinson2020)
 ```
@@
   sf::st_as_sfc()
+
+vep
 
@@
 ```{r}
-# Define the Bocinsky et al. 2016 study area
-SWUS_bbox <- sf::st_bbox(c(xmin = -113,
-                           xmax = -105,
-                           ymin = 32,
-                           ymax = 38), 
-                         crs = 4326) %>%
-  sf::st_as_sfc()
-
-# Define the Phoenix Basin
-# Defined as SW AZ counties, using the National Map small scale data
-# "https://prd-tnm.s3.amazonaws.com/StagedProducts/Small-scale/data/Boundaries/countyp010g.gdb_nt00933.tar.gz" %>%
-#   download.file(destfile="../data-raw/countyp010g.gdb_nt00933.tar.gz", 
-#                 mode='wb')
-# untar("../data-raw/countyp010g.gdb_nt00933.tar.gz",
-#       exdir="../data-raw/countyp010g")
-phx_basin <- "./data-raw/countyp010g/countyp010g.gdb/" %>%
-  sf::st_read() %>%
-  dplyr::filter(NAME %in% 
-                  c("Maricopa",
-                    "Pinal",
-                    "Pima")) %>%
-  sf::st_transform(4326) %>%
-  sf::st_union()
-
 # Read in the Bocinsky et al. 2016 tree ring data
@@
 ggplot() + 
-  geom_sf(data = four_corners_states) +
-  ggplot2::geom_raster(data = swus_hillshade_500m %>%
-                         get_df(),
-                       mapping = aes(x = x,
+  ggplot2::geom_sf(data = four_corners_states) +
+  ggplot2::geom_raster(data = Robinson2020::swus_hillshade_500m,
+                       mapping = ggplot2::aes(x = x,
                                               y = y,
@@
                        na.rm = TRUE) +
-  scale_alpha(range = c(0.8, 0),
+  ggplot2::scale_alpha(range = c(0.8, 0),
                        na.value = 0,
@@
                        guide = "none") +
-  geom_sf(data = four_corners_states,
+  ggplot2::geom_sf(data = four_corners_states,
                    fill = NA) +
-  geom_sf(data = SWUS_bbox,
+  ggplot2::geom_sf(data = SWUS_bbox,
                    fill = NA) +
-  geom_sf(data = dates,
+  ggplot2::geom_sf(data = dates,
                    mapping = aes(shape = `Date Type`,
@@
                    show.legend = "point") +
-  scale_shape(
+  ggplot2::scale_shape(
     name = 'Date Type',
     solid = FALSE,
-    guide = guide_legend(position = "top",
+    guide = ggplot2::guide_legend(position = "top",
                                   direction = "horizontal",
@@
     direction = -1,
-    guide = guide_colorbar(position = "top",
+    guide = ggplot2::guide_colorbar(position = "top",
                                     direction = "horizontal",

### 2020-01-18 analysis/paper/paper.Rmd
initial commit

---
title: "Robinson et al. 2020 Supplemental Information: Comparing the Radiocarbon and Tree-ring Records for the Upland US Southwest"
author:
  - Kyle Bocinsky:
      email: bocinsky@gmail.com
      institute: [CCAC]
      correspondence: true
  - Erick Robinson:
      email: erick.robinson@usu.edu
      institute: [USU]
      correspondence: false
institute:
  - CCAC: Crow Canyon Archaeological Center
  - USU: Utah State University
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::html_document2:
      fig_caption: yes
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
    bookdown::pdf_document2:
      fig_caption: yes
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/philosophical-transactions-of-the-royal-society-b.csl" # Insert path for the bib-style
editor_options: 
  chunk_output_type: console
---

<!-- This is the format for text comments that will be ignored during renderings. Do not put R code in these comments because it will not be ignored. -->

```{r setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = "../figures/",
  dpi = 300
)

library(robinson2020)
```

# Summary of methods
Here, we directly compare two large collections of radiocarbon and tree-ring dates drawn from the Upland US Southwest --- roughly defined as the 'Four Corners' states of Arizona, Colorado, New Mexico, and Utah, but excluding the Phoenix and Tucson basins.

After harmonizing those datasets to with one another and calibrating Radiocarbon dates are drawn from the [DARCY: DATABASE AND REFERENCE HERE], and are available [HERE]. Tree-ring data are those presented in @Bocinsky2016_SA and archived on the Digital Archaeological Record (https://doi.org/10.6067/XCV82J6D7B).

## Study Area

```{r}
four_corners_lcc_proj <- "+proj=lcc +lat_1=37 +lon_0=-109.045225"

# Define the Bocinsky et al. 2016 study area
UUSW <- robinson2020::uusw_boundary

# VEP study areas
vep_study_areas <- 
  list(CMV = robinson2020::vepii_cmv_boundary,
       NRG = robinson2020::vepii_nrg_boundary) %>%
  purrr::map(sf::st_transform,
             crs = 4326) %>%
  purrr::imap(~dplyr::mutate(.x,`Study Area` = .y)) %>%
  do.call(rbind, .)

UUSW_counties <- 
  tigris::counties(cb = TRUE, 
                   class = "sf",
                   progress_bar = FALSE) %>% 
  sf::st_transform(four_corners_lcc_proj) %>% 
  # sf::st_contains(UUSW, sparse = FALSE)
  dplyr::filter(sf::st_intersects(suppressWarnings(sf::st_centroid(.)),
                                  UUSW %>% 
                                    sf::st_transform(four_corners_lcc_proj), 
                                  sparse = FALSE)) %>%
  dplyr::select(NAME,STATEFP) %>%
  dplyr::rename(County = NAME) %>%
  dplyr::left_join(tigris::states(class = "sf",
                                  progress_bar = FALSE) %>%
                     dplyr::select(STATEFP,STUSPS) %>%
                     dplyr::rename(State = STUSPS) %>%
                     sf::st_drop_geometry(),
                   by = "STATEFP") %>%
  dplyr::select(-STATEFP) %>%
  tidyr::unite(col = County, County, State, sep = ", ")

```

## Datasets

## Preparing the radiocarbon data

```{r}
# # Read in the radiocarbon database
# radiocarbon_data <-
#   readr::read_csv(here::here("analysis/data/raw_data/0403_sbox_clean.csv")) %>%
#   # Convert to spatial object
#   sf::st_as_sf(coords = c("Long","Lat"),
#                crs = 4326) %>%
#   sf::st_transform(four_corners_lcc_proj) %>%
#   # Select salient columns
#   dplyr::select(labnumber,
#                 date,
#                 sd,
#                 Site,
#                 SiteID) %>%
#   dplyr::rename(Lab_Number = labnumber,
#                 Date_BP = date,
#                 SD = sd,
#                 Site_Number = Site) %>%
#   sf::st_intersection(UUSW_counties) %>%
#   sf::st_drop_geometry() %>%
#   dplyr::mutate(Type = "Radiocarbon",
#                 SiteID = SiteID %>% stringr::str_c("Radiocarbon - ", .)) %T>%
#   # Write the output to a new file (csv)
#   readr::write_csv(here::here("analysis/data/derived_data/radiocarbon_data.csv"))

radiocarbon_data <-
  readr::read_csv(here::here("analysis/data/derived_data/radiocarbon_data.csv"))

```

These data contain sensitive location information about cultural sites, and are restricted on tDAR. Accordingly, we pre-processed the data using the script below, and provide data as supplemental information without location information represented.
```{r}
# # Read in the Bocinsky et al. 2016 tree ring data
# # https://doi.org/10.6067/XCV82J6D7B
# dendro_data <-
#   readr::read_csv(here::here("analysis/data/raw_data/BOCINSKY_ET_AL_2015_TREE_RINGS.csv")) %>%
#   # Convert to spatial object
#   sf::st_as_sf(coords = c("LON","LAT"),
#                crs = 4326) %>%
#   sf::st_transform(four_corners_lcc_proj) %>%
#   # Convert dates to BP
#   dplyr::mutate(Date_BP = 1950 - Outer_Date_AD,
#                 Type = ifelse(C_Level %in% c(2,3),
#                               "Cutting or Near-cutting",
#                               "Non-cutting")) %>%
#   # drop C_Level
#   dplyr::select(Seq_Number,
#                 Site_Number,
#                 Lab_Number,
#                 Type,
#                 Date_BP) %>%
#   dplyr::rename(SiteID = Seq_Number) %>%
#   sf::st_intersection(UUSW_counties) %>%
#   sf::st_drop_geometry() %>%
#   dplyr::mutate(Type = Type %>% stringr::str_c("Dendro - ", .),
#                 SiteID = SiteID %>% stringr::str_c("Dendro - ", .),
#                 SD = 0) %>%
#   dplyr::filter(Type == "Dendro - Cutting or Near-cutting") %>%
#   dplyr::mutate(Type = "Tree-ring") %T>%
#   # Write the output to a new file (csv)
#   readr::write_csv(here::here("analysis/data/derived_data/dendro_data.csv"))

dendro_data <- 
  readr::read_csv(here::here("analysis/data/derived_data/dendro_data.csv"))

```

We draw on the tree-ring date database presented in @Bocinsky2016_SA and archived on the Digital Archaeological Record (https://doi.org/10.6067/XCV82J6D7B). Here, we use only the subset of those data

# Combining datasets
```{r}
all_data <- list(Radiocarbon = radiocarbon_data,
                 `Tree-ring` = dendro_data) %>%
  purrr::map(~dplyr::mutate(.x,
                            `Study Area` = ifelse(County == "Montezuma, CO",
                                                  "CMV", 
                                                  NA),
                            `Study Area` = ifelse(County %in% c("Rio Arriba, NM",
                                                                "Sandoval, NM",
                                                                "Santa Fe, NM",
                                                                "Taos, NM"),
                                                  "NRG", 
                                                  `Study Area`)))

all_data %>%
  dplyr::bind_rows() %>%
  dplyr::group_by(Type,`Study Area`) %>%
  dplyr::summarise(Count = dplyr::n()) %>%
  tidyr::spread(Type, Count)

```

# Preparing the radiocarbon SPDs

```{r}
if(!file.exists(here::here("analysis/data/derived_data/radiocarbon_spd.rds")))
  all_data$Radiocarbon %>%
  split(.,.$`Study Area`) %>%
  tibble::tibble(`Study Area` = names(.),
                 Dates = .) %>%
  dplyr::bind_rows(tibble::tibble(`Study Area` = "UUSW",
                                  Dates = list(all_data$Radiocarbon))) %>%
  dplyr::rowwise() %>%
  dplyr::mutate(Calibrated = list(Dates %$%
                                    rcarbon::calibrate(x = Date_BP,
                                                       errors = SD,
                                                       calMatrix = TRUE,
                                                       normalised = TRUE,
                                                       verbose = FALSE)),
                Bins = list(rcarbon::binPrep(sites = Dates$SiteID,
                                             ages = Calibrated,
                                             h = 100)),
                SPD = list(rcarbon::spd(x = Calibrated,
                                        bins = Bins,
                                        timeRange = c(1949,0),
                                        runm = 21,
                                        verbose = FALSE))) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(Density = SPD %>%
                  purrr::map(function(x){
                    x$grid %>%
                      tibble::as_tibble() %$%
                      tibble::tibble(Type = "Radiocarbon Density",
                                     Date_BP = calBP,
                                     Density = PrDens / sum(PrDens, na.rm = T))
                  })
  ) %>%
  readr::write_rds(here::here("analysis/data/derived_data/radiocarbon_spd.rds"),
                   compress = "gz")

radiocarbon_spd <- readr::read_rds(here::here("analysis/data/derived_data/radiocarbon_spd.rds"))

```

To prepare the radiocarbon dataset, we used functions provided by the *rcarbon* package for R [@rcarbon]. We calibrated the radiocarbon dates using the methods of @Bronk2008 and implemented in @rcarbon. To control for inconsistent sample size across the archaeological sites represented in the database, we aggregated radiocarbon dates known to be from the same phase of same archaeological site using the `rcarbon::binprep` function with an `h`-value of 100. We averaged probability densities of binned dates, and then constructed summed probability density (SPD) distributions across bins.

# 'Calibrating' tree-ring dates, and preparing tree-ring SPDs

```{r}
calibrate_tree_ring_dates <- function(x){
  out <- list(
    metadata = tibble::tibble(
      DateID = 1:length(x),
      CRA = x,
      Error = 0,
      Details = NA,
      CalCurve = "intcal13",
      ResOffsets = 0,
      ResErrors = 0,
      StartBP = 50000,
      EndBP = 0,
      Normalised = TRUE,
      F14C = FALSE,
      CalEPS = 1e-05
    ),
    grids = NA,
    calmatrix = 
      tibble::tibble(index = 1:length(x),
                     year_bp = x,
                     value = TRUE) %>%
      tidyr::pivot_wider(names_from = index,
                         values_from = value) %>%
      dplyr::bind_rows(tibble::tibble(year_bp = 50000:0)) %>%
      dplyr::filter(!duplicated(year_bp)) %>%
      dplyr::arrange(-year_bp) %>%
      dplyr::filter(year_bp >= 0) %>%
      tibble::column_to_rownames("year_bp") %>%
      as.matrix() %>%
      tidyr::replace_na(0) %>%
      magrittr::set_colnames(NULL)
  )

  attr(out,"class") <- c("CalDates", "list")

  return(out)
}

if(!file.exists(here::here("analysis/data/derived_data/dendro_spd.rds")))
  all_data$`Tree-ring` %>%
  split(.,.$`Study Area`) %>%
  tibble::tibble(`Study Area` = names(.),
                 Dates = .) %>%
  dplyr::bind_rows(tibble::tibble(`Study Area` = "UUSW",
                                  Dates = list(all_data$`Tree-ring`))) %>%
  dplyr::rowwise() %>%
  dplyr::mutate(Calibrated = list(Dates %$%
                                    calibrate_tree_ring_dates(x = Date_BP)),
                Bins = list(rcarbon::binPrep(sites = Dates$SiteID,
                                             ages = Dates$Date_BP,
                                             h = 100)),
                SPD = list(rcarbon::spd(x = Calibrated,
                                        bins = Bins,
                                        timeRange = c(1949,0),
                                        runm = 21,
                                        verbose = FALSE))) %>%

  dplyr::ungroup() %>%
  dplyr::mutate(Density = SPD %>%
                  purrr::map(function(x){
                    x$grid %>%
                      tibble::as_tibble() %$%
                      tibble::tibble(Type = "Tree-ring Density",
                                     Date_BP = calBP,
                                     Density = PrDens / sum(PrDens, na.rm = T))
                  })
  ) %>%
  readr::write_rds(here::here("analysis/data/derived_data/dendro_spd.rds"),
                   compress = "gz")

dendro_spd <- readr::read_rds(here::here("analysis/data/derived_data/dendro_spd.rds"))

```

## Comparing radiocarbon and tree-ring SPDs

```{r}
all_density <- 
  dendro_spd %>%
  dplyr::select(`Study Area`,Density) %>%
  tidyr::unnest(cols = c(Density)) %>%
  dplyr::bind_rows(radiocarbon_spd %>%
                     dplyr::select(`Study Area`,Density) %>%
                     tidyr::unnest(cols = c(Density))) %>%
  dplyr::mutate(`Year AD` = 1950 - Date_BP) %>%
  dplyr::select(`Year AD`,
                Date_BP,
                `Study Area`,
                Type, 
                Density) %>%
  dplyr::rename(Value = Density)
```

```{r}
vep_demography <- 
  robinson2020::vep_demography %>%
  dplyr::rowwise() %>%
  dplyr::mutate(`Year AD` = list((Start+1):End)) %>%
  dplyr::select(`Study Area`,`Year AD`,Population) %>%
  tidyr::unnest(cols = c(`Year AD`)) %>%
  dplyr::mutate(Type = "Population",
                Date_BP = 1950 - `Year AD`) %>%
  dplyr::rename(Value = Population)
```

```{r c14_tr_pop}
all_density %>%
  dplyr::bind_rows(vep_demography) %>%
  dplyr::mutate(Type = factor(Type, 
                              levels = c("Radiocarbon Density",
                                         "Tree-ring Density",
                                         "Population"),
                              ordered = TRUE)#,
                # `Study Area` = ifelse(`Study Area` == "CMV", "Central Mesa Verde", "Northern Rio Grande")
  ) %>%
  ggplot2::ggplot(ggplot2::aes(x = `Year AD`)) +
  ggplot2::geom_line(ggplot2::aes(y = Value,
                                  color = `Study Area`),
                     size = 1.25) +
  # coord_cartesian(xlim = c(600,1600)) +
  ggplot2::xlab("Year AD") +
  ggplot2::theme_minimal(24) +
  ggplot2::theme(legend.title = ggplot2::element_blank(),
                 legend.justification = c(0, 1), 
                 legend.position = "bottom",
                 axis.title.y = ggplot2::element_blank(),
                 axis.text.y = ggplot2::element_blank(),
                 panel.grid.major.y = ggplot2::element_blank(),
                 panel.grid.minor.y = ggplot2::element_blank(),
                 plot.margin = ggplot2::unit(c(0.25,0.5,0.25,0.25),"in")
  ) +
  ggplot2::facet_wrap("Type", 
                      nrow = 3,
                      scales = "free_y")

```

<!-- ```{r} -->
<!-- four_corners_lcc_proj <- "+proj=lcc +lat_1=37 +lon_0=-109.045225" -->

<!-- # Define the Bocinsky et al. 2016 study area -->
<!-- SWUS_bbox <- sf::st_bbox(c(xmin = -113, -->
<!--                            xmax = -105, -->
<!--                            ymin = 32, -->
<!--                            ymax = 38),  -->
<!--                          crs = 4326) %>% -->
<!--   sf::st_as_sfc() %>% -->
<!--   sf::st_set_crs(NA) %>% -->
<!--   sf::st_segmentize(dfMaxLength = 0.01) %>% -->
<!--   sf::st_set_crs(4326) %>% -->
<!--   sf::st_transform(four_corners_lcc_proj) -->

<!-- four_corners_states <- "./data-raw/statesp010g/statesp010g.gdb/" %>% -->
<!--   sf::st_read() %>% -->
<!--   dplyr::filter(NAME %in%  -->
<!--                   c("Arizona", -->
<!--                     "Colorado", -->
<!--                     "New Mexico", -->
<!--                     "Utah")) %>% -->
<!--   sf::st_transform(four_corners_lcc_proj) -->

<!-- swus_hillshade_500m <- readr::read_rds("./data-derived/swus_hillshade_500m.rds") %>% -->
<!--   raster::crop(four_corners_states) %>% -->
<!--   raster::mask(four_corners_states) -->

<!-- dendro_data <- sf::read_sf("./data-derived/dendro_data.geojson") %>% -->
<!--   sf::st_transform(four_corners_lcc_proj) %>% -->
<!--   dplyr::filter(Date_BP >= 550, -->
<!--                 Date_BP <= 1450) %>% -->
<!--   dplyr::group_by(Site_Number) %>% -->
<!--   dplyr::summarise(Date_BP = mean(Date_BP, na.rm = T), -->
<!--                    geometry = mean(geometry)) %>% -->
<!--   dplyr::mutate(`Date Type` = "Dendro") %>% -->
<!--   dplyr::select(-Site_Number) %>% -->
<!--   sf::st_centroid() -->

<!-- radiocarbon_data <- sf::read_sf("./data-derived/radiocarbon_data.geojson") %>% -->
<!--   sf::st_transform(four_corners_lcc_proj) %>% -->
<!--   dplyr::filter(Date_BP >= 550, -->
<!--                 Date_BP <= 1450) %>% -->
<!--   dplyr::group_by(Site_Number) %>% -->
<!--   dplyr::summarise(Date_BP = mean(Date_BP, na.rm = T)) %>% -->
<!--   sf::st_jitter(20000) %>% -->
<!--   dplyr::mutate(`Date Type` = "C14") %>% -->
<!--   dplyr::select(-Site_Number) %>% -->
<!--   sf::st_centroid() -->

<!-- dates <- rbind(dendro_data, -->
<!--                radiocarbon_data) -->

<!-- theme_map <- function(base_size = 6.5,  -->
<!--                       base_family = ""){ -->
<!--   ggplot2::theme_bw(base_size = base_size,  -->
<!--                     base_family = base_family) %+replace% -->
<!--     ggplot2::theme(axis.line = ggplot2::element_blank(), -->
<!--                    axis.text = ggplot2::element_blank(), -->
<!--                    axis.ticks = ggplot2::element_blank(), -->
<!--                    axis.title = ggplot2::element_blank(), -->
<!--                    panel.background = ggplot2::element_blank(), -->
<!--                    panel.border = ggplot2::element_blank(), -->
<!--                    panel.grid = ggplot2::element_blank(), -->
<!--                    panel.grid.major = ggplot2::element_line(colour = 'transparent'), -->
<!--                    panel.grid.minor = ggplot2::element_line(colour = 'transparent'), -->
<!--                    legend.key = element_blank(), -->
<!--                    legend.position = "top", -->
<!--                    # legend.justification = c(0,0), -->
<!--                    # legend.position = "top", -->
<!--                    # legend.position = c(0.5,1), -->
<!--                    # legend.background = ggplot2::element_blank(), -->
<!--                    # legend.key.width = unit(0.15,"in"), -->
<!--                    # legend.key.height = unit(0.15,"in"), -->
<!--                    # legend.text = element_text(size = ggplot2::rel(1)), -->
<!--                    plot.background = ggplot2::element_blank(), -->
<!--                    plot.margin = ggplot2::margin(t = 0, r = 0, b = 0, l = 0, unit = "npc") -->
<!--     ) -->
<!-- } -->

<!-- ggplot() +  -->
<!--   ggplot2::geom_sf(data = four_corners_states) + -->
<!--   ggplot2::geom_raster(data = Robinson2020::swus_hillshade_500m, -->
<!--                        mapping = ggplot2::aes(x = x, -->
<!--                                               y = y, -->
<!--                                               alpha = ID), -->
<!--                        na.rm = TRUE) + -->
<!--   ggplot2::scale_alpha(range = c(0.8, 0), -->
<!--                        na.value = 0, -->
<!--                        limits = c(0,255), -->
<!--                        guide = "none") + -->
<!--   ggplot2::geom_sf(data = four_corners_states, -->
<!--                    fill = NA) + -->
<!--   ggplot2::geom_sf(data = SWUS_bbox, -->
<!--                    fill = NA) + -->
<!--   ggplot2::geom_sf(data = dates, -->
<!--                    mapping = aes(shape = `Date Type`, -->
<!--                                  color = Date_BP), -->
<!--                    show.legend = "point") + -->
<!--   ggplot2::scale_shape( -->
<!--     name = 'Date Type', -->
<!--     solid = FALSE, -->
<!--     guide = ggplot2::guide_legend(position = "top", -->
<!--                                   direction = "horizontal", -->
<!--                                   title.position = "top", -->
<!--                                   label.position = "bottom") -->
<!--   ) + -->
<!--   viridis::scale_color_viridis( -->
<!--     name = 'Year BP', -->
<!--     limits = c(450,1450), -->
<!--     breaks = c(450,1450), -->
<!--     direction = -1, -->
<!--     guide = ggplot2::guide_colorbar(position = "top", -->
<!--                                     direction = "horizontal", -->
<!--                                     title.position = "top", -->
<!--                                     label.position = "bottom", -->
<!--                                     reverse = TRUE)) + -->
<!--   theme_map(16) -->

<!-- ggsave('./swus_sample_map.pdf', -->
<!--        height = 8, -->
<!--        width = 6) -->
<!-- ``` -->

<!-- The following line inserts a page break  -->
\newpage

# References 
<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->
<div id="refs"></div>

\newpage

### Colophon

This report is part of a research compendium developed using the `rrtools` package for reproducible research [@Marwick2017, @rrtools]. This report was generated on `r Sys.time()` using the following computational environment and dependencies:

```{r colophon, cache = FALSE}
# which R packages and versions?
if ("devtools" %in% installed.packages()) devtools::session_info()
```

The current Git commit details are:

```{r}
# what commit is this file at? 
if ("git2r" %in% installed.packages() & git2r::in_repository(path = ".")) git2r::repository(here::here())  
```

### 2020-01-18 R/Recursive14Ccode.R

@@ -1,175 +1,0 @@
-setwd("~/R/19_entropyregimechange/")
-
-library(plyr)
-library(searchable)
-library(reshape)
-library(sf)
-library(robustbase)
-library(rcarbon)
-#library(tidyverse)
-#remove.packages(tidyverse)
-
-
-
-#### 2.  select relevant datess -------
-Rawdata <- read.csv("14C_Raw.csv")
-Rawdata$state <- Rawdata$state
-
-# Define the Bocinsky et al. 2016 study area
-SWUS_bbox <- sf::st_bbox(c(xmin = -113,
-                           xmax = -105,
-                           ymin = 32,
-                           ymax = 38), 
-                         crs = 4326) %>%
-  sf::st_as_sfc() 
-
-Rawdata <-subset(Rawdata, Rawdata$Country == c("USA")) %>%
-  sf::st_as_sf(coords = c("Long","Lat"),
-               crs = 4326)%>%
-  #Remove SWUS bounding box
-  sf::st_difference(SWUS_bbox)
-
-Rawdata$state <- gsub("UT", "Northern Southwest", Rawdata$state, fixed= TRUE)
-Rawdata$state <- gsub("CO", "Northern Southwest", Rawdata$state, fixed= TRUE)
-Rawdata$state <- gsub("AZ", "Southern Southwest", Rawdata$state, fixed= TRUE)
-Rawdata$state <- gsub("NM", "Southern Southwest", Rawdata$state, fixed= TRUE)
-
-Rawdata$state <- as.factor(Rawdata$state)
-
-write.csv(Rawdata, "Rawdata2.csv")
-
-
-
-
-##### make directory -----
-Directory <- count(RawData, vars = "state")
-names(Directory) <- c("state", "n")
-Directory<-Directory[!(Directory$n<200),]
-write.csv(Directory, "Directory_state.csv", row.names = FALSE)
-###
-
-#### 3. Calibrate each sampling unit recursively --------
-
-#Sometimes r treats values within a dataframe in a way you cannot use. The lines ensure our calibration will work.
-Rawdata$date <- as.numeric(Rawdata$date)
-Rawdata$sd <- as.numeric(Rawdata$sd)
-Rawdata$labnumber <- as.character(Rawdata$labnumber)
-Rawdata$state <- as.character(Rawdata$state)
-
-##Turn each sampling unit into a data.frame and make a list of them. 
-stateList <- list()
-for(i in 1:length(unique(Directory$state))){
[…]

### 2020-01-18 R/get_df.R

@@ -1,21 +1,0 @@
-get_df <- function(x){
-  out <- cbind(raster::xyFromCell(x, seq_len(raster::ncell(x))),
-               tibble::tibble(ID = raster::getValues(x))) %>%
-    tibble::as_tibble()
-  
-  if(is.factor(x)){
-    
-    levels <- levels(x)[[1]] %>%
-      dplyr::mutate_all(.funs = list(ordered)) %>%
-      tibble::as_tibble()
-    
-    fact <- out$ID %>%
-      ordered(levels = levels(levels$ID))
-    
-    out %<>%
-      dplyr::mutate(ID = fact) %>%
-      dplyr::left_join(levels)
-  }
-  
-  return(out)
-}

### 2020-01-18 R/mdt_theme_map.R

@@ -1,26 +1,0 @@
-mdt_theme_map <- function(base_size = 6.5, 
-                          base_family = ""){
-  ggplot2::theme_bw(base_size = base_size, 
-                    base_family = base_family) %+replace%
-    ggplot2::theme(axis.line = ggplot2::element_blank(),
-                   axis.text = ggplot2::element_blank(),
-                   axis.ticks = ggplot2::element_blank(),
-                   axis.title = ggplot2::element_blank(),
-                   panel.background = ggplot2::element_blank(),
-                   panel.border = ggplot2::element_blank(),
-                   panel.grid = ggplot2::element_blank(),
-                   panel.grid.major = ggplot2::element_line(colour = 'transparent'),
-                   panel.grid.minor = ggplot2::element_line(colour = 'transparent'),
-                   legend.key = element_blank(),
-                   legend.position = "top",
-                   # legend.justification = c(0,0),
-                   # legend.position = "top",
-                   # legend.position = c(0.5,1),
-                   # legend.background = ggplot2::element_blank(),
-                   # legend.key.width = unit(0.15,"in"),
-                   # legend.key.height = unit(0.15,"in"),
-                   # legend.text = element_text(size = ggplot2::rel(1)),
-                   plot.background = ggplot2::element_blank(),
-                   plot.margin = ggplot2::margin(t = 0, r = 0, b = 0, l = 0, unit = "npc")
-    )
-}

### 2020-01-18 R/plot_dates_map.R

@@ -1,121 +1,0 @@
-packages <- c("magrittr",
-              "tidyverse",
-              "googledrive",
-              "sf",
-              "lwgeom",
-              "units"
-) 
-# purrr::walk(packages, devtools::install_cran) %>%
-purrr::walk(packages,
-            library, 
-            character.only = TRUE)
-
-source("./R/get_df.R")
-
-# A nice projection for the Four Corners states
-four_corners_lcc_proj <- "+proj=lcc +lat_1=37 +lon_0=-109.045225"
-
-# Define the Bocinsky et al. 2016 study area
-SWUS_bbox <- sf::st_bbox(c(xmin = -113,
-                           xmax = -105,
-                           ymin = 32,
-                           ymax = 38), 
-                         crs = 4326) %>%
-  sf::st_as_sfc() %>%
-  sf::st_set_crs(NA) %>%
-  sf::st_segmentize(dfMaxLength = 0.01) %>%
-  sf::st_set_crs(4326) %>%
-  sf::st_transform(four_corners_lcc_proj)
-
-# Define the Four Corners states
-# "https://prd-tnm.s3.amazonaws.com/StagedProducts/Small-scale/data/Boundaries/statesp010g.gdb_nt00937.tar.gz" %>%
-#   download.file(destfile="../Data/data-raw/statesp010g.gdb_nt00937.tar.gz",
-#                 mode='wb')
-# untar("../Data/data-raw/statesp010g.gdb_nt00937.tar.gz",
-#       exdir="../Data/data-raw/statesp010g")
-four_corners_states <- "./Data/data-raw/statesp010g/statesp010g.gdb/" %>%
-  sf::st_read() %>%
-  dplyr::filter(NAME %in% 
-                  c("Arizona",
-                    "Colorado",
-                    "New Mexico",
-                    "Utah")) %>%
-  sf::st_transform(four_corners_lcc_proj)
-
-swus_hillshade_500m <- readr::read_rds("./Data/data-derived/swus_hillshade_500m.rds") %>%
-  raster::crop(four_corners_states) %>%
-  raster::mask(four_corners_states)
-
-dendro_data <- sf::read_sf("./Data/data-derived/dendro_data.geojson") %>%
-  sf::st_transform(four_corners_lcc_proj) %>%
-  dplyr::filter(Date_BP >= 550,
-                Date_BP <= 1450) %>%
-  dplyr::group_by(Site_Number) %>%
[…]

### 2020-01-18 R/prepare_dendro_dates.R

@@ -1,47 +1,0 @@
-packages <- c("magrittr",
-              "tidyverse",
-              "googledrive",
-              "sf",
-              "lwgeom"
-) 
-# purrr::walk(packages, devtools::install_cran) %>%
-purrr::walk(packages,
-            library, 
-            character.only = TRUE)
-
-# Define the Bocinsky et al. 2016 study area
-SWUS_bbox <- sf::st_bbox(c(xmin = -113,
-                           xmax = -105,
-                           ymin = 32,
-                           ymax = 38), 
-                         crs = 4326) %>%
-  sf::st_as_sfc() 
-  
-
-# Read in the Bocinsky et al. 2016 tree ring data
-dendro_data <- 
-  readr::read_csv("./data-raw/TreeRing/Bocinsky_et_al_2016_dendrochronology.csv") %>%
-  # Convert to spatial object
-  sf::st_as_sf(coords = c("LON","LAT"),
-            crs = 4326) %>%
-  # Crop to SWUS bounding box
-  sf::st_intersection(SWUS_bbox) %>%
-  # Convert dates to BP
-  dplyr::mutate(Date_BP = 1950 - Outer_Date_AD,
-                Type = ifelse(C_Level %in% c(2,3),
-                              "Cutting or Near-cutting",
-                              "Non-cutting")) %>%
-  # drop C_Level
-  dplyr::select(Seq_Number,
-                Site_Number,
-                Site_Name,
-                Lab_Number,
-                Type,
-                Date_BP) %>%
-  dplyr::rename(SiteID = Seq_Number) %T>%
-  # Write the output to a new file (geojson)
-  sf::write_sf("./data-derived/dendro_data.geojson",
-               delete_dsn = TRUE)
-  
-# Check R/T
-# sf::read_sf("./data-derived/dendro_data.geojson")

### 2020-01-18 R/prepare_radiocarbon_dates.R

@@ -1,72 +1,0 @@
-packages <- c("magrittr",
-              "tidyverse",
-              "googledrive",
-              "sf",
-              "lwgeom"
-) 
-# purrr::walk(packages, devtools::install_cran) %>%
-purrr::walk(packages,
-            library, 
-            character.only = TRUE)
-
-# Define the Bocinsky et al. 2016 study area
-SWUS_bbox <- sf::st_bbox(c(xmin = -113,
-                           xmax = -105,
-                           ymin = 32,
-                           ymax = 38), 
-                         crs = 4326) %>%
-  sf::st_as_sfc() 
-
-# Define the Phoenix Basin
-# Defined as SW AZ counties, using the National Map small scale data
-# "https://prd-tnm.s3.amazonaws.com/StagedProducts/Small-scale/data/Boundaries/countyp010g.gdb_nt00933.tar.gz" %>%
-#   download.file(destfile="../data-raw/countyp010g.gdb_nt00933.tar.gz", 
-#                 mode='wb')
-# untar("../data-raw/countyp010g.gdb_nt00933.tar.gz",
-#       exdir="../data-raw/countyp010g")
-phx_basin <- "./data-raw/countyp010g/countyp010g.gdb/" %>%
-  sf::st_read() %>%
-  dplyr::filter(NAME %in% 
-                  c("Maricopa",
-                    "Pinal",
-                    "Pima")) %>%
-  sf::st_transform(4326) %>%
-  sf::st_union()
-
-# Read in the Bocinsky et al. 2016 tree ring data
-radiocarbon_data <- #readr::read_csv("./data-raw/Radiocarbon/14C_Raw.csv") %>%
-  readr::read_csv("./data-raw/Radiocarbon/0403_sbox_clean.csv") %>%
-  # Convert to spatial object
-  sf::st_as_sf(coords = c("Long","Lat"),
-               crs = 4326) %>%
-  # Crop to SWUS bounding box
-  sf::st_intersection(SWUS_bbox) %>%
-  # Remove Phoenix Basin dates
-  sf::st_difference(phx_basin) %>%
-  # Select salient columns
-  dplyr::select(labnumber,
-                date,
-                sd,
-                #diff,
-                #d13,
-                #mat,
-                #Source,
-                Site,
-                SiteID) %>%
-  dplyr::rename(Lab_Number = labnumber,
-                Date_BP = date,
-                SD = sd,
-                # Diff = diff,
-                # Material = mat,
[…]

### 2020-01-18 R/swus_hillshade.R

@@ -1,93 +1,0 @@
-packages <- c("magrittr",
-              "tidyverse",
-              "googledrive",
-              "sf",
-              "lwgeom",
-              "units",
-              "FedData",
-              "rgdal",
-              "raster"
-) 
-# purrr::walk(packages, devtools::install_cran) %>%
-purrr::walk(packages,
-            library, 
-            character.only = TRUE)
-
-# A nice projection for the Four Corners states
-four_corners_lcc_proj <- "+proj=lcc +lat_1=37 +lon_0=-109.045225"
-
-# Define the Four Corners states
-# "https://prd-tnm.s3.amazonaws.com/StagedProducts/Small-scale/data/Boundaries/statesp010g.gdb_nt00937.tar.gz" %>%
-#   download.file(destfile="../Data/data-raw/statesp010g.gdb_nt00937.tar.gz",
-#                 mode='wb')
-# untar("../Data/data-raw/statesp010g.gdb_nt00937.tar.gz",
-#       exdir="../Data/data-raw/statesp010g")
-four_corners_states <- "./data-raw/statesp010g/statesp010g.gdb/" %>%
-  sf::st_read() %>%
-  dplyr::filter(NAME %in% 
-                  c("Arizona",
-                    "Colorado",
-                    "New Mexico",
-                    "Utah")) %>%
-  sf::st_transform(four_corners_lcc_proj) %>%
-  sf::st_union()
-
-swus_rast_500 <- raster::raster(raster::extent(-566500,643000,3815000,5052500),
-                                nrows = 2475,
-                                ncols = 2419,
-                                crs = CRS(four_corners_lcc_proj))
-
-swus_ned <- FedData::get_ned(template = four_corners_states %>%
-                             sf::st_buffer(20000) %>%
-                             as("Spatial"),
-                           label = "swus_4C",
-                           raw.dir = "./data-raw/ned",
-                           extraction.dir = "./data-raw/")
-
-system("gdaldem hillshade ./data-raw/swus_4C_NED_1.tif ./data-derived/swus_4C_hillshade.tif -z 2 -s 111120 -multidirectional -co 'COMPRESS=DEFLATE' -co 'ZLEVEL=9'")
-
-## Generate a hillshade for statewide mapping
[…]

### 2020-01-18 Robinson2020.R

@@ -1,258 +1,0 @@
-packages <- c("magrittr",
-              "tidyverse",
-              "googledrive",
-              "sf",
-              "lwgeom",
-              "units",
-              "rcarbon",
-              "villager"
-)
-# devtools::install_github("benmarwick/rrtools")
-# devtools::install_github("village-ecodynamics/villager")
-
-# purrr::walk(packages, devtools::install_cran)
-purrr::walk(packages,
-            library, 
-            character.only = TRUE)
-
-# VEP study areas
-vep_study_areas <- 
-  list(CMV = villager::get_vep_boundary(version = "vepii_cmv"),
-       NRG = villager::get_vep_boundary(version = "vepii_nrg")) %>%
-  purrr::map(sf::st_transform,
-             crs = 4326) %>%
-  purrr::map(tibble::as_tibble) %>%
-  dplyr::bind_rows(.id = "Study Area") %>%
-  sf::st_as_sf(crs = 4326)
-
-#SPDs for SW region overlapping Bocinsky et al. 2016 dendro cutting date database
-radiocarbon_data <-
-  sf::read_sf("./data-derived/radiocarbon_data.geojson") %>%
-  dplyr::mutate(Type = "Radiocarbon",
-                SiteID = SiteID %>% stringr::str_c("Radiocarbon - ", .)) %>%
-  dplyr::mutate(`Study Area` = ifelse(County == "Montezuma, CO",
-                                      "CMV", 
-                                      NA),
-                `Study Area` = ifelse(County %in% c("Rio Arriba, NM",
-                                                    "Sandoval, NM",
-                                                    "Santa Fe, NM",
-                                                    "Taos, NM"),
-                                      "NRG", 
-                                      `Study Area`)) %>%
-  dplyr::select(-County)
-
-# We can do many of the same things for the tr dates
-dendro_data <- 
-  sf::read_sf("./data-derived/dendro_data.geojson") %>%
-  dplyr::mutate(Type = Type %>% stringr::str_c("Dendro - ", .),
-                SiteID = SiteID %>% stringr::str_c("Dendro - ", .),
-                SD = 0) %>%
-  dplyr::filter(Type == "Dendro - Cutting or Near-cutting") %>%
-  dplyr::mutate(Type = "Tree-ring") %>%
-  sf::st_join(vep_study_areas) %>%
-  dplyr::select(-Site_Name) %>%
-  # dplyr::rename(`Study Area` = `Study.Area`) %>%
[…]

### 2020-01-18 data-raw/DATASET.R

@@ -1,0 +1,146 @@
+library(magrittr)
+library(tidycensus)
+
+## Get the VEP II study area definitions from the `villager` package.
+## https://github.com/village-ecodynamics/villager
+vepii_cmv_boundary <- villager::vepii_cmv_boundary
+vepii_nrg_boundary <- villager::vepii_nrg_boundary
+
+usethis::use_data(vepii_cmv_boundary)
+usethis::use_data(vepii_nrg_boundary)
+
+## Create the Upland US Southwest boundary
+## The boundary is defined as a geographic rectangle sans Maricopa, Pinal, and Pima counties
+uusw_boundary <-
+  sf::st_bbox(c(xmin = -113,
+                xmax = -105,
+                ymin = 32,
+                ymax = 38),
+              crs = 4326) %>%
+  sf::st_as_sfc() %>%
+  sf::st_difference(tigris::counties(state="Arizona", cb = TRUE) %>%
+                      sf::st_as_sf() %>%
+                      dplyr::filter(NAME %in% c("Maricopa",
+                                                "Pinal",
+                                                "Pima")) %>%
+                      sf::st_union() %>%
+                      sf::st_transform(4326))
+
+usethis::use_data(uusw_boundary)
+
+## Create a nice hillshade for the Four Corners states for mapping
+# A nice projection for the Four Corners states
+four_corners_lcc_proj <- "+proj=lcc +lat_1=37 +lon_0=-109.045225"
+
+
+# Define the Four Corners states
+# "https://prd-tnm.s3.amazonaws.com/StagedProducts/Small-scale/data/Boundaries/statesp010g.gdb_nt00937.tar.gz" %>%
+#   download.file(destfile="../Data/data-raw/statesp010g.gdb_nt00937.tar.gz",
+#                 mode='wb')
+# untar("../Data/data-raw/statesp010g.gdb_nt00937.tar.gz",
+#       exdir="../Data/data-raw/statesp010g")
+four_corners_states <- "./data-raw/statesp010g/statesp010g.gdb/" %>%
+  sf::st_read() %>%
+  dplyr::filter(NAME %in%
+                  c("Arizona",
+                    "Colorado",
+                    "New Mexico",
+                    "Utah")) %>%
+  sf::st_transform(four_corners_lcc_proj) %>%
+  sf::st_union()
+
[…]

### 2020-01-18 README.Rmd
initial commit
gpl3

@@
 # Please put your title here to include it in the file below.
-Title <- "Title of your paper goes here"
+Title <- "Dendrochronological dates confirm a Late Prehistoric population decline in the American Southwest derived from radiocarbon dates"
 ```
 
-# Robinson2020
+# robinson2020
 
@@
 
-> Authors, (YYYY). _`r Title`_. Name of journal/book <https://doi.org/xxx/xxx>
+> Robinson, Erick, R. Kyle Bocinsky, Jacob Freeman, Darcy Bird, Robert L. Kelly, (2020). _`r Title`_. Under review.
 
-Our pre-print is online here:
+<!-- Our pre-print is online here: -->
 
-> Authors, (YYYY). _`r Title`_. Name of journal/book, Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>
+<!-- > Authors, (YYYY). _`r Title`_. Name of journal/book, Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx> -->
 
@@
 
-> Authors, (`r format(Sys.Date(), "%Y")`). _Compendium of R code and data for `r Title`_. Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>
+> Bocinsky, R. Kyle and Erick Robinson, (`r format(Sys.Date(), "%Y")`). _Compendium of R code and data for `r Title`_. Accessed `r format(Sys.Date(), "%d %b %Y")`.
 
@@
 # install.packages("devtools")
-remotes::install_github("/")
+remotes::install_github("bocinsky/robinson2020")
 ```

### 2020-01-18 README.Rmd
gpl3
binder

@@
 
-[![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh///master?urlpath=rstudio)
+[![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh/bocinsky/robinson2020/master?urlpath=rstudio)
 
@@
 
-**Code :** See the [DESCRIPTION](DESCRIPTION) file
+**Code :** See the [LICENSE](LICENSE.md) file
 

### 2020-01-18 analysis/paper/paper.Rmd
gpl3
binder

@@
 
-<!-- This is the format for text comments that will be ignored during renderings. Do not put R code in these comments because it will not be ignored. -->
-
 ```{r setup, echo = FALSE}
@@
 # Summary of methods
-Here, we directly compare two large collections of radiocarbon and tree-ring dates drawn from the Upland US Southwest --- roughly defined as the 'Four Corners' states of Arizona, Colorado, New Mexico, and Utah, but excluding the Phoenix and Tucson basins.
+Here, we directly compare two large collections of radiocarbon and tree-ring dates drawn from the Upland US Southwest.
 
@@
 
-## Study Area
+# Study Area
 
@@
   sf::st_transform(four_corners_lcc_proj) %>% 
-  # sf::st_contains(UUSW, sparse = FALSE)
   dplyr::filter(sf::st_intersects(suppressWarnings(sf::st_centroid(.)),
@@
 
-## Datasets
+The UUSW was geographically definied by @Bocinsky2016_SA as the region between 105--113ºW and 32--38ºN.
+Here, we refine that definition in two ways.
+First, since location data for our radiocarbon database are only resolved to the county level, we define the study region to include all counties whose geographic centroids fall within the region from @Bocinsky2016_SA.
+Second, we roughly exclude the Phoenix and Tucson basins by excluding Maricopa, Pinal, and Pima counties in Arizona.
+Figure \@ref(fig:study-area) shows the original UUSW from @Bocinsky2016_SA, and the revised study area used in this analysis.
 
-## Preparing the radiocarbon data
+```{r study-area}
+theme_map <- function(base_size = 6.5,
+                      base_family = ""){
+  ggplot2::theme_bw(base_size = base_size,
+                    base_family = base_family) %+replace%
+    ggplot2::theme(axis.line = ggplot2::element_blank(),
+                   axis.text = ggplot2::element_blank(),
+                   axis.ticks = ggplot2::element_blank(),
+                   axis.title = ggplot2::element_blank(),
+                   panel.background = ggplot2::element_blank(),
+                   panel.border = ggplot2::element_blank(),
+                   panel.grid = ggplot2::element_blank(),
+                   panel.grid.major = ggplot2::element_line(colour = 'transparent'),
+                   panel.grid.minor = ggplot2::element_line(colour = 'transparent'),
+                   legend.key = ggplot2::element_blank(),
+                   legend.position = "top",
+                   # legend.justification = c(0,0),
+                   # legend.position = "top",
+                   # legend.position = c(0.5,1),
+                   # legend.background = ggplot2::element_blank(),
+                   # legend.key.width = unit(0.15,"in"),
+                   # legend.key.height = unit(0.15,"in"),
+                   # legend.text = element_text(size = ggplot2::rel(1)),
+                   plot.background = ggplot2::element_blank(),
+                   plot.margin = ggplot2::margin(t = 0, r = 0, b = 0, l = 0, unit = "npc")
+    )
+}
+
+four_corners_states <- 
+  tigris::states(class = "sf",
+                 progress_bar = FALSE) %>%
+  dplyr::filter(NAME %in% c("Arizona", "Colorado", 
+                            "New Mexico", "Utah")) %>%
+  sf::st_transform(four_corners_lcc_proj)
+
+ggplot2::ggplot() +
+  ggplot2::geom_sf(data = four_corners_states) +
+  ggplot2::geom_raster(data = robinson2020::swus_hillshade_500m,
+                       mapping = ggplot2::aes(x = x,
+                                              y = y,
+                                              alpha = ID),
+                       na.rm = TRUE) +
+  ggplot2::scale_alpha(range = c(0.8, 0),
+                       na.value = 0,
+                       limits = c(0,255),
+                       guide = "none") +
+  ggplot2::geom_sf(data = four_corners_states,
+                   fill = NA) +
+  ggplot2::geom_sf(data = UUSW,
+                   fill = NA) +
+    ggplot2::geom_sf(data = UUSW_counties,
+                   fill = NA) +
+  # ggplot2::geom_sf(data = dates,
+  #                  mapping = aes(shape = `Date Type`,
+  #                                color = Date_BP),
+  #                  show.legend = "point") +
+  # ggplot2::scale_shape(
+  #   name = 'Date Type',
+  #   solid = FALSE,
+  #   guide = ggplot2::guide_legend(position = "top",
+  #                                 direction = "horizontal",
+  #                                 title.position = "top",
+  #                                 label.position = "bottom")
+  # ) +
+  # viridis::scale_color_viridis(
+  #   name = 'Year BP',
+  #   limits = c(450,1450),
+  #   breaks = c(450,1450),
+  #   direction = -1,
+  #   guide = ggplot2::guide_colorbar(position = "top",
+  #                                   direction = "horizontal",
+  #                                   title.position = "top",
+  #                                   label.position = "bottom",
+  #                                   reverse = TRUE)) +
+  theme_map(16)
+
+```
+
+# Datasets
+
+# Preparing the radiocarbon data
 
@@
 
-<!-- ```{r} -->
-<!-- four_corners_lcc_proj <- "+proj=lcc +lat_1=37 +lon_0=-109.045225" -->
+```{r}
+four_corners_lcc_proj <- "+proj=lcc +lat_1=37 +lon_0=-109.045225"
 
-<!-- # Define the Bocinsky et al. 2016 study area -->
-<!-- SWUS_bbox <- sf::st_bbox(c(xmin = -113, -->
-<!--                            xmax = -105, -->
-<!--                            ymin = 32, -->
-<!--                            ymax = 38),  -->
-<!--                          crs = 4326) %>% -->
-<!--   sf::st_as_sfc() %>% -->
-<!--   sf::st_set_crs(NA) %>% -->
-<!--   sf::st_segmentize(dfMaxLength = 0.01) %>% -->
-<!--   sf::st_set_crs(4326) %>% -->
-<!--   sf::st_transform(four_corners_lcc_proj) -->
+# Define the Bocinsky et al. 2016 study area
+SWUS_bbox <- sf::st_bbox(c(xmin = -113,
+                           xmax = -105,
+                           ymin = 32,
+                           ymax = 38),
+                         crs = 4326) %>%
+  sf::st_as_sfc() %>%
+  sf::st_set_crs(NA) %>%
+  sf::st_segmentize(dfMaxLength = 0.01) %>%
+  sf::st_set_crs(4326) %>%
+  sf::st_transform(four_corners_lcc_proj)
 
-<!-- four_corners_states <- "./data-raw/statesp010g/statesp010g.gdb/" %>% -->
-<!--   sf::st_read() %>% -->
-<!--   dplyr::filter(NAME %in%  -->
-<!--                   c("Arizona", -->
-<!--                     "Colorado", -->
-<!--                     "New Mexico", -->
-<!--                     "Utah")) %>% -->
-<!--   sf::st_transform(four_corners_lcc_proj) -->
+four_corners_states <- "./data-raw/statesp010g/statesp010g.gdb/" %>%
+  sf::st_read() %>%
+  dplyr::filter(NAME %in%
+                  c("Arizona",
+                    "Colorado",
+                    "New Mexico",
+                    "Utah")) %>%
+  sf::st_transform(four_corners_lcc_proj)
 
-<!-- swus_hillshade_500m <- readr::read_rds("./data-derived/swus_hillshade_500m.rds") %>% -->
-<!--   raster::crop(four_corners_states) %>% -->
-<!--   raster::mask(four_corners_states) -->
+swus_hillshade_500m <- readr::read_rds("./data-derived/swus_hillshade_500m.rds") %>%
+  raster::crop(four_corners_states) %>%
+  raster::mask(four_corners_states)
 
-<!-- dendro_data <- sf::read_sf("./data-derived/dendro_data.geojson") %>% -->
-<!--   sf::st_transform(four_corners_lcc_proj) %>% -->
-<!--   dplyr::filter(Date_BP >= 550, -->
-<!--                 Date_BP <= 1450) %>% -->
-<!--   dplyr::group_by(Site_Number) %>% -->
-<!--   dplyr::summarise(Date_BP = mean(Date_BP, na.rm = T), -->
-<!--                    geometry = mean(geometry)) %>% -->
-<!--   dplyr::mutate(`Date Type` = "Dendro") %>% -->
-<!--   dplyr::select(-Site_Number) %>% -->
-<!--   sf::st_centroid() -->
+dendro_data <- sf::read_sf("./data-derived/dendro_data.geojson") %>%
+  sf::st_transform(four_corners_lcc_proj) %>%
+  dplyr::filter(Date_BP >= 550,
+                Date_BP <= 1450) %>%
+  dplyr::group_by(Site_Number) %>%
+  dplyr::summarise(Date_BP = mean(Date_BP, na.rm = T),
+                   geometry = mean(geometry)) %>%
+  dplyr::mutate(`Date Type` = "Dendro") %>%
+  dplyr::select(-Site_Number) %>%
+  sf::st_centroid()
 
-<!-- radiocarbon_data <- sf::read_sf("./data-derived/radiocarbon_data.geojson") %>% -->
-<!--   sf::st_transform(four_corners_lcc_proj) %>% -->
-<!--   dplyr::filter(Date_BP >= 550, -->
-<!--                 Date_BP <= 1450) %>% -->
-<!--   dplyr::group_by(Site_Number) %>% -->
-<!--   dplyr::summarise(Date_BP = mean(Date_BP, na.rm = T)) %>% -->
-<!--   sf::st_jitter(20000) %>% -->
-<!--   dplyr::mutate(`Date Type` = "C14") %>% -->
-<!--   dplyr::select(-Site_Number) %>% -->
-<!--   sf::st_centroid() -->
+radiocarbon_data <- sf::read_sf("./data-derived/radiocarbon_data.geojson") %>%
+  sf::st_transform(four_corners_lcc_proj) %>%
+  dplyr::filter(Date_BP >= 550,
+                Date_BP <= 1450) %>%
+  dplyr::group_by(Site_Number) %>%
+  dplyr::summarise(Date_BP = mean(Date_BP, na.rm = T)) %>%
+  sf::st_jitter(20000) %>%
+  dplyr::mutate(`Date Type` = "C14") %>%
+  dplyr::select(-Site_Number) %>%
+  sf::st_centroid()
 
-<!-- dates <- rbind(dendro_data, -->
-<!--                radiocarbon_data) -->
+dates <- rbind(dendro_data,
+               radiocarbon_data)
 
-<!-- theme_map <- function(base_size = 6.5,  -->
-<!--                       base_family = ""){ -->
-<!--   ggplot2::theme_bw(base_size = base_size,  -->
-<!--                     base_family = base_family) %+replace% -->
-<!--     ggplot2::theme(axis.line = ggplot2::element_blank(), -->
-<!--                    axis.text = ggplot2::element_blank(), -->
-<!--                    axis.ticks = ggplot2::element_blank(), -->
-<!--                    axis.title = ggplot2::element_blank(), -->
-<!--                    panel.background = ggplot2::element_blank(), -->
-<!--                    panel.border = ggplot2::element_blank(), -->
-<!--                    panel.grid = ggplot2::element_blank(), -->
-<!--                    panel.grid.major = ggplot2::element_line(colour = 'transparent'), -->
-<!--                    panel.grid.minor = ggplot2::element_line(colour = 'transparent'), -->
-<!--                    legend.key = element_blank(), -->
-<!--                    legend.position = "top", -->
-<!--                    # legend.justification = c(0,0), -->
-<!--                    # legend.position = "top", -->
-<!--                    # legend.position = c(0.5,1), -->
-<!--                    # legend.background = ggplot2::element_blank(), -->
-<!--                    # legend.key.width = unit(0.15,"in"), -->
-<!--                    # legend.key.height = unit(0.15,"in"), -->
-<!--                    # legend.text = element_text(size = ggplot2::rel(1)), -->
-<!--                    plot.background = ggplot2::element_blank(), -->
-<!--                    plot.margin = ggplot2::margin(t = 0, r = 0, b = 0, l = 0, unit = "npc") -->
-<!--     ) -->
-<!-- } -->
+theme_map <- function(base_size = 6.5,
+                      base_family = ""){
+  ggplot2::theme_bw(base_size = base_size,
+                    base_family = base_family) %+replace%
+    ggplot2::theme(axis.line = ggplot2::element_blank(),
+                   axis.text = ggplot2::element_blank(),
+                   axis.ticks = ggplot2::element_blank(),
+                   axis.title = ggplot2::element_blank(),
+                   panel.background = ggplot2::element_blank(),
+                   panel.border = ggplot2::element_blank(),
+                   panel.grid = ggplot2::element_blank(),
+                   panel.grid.major = ggplot2::element_line(colour = 'transparent'),
+                   panel.grid.minor = ggplot2::element_line(colour = 'transparent'),
+                   legend.key = element_blank(),
+                   legend.position = "top",
+                   # legend.justification = c(0,0),
+                   # legend.position = "top",
+                   # legend.position = c(0.5,1),
+                   # legend.background = ggplot2::element_blank(),
+                   # legend.key.width = unit(0.15,"in"),
+                   # legend.key.height = unit(0.15,"in"),
+                   # legend.text = element_text(size = ggplot2::rel(1)),
+                   plot.background = ggplot2::element_blank(),
+                   plot.margin = ggplot2::margin(t = 0, r = 0, b = 0, l = 0, unit = "npc")
+    )
+}
 
-<!-- ggplot() +  -->
-<!--   ggplot2::geom_sf(data = four_corners_states) + -->
-<!--   ggplot2::geom_raster(data = Robinson2020::swus_hillshade_500m, -->
-<!--                        mapping = ggplot2::aes(x = x, -->
-<!--                                               y = y, -->
-<!--                                               alpha = ID), -->
-<!--                        na.rm = TRUE) + -->
-<!--   ggplot2::scale_alpha(range = c(0.8, 0), -->
-<!--                        na.value = 0, -->
-<!--                        limits = c(0,255), -->
-<!--                        guide = "none") + -->
-<!--   ggplot2::geom_sf(data = four_corners_states, -->
-<!--                    fill = NA) + -->
-<!--   ggplot2::geom_sf(data = SWUS_bbox, -->
-<!--                    fill = NA) + -->
-<!--   ggplot2::geom_sf(data = dates, -->
-<!--                    mapping = aes(shape = `Date Type`, -->
-<!--                                  color = Date_BP), -->
-<!--                    show.legend = "point") + -->
-<!--   ggplot2::scale_shape( -->
-<!--     name = 'Date Type', -->
-<!--     solid = FALSE, -->
-<!--     guide = ggplot2::guide_legend(position = "top", -->
-<!--                                   direction = "horizontal", -->
-<!--                                   title.position = "top", -->
-<!--                                   label.position = "bottom") -->
-<!--   ) + -->
-<!--   viridis::scale_color_viridis( -->
-<!--     name = 'Year BP', -->
-<!--     limits = c(450,1450), -->
-<!--     breaks = c(450,1450), -->
-<!--     direction = -1, -->
-<!--     guide = ggplot2::guide_colorbar(position = "top", -->
-<!--                                     direction = "horizontal", -->
-<!--                                     title.position = "top", -->
-<!--                                     label.position = "bottom", -->
-<!--                                     reverse = TRUE)) + -->
-<!--   theme_map(16) -->
+ggplot() +
+  ggplot2::geom_sf(data = four_corners_states) +
+  ggplot2::geom_raster(data = Robinson2020::swus_hillshade_500m,
+                       mapping = ggplot2::aes(x = x,
+                                              y = y,
+                                              alpha = ID),
+                       na.rm = TRUE) +
+  ggplot2::scale_alpha(range = c(0.8, 0),
+                       na.value = 0,
+                       limits = c(0,255),
+                       guide = "none") +
+  ggplot2::geom_sf(data = four_corners_states,
+                   fill = NA) +
+  ggplot2::geom_sf(data = SWUS_bbox,
+                   fill = NA) +
+  ggplot2::geom_sf(data = dates,
+                   mapping = aes(shape = `Date Type`,
+                                 color = Date_BP),
+                   show.legend = "point") +
+  ggplot2::scale_shape(
+    name = 'Date Type',
+    solid = FALSE,
+    guide = ggplot2::guide_legend(position = "top",
+                                  direction = "horizontal",
+                                  title.position = "top",
+                                  label.position = "bottom")
+  ) +
+  viridis::scale_color_viridis(
+    name = 'Year BP',
+    limits = c(450,1450),
+    breaks = c(450,1450),
+    direction = -1,
+    guide = ggplot2::guide_colorbar(position = "top",
+                                    direction = "horizontal",
+                                    title.position = "top",
+                                    label.position = "bottom",
+                                    reverse = TRUE)) +
+  theme_map(16)
 
-<!-- ggsave('./swus_sample_map.pdf', -->
-<!--        height = 8, -->
-<!--        width = 6) -->
-<!-- ``` -->
+ggsave('./swus_sample_map.pdf',
+       height = 8,
+       width = 6)
+```
 

### 2020-01-18 data-raw/DATASET.R

@@ -17,16 +17,9 @@ uusw_boundary <-
                 ymin = 32,
                 ymax = 38),
               crs = 4326) %>%
-  sf::st_as_sfc() %>%
-  sf::st_difference(tigris::counties(state="Arizona", cb = TRUE) %>%
-                      sf::st_as_sf() %>%
-                      dplyr::filter(NAME %in% c("Maricopa",
-                                                "Pinal",
-                                                "Pima")) %>%
-                      sf::st_union() %>%
-                      sf::st_transform(4326))
+  sf::st_as_sfc()
 
-usethis::use_data(uusw_boundary)
+usethis::use_data(uusw_boundary, overwrite = TRUE)
 
 ## Create a nice hillshade for the Four Corners states for mapping
 # A nice projection for the Four Corners states

### 2020-01-19 analysis/paper/paper.Rmd
finished drafting methods section

--- 2020-01-18 analysis/paper/paper.Rmd
@@
     bookdown::html_document2:
+      code_folding: hide
       fig_caption: yes
@@
       - --lua-filter=../templates/pagebreak.lua
+header-includes:
+  - \usepackage{subfig}
 bibliography: references.bib
@@
 
-<!-- This is the format for text comments that will be ignored during renderings. Do not put R code in these comments because it will not be ignored. -->
-
-```{r setup, echo = FALSE}
+```{r setup, echo = FALSE, message = FALSE}
 knitr::opts_chunk$set(
+	echo = TRUE,
+	fig.align = "center",
+	fig.path = "../figures/",
+	message = FALSE,
+	warning = FALSE,
 	collapse = TRUE,
-  warning = FALSE,
-  message = FALSE,
-  echo = FALSE,
 	comment = "#>",
-  fig.path = "../figures/",
+	dev = "png",
 	dpi = 300
 )
-
 library(robinson2020)
@@
 
-# Summary of methods
-Here, we directly compare two large collections of radiocarbon and tree-ring dates drawn from the Upland US Southwest --- roughly defined as the 'Four Corners' states of Arizona, Colorado, New Mexico, and Utah, but excluding the Phoenix and Tucson basins.
+# Methods {-}
+Here, we directly compare two large collections of radiocarbon and tree-ring dates drawn from the Upland US Southwest (UUSW).
+After harmonizing those datasets to with one another and calibrating the radiocarbon dates, we calculate binned summed probability distributions across the UUSW and two regions intensly studied during the Village Ecodynamics Project (VEP) --- the Central Mesa Verde and the Northern Rio Grande [@Schwindt2016].
+We compare the proxy records to eachother, and to population estimates primarily derived from architecture and ceramic data [@Ortman2014,@Schwindt2016].
+The radiocarbon dates we analyze here are drawn from the [DARCY: DATABASE AND REFERENCE HERE], and are available [HERE]; The tree-ring data are those presented in @Bocinsky2016_SA and archived on the Digital Archaeological Record (https://doi.org/10.6067/XCV82J6D7B).
 
-After harmonizing those datasets to with one another and calibrating Radiocarbon dates are drawn from the [DARCY: DATABASE AND REFERENCE HERE], and are available [HERE].
-Tree-ring data are those presented in @Bocinsky2016_SA and archived on the Digital Archaeological Record (https://doi.org/10.6067/XCV82J6D7B).
+## Study Area {-}
 
-## Study Area
+The UUSW was geographically defined by @Bocinsky2016_SA as the region between 105--113ºW and 32--38ºN.
+Here, we refine that definition in two ways.
+First, because location data for our radiocarbon database are only resolved to the county level for most sites, we define our study region to include all counties whose geographic centroids fall within the region from @Bocinsky2016_SA.
+Second, because of the generally poor tree-ring date record in the Sonoran Desert, we roughly exclude the Phoenix and Tucson basins by dropping Maricopa, Pinal, and Pima counties in Arizona.
+We also analyze subsets of both datasets for the Village Ecodynamic Project (VEP) study areas in the Central Mesa Verde (CMV) and Northern Rio Grande (NRG) regions [@Schwindt2016].
+Figure \@ref(fig:study-area) shows the original UUSW from @Bocinsky2016_SA and revised study area used in this analysis, as well as the two VEP study areas.
 
-```{r}
-four_corners_lcc_proj <- "+proj=lcc +lat_1=37 +lon_0=-109.045225"
+```{r study-area, fig.height=6.5, fig.width=6.5, out.width = '50%', fig.cap="The UUSW as defined in this study (region with black border).
+The white dashed area represents the UUSW as defined in @Bocinsky2016_SA.
+VEP study areas are white boxes.
+CMV: Central Mesa Verde; NRG: Northern Rio Grande."}
+# A nice projection of the Four Corners states to use for mapping
+four_corners_lcc_proj <- 
+  robinson2020::four_corners_lcc_proj
 
-# Define the Bocinsky et al. 2016 study area
-UUSW <- robinson2020::uusw_boundary
+# The Four Corners states in four_corners_lcc_proj
+four_corners_states <- 
+  robinson2020::four_corners_states
 
-# VEP study areas
+# The UUSW as defined in Bocinsky et al. 2016
+uusw_boundary <- robinson2020::uusw_boundary
+
+# The UUSW study area for this analysis
+uusw_counties <- 
+  robinson2020::uusw_counties
+
+# The VEP study areas
 vep_study_areas <- 
-  list(CMV = robinson2020::vepii_cmv_boundary,
-       NRG = robinson2020::vepii_nrg_boundary) %>%
-  purrr::map(sf::st_transform,
-             crs = 4326) %>%
-  purrr::imap(~dplyr::mutate(.x,`Study Area` = .y)) %>%
-  do.call(rbind, .)
+  robinson2020::vep_study_areas
 
-UUSW_counties <- 
-  tigris::counties(cb = TRUE, 
-                   class = "sf",
-                   progress_bar = FALSE) %>% 
-  sf::st_transform(four_corners_lcc_proj) %>% 
-  # sf::st_contains(UUSW, sparse = FALSE)
-  dplyr::filter(sf::st_intersects(suppressWarnings(sf::st_centroid(.)),
-                                  UUSW %>% 
-                                    sf::st_transform(four_corners_lcc_proj), 
-                                  sparse = FALSE)) %>%
-  dplyr::select(NAME,STATEFP) %>%
-  dplyr::rename(County = NAME) %>%
-  dplyr::left_join(tigris::states(class = "sf",
-                                  progress_bar = FALSE) %>%
-                     dplyr::select(STATEFP,STUSPS) %>%
-                     dplyr::rename(State = STUSPS) %>%
-                     sf::st_drop_geometry(),
-                   by = "STATEFP") %>%
-  dplyr::select(-STATEFP) %>%
-  tidyr::unite(col = County, County, State, sep = ", ")
+theme_map <- function(base_size = 6.5,
+                      base_family = ""){
+  theme_bw(base_size = base_size,
+           base_family = base_family) %+replace%
+    theme(axis.line = element_blank(),
+          axis.text = element_blank(),
+          axis.ticks = element_blank(),
+          axis.title = element_blank(),
+          panel.background = element_blank(),
+          panel.border = element_blank(),
+          panel.grid = element_blank(),
+          panel.grid.major = element_line(colour = 'transparent'),
+          panel.grid.minor = element_line(colour = 'transparent'),
+          legend.key = element_blank(),
+          legend.position = "top",
+          plot.background = element_blank(),
+          plot.margin = margin(t = 0, r = 0, b = 0, l = 0, unit = "npc")
+    )
+}
+
+ggplot() +
+  geom_sf(data = four_corners_states) +
+  geom_raster(data = robinson2020::swus_hillshade_500m,
+              mapping = aes(x = x,
+                            y = y,
+                            alpha = ID),
+              na.rm = TRUE) +
+  scale_alpha(range = c(0.8, 0),
+              na.value = 0,
+              limits = c(0,255),
+              guide = "none") +
+  geom_sf(data = four_corners_states,
+          fill = NA) +
+  geom_sf(data = uusw_counties %>% 
+            sf::st_union() %>%
+            rmapshaper::ms_filter_islands(),
+          fill = NA,
+          color = "black") +
+  geom_sf(data = uusw_boundary,
+          fill = NA,
+          linetype = 2,
+          color = "white") +
+  geom_sf(data = vep_study_areas,
+          fill = NA,
+          color = "white",
+          size = 1) +
+  geom_sf_text(data = vep_study_areas,
+               aes(label = `Study Area`),
+               size = 3) +
+  theme_map(16)
 
@@
 
-## Datasets
+## Datasets {-}
 
-## Preparing the radiocarbon data
+The radiocarbon data presented here are derived from the [RADIOCARBON DATA DETAILS HERE].
+Data from Arizona, Colorado, and New Mexico in the [DATABASE] are spatially resolved to only the county level; we therefor translated the Utah dates to their counties as well, and then spatially cropped the data to our study area.
 
-```{r}
+```{r prepare-radiocarbon-data}
 # # Read in the radiocarbon database
-# radiocarbon_data <-
+# # radiocarbon_data <-
 #   readr::read_csv(here::here("analysis/data/raw_data/0403_sbox_clean.csv")) %>%
@@
 #                 Site_Number = Site) %>%
-#   sf::st_intersection(UUSW_counties) %>%
+#   sf::st_intersection(uusw_counties) %>%
 #   sf::st_drop_geometry() %>%
 #   dplyr::mutate(Type = "Radiocarbon",
-#                 SiteID = SiteID %>% stringr::str_c("Radiocarbon - ", .)) %T>%
+#                 SiteID = SiteID %>% stringr::str_c("Radiocarbon - ", .)) %>%
+#   dplyr::select(Type,
+#                 SiteID,
+#                 Site_Number,
+#                 County,
+#                 Lab_Number,
+#                 Date_BP,
+#                 SD) %T>%
 #   # Write the output to a new file (csv)
@@
 
-These data contain sensitive location information about cultural sites, and are restricted on tDAR.
-Accordingly, we pre-processed the data using the script below, and provide data as supplemental information without location information represented.
-```{r}
+We draw on the tree-ring date database presented in @Bocinsky2016_SA and archived in the Digital Archaeological Record (https://doi.org/10.6067/XCV82J6D7B).
+The tree-ring data we use here were compiled by Tim Kohler and Rebecca Higgins from many contributers, and the vast majority of dates were determined by researchers in the Laboratory of Tree Ring Research at the University of Arizona.
+The original dataset contained 32,863 cutting, near-cutting, and non-cutting dates from across the US Southwest.
+Here, we only use the dates that are most likely to be accurate --- cutting dates ('B', 'G', 'L', 'c', or 'r' dates, with or without a '+') for which we know the year the tree died, and near-cutting dates ('v' or 'v+' dates) which are likely within 0--3 years of the true date of the outermost ring.
+The @Bocinsky2016_SA data contain site locations; for this analysis, we first spatially cropped the database to our study area, and then translated those locations to the county level.
+We converted the original date determinations to BP to match radiocarbon conventions.
+
+```{r prepare-treering-data}
 # # Read in the Bocinsky et al. 2016 tree ring data
@@
 #   dplyr::rename(SiteID = Seq_Number) %>%
-#   sf::st_intersection(UUSW_counties) %>%
+#   sf::st_intersection(uusw_counties) %>%
 #   sf::st_drop_geometry() %>%
@@
 #   dplyr::filter(Type == "Dendro - Cutting or Near-cutting") %>%
-#   dplyr::mutate(Type = "Tree-ring") %T>%
+#   dplyr::mutate(Type = "Tree-ring")  %>%
+#   dplyr::select(Type,
+#                 SiteID,
+#                 Site_Number,
+#                 County,
+#                 Lab_Number,
+#                 Date_BP,
+#                 SD) %T>%
 #   # Write the output to a new file (csv)
@@
 
-We draw on the tree-ring date database presented in @Bocinsky2016_SA and archived on the Digital Archaeological Record (https://doi.org/10.6067/XCV82J6D7B). Here, we use only the subset of those data
+Archaeological site locations are protected by US federal and state statutes.
+In order to protect these sensitive data, we pre-processed the both the radiocarbon and tree-ring data; data included in this compendium do not include site locations, and researchers will need to gain permission from the State Historic Preservation Office (SHPO) of each state in order to access the site location data.
 
-# Combining datasets
 ```{r}
@@
                                                   NA),
-                            `Study Area` = ifelse(County %in% c("Rio Arriba, NM",
+                            `Study Area` = ifelse(County %in% c("Los Alamos, NM",
+                                                                "Rio Arriba, NM",
                                                                 "Sandoval, NM",
@@
 
+```
+
+Finally, due to much of the radiocarbon data only being resolved to the county level, we had to approximate inclusion in the VEP CMV and NRG study areas.
+The CMV study area conforms well to Montezuma county, Colorado; the NRG overlaps with Los Alamos, Rio Arriba, Sandoval, Santa Fe, and Taos counties in New Mexico.
+Our final dataset includes `r nrow(all_data$Radiocarbon) %>% format(big.mark = ',')` radiocarbon dates from `r max(all_data$Radiocarbon$Date_BP)`--`r min(all_data$Radiocarbon$Date_BP)` BP (uncalibrated; `r -1 * (1950 - max(all_data$Radiocarbon$Date_BP)) - 1` BC--`r 1950 - min(all_data$Radiocarbon$Date_BP)` AD) and `r nrow(all_data[['Tree-ring']]) %>% format(big.mark = ',')` tree-ring dates from `r max(all_data[['Tree-ring']]$Date_BP)` -- `r min(all_data[['Tree-ring']]$Date_BP)` BP (`r 1950 - max(all_data[['Tree-ring']]$Date_BP)`--`r 1950 - min(all_data[['Tree-ring']]$Date_BP)` AD).
+Table \@ref(tab:data-summary) presents the counts of each date type in the CMV and NRG study areas.
+
+```{r data-summary}
 all_data %>%
   dplyr::bind_rows() %>%
+  dplyr::bind_rows(all_data %>%
+                     dplyr::bind_rows() %>%
+                     dplyr::mutate(`Study Area` = "UUSW Total")) %>%
+  dplyr::filter(!is.na(`Study Area`)) %>%
+  dplyr::mutate(`Study Area` = factor(`Study Area`, levels = c("UUSW Total","CMV","NRG"))) %>%
   dplyr::group_by(Type,`Study Area`) %>%
   dplyr::summarise(Count = dplyr::n()) %>%
-  tidyr::spread(Type, Count)
+  tidyr::spread(Type, Count) %>%
+  knitr::kable(caption = "Counts of radiocarbon and tree-ring dates in this study.",
+               booktabs = TRUE,
+               format.args = list(big.mark = ','))
+```
+
+Date densities across the study area were very uneven, with high counts of radiocarbon dates deriving from Navajo County, Arizona, and almost a third of the tree-ring dates in our study being from Montezuma County, Colorado.
+Below, we attempt to overcome some of these sampling issues by binning dates by site and occupation phase.
+Still, these disparities likely demonstrate more fundemental differences in research design across archaeological projects.
+Figure \@ref(fig:date-densities) presents the number of radiocarbon and tree-ring dates by county within our study area.
+Figure \@ref(fig:site-densities) presents the number of unique sites with radiocarbon and tree-ring dates by county within our study area.
+
+```{r date-densities, fig.height=9, fig.width=6.5, fig.cap='The number of radiocarbon and tree-ring dates by county in this study.', fig.show = 'hold', out.width = '50%'}
+county_counts <-
+  all_data %>%
+  dplyr::bind_rows() %>%
+  dplyr::group_by(Type, County) %>%
+  dplyr::summarise(Dates = dplyr::n()) %>%
+  dplyr::left_join(uusw_counties,
+                   by = "County") %>%
+  sf::st_as_sf()
+
+theme_map <- function(base_size = 6.5,
+                      base_family = ""){
+  theme_bw(base_size = base_size,
+           base_family = base_family) %+replace%
+    theme(axis.line = element_blank(),
+          axis.text = element_blank(),
+          axis.ticks = element_blank(),
+          axis.title = element_blank(),
+          panel.background = element_blank(),
+          panel.border = element_blank(),
+          panel.grid = element_blank(),
+          panel.grid.major = element_line(colour = 'transparent'),
+          panel.grid.minor = element_line(colour = 'transparent'),
+          legend.key = element_blank(),
+          legend.position = "bottom",
+          legend.background = element_blank(),
+          legend.key.width = unit(0.15,"npc"),
+          legend.text = element_text(size = rel(1)),
+          plot.background = element_blank(),
+          plot.margin = margin(t = 0, r = 0, b = 0, l = 0, unit = "npc")
+    )
+}
+
+county_counts %>%
+  dplyr::filter(Type == "Radiocarbon") %>%
+  ggplot() +
+  geom_sf(data = four_corners_states,
+          fill = NA) +
+  geom_sf(aes(fill = Dates),
+          color = "black") +
+  scale_fill_distiller(name = "Number of Radiocarbon Dates",
+                       palette = "Greys",
+                       direction = 1,
+                       limits = c(0,500),
+                       minor_breaks = seq(50,450,100),
+                       breaks = seq(0,500,100),
+                       expand = expand_scale(mult = 0, add = 0),
+                       guide = guide_colorbar(title.position = "top")) +
+  geom_sf(data = uusw_counties,
+          fill = NA,
+          color = "black") +
+  theme_map(16)
+
+county_counts %>%
+  dplyr::filter(Type == "Tree-ring") %>%
+  ggplot() +
+  geom_sf(data = four_corners_states,
+          fill = NA) +
+  geom_sf(aes(fill = Dates),
+          color = "black") +
+  scale_fill_distiller(name = "Number of Tree-ring Dates",
+                       palette = "Greys",
+                       direction = 1,
+                       limits = c(0,5000),
+                       minor_breaks = seq(500,4500,1000),
+                       breaks = seq(0,5000,1000),
+                       expand = expand_scale(mult = 0, add = 0),
+                       guide = guide_colorbar(title.position = "top")) +
+  geom_sf(data = uusw_counties,
+          fill = NA,
+          color = "black") +
+  theme_map(16)
 
@@
 
-# Preparing the radiocarbon SPDs
+```{r site-densities, fig.height=9, fig.width=6.5, fig.cap='The number of sites with radiocarbon and tree-ring dates by county in this study.', fig.show = 'hold', out.width = '50%'}
+county_counts <-
+  all_data %>%
+  dplyr::bind_rows() %>%
+  dplyr::select(Type, SiteID, County) %>%
+  dplyr::distinct() %>%
+  dplyr::group_by(Type, County) %>%
+  dplyr::summarise(Dates = dplyr::n()) %>%
+  dplyr::left_join(uusw_counties,
+                   by = "County") %>%
+  sf::st_as_sf()
 
-```{r}
+theme_map <- function(base_size = 6.5,
+                      base_family = ""){
+  theme_bw(base_size = base_size,
+           base_family = base_family) %+replace%
+    theme(axis.line = element_blank(),
+          axis.text = element_blank(),
+          axis.ticks = element_blank(),
+          axis.title = element_blank(),
+          panel.background = element_blank(),
+          panel.border = element_blank(),
+          panel.grid = element_blank(),
+          panel.grid.major = element_line(colour = 'transparent'),
+          panel.grid.minor = element_line(colour = 'transparent'),
+          legend.key = element_blank(),
+          # legend.position = "top",
+          # legend.justification = c(0,0),
+          legend.position = "bottom",
+          # legend.position = c(0.5,1),
+          legend.background = element_blank(),
+          legend.key.width = unit(0.15,"npc"),
+          # legend.key.height = unit(0.15,"in"),
+          legend.text = element_text(size = rel(1)),
+          plot.background = element_blank(),
+          plot.margin = margin(t = 0, r = 0, b = 0, l = 0, unit = "npc")
+    )
+}
+
+county_counts %>%
+  dplyr::filter(Type == "Radiocarbon") %>%
+  ggplot() +
+  geom_sf(data = four_corners_states,
+          fill = NA) +
+  geom_sf(aes(fill = Dates),
+          color = "black") +
+  scale_fill_distiller(name = "Number of Sites with Radiocarbon Dates",
+                       palette = "Greys",
+                       direction = 1,
+                       limits = c(0,80),
+                       minor_breaks = seq(10,70,20),
+                       breaks = seq(0,80,20),
+                       expand = expand_scale(mult = 0, add = 0),
+                       guide = guide_colorbar(title.position = "top")) +
+  geom_sf(data = uusw_counties,
+          fill = NA,
+          color = "black") +
+  theme_map(16)
+
+county_counts %>%
+  dplyr::filter(Type == "Tree-ring") %>%
+  ggplot() +
+  geom_sf(data = four_corners_states,
+          fill = NA) +
+  geom_sf(aes(fill = Dates),
+          color = "black") +
+  scale_fill_distiller(name = "Number of Sites with Tree-ring Dates",
+                       palette = "Greys",
+                       direction = 1,
+                       limits = c(0,250),
+                       minor_breaks = seq(25,225,50),
+                       breaks = seq(0,250,50),
+                       expand = expand_scale(mult = 0, add = 0),
+                       guide = guide_colorbar(title.position = "top")) +
+  geom_sf(data = uusw_counties,
+          fill = NA,
+          color = "black") +
+  theme_map(16)
+
+```
+
+## Preparing the SPDs {-}
+
+To prepare the radiocarbon dataset, we used functions provided by the `rcarbon` package for *R* [@rcarbon].
+We calibrated the radiocarbon dates using the methods of @Bronk2008, using the 'intcal13' calibration curve.
+To control for inconsistent sample size across the archaeological sites represented in the database, we aggregated calibrated radiocarbon dates known to be from the same phase of same archaeological site using the `rcarbon::binprep` function with an `h`-value of 100.
+This effectively clusters dates from the same site that are within 100 years of each other, and splits site occupations into distinct phases separated by 100 years or more.
+We constructed summed probability distributions (SPDs) from the calibrated dates using the `rcarbon::spd` function.
+We averaged probability distributions of binned dates, and then constructed SPDs from the site-phase distributions between 1949--0 BP (1--1950 AD).
+We smoothed the SPD using a 21-year moving average, and normalised the total probability mass of the smoothed SPD to sum to unity.
+
+```{r radiocarbon-spds}
 if(!file.exists(here::here("analysis/data/derived_data/radiocarbon_spd.rds")))
@@
                                                        calMatrix = TRUE,
-                                                       normalised = TRUE,
                                                        verbose = FALSE)),
@@
                                         bins = Bins,
+                                        datenormalised = TRUE,
+                                        spdnormalised = TRUE,
                                         timeRange = c(1949,0),
@@
 
-To prepare the radiocarbon dataset, we used functions provided by the *rcarbon* package for R [@rcarbon].
-We calibrated the radiocarbon dates using the methods of @Bronk2008 and implemented in @rcarbon.
-To control for inconsistent sample size across the archaeological sites represented in the database, we aggregated radiocarbon dates known to be from the same phase of same archaeological site using the `rcarbon::binprep` function with an `h`-value of 100.
-We averaged probability densities of binned dates, and then constructed summed probability density (SPD) distributions across bins.
+We treated the tree-ring dates much in the same way as the radiocarbon dates.
+We modeled faux-calibrated tree-ring date probabilities as discrete degenerate distributions ($\Pr(X=k_0)=1$) with $k_0$ equal to the tree-ring date --- in other words, a probability of $1$ at the date value, and $0$ elsewhere.
+Thinking of tree ring dates this way allows us to use the same binning and SPD algorithms as for radiocarbon dates.^[One might consider alternative probability distributions for non-cutting or near-cutting dates.
+The probability mass function in such a case would be zero for years prior to the date of the outer ring of the sample, and non-zero for subsequent years, presumably decaying quickly.
+For instance, we could imagine a study that derives empirical probability distributions for non-cutting dates based on cutting-dates from the same contexts.] Here, we did just that, binning the dates into site-phases using the `rcarbon::binprep` function, and constructing SPDs using `rcarbon::spd`, both with the same parameter selections as used for the radiocarbon samples.
+As above, we smoothed normalised the SPD.
 
-# 'Calibrating' tree-ring dates, and preparing tree-ring SPDs
-
-```{r}
+```{r treering-spds}
 calibrate_tree_ring_dates <- function(x){
@@
 
-## Comparing radiocarbon and tree-ring SPDs
+## Comparing radiocarbon and tree-ring SPDs {-}
 
-```{r}
+```{r c14-tr-pop, fig.height=6.5, fig.width=6.5, fig.cap='Radiocarbon and tree-ring density distributions in the UUSW, CMV, and NRG regions, and VEP population estimates for the CMV and NRG.
+Y-axis labels are omitted for simplicity; the maximum population in the lower panel is ~30,000 people, and the minimum is 0.
+Dates are given in AD.', fig.show = 'hold', out.width = '75%'}
 all_density <- 
@@
   dplyr::rename(Value = Density)
-```
 
-```{r}
 vep_demography <- 
@@
   dplyr::rename(Value = Population)
-```
 
-```{r c14_tr_pop}
 all_density %>%
@@
                                          "Population"),
-                              ordered = TRUE)#,
-                # `Study Area` = ifelse(`Study Area` == "CMV", "Central Mesa Verde", "Northern Rio Grande")
+                              ordered = TRUE),
+                `Study Area` = factor(`Study Area`,
+                                      levels = c( `Upland US Southwest` = "UUSW",
+                                                  `Central Mesa Verde` = "CMV",
+                                                 `Northern Rio Grande` = "NRG"))
   ) %>%
-  ggplot2::ggplot(ggplot2::aes(x = `Year AD`)) +
-  ggplot2::geom_line(ggplot2::aes(y = Value,
-                                  color = `Study Area`),
-                     size = 1.25) +
+  ggplot(aes(x = `Year AD`)) +
+  geom_line(aes(y = Value,
+                color = `Study Area`,
+                size = `Study Area`)) +
+  scale_color_manual(values = c("UUSW" = "black",
+                                "CMV" = "gray50",
+                                "NRG" = "gray80")) +
+  scale_size_manual(values = c("UUSW" = 1.25,
+                               "CMV" = 0.75,
+                               "NRG" = 0.75)) +
   # coord_cartesian(xlim = c(600,1600)) +
-  ggplot2::xlab("Year AD") +
-  ggplot2::theme_minimal(24) +
-  ggplot2::theme(legend.title = ggplot2::element_blank(),
+  xlab("Year AD") +
+  scale_x_continuous(breaks = c(1,seq(200,2000,200))) + 
+  theme_minimal(18) +
+  theme(legend.title = element_blank(),
         legend.justification = c(0, 1), 
         legend.position = "bottom",
-                 axis.title.y = ggplot2::element_blank(),
-                 axis.text.y = ggplot2::element_blank(),
-                 panel.grid.major.y = ggplot2::element_blank(),
-                 panel.grid.minor.y = ggplot2::element_blank(),
-                 plot.margin = ggplot2::unit(c(0.25,0.5,0.25,0.25),"in")
+        axis.title.x = element_text(size = 12),
+        axis.text.x = element_text(size = 10),
+        axis.title.y = element_blank(),
+        axis.text.y = element_blank(),
+        panel.grid.major.y = element_blank(),
+        panel.grid.minor.y = element_blank(),
+        plot.margin = unit(c(0.25,0.5,0.25,0.25),"in")
   ) +
-  ggplot2::facet_wrap("Type", 
+  facet_wrap("Type", 
              nrow = 3,
@@
 
-<!-- ```{r} -->
-<!-- four_corners_lcc_proj <- "+proj=lcc +lat_1=37 +lon_0=-109.045225" -->
-
-<!-- # Define the Bocinsky et al. 2016 study area -->
-<!-- SWUS_bbox <- sf::st_bbox(c(xmin = -113, -->
-<!--                            xmax = -105, -->
-<!--                            ymin = 32, -->
-<!--                            ymax = 38),  -->
-<!--                          crs = 4326) %>% -->
-<!--   sf::st_as_sfc() %>% -->
-<!--   sf::st_set_crs(NA) %>% -->
-<!--   sf::st_segmentize(dfMaxLength = 0.01) %>% -->
-<!--   sf::st_set_crs(4326) %>% -->
-<!--   sf::st_transform(four_corners_lcc_proj) -->
-
-<!-- four_corners_states <- "./data-raw/statesp010g/statesp010g.gdb/" %>% -->
-<!--   sf::st_read() %>% -->
-<!--   dplyr::filter(NAME %in%  -->
-<!--                   c("Arizona", -->
-<!--                     "Colorado", -->
-<!--                     "New Mexico", -->
-<!--                     "Utah")) %>% -->
-<!--   sf::st_transform(four_corners_lcc_proj) -->
-
-<!-- swus_hillshade_500m <- readr::read_rds("./data-derived/swus_hillshade_500m.rds") %>% -->
-<!--   raster::crop(four_corners_states) %>% -->
-<!--   raster::mask(four_corners_states) -->
-
-<!-- dendro_data <- sf::read_sf("./data-derived/dendro_data.geojson") %>% -->
-<!--   sf::st_transform(four_corners_lcc_proj) %>% -->
-<!--   dplyr::filter(Date_BP >= 550, -->
-<!--                 Date_BP <= 1450) %>% -->
-<!--   dplyr::group_by(Site_Number) %>% -->
-<!--   dplyr::summarise(Date_BP = mean(Date_BP, na.rm = T), -->
-<!--                    geometry = mean(geometry)) %>% -->
-<!--   dplyr::mutate(`Date Type` = "Dendro") %>% -->
-<!--   dplyr::select(-Site_Number) %>% -->
-<!--   sf::st_centroid() -->
-
-<!-- radiocarbon_data <- sf::read_sf("./data-derived/radiocarbon_data.geojson") %>% -->
-<!--   sf::st_transform(four_corners_lcc_proj) %>% -->
-<!--   dplyr::filter(Date_BP >= 550, -->
-<!--                 Date_BP <= 1450) %>% -->
-<!--   dplyr::group_by(Site_Number) %>% -->
-<!--   dplyr::summarise(Date_BP = mean(Date_BP, na.rm = T)) %>% -->
-<!--   sf::st_jitter(20000) %>% -->
-<!--   dplyr::mutate(`Date Type` = "C14") %>% -->
-<!--   dplyr::select(-Site_Number) %>% -->
-<!--   sf::st_centroid() -->
-
-<!-- dates <- rbind(dendro_data, -->
-<!--                radiocarbon_data) -->
-
-<!-- theme_map <- function(base_size = 6.5,  -->
-<!--                       base_family = ""){ -->
-<!--   ggplot2::theme_bw(base_size = base_size,  -->
-<!--                     base_family = base_family) %+replace% -->
-<!--     ggplot2::theme(axis.line = ggplot2::element_blank(), -->
-<!--                    axis.text = ggplot2::element_blank(), -->
-<!--                    axis.ticks = ggplot2::element_blank(), -->
-<!--                    axis.title = ggplot2::element_blank(), -->
-<!--                    panel.background = ggplot2::element_blank(), -->
-<!--                    panel.border = ggplot2::element_blank(), -->
-<!--                    panel.grid = ggplot2::element_blank(), -->
-<!--                    panel.grid.major = ggplot2::element_line(colour = 'transparent'), -->
-<!--                    panel.grid.minor = ggplot2::element_line(colour = 'transparent'), -->
-<!--                    legend.key = element_blank(), -->
-<!--                    legend.position = "top", -->
-<!--                    # legend.justification = c(0,0), -->
-<!--                    # legend.position = "top", -->
-<!--                    # legend.position = c(0.5,1), -->
-<!--                    # legend.background = ggplot2::element_blank(), -->
-<!--                    # legend.key.width = unit(0.15,"in"), -->
-<!--                    # legend.key.height = unit(0.15,"in"), -->
-<!--                    # legend.text = element_text(size = ggplot2::rel(1)), -->
-<!--                    plot.background = ggplot2::element_blank(), -->
-<!--                    plot.margin = ggplot2::margin(t = 0, r = 0, b = 0, l = 0, unit = "npc") -->
-<!--     ) -->
-<!-- } -->
-
-<!-- ggplot() +  -->
-<!--   ggplot2::geom_sf(data = four_corners_states) + -->
-<!--   ggplot2::geom_raster(data = Robinson2020::swus_hillshade_500m, -->
-<!--                        mapping = ggplot2::aes(x = x, -->
-<!--                                               y = y, -->
-<!--                                               alpha = ID), -->
-<!--                        na.rm = TRUE) + -->
-<!--   ggplot2::scale_alpha(range = c(0.8, 0), -->
-<!--                        na.value = 0, -->
-<!--                        limits = c(0,255), -->
-<!--                        guide = "none") + -->
-<!--   ggplot2::geom_sf(data = four_corners_states, -->
-<!--                    fill = NA) + -->
-<!--   ggplot2::geom_sf(data = SWUS_bbox, -->
-<!--                    fill = NA) + -->
-<!--   ggplot2::geom_sf(data = dates, -->
-<!--                    mapping = aes(shape = `Date Type`, -->
-<!--                                  color = Date_BP), -->
-<!--                    show.legend = "point") + -->
-<!--   ggplot2::scale_shape( -->
-<!--     name = 'Date Type', -->
-<!--     solid = FALSE, -->
-<!--     guide = ggplot2::guide_legend(position = "top", -->
-<!--                                   direction = "horizontal", -->
-<!--                                   title.position = "top", -->
-<!--                                   label.position = "bottom") -->
-<!--   ) + -->
-<!--   viridis::scale_color_viridis( -->
-<!--     name = 'Year BP', -->
-<!--     limits = c(450,1450), -->
-<!--     breaks = c(450,1450), -->
-<!--     direction = -1, -->
-<!--     guide = ggplot2::guide_colorbar(position = "top", -->
-<!--                                     direction = "horizontal", -->
-<!--                                     title.position = "top", -->
-<!--                                     label.position = "bottom", -->
-<!--                                     reverse = TRUE)) + -->
-<!--   theme_map(16) -->
-
-<!-- ggsave('./swus_sample_map.pdf', -->
-<!--        height = 8, -->
-<!--        width = 6) -->
-<!-- ``` -->
-
 <!-- The following line inserts a page break  -->
@@
 
-# References 
+## References {-}
 <!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->
@@
 
-### Colophon
+## Colophon {-}
 

### 2020-01-19 data-raw/DATASET.R

@@ -1,13 +1,58 @@
 library(magrittr)
 library(tidycensus)
 
+# A nice projection for the Four Corners states
+nm <- tigris::states(class = "sf",
+                     progress_bar = FALSE) %>%
+  dplyr::filter(NAME == "New Mexico") %>%
+  sf::st_transform(4326) %>%
+  sf::st_bbox()
+
+co <- tigris::states(class = "sf",
+                     progress_bar = FALSE) %>%
+  dplyr::filter(NAME == "Colorado") %>%
+  sf::st_transform(4326) %>%
+  sf::st_bbox()
+
+four_corners_lcc_proj <- paste0("+proj=lcc",
+                                " +lat_1=",co[['ymax']],
+                                " +lat_2=",nm[['ymax']] - (co[['ymax']] - co[['ymin']]),
+                                " +lat_0=",nm[['ymax']],
+                                " +lon_0=",nm[['xmin']],
+                                " +x_0=",nm[['xmin']],
+                                " +x_y=",nm[['xmin']]
+                                )
+usethis::use_data(four_corners_lcc_proj, overwrite = TRUE)
+
+four_corners_states <-
+  tigris::states(class = "sf",
+                 progress_bar = FALSE) %>%
+  dplyr::filter(NAME %in% c("Arizona", "Colorado",
+                            "New Mexico", "Utah")) %>%
+  smoothr::densify(max_distance = 0.01) %>%
+  sf::st_transform(four_corners_lcc_proj) %>%
+  dplyr::select(STATEFP,STUSPS) %>%
+  dplyr::rename(State = STUSPS)
+
+usethis::use_data(four_corners_states, overwrite = TRUE)
+
 ## Get the VEP II study area definitions from the `villager` package.
 ## https://github.com/village-ecodynamics/villager
 vepii_cmv_boundary <- villager::vepii_cmv_boundary
 vepii_nrg_boundary <- villager::vepii_nrg_boundary
 
-usethis::use_data(vepii_cmv_boundary)
-usethis::use_data(vepii_nrg_boundary)
+# VEP study areas
+vep_study_areas <-
+  list(CMV = vepii_cmv_boundary,
+       NRG = vepii_nrg_boundary) %>%
+  purrr::map(smoothr::densify,
+             max_distance = 100) %>%
+  purrr::map(sf::st_transform,
+             crs = four_corners_lcc_proj) %>%
[…]

### 2020-01-19 README.Rmd
updated readme and trying to get binder going

@@
 
-> Robinson, Erick, R. Kyle Bocinsky, Jacob Freeman, Darcy Bird, Robert L. Kelly, (2020). _`r Title`_. Under review.
+> Robinson, Erick [![](https://orcid.org/sites/default/files/images/orcid_16x16.png)](https://orcid.org/0000-0002-0789-3724), R.
+Kyle Bocinsky [![](https://orcid.org/sites/default/files/images/orcid_16x16.png)](https://orcid.org/0000-0003-1862-3428), Jacob Freeman, Darcy Bird [![](https://orcid.org/sites/default/files/images/orcid_16x16.png)](https://orcid.org/0000-0003-3466-6284), Robert L.
+Kelly, (2020).
+_`r Title`_.
+Under review.
 
@@
 
-### How to download or install
+## Contents
 
-You can download the compendium as a zip from from this URL: </archive/master.zip>
+The **analysis** directory contains:
 
-Or you can install this compendium as an R package, Robinson2020, from GitHub with:
+- [:file_folder: paper](/analysis/paper): R Markdown source document for the manuscript.
+Includes code to reproduce the figures and tables generated by the analysis.
+It also has a rendered version, `paper.docx`, suitable for reading (the code is replaced by figures and tables in this file)
+- [:file_folder: data](/analysis/data): Data used in the analysis.  
+- [:file_folder: figures](/analysis/figures): Plots and other illustrations
 
-```{r gh-installation, eval = FALSE}
-# install.packages("devtools")
-remotes::install_github("bocinsky/robinson2020")
-```
+## How to run in your broswer or download and run locally
+
+The simplest way to explore the text, code and data is to click on [binder](https://mybinder.org/v2/gh/bocinsky/robinson2020/master?urlpath=rstudio) to open an instance of RStudio in your browser, which will have the compendium files ready to work with.
+Binder uses [rocker-project.org Docker] images to ensure a consistent and reproducible computational environment.
+These Docker images can also be used locally.
+
+You can download the compendium as a zip from from this URL: [master.zip](/archive/master.zip).
+After unzipping, open the `Proj` file in RStudio, and run `devtools::install()` to ensure you have the packages this analysis depends on (also listed in the [DESCRIPTION](/DESCRIPTION) file).
+Then open `paper.Rmd` and knit to produce the `paper.docx`.
 
@@
 By participating in this project you agree to abide by its terms.
+
+### Acknowledgements
+
+This compendium was created using the `rrtools` package by Ben Marwick: []()

### 2020-01-19 README.Rmd
updated readme and trying to get binder going

@@
 Includes code to reproduce the figures and tables generated by the analysis.
-It also has a rendered version, `paper.docx`, suitable for reading (the code is replaced by figures and tables in this file)
+It also has a rendered version, `paper.html`, suitable for reading (the code is replaced by figures and tables in this file).
 - [:file_folder: data](/analysis/data): Data used in the analysis.  

### 2020-01-19 analysis/paper/paper.Rmd
updated readme and trying to get binder going

@@
   dplyr::mutate(`Year AD` = list((Start+1):End)) %>%
-  dplyr::select(`Study Area`,`Year AD`,Population) %>%
-  tidyr::unnest(cols = c(`Year AD`)) %>%
+  dplyr::select(`Study Area`,
+                `Year AD`,
+                Population) %>%
+  tidyr::unnest(cols = c("Year AD")) %>%
   dplyr::mutate(Type = "Population",

### 2020-01-19 analysis/paper/paper.Rmd
still trying to get binder working

@@
       - --lua-filter=../templates/pagebreak.lua
-header-includes:
-  - \usepackage{subfig}
+    bookdown::word_document2:
+      fig_caption: yes
+      pandoc_args:
+      - --lua-filter=../templates/scholarly-metadata.lua
+      - --lua-filter=../templates/author-info-blocks.lua
+      - --lua-filter=../templates/pagebreak.lua
 bibliography: references.bib

### 2020-01-19 analysis/paper/paper.Rmd
still trying to get binder working

--- 2020-01-19 analysis/paper/paper.Rmd
@@
       - --lua-filter=../templates/pagebreak.lua
-    bookdown::pdf_document2:
-      fig_caption: yes
-      pandoc_args:
-      - --lua-filter=../templates/scholarly-metadata.lua
-      - --lua-filter=../templates/author-info-blocks.lua
-      - --lua-filter=../templates/pagebreak.lua
-header-includes:
-  - \usepackage{subfig}
 bibliography: references.bib

### 2020-01-19 README.Rmd
updated readme and trying to get binder going
still trying to get binder working + docker

@@
 # robinson2020
-
+[![Travis-CI Build Status](https://travis-ci.org/bocinsky/robinson2020.svg?branch=master)](https://travis-ci.org/bocinsky/robinson2020)
 [![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh/bocinsky/robinson2020/master?urlpath=rstudio)
@@
 
-**Code :** See the [LICENSE](LICENSE.md) file
+**Code :** [GNU GPLv3](LICENSE.md)
 
@@
 
-This compendium was created using the `rrtools` package by Ben Marwick: []()
+This compendium was created using the [`rrtools` package by Ben Marwick](https://github.com/benmarwick/rrtools), which is ✨pure magic ✨for doing reproducible research.

### 2020-01-19 analysis/paper/paper.Rmd
still trying to get binder working + docker

@@
                 Population) %>%
-  tidyr::unnest(cols = c("Year AD")) %>%
+  tidyr::unnest(`Year AD`) %>%
   dplyr::mutate(Type = "Population",

### 2020-01-19 analysis/paper/paper.Rmd
reducing memory impact of the analysis

@@
 ```{r radiocarbon-spds}
+# Set the time range for the reconstructions
+timerange <- c(1949,0)
+
 if(!file.exists(here::here("analysis/data/derived_data/radiocarbon_spd.rds")))
@@
                                                        errors = SD,
+                                                       timerange = timerange,
                                                        calMatrix = TRUE,
@@
       ResErrors = 0,
-      StartBP = 50000,
+      StartBP = timerange[1],
       EndBP = 0,
@@
                          values_from = value) %>%
-      dplyr::bind_rows(tibble::tibble(year_bp = 50000:0)) %>%
+      dplyr::bind_rows(tibble::tibble(year_bp = timerange[1]:timerange[2])) %>%
       dplyr::filter(!duplicated(year_bp)) %>%
@@
                                         spdnormalised = TRUE,
-                                        timeRange = c(1949,0),
+                                        timeRange = timerange,
                                         runm = 21,

### 2020-01-19 analysis/paper/paper.Rmd
uploaded more memory efficient version

@@
     bookdown::html_document2:
+      code_folding: hide
+      fig_caption: yes
+      pandoc_args:
+      - --lua-filter=../templates/scholarly-metadata.lua
+      - --lua-filter=../templates/author-info-blocks.lua
+      - --lua-filter=../templates/pagebreak.lua
+    bookdown::word_document2:
+      reference_docx: "../templates/template.docx"
       code_folding: hide
@@
 # # radiocarbon_data <-
-#   readr::read_csv(here::here("analysis/data/raw_data/0403_sbox_clean.csv")) %>%
+# readr::read_csv(
+#   here::here("analysis/data/raw_data/0403_sbox_clean.csv")
+# ) %>%
 #   # Convert to spatial object
@@
 #   # Write the output to a new file (csv)
-#   readr::write_csv(here::here("analysis/data/derived_data/radiocarbon_data.csv"))
+#   readr::write_csv(
+#     here::here("analysis/data/derived_data/radiocarbon_data.csv")
+#   )
 
 radiocarbon_data <-
-  readr::read_csv(here::here("analysis/data/derived_data/radiocarbon_data.csv"))
+  readr::read_csv(
+    here::here("analysis/data/derived_data/radiocarbon_data.csv")
+  )
 
@@
 We draw on the tree-ring date database presented in @Bocinsky2016_SA and archived in the Digital Archaeological Record (https://doi.org/10.6067/XCV82J6D7B).
-The tree-ring data we use here were compiled by Tim Kohler and Rebecca Higgins from many contributers, and the vast majority of dates were determined by researchers in the Laboratory of Tree Ring Research at the University of Arizona.
+The tree-ring data we use here were compiled by Tim Kohler and Rebecca Higgins from many contributors, and the vast majority of dates were determined by researchers in the Laboratory of Tree Ring Research at the University of Arizona.
 The original dataset contained 32,863 cutting, near-cutting, and non-cutting dates from across the US Southwest.
@@
 # dendro_data <-
-#   readr::read_csv(here::here("analysis/data/raw_data/BOCINSKY_ET_AL_2015_TREE_RINGS.csv")) %>%
+#   readr::read_csv(
+#     here::here("analysis/data/raw_data/BOCINSKY_ET_AL_2015_TREE_RINGS.csv")
+#   ) %>%
 #   # Convert to spatial object
@@
 #   # Write the output to a new file (csv)
-#   readr::write_csv(here::here("analysis/data/derived_data/dendro_data.csv"))
+#   readr::write_csv(
+#     here::here("analysis/data/derived_data/dendro_data.csv")
+#   )
 
 dendro_data <- 
-  readr::read_csv(here::here("analysis/data/derived_data/dendro_data.csv"))
+  readr::read_csv(
+    here::here("analysis/data/derived_data/dendro_data.csv")
+  )
 
@@
   dplyr::filter(!is.na(`Study Area`)) %>%
-  dplyr::mutate(`Study Area` = factor(`Study Area`, levels = c("UUSW Total","CMV","NRG"))) %>%
+  dplyr::mutate(`Study Area` = factor(`Study Area`, 
+                                      levels = c("UUSW Total",
+                                                 "CMV",
+                                                 "NRG"))) %>%
   dplyr::group_by(Type,`Study Area`) %>%
@@
 
-if(!file.exists(here::here("analysis/data/derived_data/radiocarbon_spd.rds")))
+if(!file.exists(
+  here::here("analysis/data/derived_data/radiocarbon_spd.rds")
+))
   all_data$Radiocarbon %>%
@@
   ) %>%
-  readr::write_rds(here::here("analysis/data/derived_data/radiocarbon_spd.rds"),
+  readr::write_rds(
+    here::here("analysis/data/derived_data/radiocarbon_spd.rds"),
     compress = "gz")
 
-radiocarbon_spd <- readr::read_rds(here::here("analysis/data/derived_data/radiocarbon_spd.rds"))
+radiocarbon_spd <- readr::read_rds(
+  here::here("analysis/data/derived_data/radiocarbon_spd.rds")
+)
 
@@
 
-if(!file.exists(here::here("analysis/data/derived_data/dendro_spd.rds")))
+if(!file.exists(
+  here::here("analysis/data/derived_data/dendro_spd.rds")
+))
   all_data$`Tree-ring` %>%
@@
   ) %>%
-  readr::write_rds(here::here("analysis/data/derived_data/dendro_spd.rds"),
+  readr::write_rds(
+    here::here("analysis/data/derived_data/dendro_spd.rds"),
     compress = "gz")
 
-dendro_spd <- readr::read_rds(here::here("analysis/data/derived_data/dendro_spd.rds"))
+dendro_spd <- readr::read_rds(
+  here::here("analysis/data/derived_data/dendro_spd.rds")
+)
 
@@
 # what commit is this file at? 
-if ("git2r" %in% installed.packages() & git2r::in_repository(path = ".")) git2r::repository(here::here())  
+if ("git2r" %in% installed.packages() & 
+    git2r::in_repository(path = ".")) 
+  git2r::repository(here::here())  
 ```

### 2020-01-19 README.Rmd
uploaded more memory efficient version

@@
 
-This compendium was created using the [`rrtools` package by Ben Marwick](https://github.com/benmarwick/rrtools), which is ✨pure magic ✨for doing reproducible research.
+This compendium was created using the [`rrtools` package by Ben Marwick](https://github.com/benmarwick/rrtools), which is ✨ pure magic ✨ for doing reproducible research.

### 2020-01-19 analysis/paper/paper.Rmd
reduced memory overhead

@@
 ```{r prepare-radiocarbon-data}
-# # Read in the radiocarbon database
-# # radiocarbon_data <-
-# readr::read_csv(
-#   here::here("analysis/data/raw_data/0403_sbox_clean.csv")
-# ) %>%
-#   # Convert to spatial object
-#   sf::st_as_sf(coords = c("Long","Lat"),
-#                crs = 4326) %>%
-#   sf::st_transform(four_corners_lcc_proj) %>%
-#   # Select salient columns
-#   dplyr::select(labnumber,
-#                 date,
-#                 sd,
-#                 Site,
-#                 SiteID) %>%
-#   dplyr::rename(Lab_Number = labnumber,
-#                 Date_BP = date,
-#                 SD = sd,
-#                 Site_Number = Site) %>%
-#   sf::st_intersection(uusw_counties) %>%
-#   sf::st_drop_geometry() %>%
-#   dplyr::mutate(Type = "Radiocarbon",
-#                 SiteID = SiteID %>% stringr::str_c("Radiocarbon - ", .)) %>%
-#   dplyr::select(Type,
-#                 SiteID,
-#                 Site_Number,
-#                 County,
-#                 Lab_Number,
-#                 Date_BP,
-#                 SD) %T>%
-#   # Write the output to a new file (csv)
-#   readr::write_csv(
-#     here::here("analysis/data/derived_data/radiocarbon_data.csv")
-#   )
+# Read in the radiocarbon database
+if(!file.exists(
+  here::here("analysis/data/derived_data/radiocarbon_data.csv")
+))
+  readr::read_csv(
+    here::here("analysis/data/raw_data/0403_sbox_clean_sd300.csv")
+  ) %>%
+  # Convert to spatial object
+  sf::st_as_sf(coords = c("Long","Lat"),
+               crs = 4326) %>%
+  sf::st_transform(four_corners_lcc_proj) %>%
+  # Select salient columns
+  dplyr::select(labnumber,
+                date,
+                sd,
+                Site,
+                SiteID) %>%
+  dplyr::rename(Lab_Number = labnumber,
+                Date_BP = date,
+                SD = sd,
+                Site_Number = Site) %>%
+  sf::st_intersection(uusw_counties) %>%
+  sf::st_drop_geometry() %>%
+  dplyr::mutate(Type = "Radiocarbon",
+                SiteID = SiteID %>% stringr::str_c("Radiocarbon - ", .)) %>%
+  dplyr::select(Type,
+                SiteID,
+                Site_Number,
+                County,
+                Lab_Number,
+                Date_BP,
+                SD) %>%
+  # Write the output to a new file (csv)
+  readr::write_csv(
+    here::here("analysis/data/derived_data/radiocarbon_data.csv")
+  )
 
@@
 ```{r prepare-treering-data}
-# # Read in the Bocinsky et al. 2016 tree ring data
-# # https://doi.org/10.6067/XCV82J6D7B
-# dendro_data <-
-#   readr::read_csv(
-#     here::here("analysis/data/raw_data/BOCINSKY_ET_AL_2015_TREE_RINGS.csv")
-#   ) %>%
-#   # Convert to spatial object
-#   sf::st_as_sf(coords = c("LON","LAT"),
-#                crs = 4326) %>%
-#   sf::st_transform(four_corners_lcc_proj) %>%
-#   # Convert dates to BP
-#   dplyr::mutate(Date_BP = 1950 - Outer_Date_AD,
-#                 Type = ifelse(C_Level %in% c(2,3),
-#                               "Cutting or Near-cutting",
-#                               "Non-cutting")) %>%
-#   # drop C_Level
-#   dplyr::select(Seq_Number,
-#                 Site_Number,
-#                 Lab_Number,
-#                 Type,
-#                 Date_BP) %>%
-#   dplyr::rename(SiteID = Seq_Number) %>%
-#   sf::st_intersection(uusw_counties) %>%
-#   sf::st_drop_geometry() %>%
-#   dplyr::mutate(Type = Type %>% stringr::str_c("Dendro - ", .),
-#                 SiteID = SiteID %>% stringr::str_c("Dendro - ", .),
-#                 SD = 0) %>%
-#   dplyr::filter(Type == "Dendro - Cutting or Near-cutting") %>%
-#   dplyr::mutate(Type = "Tree-ring")  %>%
-#   dplyr::select(Type,
-#                 SiteID,
-#                 Site_Number,
-#                 County,
-#                 Lab_Number,
-#                 Date_BP,
-#                 SD) %T>%
-#   # Write the output to a new file (csv)
-#   readr::write_csv(
-#     here::here("analysis/data/derived_data/dendro_data.csv")
-#   )
+# Read in the Bocinsky et al. 2016 tree ring data
+# https://doi.org/10.6067/XCV82J6D7B
+
+if(!file.exists(
+  here::here("analysis/data/derived_data/dendro_data.csv")
+))
+  readr::read_csv(
+    here::here("analysis/data/raw_data/BOCINSKY_ET_AL_2015_TREE_RINGS.csv")
+  ) %>%
+  # Convert to spatial object
+  sf::st_as_sf(coords = c("LON","LAT"),
+               crs = 4326) %>%
+  sf::st_transform(four_corners_lcc_proj) %>%
+  # Convert dates to BP
+  dplyr::mutate(Date_BP = 1950 - Outer_Date_AD,
+                Type = ifelse(C_Level %in% c(2,3),
+                              "Cutting or Near-cutting",
+                              "Non-cutting")) %>%
+  # drop C_Level
+  dplyr::select(Seq_Number,
+                Site_Number,
+                Lab_Number,
+                Type,
+                Date_BP) %>%
+  dplyr::rename(SiteID = Seq_Number) %>%
+  sf::st_intersection(uusw_counties) %>%
+  sf::st_drop_geometry() %>%
+  dplyr::mutate(Type = Type %>% stringr::str_c("Dendro - ", .),
+                SiteID = SiteID %>% stringr::str_c("Dendro - ", .),
+                SD = 0) %>%
+  dplyr::filter(Type == "Dendro - Cutting or Near-cutting") %>%
+  dplyr::mutate(Type = "Tree-ring")  %>%
+  dplyr::select(Type,
+                SiteID,
+                Site_Number,
+                County,
+                Lab_Number,
+                Date_BP,
+                SD) %>%
+  # Write the output to a new file (csv)
+  readr::write_csv(
+    here::here("analysis/data/derived_data/dendro_data.csv")
+  )
 
@@
 # Set the time range for the reconstructions
-timerange <- c(1949,0)
+timerange <- c(1949, 0)
 
-if(!file.exists(
-  here::here("analysis/data/derived_data/radiocarbon_spd.rds")
-))
 all_data$Radiocarbon %>%
@@
                                                        errors = SD,
-                                                       timerange = timerange,
+                                                       timeRange = timerange,
                                                        calMatrix = TRUE,
@@
                                         bins = Bins,
-                                        datenormalised = TRUE,
                                         spdnormalised = TRUE,
-                                        timeRange = c(1949,0),
+                                        timeRange = timerange,
                                         runm = 21,
@@
       StartBP = timerange[1],
-      EndBP = 0,
+      EndBP = timerange[2],
       Normalised = TRUE,
@@
 
-if(!file.exists(
-  here::here("analysis/data/derived_data/dendro_spd.rds")
-))
 all_data$`Tree-ring` %>%
@@
                                         bins = Bins,
-                                        datenormalised = TRUE,
                                         spdnormalised = TRUE,

### 2020-01-19 data-raw/DATASET.R

@@ -2,36 +2,45 @@ library(magrittr)
 library(tidycensus)
 
 # A nice projection for the Four Corners states
-nm <- tigris::states(class = "sf",
-                     progress_bar = FALSE) %>%
+nm <- tigris::states(
+  class = "sf",
+  progress_bar = FALSE
+) %>%
   dplyr::filter(NAME == "New Mexico") %>%
   sf::st_transform(4326) %>%
   sf::st_bbox()
 
-co <- tigris::states(class = "sf",
-                     progress_bar = FALSE) %>%
+co <- tigris::states(
+  class = "sf",
+  progress_bar = FALSE
+) %>%
   dplyr::filter(NAME == "Colorado") %>%
   sf::st_transform(4326) %>%
   sf::st_bbox()
 
-four_corners_lcc_proj <- paste0("+proj=lcc",
-                                " +lat_1=",co[['ymax']],
-                                " +lat_2=",nm[['ymax']] - (co[['ymax']] - co[['ymin']]),
-                                " +lat_0=",nm[['ymax']],
-                                " +lon_0=",nm[['xmin']],
-                                " +x_0=",nm[['xmin']],
-                                " +x_y=",nm[['xmin']]
-                                )
+four_corners_lcc_proj <- paste0(
+  "+proj=lcc",
+  " +lat_1=", co[["ymax"]],
+  " +lat_2=", nm[["ymax"]] - (co[["ymax"]] - co[["ymin"]]),
+  " +lat_0=", nm[["ymax"]],
+  " +lon_0=", nm[["xmin"]],
+  " +x_0=", nm[["xmin"]],
+  " +x_y=", nm[["xmin"]]
+)
 usethis::use_data(four_corners_lcc_proj, overwrite = TRUE)
 
 four_corners_states <-
-  tigris::states(class = "sf",
-                 progress_bar = FALSE) %>%
-  dplyr::filter(NAME %in% c("Arizona", "Colorado",
-                            "New Mexico", "Utah")) %>%
+  tigris::states(
+    class = "sf",
+    progress_bar = FALSE
+  ) %>%
+  dplyr::filter(NAME %in% c(
+    "Arizona", "Colorado",
+    "New Mexico", "Utah"
+  )) %>%
   smoothr::densify(max_distance = 0.01) %>%
   sf::st_transform(four_corners_lcc_proj) %>%
[…]

### 2020-01-19 analysis/paper/paper.Rmd
trying precommit

@@
 theme_map <- function(base_size = 6.5,
-                      base_family = ""){
-  theme_bw(base_size = base_size,
-           base_family = base_family) %+replace%
-    theme(axis.line = element_blank(),
+                      base_family = "") {
+  theme_bw(
+    base_size = base_size,
+    base_family = base_family
+  ) %+replace%
+    theme(
+      axis.line = element_blank(),
       axis.text = element_blank(),
@@
       panel.grid = element_blank(),
-          panel.grid.major = element_line(colour = 'transparent'),
-          panel.grid.minor = element_line(colour = 'transparent'),
+      panel.grid.major = element_line(colour = "transparent"),
+      panel.grid.minor = element_line(colour = "transparent"),
       legend.key = element_blank(),
@@
   geom_sf(data = four_corners_states) +
-  geom_raster(data = robinson2020::swus_hillshade_500m,
-              mapping = aes(x = x,
+  geom_raster(
+    data = robinson2020::swus_hillshade_500m,
+    mapping = aes(
+      x = x,
       y = y,
-                            alpha = ID),
-              na.rm = TRUE) +
-  scale_alpha(range = c(0.8, 0),
+      alpha = ID
+    ),
+    na.rm = TRUE
+  ) +
+  scale_alpha(
+    range = c(0.8, 0),
     na.value = 0,
-              limits = c(0,255),
-              guide = "none") +
-  geom_sf(data = four_corners_states,
-          fill = NA) +
-  geom_sf(data = uusw_counties %>% 
+    limits = c(0, 255),
+    guide = "none"
+  ) +
+  geom_sf(
+    data = four_corners_states,
+    fill = NA
+  ) +
+  geom_sf(
+    data = uusw_counties %>%
       sf::st_union() %>%
@@
     fill = NA,
-          color = "black") +
-  geom_sf(data = uusw_boundary,
+    color = "black"
+  ) +
+  geom_sf(
+    data = uusw_boundary,
     fill = NA,
     linetype = 2,
-          color = "white") +
-  geom_sf(data = vep_study_areas,
+    color = "white"
+  ) +
+  geom_sf(
+    data = vep_study_areas,
     fill = NA,
     color = "white",
-          size = 1) +
-  geom_sf_text(data = vep_study_areas,
+    size = 1
+  ) +
+  geom_sf_text(
+    data = vep_study_areas,
     aes(label = `Study Area`),
-               size = 3) +
+    size = 3
+  ) +
   theme_map(16)
-
 ```
@@
 # Read in the radiocarbon database
-if(!file.exists(
+if (!file.exists(
   here::here("analysis/data/derived_data/radiocarbon_data.csv")
-))
+)) {
   readr::read_csv(
@@
     # Convert to spatial object
-  sf::st_as_sf(coords = c("Long","Lat"),
-               crs = 4326) %>%
+    sf::st_as_sf(
+      coords = c("Long", "Lat"),
+      crs = 4326
+    ) %>%
     sf::st_transform(four_corners_lcc_proj) %>%
     # Select salient columns
-  dplyr::select(labnumber,
+    dplyr::select(
+      labnumber,
       date,
@@
       Site,
-                SiteID) %>%
-  dplyr::rename(Lab_Number = labnumber,
+      SiteID
+    ) %>%
+    dplyr::rename(
+      Lab_Number = labnumber,
       Date_BP = date,
       SD = sd,
-                Site_Number = Site) %>%
+      Site_Number = Site
+    ) %>%
     sf::st_intersection(uusw_counties) %>%
     sf::st_drop_geometry() %>%
-  dplyr::mutate(Type = "Radiocarbon",
-                SiteID = SiteID %>% stringr::str_c("Radiocarbon - ", .)) %>%
-  dplyr::select(Type,
+    dplyr::mutate(
+      Type = "Radiocarbon",
+      SiteID = SiteID %>% stringr::str_c("Radiocarbon - ", .)
+    ) %>%
+    dplyr::select(
+      Type,
       SiteID,
@@
       Date_BP,
-                SD) %>%
+      SD
+    ) %>%
     # Write the output to a new file (csv)
@@
     )
+}
 
@@
   )
-
 ```
@@
 
-if(!file.exists(
+if (!file.exists(
   here::here("analysis/data/derived_data/dendro_data.csv")
-))
+)) {
   readr::read_csv(
@@
     # Convert to spatial object
-  sf::st_as_sf(coords = c("LON","LAT"),
-               crs = 4326) %>%
+    sf::st_as_sf(
+      coords = c("LON", "LAT"),
+      crs = 4326
+    ) %>%
     sf::st_transform(four_corners_lcc_proj) %>%
     # Convert dates to BP
-  dplyr::mutate(Date_BP = 1950 - Outer_Date_AD,
-                Type = ifelse(C_Level %in% c(2,3),
+    dplyr::mutate(
+      Date_BP = 1950 - Outer_Date_AD,
+      Type = ifelse(C_Level %in% c(2, 3),
         "Cutting or Near-cutting",
-                              "Non-cutting")) %>%
+        "Non-cutting"
+      )
+    ) %>%
     # drop C_Level
-  dplyr::select(Seq_Number,
+    dplyr::select(
+      Seq_Number,
       Site_Number,
@@
       Type,
-                Date_BP) %>%
+      Date_BP
+    ) %>%
     dplyr::rename(SiteID = Seq_Number) %>%
@@
     sf::st_drop_geometry() %>%
-  dplyr::mutate(Type = Type %>% stringr::str_c("Dendro - ", .),
+    dplyr::mutate(
+      Type = Type %>% stringr::str_c("Dendro - ", .),
       SiteID = SiteID %>% stringr::str_c("Dendro - ", .),
-                SD = 0) %>%
+      SD = 0
+    ) %>%
     dplyr::filter(Type == "Dendro - Cutting or Near-cutting") %>%
     dplyr::mutate(Type = "Tree-ring") %>%
-  dplyr::select(Type,
+    dplyr::select(
+      Type,
       SiteID,
@@
 ```{r}
-all_data <- list(Radiocarbon = radiocarbon_data,
-                 `Tree-ring` = dendro_data) %>%
-  purrr::map(~dplyr::mutate(.x,
+all_data <- list(
+  Radiocarbon = radiocarbon_data,
+  `Tree-ring` = dendro_data
+) %>%
+  purrr::map(~ dplyr::mutate(.x,
     `Study Area` = ifelse(County == "Montezuma, CO",
       "CMV",
-                                                  NA),
-                            `Study Area` = ifelse(County %in% c("Los Alamos, NM",
+      NA
+    ),
+    `Study Area` = ifelse(County %in% c(
+      "Los Alamos, NM",
       "Rio Arriba, NM",
@@
       "Santa Fe, NM",
-                                                                "Taos, NM"),
+      "Taos, NM"
+    ),
     "NRG",
-                                                  `Study Area`)))
-
+    `Study Area`
+    )
+  ))
 ```
@@
   dplyr::mutate(`Study Area` = factor(`Study Area`,
-                                      levels = c("UUSW Total",
+    levels = c(
+      "UUSW Total",
       "CMV",
-                                                 "NRG"))) %>%
-  dplyr::group_by(Type,`Study Area`) %>%
+      "NRG"
+    )
+  )) %>%
+  dplyr::group_by(Type, `Study Area`) %>%
   dplyr::summarise(Count = dplyr::n()) %>%
   tidyr::spread(Type, Count) %>%
-  knitr::kable(caption = "Counts of radiocarbon and tree-ring dates in this study.",
+  knitr::kable(
+    caption = "Counts of radiocarbon and tree-ring dates in this study.",
     booktabs = TRUE,
-               format.args = list(big.mark = ','))
+    format.args = list(big.mark = ",")
+  )
 ```
@@
   dplyr::left_join(uusw_counties,
-                   by = "County") %>%
+    by = "County"
+  ) %>%
   sf::st_as_sf()
@@
       legend.background = element_blank(),
-          legend.key.width = unit(0.15,"npc"),
+      legend.key.width = unit(0.15, "npc"),
       legend.text = element_text(size = rel(1)),
@@
   ggplot() +
-  geom_sf(data = four_corners_states,
-          fill = NA) +
+  geom_sf(
+    data = four_corners_states,
+    fill = NA
+  ) +
   geom_sf(aes(fill = Dates),
-          color = "black") +
-  scale_fill_distiller(name = "Number of Radiocarbon Dates",
+    color = "black"
+  ) +
+  scale_fill_distiller(
+    name = "Number of Radiocarbon Dates",
     palette = "Greys",
     direction = 1,
-                       limits = c(0,500),
-                       minor_breaks = seq(50,450,100),
-                       breaks = seq(0,500,100),
+    limits = c(0, 500),
+    minor_breaks = seq(50, 450, 100),
+    breaks = seq(0, 500, 100),
     expand = expand_scale(mult = 0, add = 0),
-                       guide = guide_colorbar(title.position = "top")) +
-  geom_sf(data = uusw_counties,
+    guide = guide_colorbar(title.position = "top")
+  ) +
+  geom_sf(
+    data = uusw_counties,
     fill = NA,
-          color = "black") +
+    color = "black"
+  ) +
   theme_map(16)
@@
   ggplot() +
-  geom_sf(data = four_corners_states,
-          fill = NA) +
+  geom_sf(
+    data = four_corners_states,
+    fill = NA
+  ) +
   geom_sf(aes(fill = Dates),
-          color = "black") +
-  scale_fill_distiller(name = "Number of Tree-ring Dates",
+    color = "black"
+  ) +
+  scale_fill_distiller(
+    name = "Number of Tree-ring Dates",
     palette = "Greys",
     direction = 1,
-                       limits = c(0,5000),
-                       minor_breaks = seq(500,4500,1000),
-                       breaks = seq(0,5000,1000),
+    limits = c(0, 5000),
+    minor_breaks = seq(500, 4500, 1000),
+    breaks = seq(0, 5000, 1000),
     expand = expand_scale(mult = 0, add = 0),
-                       guide = guide_colorbar(title.position = "top")) +
-  geom_sf(data = uusw_counties,
+    guide = guide_colorbar(title.position = "top")
+  ) +
+  geom_sf(
+    data = uusw_counties,
     fill = NA,
-          color = "black") +
+    color = "black"
+  ) +
   theme_map(16)
-
 ```
@@
       panel.grid = element_blank(),
-          panel.grid.major = element_line(colour = 'transparent'),
-          panel.grid.minor = element_line(colour = 'transparent'),
+      panel.grid.major = element_line(colour = "transparent"),
+      panel.grid.minor = element_line(colour = "transparent"),
       legend.key = element_blank(),
-          # legend.position = "top",
-          # legend.justification = c(0,0),
       legend.position = "bottom",
-          # legend.position = c(0.5,1),
       legend.background = element_blank(),
-          legend.key.width = unit(0.15,"npc"),
-          # legend.key.height = unit(0.15,"in"),
+      legend.key.width = unit(0.15, "npc"),
       legend.text = element_text(size = rel(1)),
@@
   ggplot() +
-  geom_sf(data = four_corners_states,
-          fill = NA) +
+  geom_sf(
+    data = four_corners_states,
+    fill = NA
+  ) +
   geom_sf(aes(fill = Dates),
-          color = "black") +
-  scale_fill_distiller(name = "Number of Sites with Radiocarbon Dates",
+    color = "black"
+  ) +
+  scale_fill_distiller(
+    name = "Number of Sites with Radiocarbon Dates",
     palette = "Greys",
     direction = 1,
-                       limits = c(0,80),
-                       minor_breaks = seq(10,70,20),
-                       breaks = seq(0,80,20),
+    limits = c(0, 80),
+    minor_breaks = seq(10, 70, 20),
+    breaks = seq(0, 80, 20),
     expand = expand_scale(mult = 0, add = 0),
-                       guide = guide_colorbar(title.position = "top")) +
-  geom_sf(data = uusw_counties,
+    guide = guide_colorbar(title.position = "top")
+  ) +
+  geom_sf(
+    data = uusw_counties,
     fill = NA,
-          color = "black") +
+    color = "black"
+  ) +
   theme_map(16)
@@
   ggplot() +
-  geom_sf(data = four_corners_states,
-          fill = NA) +
+  geom_sf(
+    data = four_corners_states,
+    fill = NA
+  ) +
   geom_sf(aes(fill = Dates),
-          color = "black") +
-  scale_fill_distiller(name = "Number of Sites with Tree-ring Dates",
+    color = "black"
+  ) +
+  scale_fill_distiller(
+    name = "Number of Sites with Tree-ring Dates",
     palette = "Greys",
     direction = 1,
-                       limits = c(0,250),
-                       minor_breaks = seq(25,225,50),
-                       breaks = seq(0,250,50),
+    limits = c(0, 250),
+    minor_breaks = seq(25, 225, 50),
+    breaks = seq(0, 250, 50),
     expand = expand_scale(mult = 0, add = 0),
-                       guide = guide_colorbar(title.position = "top")) +
-  geom_sf(data = uusw_counties,
+    guide = guide_colorbar(title.position = "top")
+  ) +
+  geom_sf(
+    data = uusw_counties,
     fill = NA,
-          color = "black") +
+    color = "black"
+  ) +
   theme_map(16)
-
 ```
@@
 all_data$Radiocarbon %>%
-  split(.,.$`Study Area`) %>%
-  tibble::tibble(`Study Area` = names(.),
-                 Dates = .) %>%
-  dplyr::bind_rows(tibble::tibble(`Study Area` = "UUSW",
-                                  Dates = list(all_data$Radiocarbon))) %>%
+  split(., .$`Study Area`) %>%
+  tibble::tibble(
+    `Study Area` = names(.),
+    Dates = .
+  ) %>%
+  dplyr::bind_rows(tibble::tibble(
+    `Study Area` = "UUSW",
+    Dates = list(all_data$Radiocarbon)
+  )) %>%
   dplyr::rowwise() %>%
-  dplyr::mutate(Calibrated = list(Dates %$%
-                                    rcarbon::calibrate(x = Date_BP,
+  dplyr::mutate(
+    Calibrated = list(Dates %$%
+      rcarbon::calibrate(
+        x = Date_BP,
         errors = SD,
@@
         calMatrix = TRUE,
-                                                       verbose = FALSE)),
-                Bins = list(rcarbon::binPrep(sites = Dates$SiteID,
+        verbose = FALSE
+      )),
+    Bins = list(rcarbon::binPrep(
+      sites = Dates$SiteID,
       ages = Calibrated,
-                                             h = 100)),
-                SPD = list(rcarbon::spd(x = Calibrated,
+      h = 100
+    )),
+    SPD = list(rcarbon::spd(
+      x = Calibrated,
       bins = Bins,
@@
       runm = 21,
-                                        verbose = FALSE))) %>%
+      verbose = FALSE
+    ))
+  ) %>%
   dplyr::ungroup() %>%
   dplyr::mutate(Density = SPD %>%
-                  purrr::map(function(x){
+    purrr::map(function(x) {
       x$grid %>%
         tibble::as_tibble() %$%
-                      tibble::tibble(Type = "Radiocarbon Density",
+        tibble::tibble(
+          Type = "Radiocarbon Density",
           Date_BP = calBP,
-                                     Density = PrDens / sum(PrDens, na.rm = T))
-                  })
-  ) %>%
+          Density = PrDens / sum(PrDens, na.rm = T)
+        )
+    })) %>%
   readr::write_rds(
     here::here("analysis/data/derived_data/radiocarbon_spd.rds"),
-    compress = "gz")
+    compress = "gz"
+  )
 
@@
 )
-
 ```
@@
 ```{r treering-spds}
-calibrate_tree_ring_dates <- function(x){
+calibrate_tree_ring_dates <- function(x) {
   out <- list(
@@
     calmatrix =
-      tibble::tibble(index = 1:length(x),
+      tibble::tibble(
+        index = 1:length(x),
         year_bp = x,
-                     value = TRUE) %>%
-      tidyr::pivot_wider(names_from = index,
-                         values_from = value) %>%
+        value = TRUE
+      ) %>%
+        tidyr::pivot_wider(
+          names_from = index,
+          values_from = value
+        ) %>%
         dplyr::bind_rows(tibble::tibble(year_bp = timerange[1]:timerange[2])) %>%
@@
 
-  attr(out,"class") <- c("CalDates", "list")
+  attr(out, "class") <- c("CalDates", "list")
 
@@
 all_data$`Tree-ring` %>%
-  split(.,.$`Study Area`) %>%
-  tibble::tibble(`Study Area` = names(.),
-                 Dates = .) %>%
-  dplyr::bind_rows(tibble::tibble(`Study Area` = "UUSW",
-                                  Dates = list(all_data$`Tree-ring`))) %>%
+  split(., .$`Study Area`) %>%
+  tibble::tibble(
+    `Study Area` = names(.),
+    Dates = .
+  ) %>%
+  dplyr::bind_rows(tibble::tibble(
+    `Study Area` = "UUSW",
+    Dates = list(all_data$`Tree-ring`)
+  )) %>%
   dplyr::rowwise() %>%
-  dplyr::mutate(Calibrated = list(Dates %$%
+  dplyr::mutate(
+    Calibrated = list(Dates %$%
       calibrate_tree_ring_dates(x = Date_BP)),
-                Bins = list(rcarbon::binPrep(sites = Dates$SiteID,
+    Bins = list(rcarbon::binPrep(
+      sites = Dates$SiteID,
       ages = Dates$Date_BP,
-                                             h = 100)),
-                SPD = list(rcarbon::spd(x = Calibrated,
+      h = 100
+    )),
+    SPD = list(rcarbon::spd(
+      x = Calibrated,
       bins = Bins,
@@
       runm = 21,
-                                        verbose = FALSE))) %>%
-
+      verbose = FALSE
+    ))
+  ) %>%
   dplyr::ungroup() %>%
   dplyr::mutate(Density = SPD %>%
-                  purrr::map(function(x){
+    purrr::map(function(x) {
       x$grid %>%
         tibble::as_tibble() %$%
-                      tibble::tibble(Type = "Tree-ring Density",
+        tibble::tibble(
+          Type = "Tree-ring Density",
           Date_BP = calBP,
-                                     Density = PrDens / sum(PrDens, na.rm = T))
-                  })
-  ) %>%
+          Density = PrDens / sum(PrDens, na.rm = T)
+        )
+    })) %>%
   readr::write_rds(
     here::here("analysis/data/derived_data/dendro_spd.rds"),
-    compress = "gz")
+    compress = "gz"
+  )
 
@@
   dendro_spd %>%
-  dplyr::select(`Study Area`,Density) %>%
+  dplyr::select(`Study Area`, Density) %>%
   tidyr::unnest(cols = c(Density)) %>%
   dplyr::bind_rows(radiocarbon_spd %>%
-                     dplyr::select(`Study Area`,Density) %>%
+    dplyr::select(`Study Area`, Density) %>%
     tidyr::unnest(cols = c(Density))) %>%
   dplyr::mutate(`Year AD` = 1950 - Date_BP) %>%
-  dplyr::select(`Year AD`,
+  dplyr::select(
+    `Year AD`,
     Date_BP,
@@
     Type,
-                Density) %>%
+    Density
+  ) %>%
   dplyr::rename(Value = Density)
@@
   dplyr::rowwise() %>%
-  dplyr::mutate(`Year AD` = list((Start+1):End)) %>%
-  dplyr::select(`Study Area`,
+  dplyr::mutate(`Year AD` = list((Start + 1):End)) %>%
+  dplyr::select(
+    `Study Area`,
     `Year AD`,
-                Population) %>%
+    Population
+  ) %>%
   tidyr::unnest(`Year AD`) %>%
-  dplyr::mutate(Type = "Population",
-                Date_BP = 1950 - `Year AD`) %>%
+  dplyr::mutate(
+    Type = "Population",
+    Date_BP = 1950 - `Year AD`
+  ) %>%
   dplyr::rename(Value = Population)
@@
   dplyr::bind_rows(vep_demography) %>%
-  dplyr::mutate(Type = factor(Type, 
-                              levels = c("Radiocarbon Density",
+  dplyr::mutate(
+    Type = factor(Type,
+      levels = c(
+        "Radiocarbon Density",
         "Tree-ring Density",
-                                         "Population"),
-                              ordered = TRUE),
+        "Population"
+      ),
+      ordered = TRUE
+    ),
     `Study Area` = factor(`Study Area`,
-                                      levels = c( `Upland US Southwest` = "UUSW",
+      levels = c(
+        `Upland US Southwest` = "UUSW",
         `Central Mesa Verde` = "CMV",
-                                                  `Northern Rio Grande` = "NRG"))
+        `Northern Rio Grande` = "NRG"
+      )
+    )
   ) %>%
   ggplot(aes(x = `Year AD`)) +
-  geom_line(aes(y = Value,
+  geom_line(aes(
+    y = Value,
     color = `Study Area`,
-                size = `Study Area`)) +
-  scale_color_manual(values = c("UUSW" = "black",
+    size = `Study Area`
+  )) +
+  scale_color_manual(values = c(
+    "UUSW" = "black",
     "CMV" = "gray50",
-                                "NRG" = "gray80")) +
-  scale_size_manual(values = c("UUSW" = 1.25,
+    "NRG" = "gray80"
+  )) +
+  scale_size_manual(values = c(
+    "UUSW" = 1.25,
     "CMV" = 0.75,
-                               "NRG" = 0.75)) +
+    "NRG" = 0.75
+  )) +
   # coord_cartesian(xlim = c(600,1600)) +
   xlab("Year AD") +
-  scale_x_continuous(breaks = c(1,seq(200,2000,200))) + 
+  scale_x_continuous(breaks = c(1, seq(200, 2000, 200))) +
   theme_minimal(18) +
-  theme(legend.title = element_blank(),
+  theme(
+    legend.title = element_blank(),
     legend.justification = c(0, 1),
@@
     panel.grid.minor.y = element_blank(),
-        plot.margin = unit(c(0.25,0.5,0.25,0.25),"in")
+    plot.margin = unit(c(0.25, 0.5, 0.25, 0.25), "in")
   ) +
@@
     nrow = 3,
-             scales = "free_y")
-
+    scales = "free_y"
+  )
 ```
@@
 if ("git2r" %in% installed.packages() &
-    git2r::in_repository(path = ".")) 
+  git2r::in_repository(path = ".")) {
   git2r::repository(here::here())
+}
 ```

### 2020-02-05 README.Rmd
color figs

@@
 
-## How to run in your broswer or download and run locally
+## How to run in your browser or download and run locally
 
 The simplest way to explore the text, code and data is to click on [binder](https://mybinder.org/v2/gh/bocinsky/robinson2020/master?urlpath=rstudio) to open an instance of RStudio in your browser, which will have the compendium files ready to work with.
-Binder uses [rocker-project.org Docker] images to ensure a consistent and reproducible computational environment.
+Binder uses (rocker-project.org)[rocker-project.org Docker] images to ensure a consistent and reproducible computational environment.
 These Docker images can also be used locally.

### 2020-02-05 analysis/paper/paper.Rmd
color figs

@@
   scale_color_manual(values = c(
-    "UUSW" = "black",
-    "CMV" = "gray50",
-    "NRG" = "gray80"
+    "UUSW" = "#1b9e77",
+    "CMV" = "#d95f02",
+    "NRG" = "#7570b3"
   )) +

### 2020-02-05 analysis/paper/paper.Rmd
documented data

@@
 library(robinson2020)
+library(ggplot2)
 ```

### 2020-02-05 R/data.R

@@ -1,0 +1,78 @@
+#' A nice projection for the Four Corners states
+#'
+#' In PROJ.4:
+#' "+proj=lcc +lat_1=41.0034439990657 +lat_2=32.9892379991201 +lat_0=37.0002329990929
+#' +lon_0=-109.050431 +x_0=-109.050431 +x_y=-109.050431"
+#'
+#' @format A character string with the PROJ.4 projection information
+"four_corners_lcc_proj"
+
+#' The boundaries of the Four Corners states
+#'
+#' These data are derived from the US Census TIGRIS database.
+#'
+#' @format Simple feature collection with 4 features and 2 fields:
+#' \describe{
+#'   \item{STATEFP}{the state's FIPS code}
+#'   \item{geometry}{the state's boundary}
+#'  }
+"four_corners_states"
+
+#' The boundaries of the Upland US Southwest
+#'
+#' The UUSW was geographically defined by Bocinsky et al. 2016 as the region between 105–113ºW and 32–38ºN.
+#'
+#' @format Geometry set for 1 feature
+"uusw_boundary"
+
+#' The boundaries of the Upland US Southwest counties used in Robinson et al. 2020
+#'
+#' The UUSW was geographically defined by Bocinsky et al. 2016 as the region between 105–113ºW and 32–38ºN.
+#' Robinson et al. 2020 refined that definition in two ways.
+#' First, because location data for the radiocarbon database are only resolved to the county
+#' level for most sites, the authors defined their study region to include all counties whose
+#' geographic centroids fall within the region from Bocinsky et al. 2016.
+#' Second, because of the generally poor tree-ring date record in the Sonoran Desert,
+#' the authors roughly exclude the Phoenix and Tucson basins by dropping
+#' Maricopa, Pinal, and Pima counties in Arizona.
+#'
+#' @format SSimple feature collection with 40 features and 2 field:
+#' \describe{
+#'   \item{County}{the county's name and state}
+#'   \item{geometry}{the county's boundary}
+#'  }
+"uusw_counties"
+
+#' The boundaries of the Village Ecodynamics Project (VEP) study areas
+#'
+#' @format Simple feature collection with 2 features and 1 field:
+#' \describe{
+#'   \item{Study Area}{the VEP study area's name}
+#'   \item{geometry}{the VEP study area's boundary}
+#'  }
+"vep_study_areas"
+
[…]

### 2020-02-05 README.Rmd
Added Bob's ORCID
getting travis up and running again

@@
 # robinson2020
-[![Travis-CI Build Status](https://travis-ci.org/bocinsky/robinson2020.svg?branch=master)](https://travis-ci.org/bocinsky/robinson2020)
-[![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh/bocinsky/robinson2020/master?urlpath=rstudio)
+[![Travis build status](https://travis-ci.org/people3k/robinson2020.svg?branch=master)](https://travis-ci.org/people3k/robinson2020)
+[![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh/people3k/robinson2020/master?urlpath=rstudio)
 
@@
 Kyle Bocinsky [![](https://orcid.org/sites/default/files/images/orcid_16x16.png)](https://orcid.org/0000-0003-1862-3428), Jacob Freeman, Darcy Bird [![](https://orcid.org/sites/default/files/images/orcid_16x16.png)](https://orcid.org/0000-0003-3466-6284), Robert L.
-Kelly, (2020).
+Kelly[![](https://orcid.org/sites/default/files/images/orcid_16x16.png)](https://orcid.org/0000-0001-9737-0152), (2020).
 _`r Title`_.
@@
 
-The simplest way to explore the text, code and data is to click on [binder](https://mybinder.org/v2/gh/bocinsky/robinson2020/master?urlpath=rstudio) to open an instance of RStudio in your browser, which will have the compendium files ready to work with.
+The simplest way to explore the text, code and data is to click on [binder](https://mybinder.org/v2/gh/people3k/robinson2020/master?urlpath=rstudio) to open an instance of RStudio in your browser, which will have the compendium files ready to work with.
 Binder uses (rocker-project.org)[rocker-project.org Docker] images to ensure a consistent and reproducible computational environment.
@@
 Then open `paper.Rmd` and knit to produce the `paper.docx`.
+
+Or, simply install straight from Github:
+
+```{r, eval = FALSE}
+# install.packages("devtools")
+devtools::install_github("people3k/robinson2020")
+```
 

### 2020-02-05 README.Rmd
updated readme

@@
 
-> Robinson, Erick [![](https://orcid.org/sites/default/files/images/orcid_16x16.png)](https://orcid.org/0000-0002-0789-3724), R.
-Kyle Bocinsky [![](https://orcid.org/sites/default/files/images/orcid_16x16.png)](https://orcid.org/0000-0003-1862-3428), Jacob Freeman, Darcy Bird [![](https://orcid.org/sites/default/files/images/orcid_16x16.png)](https://orcid.org/0000-0003-3466-6284), Robert L.
-Kelly[![](https://orcid.org/sites/default/files/images/orcid_16x16.png)](https://orcid.org/0000-0001-9737-0152), (2020).
-_`r Title`_.
-Under review.
+> Robinson, Erick [![](https://orcid.org/sites/default/files/images/orcid_16x16.png)](https://orcid.org/0000-0002-0789-3724), 
+R. Kyle Bocinsky [![](https://orcid.org/sites/default/files/images/orcid_16x16.png)](https://orcid.org/0000-0003-1862-3428),
+Jacob Freeman [![](https://orcid.org/sites/default/files/images/orcid_16x16.png)](https://orcid.org/0000-0001-7402-8450), 
+Darcy Bird [![](https://orcid.org/sites/default/files/images/orcid_16x16.png)](https://orcid.org/0000-0003-3466-6284), 
+Robert L. Kelly[![](https://orcid.org/sites/default/files/images/orcid_16x16.png)](https://orcid.org/0000-0001-9737-0152), 
+(2020). _`r Title`_. Under review.
 

### 2020-02-05 README.Rmd
added Zenodo badge

@@
 # robinson2020
+[![DOI](https://zenodo.org/badge/234769845.svg)](https://zenodo.org/badge/latestdoi/234769845)
 [![Travis build status](https://travis-ci.org/people3k/robinson2020.svg?branch=master)](https://travis-ci.org/people3k/robinson2020)

### 2020-02-05 README.Rmd
added Zenodo badge

--- 2020-02-05 README.Rmd
@@
 # robinson2020
+[![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.3647599.svg)](https://doi.org/10.5281/zenodo.3647599)
 [![Travis build status](https://travis-ci.org/people3k/robinson2020.svg?branch=master)](https://travis-ci.org/people3k/robinson2020)

### 2020-03-11 analysis/paper/paper.Rmd
refining figures

--- 2020-01-19 analysis/paper/paper.Rmd
@@
 knitr::opts_chunk$set(
+  cache = FALSE,
   echo = TRUE,
@@
     ) %>%
+    # Get only raw dates between 1950 and 150 BP
+    dplyr::filter(
+      Date_BP > 150,
+      Date_BP <= 1950
+    ) %>%
     # Write the output to a new file (csv)
@@
       SD
+    ) %>%
+    # Get only raw dates between 1950 and 150 BP
+    dplyr::filter(
+      Date_BP > 150,
+      Date_BP <= 1950
     ) %>%
@@
     breaks = seq(0, 500, 100),
-    expand = expand_scale(mult = 0, add = 0),
+    expand = expansion(mult = 0, add = 0),
     guide = guide_colorbar(title.position = "top")
@@
     breaks = seq(0, 5000, 1000),
-    expand = expand_scale(mult = 0, add = 0),
+    expand = expansion(mult = 0, add = 0),
     guide = guide_colorbar(title.position = "top")
@@
     breaks = seq(0, 80, 20),
-    expand = expand_scale(mult = 0, add = 0),
+    expand = expansion(mult = 0, add = 0),
     guide = guide_colorbar(title.position = "top")
@@
     breaks = seq(0, 250, 50),
-    expand = expand_scale(mult = 0, add = 0),
+    expand = expansion(mult = 0, add = 0),
     guide = guide_colorbar(title.position = "top")
@@
 We calibrated the radiocarbon dates using the methods of @Bronk2008, using the 'intcal13' calibration curve.
-To control for inconsistent sample size across the archaeological sites represented in the database, we aggregated calibrated radiocarbon dates known to be from the same phase of same archaeological site using the `rcarbon::binprep` function with an `h`-value of 100.
-This effectively clusters dates from the same site that are within 100 years of each other, and splits site occupations into distinct phases separated by 100 years or more.
+To control for inconsistent sample size across the archaeological sites represented in the database, we aggregated calibrated radiocarbon dates known to be from the same phase of same archaeological site using the `robinson2020::binPrep` function using the "single linkage" agglomeration method with an `h`-value of 100 .
+This clusters dates from the same site whose mean values (uncalibrated) are within 100 years of each other, and splits site occupations into distinct phases separated by 100 years or more.
 We constructed summed probability distributions (SPDs) from the calibrated dates using the `rcarbon::spd` function.
-We averaged probability distributions of binned dates, and then constructed SPDs from the site-phase distributions between 1949--0 BP (1--1950 AD).
-We smoothed the SPD using a 21-year moving average, and normalised the total probability mass of the smoothed SPD to sum to unity.
+We averaged probability distributions of binned dates, and then constructed SPDs from the site-phase distributions.
 
 ```{r radiocarbon-spds}
-# Set the time range for the reconstructions
-timerange <- c(1949, 0)
+timerange <- c(10000, 0)
 
+radiocarbon_spd <-
   all_data$Radiocarbon %>%
@@
       )),
-    Bins = list(rcarbon::binPrep(
+    Bins = list(robinson2020::binPrep(
       sites = Dates$SiteID,
-      ages = Calibrated,
+      ages = Dates$Date_BP,
       h = 100
@@
       bins = Bins,
+      timeRange = timerange,
       spdnormalised = TRUE,
-      timeRange = timerange,
-      runm = 21,
       verbose = FALSE
@@
           Date_BP = calBP,
-          Density = PrDens / sum(PrDens, na.rm = T)
+          Density = PrDens
         )
-    })) %>%
+    })) %T>%
   readr::write_rds(
@@
     compress = "gz"
-  )
-
-radiocarbon_spd <- readr::read_rds(
-  here::here("analysis/data/derived_data/radiocarbon_spd.rds")
   )
@@
 The probability mass function in such a case would be zero for years prior to the date of the outer ring of the sample, and non-zero for subsequent years, presumably decaying quickly.
-For instance, we could imagine a study that derives empirical probability distributions for non-cutting dates based on cutting-dates from the same contexts.] Here, we did just that, binning the dates into site-phases using the `rcarbon::binprep` function, and constructing SPDs using `rcarbon::spd`, both with the same parameter selections as used for the radiocarbon samples.
-As above, we smoothed normalised the SPD.
+For instance, we could imagine a study that derives empirical probability distributions for non-cutting dates based on cutting-dates from the same contexts.] Here, we did just that, binning the dates into site-phases using the `robinson2020::binPrep` function, and constructing SPDs using `rcarbon::spd`, both with the same parameter selections as used for the radiocarbon samples.
 
 ```{r treering-spds}
-calibrate_tree_ring_dates <- function(x) {
+calibrate_tree_ring_dates <- function(x, timeRange = c(50000, 0)) {
   out <- list(
@@
       ResErrors = 0,
-      StartBP = timerange[1],
-      EndBP = timerange[2],
+      StartBP = timeRange[1],
+      EndBP = timeRange[2],
       Normalised = TRUE,
@@
         ) %>%
-        dplyr::bind_rows(tibble::tibble(year_bp = timerange[1]:timerange[2])) %>%
+        dplyr::bind_rows(tibble::tibble(year_bp = timeRange[1]:timeRange[2])) %>%
         dplyr::filter(!duplicated(year_bp)) %>%
@@
 
+dendro_spd <-
   all_data$`Tree-ring` %>%
@@
     Calibrated = list(Dates %$%
-      calibrate_tree_ring_dates(x = Date_BP)),
-    Bins = list(rcarbon::binPrep(
+      calibrate_tree_ring_dates(
+        x = Date_BP,
+        timeRange = timerange
+      )),
+    Bins = list(robinson2020::binPrep(
       sites = Dates$SiteID,
@@
     compress = "gz"
-  )
-
-dendro_spd <- readr::read_rds(
-  here::here("analysis/data/derived_data/dendro_spd.rds")
   )
@@
 ```{r c14-tr-pop, fig.height=6.5, fig.width=6.5, fig.cap='Radiocarbon and tree-ring density distributions in the UUSW, CMV, and NRG regions, and VEP population estimates for the CMV and NRG.
-Y-axis labels are omitted for simplicity; the maximum population in the lower panel is ~30,000 people, and the minimum is 0.
-Dates are given in AD.', fig.show = 'hold', out.width = '75%'}
+The radiocarbon and tree-ring density distributions for the UUSW and CMV are normalized such that the area under their densities is equal to the area under the VEP population estimates (i.e., the cumulative population).
+Years are given in AD.', fig.show = 'hold', out.width = '75%'}
+
+# A smoothing distribution to use throughout the analyses
+# This is a 21-year wide Gaussian distribution with a
+# mean of 0 and standard deviation of 5 years.
+# This is the smoother used in Bocinsky et al. 2016.
+dist <- dnorm(seq(-10, 10, 1), sd = 5)
+
 all_density <-
@@
   dplyr::select(
-    `Year AD`,
-    Date_BP,
     `Study Area`,
     Type,
+    `Year AD`,
     Density
-  ) %>%
-  dplyr::rename(Value = Density)
+  )
 
@@
   robinson2020::vep_demography %>%
+  dplyr::mutate(End = ifelse(`Study Area` == "CMV" & Period == 20, 1300, End)) %>%
   dplyr::rowwise() %>%
-  dplyr::mutate(`Year AD` = list((Start + 1):End)) %>%
+  dplyr::mutate(`Year AD` = list(Start:(End - 1))) %>%
   dplyr::select(
@@
     `Year AD`,
+    `Period`,
     Population
   ) %>%
-  tidyr::unnest(`Year AD`) %>%
+  tidyr::unnest(`Year AD`)
+
+dplyr::left_join(all_density,
+  vep_demography,
+  by = c(
+    "Study Area",
+    "Year AD"
+  )
+) %>%
+  dplyr::group_by(
+    `Study Area`,
+    Type
+  ) %>%
+  dplyr::mutate(`Scaled Density` = Density * (sum(Population, na.rm = T) / sum(Density * !is.na(Population), na.rm = TRUE))) %>%
+  dplyr::group_by(
+    `Study Area`,
+    Type,
+    Period
+  ) %>%
+  # dplyr::mutate(`Binned Density` = ifelse(is.na(Population), NA, rep(mean(`Scaled Density`), dplyr::n()))) %>%
+  dplyr::group_by(
+    `Study Area`,
+    Type
+  ) %>%
   dplyr::mutate(
-    Type = "Population",
-    Date_BP = 1950 - `Year AD`
+    `Raw Density` = ifelse(`Study Area` == "UUSW", Density, `Scaled Density`),
+    `Smoothed Density` = ifelse(Type == "Tree-ring Density", stats::filter(`Raw Density`, filter = dist), `Raw Density`),
+    `Raw Density` = ifelse(`Study Area` == "UUSW", `Raw Density`, NA)
   ) %>%
-  dplyr::rename(Value = Population)
-
-all_density %>%
-  dplyr::bind_rows(vep_demography) %>%
+  dplyr::select(
+    `Study Area`,
+    Type,
+    `Year AD`,
+    Population,
+    `Raw Density`,
+    `Smoothed Density`,
+    # `Binned Density`
+  ) %>%
+  dplyr::ungroup() %>%
   dplyr::mutate(
@@
       levels = c(
+        "Tree-ring Density",
         "Radiocarbon Density",
-        "Tree-ring Density",
-        "Population"
+        "VEP Population"
       ),
@@
   geom_line(aes(
-    y = Value,
-    color = `Study Area`,
-    size = `Study Area`
+    y = `Raw Density`,
+    color = Type
+  ),
+  size = 0.5,
+  alpha = 0.5
+  ) +
+  geom_line(aes(
+    y = `Population`
+  ),
+  size = 1.25
+  ) +
+  geom_line(aes(
+    y = `Smoothed Density`,
+    color = Type
+  ),
+  size = 0.75
+  ) +
+  # geom_line(aes(
+  #   y = `Binned Density`,
+  #   color = Type
+  # ),
+  # size = 1.25) +
+
+  scale_color_manual(values = c(
+    "Tree-ring Density" = "#d95f02",
+    "Radiocarbon Density" = "#1b9e77",
+    "VEP Population" = "black"
   )) +
-  scale_color_manual(values = c(
-    "UUSW" = "black",
-    "CMV" = "gray50",
-    "NRG" = "gray80"
-  )) +
-  scale_size_manual(values = c(
-    "UUSW" = 1.25,
-    "CMV" = 0.75,
-    "NRG" = 0.75
-  )) +
-  # coord_cartesian(xlim = c(600,1600)) +
   xlab("Year AD") +
-  scale_x_continuous(breaks = c(1, seq(200, 2000, 200))) +
+  scale_x_continuous(
+    limits = c(200, 1800),
+    breaks = seq(200, 1800, 200),
+    expand = expansion(0, 0)
+  ) +
   theme_minimal(18) +
@@
     axis.title.y = element_blank(),
-    axis.text.y = element_blank(),
-    panel.grid.major.y = element_blank(),
-    panel.grid.minor.y = element_blank(),
+    # axis.text.y = element_blank(),
+    # panel.grid.major.y = element_blank(),
+    # panel.grid.minor.y = element_blank(),
     plot.margin = unit(c(0.25, 0.5, 0.25, 0.25), "in")
   ) +
-  facet_wrap("Type",
-    nrow = 3,
+  facet_grid(
+    rows = vars(`Study Area`),
     scales = "free_y"

### 2020-03-11 R/binPrep.R

@@ -1,0 +1,41 @@
+#' @title Binning function of radiocarbon dates.
+#'
+#' @description Prepare a set of bins for controlling the aggregation of radiocarbon dates
+#' known to be from the same phase of same archaeological site (for use with \code{\link{spd}}). Used in cases where there is a concern that unusually high levels of sampling for radiocarbon at a given site or in a given site phase will impede comparison between sites or phases.
+#'
+#' @param sites a vector of character strings (or number to coerce to character) of all sites or site phases. If character strings are used these should not contain underscores (see also below)
+#' @param ages a vector of uncalibrated conventional radiocarbon ages or a \code{CalDates} class object obtained using the \code{\link{calibrate}} function.
+#' @param h a single numeric value passed to \code{\link{cutree}} control degree of grouping of similar ages in a phase site.
+#' @param method the agglomeration method to be used, passed on to \code{\link{hclust}}. Defaults to "complete" as in \code{\link{hclust}}.
+#'
+#' @details If \code{ages} is a \code{CalDates} class object, median dates are used for the clustering.
+#'
+#' @return A vector of character strings with the same length of the object supplied for the argument \code{ages} identifying intra-site or intra-phase grouping, for use with \code{\link{spd}}.The character strings effectively provide a "name" for each "phase" within a "site", using sequential integers after an underscore. For example if a site named "S001" had four dates grouped into two bins with two dates each, the resulting vector would be "S001_1", "S001_1", "S001_2", and "S001_2".
+
+#' @seealso \code{\link{spd}} for generating SPD; \code{\link{binsense}} for sensitivity analysis pertaining the choice of the parameter \code{h}.
+#' @import stats
+#' @import utils
+#' @export
+binPrep <- function(sites, ages, h, method = "complete") {
+  if (any(grepl("_", sites))) {
+    stop("sites should not contain elements with underscores")
+  }
+
+  clusters <- rep(NA, length(sites))
+
+  if (any(class(ages) %in% "CalDates")) {
+    ages <- medCal(ages)
+  }
+
+  for  (x in 1:length(unique(sites))) {
+    index <- which(sites == unique(sites)[x])
+    if (length(index) > 1) {
+      clusters[index] <- paste(unique(sites)[x], cutree(hclust(dist(ages[index]), method = method), h = h), sep = "_")
+    }
+    if (length(index) == 1) {
+      clusters[index] <- paste(unique(sites)[x], "1", sep = "_")
+    }
+  }
+
+  return(clusters)
+}

### 2020-03-11 analysis/paper/paper.Rmd
removing density binning

@@
   ) %>%
-  # dplyr::mutate(`Binned Density` = ifelse(is.na(Population), NA, rep(mean(`Scaled Density`), dplyr::n()))) %>%
   dplyr::group_by(
@@
   ) +
-  # geom_line(aes(
-  #   y = `Binned Density`,
-  #   color = Type
-  # ),
-  # size = 1.25) +
-
   scale_color_manual(values = c(

### 2020-03-11 analysis/paper/paper.Rmd
refining figures

@@
 
-# A smoothing distribution to use throughout the analyses
 # This is a 21-year wide Gaussian distribution with a
@@
       levels = c(
-        `Upland US Southwest` = "UUSW",
-        `Central Mesa Verde` = "CMV",
-        `Northern Rio Grande` = "NRG"
-      )
+        "UUSW",
+        "CMV",
+        "NRG"
+      ),
+      labels = c(
+        "Upland US Southwest Density",
+        "Central Mesa Verde Population",
+        "Northern Rio Grande Population"
+      ),
+      ordered = TRUE
     )
@@
   ) +
-  scale_color_manual(values = c(
+  scale_color_manual(
+    values = c(
       "Tree-ring Density" = "#d95f02",
@@
       "VEP Population" = "black"
-  )) +
+    ),
+    limits = c(
+      "Tree-ring Density",
+      "Radiocarbon Density",
+      "VEP Population"
+    )
+  ) +
   xlab("Year AD") +
@@
   ) +
-  facet_grid(
-    rows = vars(`Study Area`),
+  facet_wrap(
+    vars(`Study Area`),
+    ncol = 1,
     scales = "free_y"

### 2020-03-11 analysis/paper/paper.Rmd
refining figures

@@
       email: bocinsky@gmail.com
-      institute: [CCAC]
+      institute: [CCAC,UM]
       correspondence: true
+  - Darcy Bird:
+      email: darcy.bird@wsu.edu
+      institute: [WSU]
+      correspondence: false
   - Erick Robinson:
@@
   - CCAC: Crow Canyon Archaeological Center
+  - UM: University of Montana
+  - WSU: Washington State University
   - USU: Utah State University
@@
   ) %>%
-  ggplot(aes(x = `Year AD`)) +
+  {
+    ggplot(
+      .,
+      aes(x = `Year AD`)
+    ) +
       geom_line(aes(
@@
         axis.title.y = element_blank(),
+        strip.text = element_text(size = 12),
         # axis.text.y = element_blank(),
@@
       facet_wrap(
+        # rows =
         vars(`Study Area`),
@@
         scales = "free_y"
+      )
+  } %>%
+  egg::tag_facet(
+    open = "",
+    close = "",
+    tag_pool = paste0(LETTERS, ". ", c(
+      "Upland US Southwest Density",
+      "Central Mesa Verde Population",
+      "Northern Rio Grande Population"
+    )),
+    hjust = 0,
+    vjust = 1.5,
+    size = 5
   )

### 2020-03-11 analysis/paper/paper.Rmd
re-running everything

@@
       ages = Dates$Date_BP,
-      h = 100
+      h = 100,
+      method = "single"
     )),
@@
         legend.position = "bottom",
+        legend.text = element_text(size = 12),
         axis.title.x = element_text(size = 12),
@@
         axis.title.y = element_blank(),
-        strip.text = element_text(size = 12),
-        # axis.text.y = element_blank(),
+        axis.text.y = element_text(size = 10),
         # panel.grid.major.y = element_blank(),
@@
     vjust = 1.5,
-    size = 5
+    size = 4
   )

### 2020-03-11 R/binPrep.R
referencing package in binPrep

@@ -24,7 +24,7 @@ binPrep <- function(sites, ages, h, method = "complete") {
   clusters <- rep(NA, length(sites))
 
   if (any(class(ages) %in% "CalDates")) {
-    ages <- medCal(ages)
+    ages <- rcarbon::medCal(ages)
   }
 
   for  (x in 1:length(unique(sites))) {

### 2020-03-18 analysis/paper/paper.Rmd
added bin-width sensitivity

@@
   cache = FALSE,
-  echo = TRUE,
+  echo = FALSE,
   fig.align = "center",
@@
   comment = "#>",
-  dev = "png",
-  dpi = 300
+  dev = "pdf" # ,
+  # dpi = 300
 )
@@
   robinson2020::vep_study_areas
+
+# Truncate calendar dates to AD 200 - 1800
+trunc_dates <- c(1750, 150)
 
@@
 # Read in the radiocarbon database
+# Filter the 2-sigma calibrated ranges to the trunc_dates range
 if (!file.exists(
@@
     ) %>%
-    # Get only raw dates between 1950 and 150 BP
+    # retain dates whose calibrated 95% probability mass overlaps trunc_dates
     dplyr::filter(
-      Date_BP > 150,
-      Date_BP <= 1950
+      rcarbon::calibrate(
+        x = Date_BP,
+        errors = SD
+      ) %>%
+        rcarbon:::hpdi(credMass = 0.95) %>%
+        purrr::map_lgl(function(x) {
+          startdate <- max(x[, "startCalBP"])
+          enddate <- min(x[, "endCalBP"])
+          startdate > trunc_dates[[2]] && enddate <= trunc_dates[[1]]
+        })
     ) %>%
@@
     ) %>%
-    # Get only raw dates between 1950 and 150 BP
+    # Get only raw dates within trunc_dates
     dplyr::filter(
-      Date_BP > 150,
-      Date_BP <= 1950
+      Date_BP > trunc_dates[[2]],
+      Date_BP <= trunc_dates[[1]]
     ) %>%
@@
     direction = 1,
-    limits = c(0, 500),
-    minor_breaks = seq(50, 450, 100),
-    breaks = seq(0, 500, 100),
+    limits = c(0, 200),
+    minor_breaks = seq(25, 175, 50),
+    breaks = seq(0, 200, 50),
     expand = expansion(mult = 0, add = 0),
@@
 We calibrated the radiocarbon dates using the methods of @Bronk2008, using the 'intcal13' calibration curve.
-To control for inconsistent sample size across the archaeological sites represented in the database, we aggregated calibrated radiocarbon dates known to be from the same phase of same archaeological site using the `robinson2020::binPrep` function using the "single linkage" agglomeration method with an `h`-value of 100 .
+To control for inconsistent sample size across the archaeological sites represented in the database, we aggregated calibrated radiocarbon dates known to be from the same phase of same archaeological site using the `robinson2020::binPrep` function using the "single linkage" agglomeration method with an `h`-value of 100.
 This clusters dates from the same site whose mean values (uncalibrated) are within 100 years of each other, and splits site occupations into distinct phases separated by 100 years or more.
@@
 # This is the smoother used in Bocinsky et al. 2016.
-dist <- dnorm(seq(-10, 10, 1), sd = 5)
+smoother <- dnorm(seq(-10, 10, 1), sd = 5)
 
@@
     `Raw Density` = ifelse(`Study Area` == "UUSW", Density, `Scaled Density`),
-    `Smoothed Density` = ifelse(Type == "Tree-ring Density", stats::filter(`Raw Density`, filter = dist), `Raw Density`),
-    `Raw Density` = ifelse(`Study Area` == "UUSW", `Raw Density`, NA)
+    `Smoothed Density` = ifelse(Type == "Tree-ring Density", stats::filter(`Raw Density`, filter = smoother), `Raw Density`),
+    `Raw Density` = ifelse(`Study Area` == "UUSW", `Raw Density`, NA),
+    `Raw Density` = ifelse(Type == "Tree-ring Density", `Raw Density`, NA)
   ) %>%
@@
   ) %>%
+  dplyr::mutate(`Smoothed Density` = ifelse(`Study Area` == "Upland US Southwest Density" & Type == "Radiocarbon Density",
+    (`Smoothed Density` * 3),
+    `Smoothed Density`
+  )) %>%
   {
@@
 
+## Supplementary material: Bin-width sensitivity {-}
+```{r bin-sensitivity, fig.height=6.5, fig.width=6.5, fig.cap='Radiocarbon (A) and tree-ring (B) density distributions in the UUSW, given varying widths for binning dates within sites.
+The density distributions for the UUSW are normalized.
+Years are given in AD.', fig.show = 'hold', out.width = '75%'}
+densities <- c(0, 25, 50, 100, 150, 200)
+
+spds <-
+  radiocarbon_spd %>%
+  dplyr::filter(`Study Area` == "UUSW") %>%
+  dplyr::bind_rows(dendro_spd %>%
+    dplyr::filter(`Study Area` == "UUSW")) %>%
+  dplyr::mutate(Type = c(
+    "Radiocarbon",
+    "Tree-ring"
+  )) %>%
+  dplyr::select(Type, Dates, Calibrated) %>%
+  dplyr::rowwise() %>%
+  dplyr::mutate(
+    Densities =
+      list(
+        densities %>%
+          magrittr::set_names(., .) %>%
+          purrr::map_dfr(~ rcarbon::spd(
+            x = Calibrated,
+            timeRange = timerange,
+            bins =
+              robinson2020::binPrep(
+                sites = Dates$SiteID,
+                ages = Dates$Date_BP,
+                h = .x,
+                method = "single"
+              ),
+            spdnormalised = TRUE,
+            verbose = FALSE
+          ) %$%
+            grid %>%
+            tibble::as_tibble() %$%
+            tibble::tibble(
+              Date_AD = 1950 - calBP,
+              Density = PrDens
+            ),
+          .id = "h"
+          ) %>%
+          dplyr::mutate(
+            `Bin Width (h)` =
+              factor(h,
+                levels = densities,
+                ordered = TRUE
+              )
+          ) %>%
+          dplyr::filter(Date_AD > 0) %>%
+          dplyr::select(`Bin Width (h)`,
+            `Year AD` = Date_AD,
+            Density
+          )
+      )
+  ) %>%
+  dplyr::select(Type, Densities) %>%
+  tidyr::unnest(cols = c(Densities))
+
+spds %>%
+  dplyr::group_by(Type, `Bin Width (h)`) %>%
+  dplyr::mutate(
+    Density =
+      ifelse(Type == "Tree-ring",
+        stats::filter(Density,
+          filter = smoother
+        ),
+        Density
+      )
+  ) %>%
+  dplyr::ungroup() %>%
+  {
+    ggplot2::ggplot(
+      .,
+      ggplot2::aes(
+        x = `Year AD`,
+        y = Density,
+        color = `Bin Width (h)`
+      )
+    ) +
+      ggplot2::geom_line() +
+      ggplot2::facet_wrap("Type",
+        nrow = 2,
+        scales = "free_y"
+      ) +
+      xlab("Year AD") +
+      scale_x_continuous(
+        limits = c(200, 1800),
+        breaks = seq(200, 1800, 200),
+        expand = expansion(0, 0)
+      ) +
+      scale_y_continuous(
+        limits = c(0, NA),
+        expand = expansion(0, 0)
+      ) +
+      theme_minimal(18) +
+      theme(
+        legend.title = element_text(size = 12),
+        # legend.justification = c(0, 1),
+        legend.position = "bottom",
+        legend.text = element_text(size = 12),
+        axis.title.x = element_text(size = 12),
+        axis.text.x = element_text(size = 10),
+        axis.title.y = element_blank(),
+        axis.text.y = element_text(size = 10),
+        # panel.grid.major.y = element_blank(),
+        # panel.grid.minor.y = element_blank(),
+        plot.margin = unit(c(0.25, 0.5, 0.25, 0.25), "in")
+      ) +
+      guides(color = guide_legend(nrow = 1))
+  } %>%
+  egg::tag_facet(
+    open = "",
+    close = "",
+    tag_pool = paste0(LETTERS, ". ", c(
+      "Radiocarbon Dates",
+      "Tree-ring Dates"
+    )),
+    hjust = 0,
+    vjust = 1.5,
+    size = 4
+  )
+```
+
 <!-- The following line inserts a page break  -->

### 2020-03-24 Untitled/Untitled.Rmd
working towards resubmission

---
title: Template for preparing your research report submission to Royal Society Open Science using RMarkdown
author:
  - name: Alice Anonymous
    affiliation: "1,2"
  - name: Bob Security
    affiliation: "2"
address:
  - code: "1"
    address: Some Institute of Technology, Department, Street, City, State, Zip
  - code: "2"
    address: Another University Department, Street, City, State, Zip

corresp_author_name:  "B. Security"
corresp_author_email: "bob@email.com"

subject:
  - "subject 1"
  - "subject 2"
  - "subject 3"

keywords:
  - one
  - two
  - optional
  - optional
  - optional

abstract: |
  The abstract text goes here. The abstract text goes here. The abstract text goes here. The abstract text goes here. The abstract text goes here. The abstract text goes here. The abstract text goes here. The abstract text goes here.

## Remove this if not required
ethics: |
  Please provide details on the ethics.

data_accessibility: |
  Please provide details on the data availability.

author_contributions: |
  Please provide details of author contributions here.

## Remove this if not required
conflict_of_interest: |
  Please declare any conflict of interest here.

## Remove this if not required
funding: |
  Please provide details on funding

## Remove this if not required
disclaimer: |
  Please provide disclaimer text here.

## Remove this if not required
acknowledgements: |
  Please include your acknowledgments here, set in a single paragraph. Please do not include any acknowledgments in the Supporting Information, or anywhere else in the manuscript.

bibliography: sample.bib

## change to true to add optional line numbering
lineno: false

site: bookdown::bookdown_site
output: 
  bookdown::pdf_book:
    base_format: rticles::rsos_article
---

# Insert A head here

This demo file is intended to serve as a "starter file"" for articles submitted to the [Royal Society Open Science](http://rsos.royalsocietypublishing.org/) journal using `RMarkdown`.

Place `\EndFirstPage` at the point where the plain text on the first page stops. Warning: excess text will be hidden behind the copyright box. The example below contains line 1 to 19 in the code. Lines 14 to 17 are hidden behind the copyright box.

## Insert B head here

Subsection text here.

### Insert C head here

Subsubsection text here.

Line 1

Line 2

Line 3

Line 4

Line 5

Line 6

Line 7

Line 8

Line 9

Line 10

Line 11

Line 12

Line 13

Line 14

Line 15

Line 16

Line 17

Line 18

Line 19

\EndFirstPage

# Lists

* one
* two
* three

* fruits
    + apples
        - macintosh
        - red delicious
    + pears
    + peaches
* vegetables
    + broccoli
    + chard

## Citations

Blabla \cite{Lannes} blabla. Blabla \cite{HJ2} blabla. Blabla \cite{BF, Lannes} blabla. Blabla \cite{Benjamin1967, HJ2, HJ3, HP2} blabla.

### Headling level 3

Subsubsection text here.

# R code

R code can be added as usual. Note that syntax highlighting is not available. Fig. \@ref(fig:nice-plot) is a an example.

```{r echo = TRUE}
# 23456789012345678901234567890123456789012345678901234567890123456789012345
#       10        20        30        40        50        60        70
summary(lm(mpg ~ disp, data = mtcars))
```

```{r}
summary(mtcars)
```

```{r nice-plot, fig.cap="The caption"}
plot(mtcars[, 1:3])
```

### 2020-03-24 analysis/robinson_et_al_2020.Rmd
working towards resubmission

--- 2020-03-18 analysis/paper/paper.Rmd
@@
 ---
-title: "Robinson et al. 2020 Supplemental Information: Comparing the Radiocarbon and Tree-ring Records for the Upland US Southwest"
+title: "Dendrochronological dates confirm a Late Prehistoric population decline in the American Southwest derived from radiocarbon dates"
 author:
+  - Erick Robinson:
+      email: erick.robinson@usu.edu
+      institute: [USU]
+      correspondence: true
   - Kyle Bocinsky:
@@
       institute: [CCAC,UM]
-      correspondence: true
+      correspondence: false
   - Darcy Bird:
@@
       correspondence: false
-  - Erick Robinson:
-      email: erick.robinson@usu.edu
+  - Jacob Freeman:
+      email: jacob.freeman@usu.edu
       institute: [USU]
+      correspondence: false
+  - Robert L. Kelly:
+      email: rlkelly@uwyo.edu 
+      institute: [UW]
       correspondence: false
@@
   - USU: Utah State University
+  - UW: University of Wyoming
 date: "`r format(Sys.time(), '%d %B, %Y')`"
@@
       pandoc_args:
-      - --lua-filter=../templates/scholarly-metadata.lua
-      - --lua-filter=../templates/author-info-blocks.lua
-      - --lua-filter=../templates/pagebreak.lua
-    bookdown::word_document2:
-      reference_docx: "../templates/template.docx"
-      code_folding: hide
+      - --lua-filter=templates/scholarly-metadata.lua
+      - --lua-filter=templates/author-info-blocks.lua
+      - --lua-filter=templates/pagebreak.lua
+    bookdown::pdf_document2:
+      # code_folding: hide
       fig_caption: yes
       pandoc_args:
-      - --lua-filter=../templates/scholarly-metadata.lua
-      - --lua-filter=../templates/author-info-blocks.lua
-      - --lua-filter=../templates/pagebreak.lua
-bibliography: references.bib
-csl: "../templates/philosophical-transactions-of-the-royal-society-b.csl" # Insert path for the bib-style
+      - --lua-filter=templates/scholarly-metadata.lua
+      - --lua-filter=templates/author-info-blocks.lua
+      - --lua-filter=templates/pagebreak.lua
+bibliography: robinson_et_al_references.bib
+csl: "templates/philosophical-transactions-of-the-royal-society-b.csl"
 editor_options: 
@@
   fig.align = "center",
-  fig.path = "../figures/",
+  fig.path = here::here("analysis/figures/"),
   message = FALSE,
@@
 We also analyze subsets of both datasets for the Village Ecodynamic Project (VEP) study areas in the Central Mesa Verde (CMV) and Northern Rio Grande (NRG) regions [@Schwindt2016].
-Figure \@ref(fig:study-area) shows the original UUSW from @Bocinsky2016_SA and revised study area used in this analysis, as well as the two VEP study areas.
-
-```{r study-area, fig.height=6.5, fig.width=6.5, out.width = '50%', fig.cap="The UUSW as defined in this study (region with black border).
+Figure \@ref(fig:Figure1) shows the original UUSW from @Bocinsky2016_SA and revised study area used in this analysis, as well as the two VEP study areas.
+
+```{r Figure1, fig.height=6.5, fig.width=6.5, out.width = '50%', fig.cap="The UUSW as defined in this study (region with black border).
 The white dashed area represents the UUSW as defined in @Bocinsky2016_SA.
@@
   readr::read_csv(
-    here::here("analysis/data/raw_data/0403_sbox_clean_sd300.csv")
-  ) %>%
-    # Convert to spatial object
-    sf::st_as_sf(
-      coords = c("Long", "Lat"),
-      crs = 4326
-    ) %>%
-    sf::st_transform(four_corners_lcc_proj) %>%
-    # Select salient columns
-    dplyr::select(
-      labnumber,
-      date,
-      sd,
-      Site,
-      SiteID
-    ) %>%
-    dplyr::rename(
-      Lab_Number = labnumber,
-      Date_BP = date,
-      SD = sd,
-      Site_Number = Site
-    ) %>%
-    sf::st_intersection(uusw_counties) %>%
-    sf::st_drop_geometry() %>%
+    here::here("analysis/data/raw_data/robinson_et_al_2020_radiocarbon.csv")
+  ) %>%
     dplyr::mutate(
       Type = "Radiocarbon",
-      SiteID = SiteID %>% stringr::str_c("Radiocarbon - ", .)
+      Site_ID = Site_ID %>% stringr::str_c("Radiocarbon - ", .)
     ) %>%
@@
       Type,
-      SiteID,
+      SiteID = Site_ID,
       Site_Number,
@@
 
-```{r}
+```{r all-data}
 all_data <- list(
@@
 Our final dataset includes `r nrow(all_data$Radiocarbon) %>% format(big.mark = ',')` radiocarbon dates from `r max(all_data$Radiocarbon$Date_BP)`--`r min(all_data$Radiocarbon$Date_BP)` BP (uncalibrated; `r -1 * (1950 - max(all_data$Radiocarbon$Date_BP)) - 1` BC--`r 1950 - min(all_data$Radiocarbon$Date_BP)` AD) and `r nrow(all_data[['Tree-ring']]) %>% format(big.mark = ',')` tree-ring dates from `r max(all_data[['Tree-ring']]$Date_BP)` -- `r min(all_data[['Tree-ring']]$Date_BP)` BP (`r 1950 - max(all_data[['Tree-ring']]$Date_BP)`--`r 1950 - min(all_data[['Tree-ring']]$Date_BP)` AD).
-Table \@ref(tab:data-summary) presents the counts of each date type in the CMV and NRG study areas.
-
-```{r data-summary}
+Table \@ref(tab:Table1) presents the counts of each date type in the CMV and NRG study areas.
+
+```{r Table1}
 all_data %>%
@@
   )) %>%
+  dplyr::group_by(Type, SiteID, `Study Area`) %>%
+  dplyr::summarise(Dates = dplyr::n()) %>%
   dplyr::group_by(Type, `Study Area`) %>%
-  dplyr::summarise(Count = dplyr::n()) %>%
-  tidyr::spread(Type, Count) %>%
+  dplyr::summarise(
+    Dates = sum(Dates),
+    Sites = dplyr::n()
+  ) %>%
+  tidyr::pivot_wider(names_from = Type, values_from = Dates:Sites) %>%
+  dplyr::select(`Study Area`, Dates_Radiocarbon, Sites_Radiocarbon, `Dates_Tree-ring`, `Sites_Tree-ring`) %>%
   knitr::kable(
-    caption = "Counts of radiocarbon and tree-ring dates in this study.",
+    caption = "Counts of radiocarbon and tree-ring dates and sites in this study.",
     booktabs = TRUE,
@@
 Still, these disparities likely demonstrate more fundemental differences in research design across archaeological projects.
-Figure \@ref(fig:date-densities) presents the number of radiocarbon and tree-ring dates by county within our study area.
-Figure \@ref(fig:site-densities) presents the number of unique sites with radiocarbon and tree-ring dates by county within our study area.
-
-```{r date-densities, fig.height=9, fig.width=6.5, fig.cap='The number of radiocarbon and tree-ring dates by county in this study.', fig.show = 'hold', out.width = '50%'}
+Figure \@ref(fig:Figure2) presents the number of radiocarbon and tree-ring dates by county within our study area.
+Figure \@ref(fig:Figure3) presents the number of unique sites with radiocarbon and tree-ring dates by county within our study area.
+
+```{r Figure2, fig.height=9, fig.width=6.5, fig.cap='The number of radiocarbon and tree-ring dates by county in this study.', fig.show = 'hold', out.width = '50%'}
 county_counts <-
@@
 
-```{r site-densities, fig.height=9, fig.width=6.5, fig.cap='The number of sites with radiocarbon and tree-ring dates by county in this study.', fig.show = 'hold', out.width = '50%'}
+```{r Figure3, fig.height=9, fig.width=6.5, fig.cap='The number of sites with radiocarbon and tree-ring dates by county in this study.', fig.show = 'hold', out.width = '50%'}
 county_counts <-
@@
 We calibrated the radiocarbon dates using the methods of @Bronk2008, using the 'intcal13' calibration curve.
-To control for inconsistent sample size across the archaeological sites represented in the database, we aggregated calibrated radiocarbon dates known to be from the same phase of same archaeological site using the `robinson2020::binPrep` function using the "single linkage" agglomeration method with an `h`-value of 100.
+To control for inconsistent sample size across the archaeological sites represented in the database, we aggregated calibrated radiocarbon dates known to be from the same phase of same archaeological site using the `rcarbon::binPrep` function using the "single linkage" agglomeration method with an `h`-value of 100.
 This clusters dates from the same site whose mean values (uncalibrated) are within 100 years of each other, and splits site occupations into distinct phases separated by 100 years or more.
@@
       )),
-    Bins = list(robinson2020::binPrep(
+    Bins = list(rcarbon::binPrep(
       sites = Dates$SiteID,
@@
 The probability mass function in such a case would be zero for years prior to the date of the outer ring of the sample, and non-zero for subsequent years, presumably decaying quickly.
-For instance, we could imagine a study that derives empirical probability distributions for non-cutting dates based on cutting-dates from the same contexts.] Here, we did just that, binning the dates into site-phases using the `robinson2020::binPrep` function, and constructing SPDs using `rcarbon::spd`, both with the same parameter selections as used for the radiocarbon samples.
+For instance, we could imagine a study that derives empirical probability distributions for non-cutting dates based on cutting-dates from the same contexts.] Here, we did just that, binning the dates into site-phases using the `rcarbon::binPrep` function, and constructing SPDs using `rcarbon::spd`, both with the same parameter selections as used for the radiocarbon samples.
 
@@
 
-```{r c14-tr-pop, fig.height=6.5, fig.width=6.5, fig.cap='Radiocarbon and tree-ring density distributions in the UUSW, CMV, and NRG regions, and VEP population estimates for the CMV and NRG.
+```{r Figure4, fig.height=6.5, fig.width=6.5, fig.cap='Radiocarbon and tree-ring density distributions in the UUSW, CMV, and NRG regions, and VEP population estimates for the CMV and NRG.
 The radiocarbon and tree-ring density distributions for the UUSW and CMV are normalized such that the area under their densities is equal to the area under the VEP population estimates (i.e., the cumulative population).
@@
 
-## Supplementary material: Bin-width sensitivity {-}
-```{r bin-sensitivity, fig.height=6.5, fig.width=6.5, fig.cap='Radiocarbon (A) and tree-ring (B) density distributions in the UUSW, given varying widths for binning dates within sites.
+\beginsupplement
+# Cut-width sensitivity for temporal binning {-}
+Uneven sampling across archaeological contexts is a persistant problem when constructing summed probability distributions from radiocarbon dates, and perhaps even a larger issue for the more ubiquitous tree-ring dates in regions like the Upland US Southwest.
+In SPDs of radiocarbon dates, it is conventional to combine dates from single sites prior to creating the SPDs.
+Dates from separate components or phases of multi-component sites should be counted separately, however.
+
+To control for inconsistent sample size across the archaeological sites represented in the database, we aggregated dates known to be from the same phase of same archaeological site using the `rcarbon::binPrep` function using the "single linkage" agglomeration method with a cut width (or $h$-value) of 100 [@rcarbon].
+This clusters dates from the same site that are within 100 years of each other, and splits site occupations into distinct phases separated by 100 years or more.
+Binned calibrated dates are then summed, and the resulting probability mass function is divided by the number of dates in the site-phase.
+
+The choice of $h$ tends to be more art than science, but here we perform a sensitivity analysis to demonstrate that the resulting SPDs for radiocarbon and tree-ring dates stabilize at a cut width of 100 or so.
+We calculated SPDs at 25-year intervals from 0 (only duplicate dates are binned) to 150 years (effectively binning by site).
+
+As can be seen in Figure \@ref(fig:FigureS1), binning in general has little impact on the resulting SPDs, but cut widths above $h=100$ years are virtually indistinguishable.
+The shape of the radiocarbon SPD (panel A) stabilizes above $h=0$ and its values stabilize by $h=100$.
+The tree-ring SPD arguably stabilizes in both shape and value above $h=75$ or so, but in the analysis we use $h=100$ to keep our methods consistent between the radiocarbon and tree-ring SPDs.
+
+```{r FigureS1, fig.height=6.5, fig.width=6.5, fig.cap='Radiocarbon (A) and tree-ring (B) density distributions in the UUSW, given varying cut widths for binning dates within sites.
 The density distributions for the UUSW are normalized.
 Years are given in AD.', fig.show = 'hold', out.width = '75%'}
-densities <- c(0, 25, 50, 100, 150, 200)
+cut_widths <- seq(0, 150, 25)
+
+radiocarbon_spd <-
+  here::here("analysis/data/derived_data/radiocarbon_spd.rds") %>%
+  readr::read_rds()
+
+dendro_spd <-
+  here::here("analysis/data/derived_data/dendro_spd.rds") %>%
+  readr::read_rds()
+
+# This is a 21-year wide Gaussian distribution with a
+# mean of 0 and standard deviation of 5 years.
+# This is the smoother used in Bocinsky et al. 2016.
+smoother <- dnorm(seq(-10, 10, 1), sd = 5)
+
+timerange <- c(10000, 0)
 
@@
       list(
-        densities %>%
+        cut_widths %>%
           magrittr::set_names(., .) %>%
@@
             bins =
-              robinson2020::binPrep(
+              rcarbon::binPrep(
                 sites = Dates$SiteID,
@@
           dplyr::mutate(
-            `Bin Width (h)` =
+            `Cut Width (h)` =
               factor(h,
-                levels = densities,
+                levels = cut_widths,
                 ordered = TRUE
@@
           dplyr::filter(Date_AD > 0) %>%
-          dplyr::select(`Bin Width (h)`,
+          dplyr::select(`Cut Width (h)`,
             `Year AD` = Date_AD,
@@
 spds %>%
-  dplyr::group_by(Type, `Bin Width (h)`) %>%
+  dplyr::group_by(Type, `Cut Width (h)`) %>%
   dplyr::mutate(
@@
         y = Density,
-        color = `Bin Width (h)`
+        color = `Cut Width (h)`
       )
@@
         legend.title = element_text(size = 12),
-        # legend.justification = c(0, 1),
         legend.position = "bottom",
@@
         axis.text.y = element_text(size = 10),
-        # panel.grid.major.y = element_blank(),
-        # panel.grid.minor.y = element_blank(),
         plot.margin = unit(c(0.25, 0.5, 0.25, 0.25), "in")
@@
 
-## References {-}
+# References {-}
 <!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->
@@
 
-## Colophon {-}
+# Colophon {-}
 

### 2020-03-24 R/binPrep.R

@@ -1,41 +1,0 @@
-#' @title Binning function of radiocarbon dates.
-#'
-#' @description Prepare a set of bins for controlling the aggregation of radiocarbon dates
-#' known to be from the same phase of same archaeological site (for use with \code{\link{spd}}). Used in cases where there is a concern that unusually high levels of sampling for radiocarbon at a given site or in a given site phase will impede comparison between sites or phases.
-#'
-#' @param sites a vector of character strings (or number to coerce to character) of all sites or site phases. If character strings are used these should not contain underscores (see also below)
-#' @param ages a vector of uncalibrated conventional radiocarbon ages or a \code{CalDates} class object obtained using the \code{\link{calibrate}} function.
-#' @param h a single numeric value passed to \code{\link{cutree}} control degree of grouping of similar ages in a phase site.
-#' @param method the agglomeration method to be used, passed on to \code{\link{hclust}}. Defaults to "complete" as in \code{\link{hclust}}.
-#'
-#' @details If \code{ages} is a \code{CalDates} class object, median dates are used for the clustering.
-#'
-#' @return A vector of character strings with the same length of the object supplied for the argument \code{ages} identifying intra-site or intra-phase grouping, for use with \code{\link{spd}}.The character strings effectively provide a "name" for each "phase" within a "site", using sequential integers after an underscore. For example if a site named "S001" had four dates grouped into two bins with two dates each, the resulting vector would be "S001_1", "S001_1", "S001_2", and "S001_2".
-
-#' @seealso \code{\link{spd}} for generating SPD; \code{\link{binsense}} for sensitivity analysis pertaining the choice of the parameter \code{h}.
-#' @import stats
-#' @import utils
-#' @export
-binPrep <- function(sites, ages, h, method = "complete") {
-  if (any(grepl("_", sites))) {
-    stop("sites should not contain elements with underscores")
-  }
-
-  clusters <- rep(NA, length(sites))
-
-  if (any(class(ages) %in% "CalDates")) {
-    ages <- rcarbon::medCal(ages)
-  }
-
-  for  (x in 1:length(unique(sites))) {
-    index <- which(sites == unique(sites)[x])
-    if (length(index) > 1) {
-      clusters[index] <- paste(unique(sites)[x], cutree(hclust(dist(ages[index]), method = method), h = h), sep = "_")
-    }
-    if (length(index) == 1) {
-      clusters[index] <- paste(unique(sites)[x], "1", sep = "_")
-    }
-  }
-
-  return(clusters)
-}

### 2020-04-17 README.Rmd
updated readme
preparing preprint

@@
 R. Kyle Bocinsky [![](https://orcid.org/sites/default/files/images/orcid_16x16.png)](https://orcid.org/0000-0003-1862-3428),
+Darcy Bird [![](https://orcid.org/sites/default/files/images/orcid_16x16.png)](https://orcid.org/0000-0003-3466-6284), 
 Jacob Freeman [![](https://orcid.org/sites/default/files/images/orcid_16x16.png)](https://orcid.org/0000-0001-7402-8450), 
-Darcy Bird [![](https://orcid.org/sites/default/files/images/orcid_16x16.png)](https://orcid.org/0000-0003-3466-6284), 
 Robert L. Kelly[![](https://orcid.org/sites/default/files/images/orcid_16x16.png)](https://orcid.org/0000-0001-9737-0152), 
-(2020). _`r Title`_. Under review.
-
-<!-- Our pre-print is online here: -->
-
-<!-- > Authors, (YYYY). _`r Title`_. Name of journal/book, Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx> -->
+(2020). _`r Title`_. *Philosophical Transactions B*. In Press.
 
@@
 
-> Bocinsky, R. Kyle and Erick Robinson, (`r format(Sys.Date(), "%Y")`). _Compendium of R code and data for `r Title`_. Accessed `r format(Sys.Date(), "%d %b %Y")`.
+> Bocinsky, R. Kyle, Darcy Bird, and Erick Robinson, (`r format(Sys.Date(), "%Y")`). _Compendium of R code and data for `r Title`_. Accessed `r format(Sys.Date(), "%d %b %Y")`.
 
@@
 
-The **analysis** directory contains:
+The [:file_folder: analysis](/analysis) directory contains:
 
@@
 After unzipping, open the `Proj` file in RStudio, and run `devtools::install()` to ensure you have the packages this analysis depends on (also listed in the [DESCRIPTION](/DESCRIPTION) file).
-Then open `paper.Rmd` and knit to produce the `paper.docx`.
+Then open `robinson2020.Rmd` and knit to produce the paper in PDF format.
 

### 2020-04-17 analysis/robinson2020.Rmd
preparing preprint

---
title: 
    main: Dendrochronological dates confirm a Late Prehistoric population decline in the American Southwest derived from radiocarbon dates
    short: Dendrochronological and radiocarbon dates in the American Southwest
date: "`r format(lubridate::as_date('2020-04-16'), '%d %B, %Y')`"
author:
  - Erick Robinson:
      email: erick.robinson@usu.edu
      institute: [USU]
      correspondence: true
  - Kyle Bocinsky:
      email: kbocinsky@crowcanyon.org
      institute: [CCAC,UM]
      correspondence: false
  - Darcy Bird:
      email: darcy.bird@wsu.edu
      institute: [WSU]
      correspondence: false
  - Jacob Freeman:
      email: jacob.freeman@usu.edu
      institute: [USU]
      correspondence: false
  - Robert L. Kelly:
      email: rlkelly@uwyo.edu 
      institute: [UW]
      correspondence: false
institute:
  - USU: Utah State University
  - CCAC: Crow Canyon Archaeological Center
  - UM: University of Montana
  - WSU: Washington State University
  - UW: University of Wyoming
abstract: >
  The northern American Southwest provides one of the most well-documented cases of human population growth and decline in the world. The geographic extent of this decline in North America is unknown due to the lack of high-resolution palaeodemographic data from regions across and beyond the greater Southwest, where archaeological radiocarbon data is often the only available proxy for investigating these palaeodemographic processes. Radiocarbon time series across and beyond the greater Southwest suggest widespread population collapses from AD 1300--1600. However, radiocarbon data have potential biases caused by variable radiocarbon sample preservation, sample collection, and the non-linearity of the radiocarbon calibration curve. In order to be confident in the wider trends seen in radiocarbon time series across and beyond the greater Southwest, here we focus on regions that have multiple palaeodemographic proxies and compare those proxies to radiocarbon time series. We develop a new method for time series analysis and comparison between dendrochronological data and radiocarbon data. Results confirm a multiple proxy decline in human populations across the Upland US Southwest, Central Mesa Verde, and Northern Rio Grande from AD 1300--1600. These results lend confidence to single proxy radiocarbon-based reconstructions of paleodemography outside the Southwest that suggest post-AD 1300 population declines in many parts of North America.
keywords:
    formatted:
      - dates-as-data
      - radiocarbon summed probability distributions
      - dendrochronology
      - palaeodemography
      - southwestern United States
doi: [DOI HERE]
journal: Preprint
documentclass: sa
bibliography: robinson2020_references.bib
csl: "philosophical-transactions-of-the-royal-society-b.csl"
link-citations: true
output: 
    bookdown::pdf_document2:
      template: templates/template.tex
      latex_engine: xelatex
      keep_tex: true
      fig_caption: yes
      pandoc_args:
      - --lua-filter=templates/scholarly-metadata.lua
      - --lua-filter=templates/author-info-blocks.lua
      - --lua-filter=templates/pagebreak.lua
editor_options: 
  chunk_output_type: console
---

```{r setup, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(
  cache = TRUE,
  echo = FALSE,
  fig.align = "center",
  fig.path = here::here("analysis/figures/"),
  message = FALSE,
  warning = FALSE,
  collapse = TRUE,
  comment = "#>",
  dev = "pdf"
)
library(robinson2020)
library(ggplot2)
```

\newpage
<!-- \raggedright -->
# Introduction {-}
The spatial and temporal ubiquity of archaeological radiocarbon dates makes them an unprecedented proxy for human paleodemography. The past decade has witnessed a rapid increase in the use of archaeological radiocarbon time series to reconstruct prehistoric population growth and decline in various regions of the world [@Bevan2019; @Crema2016; @Goldberg2016; @Palmisano2019; @Peros2010; @Russell2009; @Shennan2013; @Wang2014; @Williams2012]. One of the first aggregations of radiocarbon dates for palaeodemographic analysis was by Berry [@Berry1982] in his study of population growth, decline and migration across the American Southwest. Radiocarbon time series were used as a complement to earlier periods that lacked robust dendrochronological records. Berry [@Berry1982: 120] noted that they were “not even remotely related to population size”, but rather the “relative probability of occupation through time” and “the major trends in these charts (i.e., peaks and troughs) should provide a reliable indication of the direction of change”. Later, Rick [@Rick1987: 56] more formally outlined the use of radiocarbon time series as a method for paleodemography by developing an ‘inference chain’ that detailed the various biases involved in their use and interpretation. Rick identified three biases [@Rick1987]: creation bias, preservation bias, and investigation bias. Over the past decade, as archaeologists have assembled and analysed large radiocarbon datasets from various regions of the world, there has been considerable focus on these biases, notably creation [@Crombe2014; @Freeman2018; @Naudinot2014] and preservation bias [@Peros2010; @Williams2012; @Bluhm2019; @Surovell2007; @Surovell2009], as well as bias caused by the radiocarbon calibration process [@Williams2012; @Armit2013; @Bamforth2012; @Contreras2014; @Michczynski2006; @Timpson2014]. Of the three biases Rick identified, investigation bias has been the most understudied thus far. Investigation bias occurs when investigators gather more radiocarbon samples for particular periods of time or specific geographic regions, which leads to periods or regions with more or less radiocarbon representation than the average time period or region. The biased selection of radiocarbon samples would then appear to be relative increases or decreases in population when they are merely artefacts of research history and fieldwork practice. Investigation bias has been understudied because for many regions of the world radiocarbon data are the only currently available proxy evidence enabling reconstructions of paleodemography in continuous time series spanning millennia. Assessing the impact of investigation bias on radiocarbon-based reconstructions of paleodemography requires comparison with other paleodemographic proxies.

The American Southwest provides one the best available case studies for assessing the role of investigation bias in radiocarbon-based reconstructions of paleodemography. It also has one of the most well-documented cases of population growth and decline in the world [@Bocinsky2016; @Hill2004; @Kohler2010b; @Robinson2019; @Freeman2018PNAS]. For almost a century, researchers have collected multiple proxies, such as dendrochronological records, settlement size records, and ceramic seriation records to address the magnitude of this population growth and decline. However, a critical knowledge gap still exists for the exact scale of population decline across and beyond the Southwest [@Cameron2010; @Boyer2010]. This gap is because the less intensively researched regions lack multiple proxies for robust reconstructions of paleodemography. In these regions, radiocarbon data are the only available proxy time series for reconstructing paleodemography. Over the past six years, we have collected all available radiocarbon data across and beyond the Southwest [@Robinson2019]. Initial analyses of this new radiocarbon dataset have revealed extensive population decline from AD 1300--1600 across and beyond the Southwest [@Robinson2019; @Freeman2018PNAS]. Therefore, the geographic extent of Late Prehistoric population decline could be much broader than previously expected.

The potential importance of this widespread population decline for research on Late Prehistoric archaeology of the American West requires comparison of the sensitivity of lower-resolution radiocarbon data against other higher-resolution palaeodemographic proxies. In the most well-documented regions (such as the northern American Southwest) we expect radiocarbon records to reflect the highest amount of investigation bias, as researchers have the option of selecting from a suite of other higher-resolution proxies, including dendrochronology and ceramic seriation. Investigation bias can create an artificial ‘edge effect’ spuriously appearing as a decrease in population. Here we develop a new method for arraying dendrochronological records as time series for comparison against radiocarbon time series. We focus on the broader region of the Upland US Southwest (UUSW) for which there is a robust dendrochronological record [@Bocinsky2016].

We also focus on the more concentrated and intensively investigated Central Mesa Verde (CMV) and Northern Rio Grande (NRG) regions of the UUSW (Figure \@ref(fig:Figure1)), for which multiproxy palaeodemographic reconstructions were produced by the Village Ecodynamics Project (VEP) [@Ortman2016; @Schwindt2016]. The VEP is an interdisciplinary, multi-method research initiative aimed at better understanding demography of ancestral Pueblo farmers and human-environment interaction across the UUSW. As part of the project, VEP researchers cataloged data on all known archaeological habitation sites within a 4,600-km2 area in the CMV [@Schwindt2016] and a 6,900-km2 area in the NRG [@Ortman2016], to which most of the CMV population migrated during the thirteenth century AD [@Ortman2016]. They integrated data including pitstructure and room count estimates, the presence of diagnostic architectural features, surface ceramic tallies, surveyor assessments, and limited dendrochronological data in an empirical Bayesian framework to develop population estimates for each region resolved to periods ranging from 20 to 125 years in length [@Ortman2016; @Schwindt2016]. The VEP estimates are the highest quality regional palaeodemographic reconstructions currently available in the UUSW, and are completely independent of the radiocarbon records for the region. They are therefore useful in exploring the impact of investigation bias in the radiocarbon record.

(ref:fig1) The UUSW as defined in this study (region with black border). The white dashed area represents the UUSW as defined in [@Bocinsky2016]. VEP study areas are white boxes. CMV: Central Mesa Verde; NRG: Northern Rio Grande.

```{r Figure1, fig.height=6.5, fig.width=6.5, out.width = '50%', fig.cap="(ref:fig1)"}
# A nice projection of the Four Corners states to use for mapping
four_corners_lcc_proj <-
  robinson2020::four_corners_lcc_proj

# The Four Corners states in four_corners_lcc_proj
four_corners_states <-
  robinson2020::four_corners_states

# The UUSW as defined in Bocinsky et al. 2016
uusw_boundary <- robinson2020::uusw_boundary

# The UUSW study area for this analysis
uusw_counties <-
  robinson2020::uusw_counties

# The VEP study areas
vep_study_areas <-
  robinson2020::vep_study_areas

# Truncate calendar dates to AD 200 - 1800
trunc_dates <- c(1750, 150)

theme_map <- function(base_size = 6.5,
                      base_family = "") {
  theme_bw(
    base_size = base_size,
    base_family = base_family
  ) %+replace%
    theme(
      axis.line = element_blank(),
      axis.text = element_blank(),
      axis.ticks = element_blank(),
      axis.title = element_blank(),
      panel.background = element_blank(),
      panel.border = element_blank(),
      panel.grid = element_blank(),
      panel.grid.major = element_line(colour = "transparent"),
      panel.grid.minor = element_line(colour = "transparent"),
      legend.key = element_blank(),
      legend.position = "top",
      plot.background = element_blank(),
      plot.margin = margin(t = 0, r = 0, b = 0, l = 0, unit = "npc")
    )
}

ggplot() +
  geom_sf(data = four_corners_states) +
  geom_raster(
    data = robinson2020::swus_hillshade_500m,
    mapping = aes(
      x = x,
      y = y,
      alpha = ID
    ),
    na.rm = TRUE
  ) +
  scale_alpha(
    range = c(0.8, 0),
    na.value = 0,
    limits = c(0, 255),
    guide = "none"
  ) +
  geom_sf(
    data = four_corners_states,
    fill = NA
  ) +
  geom_sf(
    data = uusw_counties %>%
      sf::st_union() %>%
      rmapshaper::ms_filter_islands(),
    fill = NA,
    color = "black"
  ) +
  geom_sf(
    data = uusw_boundary,
    fill = NA,
    linetype = 2,
    color = "white"
  ) +
  geom_sf(
    data = vep_study_areas,
    fill = NA,
    color = "white",
    size = 1
  ) +
  geom_sf_text(
    data = vep_study_areas,
    aes(label = `Study Area`),
    size = 3
  ) +
  theme_map(16)
```

We expect that substantial investigation bias is present in radiocarbon records if those records substantially diverge from the dendrochronological and multiproxy reconstructions. If the lower-resolution radiocarbon records reveal similar trends to the two other records, we expect investigation bias does not substantially impact the ability of radiocarbon records to be used as relative measures of the directionality of population change through time. This would, in turn, lend more confidence to radiocarbon records across the American West that reveal extensive post AD 1300 population declines but lack other palaeodemographic proxies.

# Methods {-}
Here we compare two large collections of radiocarbon and tree-ring dates drawn from the UUSW. After harmonizing those datasets with one another and calibrating the radiocarbon dates, we calculate binned summed probability distributions across the UUSW, CMV, and NRG. The radiocarbon dates we analyse here are drawn from the National Science Foundation supported project, “Populating a Radiocarbon Database for North America,” which have been uploaded to the Canadian Archaeological Radiocarbon Database (CARD; https://www.canadianarchaeology.ca). Tree-ring data are those presented in [@Bocinsky2016]. Both datasets are available on the Digital Archaeological Record (see Data Accessibility).

## Study Area {-}
The UUSW was geographically defined by [@Bocinsky2016] as the region between 105--113°W and 32--38°N. Here, we refine that definition in two ways. First, because location data for our radiocarbon database are only resolved to the county level for most sites [@Robinson2019], we define our study region to include all counties whose geographic centroids fall within the region from [@Bocinsky2016]. Second, because of the generally poor tree-ring date record in the Sonoran Desert, we exclude the Phoenix and Tucson basins by dropping Maricopa, Pinal, and Pima counties in Arizona. Figure \@ref(fig:Figure1) shows the original UUSW from [@Bocinsky2016] and revised study area used in this analysis, as well as the two VEP study areas.

## Period of Interest {-}
Here, we focus our analyses on the period for which data are most abundant: 1750--150 BP, or AD 200--1800. This period includes the periods of the VEP population estimates for the CMV (AD 600--1280) and NRG (AD 900--1760), and extends 400 years earlier than the CMV estimates to allow us to assess trends in the radiocarbon record leading up to the VEP reconstructions, and to control for edge effects. While the VEP demographic estimates are for ancestral Pueblo populations only, it is likely that the tree-ring records include samples from Spanish colonial contexts after AD 1600 or so.

## Datasets {-}
```{r prepare-radiocarbon-data}
# Read in the radiocarbon database
# Filter the 2-sigma calibrated ranges to the trunc_dates range
if (!file.exists(
  here::here("analysis/data/derived_data/radiocarbon_data.csv")
)) {
  readr::read_csv(
    here::here("analysis/data/raw_data/robinson_et_al_2020_radiocarbon.csv")
  ) %>%
    dplyr::mutate(
      Type = "Radiocarbon",
      Site_ID = Site_ID %>% stringr::str_c("Radiocarbon - ", .)
    ) %>%
    dplyr::select(
      Type,
      SiteID = Site_ID,
      Site_Number,
      County,
      Lab_Number,
      Date_BP,
      SD
    ) %>%
    # retain dates whose calibrated 95% probability mass overlaps trunc_dates
    dplyr::filter(
      rcarbon::calibrate(
        x = Date_BP,
        errors = SD
      ) %>%
        rcarbon:::hpdi(credMass = 0.95) %>%
        purrr::map_lgl(function(x) {
          startdate <- max(x[, "startCalBP"])
          enddate <- min(x[, "endCalBP"])
          startdate > trunc_dates[[2]] && enddate <= trunc_dates[[1]]
        })
    ) %>%
    # Write the output to a new file (csv)
    readr::write_csv(
      here::here("analysis/data/derived_data/radiocarbon_data.csv")
    )
}

radiocarbon_data <-
  readr::read_csv(
    here::here("analysis/data/derived_data/radiocarbon_data.csv")
  )
```

We cleaned the radiocarbon database by first removing all non-archaeological dates. To ensure our data are reasonably precise, we removed all radiocarbon dates with 1σ errors greater than 300 years and more than 25% of the uncalibrated age. We removed dates with duplicate lab numbers whose radiocarbon ages, 1σ errors, or county information did not match. We then trimmed these data to counties within the UUSW. We retained radiocarbon dates that likely derive from our period of interest by first calibrating all of the dates, then retaining dates whose 95% calibrated probability masses overlapped 1750--150 BP.

```{r prepare-treering-data}
# Read in the Bocinsky et al. 2016 tree ring data
# https://doi.org/10.6067/XCV82J6D7B

if (!file.exists(
  here::here("analysis/data/derived_data/dendro_data.csv")
)) {
  readr::read_csv(
    here::here("analysis/data/raw_data/BOCINSKY_ET_AL_2015_TREE_RINGS.csv")
  ) %>%
    # Convert to spatial object
    sf::st_as_sf(
      coords = c("LON", "LAT"),
      crs = 4326
    ) %>%
    sf::st_transform(four_corners_lcc_proj) %>%
    # Convert dates to BP
    dplyr::mutate(
      Date_BP = 1950 - Outer_Date_AD,
      Type = ifelse(C_Level %in% c(2, 3),
        "Cutting or Near-cutting",
        "Non-cutting"
      )
    ) %>%
    # drop C_Level
    dplyr::select(
      Seq_Number,
      Site_Number,
      Lab_Number,
      Type,
      Date_BP
    ) %>%
    dplyr::rename(SiteID = Seq_Number) %>%
    sf::st_intersection(uusw_counties) %>%
    sf::st_drop_geometry() %>%
    dplyr::mutate(
      Type = Type %>% stringr::str_c("Dendro - ", .),
      SiteID = SiteID %>% stringr::str_c("Dendro - ", .),
      SD = 0
    ) %>%
    dplyr::filter(Type == "Dendro - Cutting or Near-cutting") %>%
    dplyr::mutate(Type = "Tree-ring") %>%
    dplyr::select(
      Type,
      SiteID,
      Site_Number,
      County,
      Lab_Number,
      Date_BP,
      SD
    ) %>%
    # Get only raw dates within trunc_dates
    dplyr::filter(
      Date_BP > trunc_dates[[2]],
      Date_BP <= trunc_dates[[1]]
    ) %>%
    # Write the output to a new file (csv)
    readr::write_csv(
      here::here("analysis/data/derived_data/dendro_data.csv")
    )
}

dendro_data <-
  readr::read_csv(
    here::here("analysis/data/derived_data/dendro_data.csv")
  )
```

We draw on the tree-ring date database presented in [@Bocinsky2016]. These data were compiled by Tim Kohler and Rebecca Higgins from many contributors, and the vast majority of dates were produced by the Laboratory of Tree Ring Research at the University of Arizona. The original dataset contained 32,863 cutting, near-cutting, and non-cutting dates from across the US Southwest. Here, we only use the dates that are most likely to be accurate — cutting dates (‘B’, ‘G’, ‘L’, ‘c’, or ‘r’ dates, with or without a ‘+’) for which we know the year the tree died, and near-cutting dates (‘v’ or ‘v+’ dates) which are likely within 0--3 years of the true date of the outermost ring. The [@Bocinsky2016] data contain site locations; for this analysis, we first spatially cropped the database to our study area, and then translated those locations to the county level. We converted the original date determinations to BP to match calibrated radiocarbon year conventions, and retained dates within 1750--150 BP.

Finally, due to much of the radiocarbon data only being resolved to the county level, we had to approximate inclusion in the VEP, CMV and NRG study areas. The CMV study area conforms well to Montezuma county, Colorado; the NRG overlaps with Los Alamos, Rio Arriba, Sandoval, Santa Fe, and Taos counties in New Mexico. In the analyses below, dates were included as “within” each study area if they derive from those respective counties.

```{r all-data}
all_data <- list(
  Radiocarbon = radiocarbon_data,
  `Tree-ring` = dendro_data
) %>%
  purrr::map(~ dplyr::mutate(.x,
    `Study Area` = ifelse(County == "Montezuma, CO",
      "CMV",
      NA
    ),
    `Study Area` = ifelse(County %in% c(
      "Los Alamos, NM",
      "Rio Arriba, NM",
      "Sandoval, NM",
      "Santa Fe, NM",
      "Taos, NM"
    ),
    "NRG",
    `Study Area`
    )
  ))
```

Our final dataset includes `r nrow(all_data$Radiocarbon) %>% format(big.mark = ',')` radiocarbon dates from `r all_data$Radiocarbon$SiteID %>% unique() %>% length %>% format(big.mark = ',')` unique sites, dating from `r max(all_data$Radiocarbon$Date_BP)`--`r min(all_data$Radiocarbon$Date_BP)` BP (uncalibrated; `r -1 * (1950 - max(all_data$Radiocarbon$Date_BP)) - 1` BC--`r 1950 - min(all_data$Radiocarbon$Date_BP)` AD) and `r nrow(all_data[['Tree-ring']]) %>% format(big.mark = ',')` tree-ring dates from `r all_data[["Tree-ring"]]$SiteID %>% unique() %>% length %>% format(big.mark = ',')` unique sites, dating from `r max(all_data[['Tree-ring']]$Date_BP)` -- `r min(all_data[['Tree-ring']]$Date_BP)` BP (`r 1950 - max(all_data[['Tree-ring']]$Date_BP)`--`r 1950 - min(all_data[['Tree-ring']]$Date_BP)` AD). Table \@ref(tab:Table1) presents the counts of dates and sites of each date type in the CMV and NRG study areas.

```{r Table1}
all_data %>%
  dplyr::bind_rows() %>%
  dplyr::bind_rows(all_data %>%
    dplyr::bind_rows() %>%
    dplyr::mutate(`Study Area` = "UUSW Total")) %>%
  dplyr::filter(!is.na(`Study Area`)) %>%
  dplyr::mutate(`Study Area` = factor(`Study Area`,
    levels = c(
      "UUSW Total",
      "CMV",
      "NRG"
    )
  )) %>%
  dplyr::group_by(Type, SiteID, `Study Area`) %>%
  dplyr::summarise(Dates = dplyr::n()) %>%
  dplyr::group_by(Type, `Study Area`) %>%
  dplyr::summarise(
    Dates = sum(Dates),
    Sites = dplyr::n()
  ) %>%
  tidyr::pivot_wider(names_from = Type, values_from = Dates:Sites) %>%
  dplyr::select(`Study Area`, Dates_Radiocarbon, Sites_Radiocarbon, `Dates_Tree-ring`, `Sites_Tree-ring`) %>%
  knitr::kable(
    caption = "Counts of radiocarbon and tree-ring dates and sites in this study.",
    booktabs = TRUE,
    format.args = list(big.mark = ","),
    col.names = c("Study Area", "Dates", "Sites", "Dates", "Sites")
  ) %>%
  kableExtra::add_header_above(c(" " = 1, "Radiocarbon" = 2, "Tree-ring" = 2))
```

Date densities across the study area were very uneven, with high counts of radiocarbon dates deriving from Navajo County, Arizona, and almost a third of the tree-ring dates in our study being from Montezuma County, Colorado. Below, we overcome some of these sampling issues by binning dates by site and occupation phase. Still, these disparities likely demonstrate more fundamental differences in research design across archaeological projects. There are very few radiocarbon dates from the two VEP study areas, indicating a clear bias towards using tree-ring dating where (and when) possible. Figure \@ref(fig:Figure2) presents the number of radiocarbon and tree-ring dates by county within our study area. Figure \@ref(fig:Figure3) presents the number of unique sites with radiocarbon and tree-ring dates by county within our study area. There are similar ratios of radiocarbon dates to sites with radiocarbon dates for the CMV and NRG (roughly 2:1), but sites in the CMV tend to have a third more tree-ring dates than those in the NRG (20:1 versus 15:1, respectively).

```{r Figure2, fig.height=9, fig.width=6.5, fig.cap='The number of radiocarbon and tree-ring dates by county in this study.', fig.show = 'hold', out.width = '45%'}
county_counts <-
  all_data %>%
  dplyr::bind_rows() %>%
  dplyr::group_by(Type, County) %>%
  dplyr::summarise(Dates = dplyr::n()) %>%
  dplyr::left_join(uusw_counties,
    by = "County"
  ) %>%
  sf::st_as_sf()

theme_map <- function(base_size = 6.5,
                      base_family = "") {
  theme_bw(
    base_size = base_size,
    base_family = base_family
  ) %+replace%
    theme(
      axis.line = element_blank(),
      axis.text = element_blank(),
      axis.ticks = element_blank(),
      axis.title = element_blank(),
      panel.background = element_blank(),
      panel.border = element_blank(),
      panel.grid = element_blank(),
      panel.grid.major = element_line(colour = "transparent"),
      panel.grid.minor = element_line(colour = "transparent"),
      legend.key = element_blank(),
      legend.position = "bottom",
      legend.background = element_blank(),
      legend.key.width = unit(0.15, "npc"),
      legend.text = element_text(size = rel(1)),
      plot.background = element_blank(),
      plot.margin = margin(t = 0, r = 0, b = 0, l = 0, unit = "npc")
    )
}

county_counts %>%
  dplyr::filter(Type == "Radiocarbon") %>%
  ggplot() +
  geom_sf(
    data = four_corners_states,
    fill = NA
  ) +
  geom_sf(aes(fill = Dates),
    color = "black"
  ) +
  scale_fill_distiller(
    name = "Number of Radiocarbon Dates",
    palette = "Greys",
    direction = 1,
    limits = c(0, 200),
    minor_breaks = seq(25, 175, 50),
    breaks = seq(0, 200, 50),
    expand = expansion(mult = 0, add = 0),
    guide = guide_colorbar(title.position = "top")
  ) +
  geom_sf(
    data = uusw_counties,
    fill = NA,
    color = "black"
  ) +
  theme_map(16)

county_counts %>%
  dplyr::filter(Type == "Tree-ring") %>%
  ggplot() +
  geom_sf(
    data = four_corners_states,
    fill = NA
  ) +
  geom_sf(aes(fill = Dates),
    color = "black"
  ) +
  scale_fill_distiller(
    name = "Number of Tree-ring Dates",
    palette = "Greys",
    direction = 1,
    limits = c(0, 5000),
    minor_breaks = seq(500, 4500, 1000),
    breaks = seq(0, 5000, 1000),
    expand = expansion(mult = 0, add = 0),
    guide = guide_colorbar(title.position = "top")
  ) +
  geom_sf(
    data = uusw_counties,
    fill = NA,
    color = "black"
  ) +
  theme_map(16)
```

```{r Figure3, fig.height=9, fig.width=6.5, fig.cap='The number of sites with radiocarbon and tree-ring dates by county in this study.', fig.show = 'hold', out.width = '45%'}
county_counts <-
  all_data %>%
  dplyr::bind_rows() %>%
  dplyr::select(Type, SiteID, County) %>%
  dplyr::distinct() %>%
  dplyr::group_by(Type, County) %>%
  dplyr::summarise(Dates = dplyr::n()) %>%
  dplyr::left_join(uusw_counties,
    by = "County"
  ) %>%
  sf::st_as_sf()

theme_map <- function(base_size = 6.5,
                      base_family = "") {
  theme_bw(
    base_size = base_size,
    base_family = base_family
  ) %+replace%
    theme(
      axis.line = element_blank(),
      axis.text = element_blank(),
      axis.ticks = element_blank(),
      axis.title = element_blank(),
      panel.background = element_blank(),
      panel.border = element_blank(),
      panel.grid = element_blank(),
      panel.grid.major = element_line(colour = "transparent"),
      panel.grid.minor = element_line(colour = "transparent"),
      legend.key = element_blank(),
      legend.position = "bottom",
      legend.background = element_blank(),
      legend.key.width = unit(0.15, "npc"),
      legend.text = element_text(size = rel(1)),
      plot.background = element_blank(),
      plot.margin = margin(t = 0, r = 0, b = 0, l = 0, unit = "npc")
    )
}

county_counts %>%
  dplyr::filter(Type == "Radiocarbon") %>%
  ggplot() +
  geom_sf(
    data = four_corners_states,
    fill = NA
  ) +
  geom_sf(aes(fill = Dates),
    color = "black"
  ) +
  scale_fill_distiller(
    name = "Number of Sites with Radiocarbon Dates",
    palette = "Greys",
    direction = 1,
    limits = c(0, 80),
    minor_breaks = seq(10, 70, 20),
    breaks = seq(0, 80, 20),
    expand = expansion(mult = 0, add = 0),
    guide = guide_colorbar(title.position = "top")
  ) +
  geom_sf(
    data = uusw_counties,
    fill = NA,
    color = "black"
  ) +
  theme_map(16)

county_counts %>%
  dplyr::filter(Type == "Tree-ring") %>%
  ggplot() +
  geom_sf(
    data = four_corners_states,
    fill = NA
  ) +
  geom_sf(aes(fill = Dates),
    color = "black"
  ) +
  scale_fill_distiller(
    name = "Number of Sites with Tree-ring Dates",
    palette = "Greys",
    direction = 1,
    limits = c(0, 250),
    minor_breaks = seq(25, 225, 50),
    breaks = seq(0, 250, 50),
    expand = expansion(mult = 0, add = 0),
    guide = guide_colorbar(title.position = "top")
  ) +
  geom_sf(
    data = uusw_counties,
    fill = NA,
    color = "black"
  ) +
  theme_map(16)
```

## Preparing the SPDs {-}
To prepare the radiocarbon dataset, we used functions provided by the rcarbon package for R [@Bevan2019]. We calibrated the radiocarbon dates using the methods of [@Bronk2008], using the ‘intcal13’ calibration curve [@Reimer2013]. To control for the inconsistent number of dates among  the archaeological sites represented in the database (a form of investigation bias), we aggregated radiocarbon dates known to be from the same phase of an archaeological site using the rcarbon::binPrep function using the single linkage agglomeration method with an h-value of 100 (see [Supplementary Material] for a discussion of various h-values). This creates clusters of dates from the same site whose mean values (uncalibrated) are within 100 years of each other, and splits site occupations into distinct phases separated by 100 years or more. We constructed summed probability distributions (SPDs) from the calibrated dates using the `rcarbon::spd` function. We averaged probability distributions of binned dates, and then constructed SPDs from the resulting site-phase probability distributions.

```{r radiocarbon-spds}
timerange <- c(10000, 0)

radiocarbon_spd <-
  all_data$Radiocarbon %>%
  split(., .$`Study Area`) %>%
  tibble::tibble(
    `Study Area` = names(.),
    Dates = .
  ) %>%
  dplyr::bind_rows(tibble::tibble(
    `Study Area` = "UUSW",
    Dates = list(all_data$Radiocarbon)
  )) %>%
  dplyr::rowwise() %>%
  dplyr::mutate(
    Calibrated = list(Dates %$%
      rcarbon::calibrate(
        x = Date_BP,
        errors = SD,
        timeRange = timerange,
        calMatrix = TRUE,
        verbose = FALSE
      )),
    Bins = list(rcarbon::binPrep(
      sites = Dates$SiteID,
      ages = Dates$Date_BP,
      h = 100,
      method = "single"
    )),
    SPD = list(rcarbon::spd(
      x = Calibrated,
      bins = Bins,
      timeRange = timerange,
      spdnormalised = TRUE,
      verbose = FALSE
    ))
  ) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(Density = SPD %>%
    purrr::map(function(x) {
      x$grid %>%
        tibble::as_tibble() %$%
        tibble::tibble(
          Type = "Radiocarbon Density",
          Date_BP = calBP,
          Density = PrDens
        )
    })) %T>%
  readr::write_rds(
    here::here("analysis/data/derived_data/radiocarbon_spd.rds"),
    compress = "gz"
  )
```

We treated the tree-ring dates in as similar a way possible as the radiocarbon dates. We modeled faux-calibrated tree-ring date probabilities as discrete degenerate distributions ($\Pr(X=k_0)=1$) with $k_0$ equal to the tree-ring date — in other words, a probability of $1$ at the date value, and $0$ elsewhere. This is akin to modeling the tree ring date as a normal distribution with a mean of the date value, and a standard deviation of zero. Thinking of tree ring dates this way allows us to use the same binning and SPD algorithms as for radiocarbon dates.^[One might consider alternative probability distributions for non-cutting or near-cutting dates. The probability mass function in such a case would be zero for years prior to the date of the outer ring of the sample, and non-zero for subsequent years, presumably decaying quickly. For instance, we could imagine a study that derives empirical probability distributions for non-cutting dates based on cutting-dates from the same contexts.] Here, we binned the tree-ring dates into site-phases using the rcarbon::binPrep function and constructing SPDs using `rcarbon::spd`, both with the same parameter selections as used for the radiocarbon samples. As above, we averaged probability distributions of binned dates, and then constructed SPDs from the site-phase distributions.

```{r treering-spds}
calibrate_tree_ring_dates <- function(x, timeRange = c(50000, 0)) {
  out <- list(
    metadata = tibble::tibble(
      DateID = 1:length(x),
      CRA = x,
      Error = 0,
      Details = NA,
      CalCurve = "intcal13",
      ResOffsets = 0,
      ResErrors = 0,
      StartBP = timeRange[1],
      EndBP = timeRange[2],
      Normalised = TRUE,
      F14C = FALSE,
      CalEPS = 1e-05
    ),
    grids = NA,
    calmatrix =
      tibble::tibble(
        index = 1:length(x),
        year_bp = x,
        value = TRUE
      ) %>%
        tidyr::pivot_wider(
          names_from = index,
          values_from = value
        ) %>%
        dplyr::bind_rows(tibble::tibble(year_bp = timeRange[1]:timeRange[2])) %>%
        dplyr::filter(!duplicated(year_bp)) %>%
        dplyr::arrange(-year_bp) %>%
        dplyr::filter(year_bp >= 0) %>%
        tibble::column_to_rownames("year_bp") %>%
        as.matrix() %>%
        tidyr::replace_na(0) %>%
        magrittr::set_colnames(NULL)
  )

  attr(out, "class") <- c("CalDates", "list")

  return(out)
}

dendro_spd <-
  all_data$`Tree-ring` %>%
  split(., .$`Study Area`) %>%
  tibble::tibble(
    `Study Area` = names(.),
    Dates = .
  ) %>%
  dplyr::bind_rows(tibble::tibble(
    `Study Area` = "UUSW",
    Dates = list(all_data$`Tree-ring`)
  )) %>%
  dplyr::rowwise() %>%
  dplyr::mutate(
    Calibrated = list(Dates %$%
      calibrate_tree_ring_dates(
        x = Date_BP,
        timeRange = timerange
      )),
    Bins = list(rcarbon::binPrep(
      sites = Dates$SiteID,
      ages = Dates$Date_BP,
      h = 100,
      method = "single"
    )),
    SPD = list(rcarbon::spd(
      x = Calibrated,
      bins = Bins,
      timeRange = timerange,
      spdnormalised = TRUE,
      verbose = FALSE
    ))
  ) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(Density = SPD %>%
    purrr::map(function(x) {
      x$grid %>%
        tibble::as_tibble() %$%
        tibble::tibble(
          Type = "Tree-ring Density",
          Date_BP = calBP,
          Density = PrDens
        )
    })) %T>%
  readr::write_rds(
    here::here("analysis/data/derived_data/dendro_spd.rds"),
    compress = "gz"
  )
```

# Results {-}
Figure \@ref(fig:Figure4) presents radiocarbon and tree-ring SPDs for the UUSW, CMV, and NRG regions, as well as VEP population reconstructions for the CMV and NRG regions. Across the UUSW, we see striking agreement in directionality and timing between the radiocarbon and tree-ring series (Figure \@ref(fig:Figure4)A); both series trend upwards until the late thirteenth century, and then decline rapidly. The tree-ring series — the light orange line is the raw SPD, and the dark line is smoothed — shows more variation on annual to decadal timescales, and yet matches the broad trends of the radiocarbon SPD. Roughly 150-year periodic fluctuations in the tree-ring SPD, described as periods of “exploration” and “exploitation” by [@Bocinsky2016], are not present in the radiocarbon SPD, the result of a much smaller sample size of site-phases and lower temporal resolution of radiocarbon dating relative to tree-ring dating.

(ref:fig4) Radiocarbon and tree-ring density distributions in the UUSW (A), CMV (B), and NRG (C) regions, and VEP population estimates for the CMV and NRG. The dark orange tree-ring density curve smooths the raw tree-ring density (light orange) using a 21-year center-aligned Gaussian kernel with a 5-year standard deviation, following [@Bocinsky2016]. In Panel A, the radiocarbon density (y-axis) is scaled by a factor of 3 to facilitate comparison with the tree-ring density over the target time period. The radiocarbon and tree-ring densities (y-axis) in panels B and C are scaled such that the area under their densities is equal to the area under the VEP population estimates (i.e., the cumulative population).

```{r Figure4, fig.height=6.5, fig.width=6.5, fig.cap='(ref:fig4)', fig.show = 'hold', out.width = '75%'}

# This is a 21-year wide Gaussian distribution with a
# mean of 0 and standard deviation of 5 years.
# This is the smoother used in Bocinsky et al. 2016.
smoother <- dnorm(seq(-10, 10, 1), sd = 5)

all_density <-
  dendro_spd %>%
  dplyr::select(`Study Area`, Density) %>%
  tidyr::unnest(cols = c(Density)) %>%
  dplyr::bind_rows(radiocarbon_spd %>%
    dplyr::select(`Study Area`, Density) %>%
    tidyr::unnest(cols = c(Density))) %>%
  dplyr::mutate(`Year AD` = 1950 - Date_BP) %>%
  dplyr::select(
    `Study Area`,
    Type,
    `Year AD`,
    Density
  )

vep_demography <-
  robinson2020::vep_demography %>%
  dplyr::mutate(End = ifelse(`Study Area` == "CMV" & Period == 20, 1300, End)) %>%
  dplyr::rowwise() %>%
  dplyr::mutate(`Year AD` = list(Start:(End - 1))) %>%
  dplyr::select(
    `Study Area`,
    `Year AD`,
    `Period`,
    Population
  ) %>%
  tidyr::unnest(`Year AD`)

dplyr::left_join(all_density,
  vep_demography,
  by = c(
    "Study Area",
    "Year AD"
  )
) %>%
  dplyr::group_by(
    `Study Area`,
    Type
  ) %>%
  dplyr::mutate(`Scaled Density` = Density * (sum(Population, na.rm = T) / sum(Density * !is.na(Population), na.rm = TRUE))) %>%
  dplyr::group_by(
    `Study Area`,
    Type,
    Period
  ) %>%
  dplyr::group_by(
    `Study Area`,
    Type
  ) %>%
  dplyr::mutate(
    `Raw Density` = ifelse(`Study Area` == "UUSW", Density, `Scaled Density`),
    `Smoothed Density` = ifelse(Type == "Tree-ring Density", stats::filter(`Raw Density`, filter = smoother), `Raw Density`),
    `Raw Density` = ifelse(`Study Area` == "UUSW", `Raw Density`, NA),
    `Raw Density` = ifelse(Type == "Tree-ring Density", `Raw Density`, NA)
  ) %>%
  dplyr::select(
    `Study Area`,
    Type,
    `Year AD`,
    Population,
    `Raw Density`,
    `Smoothed Density`,
    # `Binned Density`
  ) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(
    Type = factor(Type,
      levels = c(
        "Tree-ring Density",
        "Radiocarbon Density",
        "VEP Population"
      ),
      ordered = TRUE
    ),
    `Study Area` = factor(`Study Area`,
      levels = c(
        "UUSW",
        "CMV",
        "NRG"
      ),
      labels = c(
        "Upland US Southwest Density",
        "Central Mesa Verde Population",
        "Northern Rio Grande Population"
      ),
      ordered = TRUE
    )
  ) %>%
  dplyr::mutate(`Smoothed Density` = ifelse(`Study Area` == "Upland US Southwest Density" & Type == "Radiocarbon Density",
    (`Smoothed Density` * 3),
    `Smoothed Density`
  )) %>%
  {
    ggplot(
      .,
      aes(x = `Year AD`)
    ) +
      geom_line(aes(
        y = `Raw Density`,
        color = Type
      ),
      size = 0.5,
      alpha = 0.5
      ) +
      geom_line(aes(
        y = `Population`
      ),
      size = 1.25
      ) +
      geom_line(aes(
        y = `Smoothed Density`,
        color = Type
      ),
      size = 0.75
      ) +
      scale_color_manual(
        values = c(
          "Tree-ring Density" = "#d95f02",
          "Radiocarbon Density" = "#1b9e77",
          "VEP Population" = "black"
        ),
        limits = c(
          "Tree-ring Density",
          "Radiocarbon Density",
          "VEP Population"
        )
      ) +
      xlab("Year AD") +
      scale_x_continuous(
        limits = c(200, 1800),
        breaks = seq(200, 1800, 200),
        expand = expansion(0, 0)
      ) +
      theme_minimal(18) +
      theme(
        legend.title = element_blank(),
        legend.justification = c(0, 1),
        legend.position = "bottom",
        legend.text = element_text(size = 12),
        axis.title.x = element_text(size = 12),
        axis.text.x = element_text(size = 10),
        axis.title.y = element_blank(),
        axis.text.y = element_text(size = 10),
        # panel.grid.major.y = element_blank(),
        # panel.grid.minor.y = element_blank(),
        plot.margin = unit(c(0.25, 0.5, 0.25, 0.25), "in")
      ) +
      facet_wrap(
        # rows =
        vars(`Study Area`),
        ncol = 1,
        scales = "free_y"
      )
  } %>%
  egg::tag_facet(
    open = "",
    close = "",
    tag_pool = paste0(LETTERS, ". ", c(
      "Upland US Southwest Density",
      "Central Mesa Verde Population",
      "Northern Rio Grande Population"
    )),
    hjust = 0,
    vjust = 1.5,
    size = 4
  )
```

Comparisons between the CMV and NRG regional SPDs and the VEP population estimates are also revealing. In Figures \@ref(fig:Figure4)B and \@ref(fig:Figure4)C, the regional SPDs have been scaled such that the areas under their densities (the orange and green lines) are equal to the area under the VEP population estimates (the black lines). For both regions, the radiocarbon and tree-ring SPDs capture the broad directionality of the VEP population estimates. In the CMV region (Figure \@ref(fig:Figure4)B), the radiocarbon SPD has substantial probability mass prior to the VEP reconstruction, which is likely due to sampling bias in the radiocarbon record that is skewed in this region towards older (pre-Pueblo) sites. The radiocarbon SPD also does not capture the Pueblo I (AD 750--900) population peak (which may be too short-lived to be captured by such a small sample of radiocarbon dates), and seems to begin declining by the early thirteenth century AD. The tree-ring SPDs better capture the variability in population, though include a decline in the mid AD 1100s that is not in the VEP population estimates. In the NRG region (Figure \@ref(fig:Figure4)C), both SPDs rise until about AD 1250, then decline thereafter. The high peak in the tree-ring record at AD 1250, far in excess of the VEP population estimate, is particularly interesting — it likely signals the net impact of the immigration of many thousands of people to the NRG region, and the resulting (and simultaneous) burst in construction. Such a sudden and synchronous demographic and architectural event is hard to capture in the relatively lower resolution and sparsely sampled radiocarbon record, which tends to be smeared across sub-centennial scales courtesy of the calibration process [@Baillie1991].

These results yield key insights on the relative sensitivity of the radiocarbon record to capture long-term, coarse-grained demographic change across the Southwest. As stated in the introduction, our aim is to understand how well the radiocarbon record captures population declines across the Southwest, as we need to develop a measure of confidence in radiocarbon records from regions across and beyond the Southwest that lack other demographic proxies. Both the radiocarbon and tree-ring SPDs show declines post-AD 1300 across all regions. In both the CMV and NRG the decline in the radiocarbon SPD begins earlier than in the VEP population series. This may indicate that radiocarbon time series make the decline seem to begin earlier than it actually does, which could be caused by a ‘cliff’ in the radiocarbon calibration curve from AD 1200--1300 that is surrounded by two plateaus at AD 1100--1200 and AD 1400 [@Reimer2013].

Comparisons of the radiocarbon and dendrochronological record with the architecture/ceramic record show similarities and differences that reinforce the need for multiple-proxy studies of palaeodemography when moving between broader and finer spatial scales. While the multi-proxy population reconstruction for the CMV region was similar to the radiocarbon and dendrochronological records, it showed a different trend for the NRG region (lower panel, Figure \@ref(fig:Figure4)). There, the population reconstruction peaks from AD 1300--1400, whereas the radiocarbon record declines, at first quickly and then slowly. These different trends between the VEP population reconstructions and dendrochronological and radiocarbon proxies reflect some investigation bias in both the radiocarbon and dendrochronological records at smaller, intra-regional scales, which calls for caution in the use of radiocarbon records to assess possible paleo-migration scenarios at finer spatial scales. This result reinforces other work that has called for the use of multiple proxies as cross-checks on radiocarbon time series [@Williams2012, @Crombe2014].

The central aim in comparing radiocarbon to dendrochronological records here was to assess the potential impacts of investigation bias in radiocarbon records, caused by researchers not taking radiocarbon samples in favour of other types of chronometric techniques (e.g., diagnostic artefact counts, tree-ring dates). Specifically, we are interested in the potential impact of this kind of bias in the post AD 1300 decline of radiocarbon SPDs across and beyond the Southwest. Similarities in the decline of both radiocarbon and dendrochronological records across the UUSW suggest that, at this particular regional scale, radiocarbon records reflect long-term, coarse-grained trends in paleodemography. Indeed, not only do we see corresponding declines in the radiocarbon and dendrochronological records at this large scale, but these patterns are also consistent with previous work on demographic and social change in the American Southwest: the Coalescent Communities Database, a Southwest-wide database of room counts and occupation spans of habitation sites with 12 rooms or more, supports a “population decline that began between AD 1300 and 1350 and continued to European contact (Hill et al. 2004:Figures 2,4)” [@Phillips2018; @Hill2004].

# Conclusion {-}
Understanding the geographic scale of post-AD 1300 human population declines in the American Southwest requires a reliance on radiocarbon-based reconstructions of palaeodemography, as radiocarbon is often the only proxy record available in many regions for developing continuous time series. However, radiocarbon records may include investigation bias caused by researchers not selecting radiocarbon samples in favour of Late Prehistoric diagnostic artefacts or other dating methods. In this paper we develop a new method for systematically comparing radiocarbon time series against multiple palaeodemographic proxies. We selected three regions—the Upland US Southwest, Central Mesa Verde, and the Northern Rio Grande—where we expected the most investigation bias to occur, because in these regions multiple higher resolution dating techniques are available. Results indicated that all three palaeodemographic proxies reveal post-AD 1300 population declines. While there is some variability at finer spatial and temporal scales that warrant the collection of other data that can be used to cross-check and validate the resolution of radiocarbon records, at coarse-grained, large spatial and temporal scales radiocarbon records provide a robust relative measure of long-term palaeodemographic trends. These results lend confidence to single proxy radiocarbon-based reconstructions of palaeodemography outside the Southwest that suggest post-AD 1300 population declines in many parts of North America.

Investigation bias is an inevitable issue that all researchers using radiocarbon records have to address. The results of this paper advance our confidence in radiocarbon records as a coarse-grained measure of the directionality of human paleodemographic trends. The methods developed can be used for other regions of the world that have dendrochronological records, such as Europe [@Kuniholm2001; @Haneca2009; @Hillam1990], where comparisons with radiocarbon records could help to advance knowledge of Neolithic ‘booms and busts’ of populations [@Shennan2013; @Whitehouse2014] and to overcome limitations of the Bronze Age radiocarbon calibration plateau [@Armit2013].

#### Acknowledgements {-}
We thank the guest editorial board for inviting us to contribute to this volume. We thank two anonymous reviewers for their helpful suggestions for improvement. We thank our colleagues in the PAGES People3000 working group, and PAGES for funding this working group. We thank the National Foundation for supporting this research. Lastly, we thank the generations of different stakeholders working across the Southwest for their collection, storage and publication of the data presented here.

#### Data Accessibility {-}
A compendium of data and code used in this analysis is available on Github at https://github.com/people3k/robinson2020 and is archived with Zenodo at https://doi.org/10.5281/zenodo.3647599. Radiocarbon data used in this analysis are archived with the Digital Archaeological Record (tDAR) and are available at https://doi.org/10.6067/XCV8455460. Tree-ring data used here are archived with tDAR at https://doi.org/10.6067/XCV82J6D7B. Archaeological site locations are protected by US federal and state statutes. In order to protect these sensitive data, we pre-processed both the radiocarbon and tree-ring data; data included in the Github repository do not include site locations, and researchers will need to gain permission from the State Historic Preservation Office (SHPO) of each state in order to access the site location data.

#### Ethics {-}
No ethical issues are associated with this paper.

#### Authors’ contributions {-}
E.R. and R.K.B designed the study, collected and analysed the data, and wrote the first draft. J.F. and D.B collected and organized data. R.L.K collected data. All authors revised/improved the manuscript.

#### Competing interests {-}
We have no competing interests.

#### Funding {-}
Funding for this research was provided by NSF grants BCS-1418858, BCS-1624061, BCS-1822033 and PAGES (Past Global Changes) funding to the Working Group [Paleoclimate and the Peopling of the Earth (PEOPLE 3000)](http://www.pastglobalchanges.org/ini/wg/arctic2k/170-initiatives/working-group/people-3000/1732-people-3000).

<!-- The following line inserts a page break  -->
\newpage

# References {-}
<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->
<div id="refs"></div>

<!-- The following line inserts a page break  -->
\newpage

# Supplementary Material: Cut-width sensitivity for temporal binning {-}
Uneven sampling across archaeological contexts is a persistant problem when constructing summed probability distributions from radiocarbon dates, and perhaps even a larger issue for the more ubiquitous tree-ring dates in regions like the Upland US Southwest. In SPDs of radiocarbon dates, it is conventional to combine dates from single sites prior to creating the SPDs. Dates from separate components or phases of multi-component sites should be counted separately, however.

To control for inconsistent sample size across the archaeological sites represented in the database, we aggregated dates known to be from the same phase of same archaeological site using the `rcarbon::binPrep` function using the “single linkage” agglomeration method with a cut width (or $h$-value) of 100 [@rcarbon]. This clusters dates from the same site that are within 100 years of each other, and splits site occupations into distinct phases separated by 100 years or more. Binned calibrated dates are then summed, and the resulting probability mass function is divided by the number of dates in the site-phase.

Because the wider archaeological SPD research community still lacks a standard protocol for determining $h$, we perform a sensitivity analysis to demonstrate that the resulting SPDs for radiocarbon and tree-ring dates stabilize at a cut width of 100 or so. We calculated SPDs at 25-year intervals from 0 (only duplicate dates are binned) to 150 years (effectively binning by site).

As can be seen in Figure \@ref(fig:FigureS1), binning in general has little impact on the resulting SPDs, but cut widths above $h=100$ years are virtually indistinguishable. The shape of the radiocarbon SPD (panel A) stabilizes above $h=0$ and its values stabilize by $h=100$. The tree-ring SPD arguably stabilizes in both shape and value above $h=75$ or so, but in the analysis we use $h=100$ to keep our methods consistent between the radiocarbon and tree-ring SPDs.

```{r FigureS1, fig.height=6.5, fig.width=6.5, fig.cap='Radiocarbon (A) and tree-ring (B) density distributions in the UUSW, given varying cut widths for binning dates within sites. The density distributions for the UUSW are normalized. Years are given in AD.', fig.show = 'hold', out.width = '75%'}
cut_widths <- seq(0, 150, 25)

radiocarbon_spd <-
  here::here("analysis/data/derived_data/radiocarbon_spd.rds") %>%
  readr::read_rds()

dendro_spd <-
  here::here("analysis/data/derived_data/dendro_spd.rds") %>%
  readr::read_rds()

# This is a 21-year wide Gaussian distribution with a
# mean of 0 and standard deviation of 5 years.
# This is the smoother used in Bocinsky et al. 2016.
smoother <- dnorm(seq(-10, 10, 1), sd = 5)

timerange <- c(10000, 0)

spds <-
  radiocarbon_spd %>%
  dplyr::filter(`Study Area` == "UUSW") %>%
  dplyr::bind_rows(dendro_spd %>%
    dplyr::filter(`Study Area` == "UUSW")) %>%
  dplyr::mutate(Type = c(
    "Radiocarbon",
    "Tree-ring"
  )) %>%
  dplyr::select(Type, Dates, Calibrated) %>%
  dplyr::rowwise() %>%
  dplyr::mutate(
    Densities =
      list(
        cut_widths %>%
          magrittr::set_names(., .) %>%
          purrr::map_dfr(~ rcarbon::spd(
            x = Calibrated,
            timeRange = timerange,
            bins =
              rcarbon::binPrep(
                sites = Dates$SiteID,
                ages = Dates$Date_BP,
                h = .x,
                method = "single"
              ),
            spdnormalised = TRUE,
            verbose = FALSE
          ) %$%
            grid %>%
            tibble::as_tibble() %$%
            tibble::tibble(
              Date_AD = 1950 - calBP,
              Density = PrDens
            ),
          .id = "h"
          ) %>%
          dplyr::mutate(
            `Cut Width (h)` =
              factor(h,
                levels = cut_widths,
                ordered = TRUE
              )
          ) %>%
          dplyr::filter(Date_AD > 0) %>%
          dplyr::select(`Cut Width (h)`,
            `Year AD` = Date_AD,
            Density
          )
      )
  ) %>%
  dplyr::select(Type, Densities) %>%
  tidyr::unnest(cols = c(Densities))

spds %>%
  dplyr::group_by(Type, `Cut Width (h)`) %>%
  dplyr::mutate(
    Density =
      ifelse(Type == "Tree-ring",
        stats::filter(Density,
          filter = smoother
        ),
        Density
      )
  ) %>%
  dplyr::ungroup() %>%
  {
    ggplot2::ggplot(
      .,
      ggplot2::aes(
        x = `Year AD`,
        y = Density,
        color = `Cut Width (h)`
      )
    ) +
      ggplot2::geom_line() +
      ggplot2::facet_wrap("Type",
        nrow = 2,
        scales = "free_y"
      ) +
      xlab("Year AD") +
      scale_x_continuous(
        limits = c(200, 1800),
        breaks = seq(200, 1800, 200),
        expand = expansion(0, 0)
      ) +
      scale_y_continuous(
        limits = c(0, NA),
        expand = expansion(0, 0)
      ) +
      theme_minimal(18) +
      theme(
        legend.title = element_text(size = 12),
        legend.position = "bottom",
        legend.text = element_text(size = 12),
        axis.title.x = element_text(size = 12),
        axis.text.x = element_text(size = 10),
        axis.title.y = element_blank(),
        axis.text.y = element_text(size = 10),
        plot.margin = unit(c(0.25, 0.5, 0.25, 0.25), "in")
      ) +
      guides(color = guide_legend(nrow = 1))
  } %>%
  egg::tag_facet(
    open = "",
    close = "",
    tag_pool = paste0(LETTERS, ". ", c(
      "Radiocarbon Dates",
      "Tree-ring Dates"
    )),
    hjust = 0,
    vjust = 1.5,
    size = 4
  )
```

\newpage

# Colophon {-}

This report is part of a research compendium developed using the `rrtools` package for reproducible research [@Marwick2017, @rrtools]. This report was generated on `r Sys.time()` using the following computational environment and dependencies:

```{r colophon, cache = FALSE}
# which R packages and versions?
if ("devtools" %in% installed.packages()) devtools::session_info()
```

The current Git commit details are:

```{r}
# what commit is this file at?
if ("git2r" %in% installed.packages() &
  git2r::in_repository(path = ".")) {
  git2r::repository(here::here())
}
```

### 2020-04-17 analysis/robinson2020.tex
preparing preprint

\documentclass[
]{sa}

% \usepackage{graphicx}

% \usepackage{fontspec,xunicode}

% \usepackage{polyglossia}
% \setdefaultlanguage[variant=usmax]{english}

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\author{
Erick Robinson\textsuperscript{1,*}

\And
Kyle Bocinsky\textsuperscript{2,3}

\And
Darcy Bird\textsuperscript{4}

\And
Jacob Freeman\textsuperscript{1}

\And
Robert L. Kelly\textsuperscript{5}

}

\title{Dendrochronological dates confirm a Late Prehistoric population decline in the American Southwest derived from radiocarbon dates}

\Plainauthor{Erick Robinson\textsuperscript{1,*}, Kyle Bocinsky\textsuperscript{2,3}, Darcy Bird\textsuperscript{4}, Jacob Freeman\textsuperscript{1}, Robert L. Kelly\textsuperscript{5}}

\Shorttitle{Dendrochronological and radiocarbon dates in the American Southwest}

\Abstract{
The northern American Southwest provides one of the most well-documented cases of human population growth and decline in the world. The geographic extent of this decline in North America is unknown due to the lack of high-resolution palaeodemographic data from regions across and beyond the greater Southwest, where archaeological radiocarbon data is often the only available proxy for investigating these palaeodemographic processes. Radiocarbon time series across and beyond the greater Southwest suggest widespread population collapses from AD 1300--1600. However, radiocarbon data have potential biases caused by variable radiocarbon sample preservation, sample collection, and the non-linearity of the radiocarbon calibration curve. In order to be confident in the wider trends seen in radiocarbon time series across and beyond the greater Southwest, here we focus on regions that have multiple palaeodemographic proxies and compare those proxies to radiocarbon time series. We develop a new method for time series analysis and comparison between dendrochronological data and radiocarbon data. Results confirm a multiple proxy decline in human populations across the Upland US Southwest, Central Mesa Verde, and Northern Rio Grande from AD 1300--1600. These results lend confidence to single proxy radiocarbon-based reconstructions of paleodemography outside the Southwest that suggest post-AD 1300 population declines in many parts of North America.
}

\Keywords{dates-as-data, radiocarbon summed probability distributions, dendrochronology, palaeodemography, southwestern United States}

%% publication information
%% \Volume{50}
%% \Issue{9}
%% \Month{June}
%% \Year{2012}
%% \Submitdate{}
%% \Acceptdate{2012-06-04}

\Address{
          }

%% Publication information
%% Edit as needed depending on status of manuscript.
%% Replace Preprint with journal short name of accepted
%% Pubs if copyright is open compliant. Add punctuation
%% for formatting as desired.
\Journal{Preprint}
% \Volume{}
% \Issue{} %Enter as "(Issue)" for APA
% \Pages{} %Enter as "pp. m---n" for APA
% \Year{}
\Submitdate{\today} %Enter date you submitted preprint for review
\Acceptdate{\today} %Enter date your paper was accepted for publication
\DOI{
DOI HERE
} %Enter final DOI here

\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newenvironment{cslreferences}%
  {}%
  {\par}

% Pandoc header

\begin{document}

\textsuperscript{1} Utah State University\\
\textsuperscript{2} Crow Canyon Archaeological Center\\
\textsuperscript{3} University of Montana\\
\textsuperscript{4} Washington State University\\
\textsuperscript{5} University of Wyoming

\textsuperscript{*} Correspondence: \href{mailto:erick.robinson@usu.edu}{Erick Robinson \textless{}\href{mailto:erick.robinson@usu.edu}{\nolinkurl{erick.robinson@usu.edu}}\textgreater{}}

\newpage

\hypertarget{introduction}{%
\section*{Introduction}\label{introduction}}
\addcontentsline{toc}{section}{Introduction}

The spatial and temporal ubiquity of archaeological radiocarbon dates makes them an unprecedented proxy for human paleodemography. The past decade has witnessed a rapid increase in the use of archaeological radiocarbon time series to reconstruct prehistoric population growth and decline in various regions of the world {[}\protect\hyperlink{ref-Bevan2019}{1}--\protect\hyperlink{ref-Williams2012}{9}{]}. One of the first aggregations of radiocarbon dates for palaeodemographic analysis was by Berry {[}\protect\hyperlink{ref-Berry1982}{10}{]} in his study of population growth, decline and migration across the American Southwest. Radiocarbon time series were used as a complement to earlier periods that lacked robust dendrochronological records. Berry {[}\protect\hyperlink{ref-Berry1982}{10}{]} noted that they were ``not even remotely related to population size'', but rather the ``relative probability of occupation through time'' and ``the major trends in these charts (i.e., peaks and troughs) should provide a reliable indication of the direction of change''. Later, Rick {[}\protect\hyperlink{ref-Rick1987}{11}{]} more formally outlined the use of radiocarbon time series as a method for paleodemography by developing an `inference chain' that detailed the various biases involved in their use and interpretation. Rick identified three biases {[}\protect\hyperlink{ref-Rick1987}{11}{]}: creation bias, preservation bias, and investigation bias. Over the past decade, as archaeologists have assembled and analysed large radiocarbon datasets from various regions of the world, there has been considerable focus on these biases, notably creation {[}\protect\hyperlink{ref-Crombe2014}{12}--\protect\hyperlink{ref-Naudinot2014}{14}{]} and preservation bias {[}\protect\hyperlink{ref-Peros2010}{5},\protect\hyperlink{ref-Williams2012}{9},\protect\hyperlink{ref-Bluhm2019}{15}--\protect\hyperlink{ref-Surovell2009}{17}{]}, as well as bias caused by the radiocarbon calibration process {[}\protect\hyperlink{ref-Williams2012}{9},\protect\hyperlink{ref-Armit2013}{18}--\protect\hyperlink{ref-Timpson2014}{22}{]}. Of the three biases Rick identified, investigation bias has been the most understudied thus far. Investigation bias occurs when investigators gather more radiocarbon samples for particular periods of time or specific geographic regions, which leads to periods or regions with more or less radiocarbon representation than the average time period or region. The biased selection of radiocarbon samples would then appear to be relative increases or decreases in population when they are merely artefacts of research history and fieldwork practice. Investigation bias has been understudied because for many regions of the world radiocarbon data are the only currently available proxy evidence enabling reconstructions of paleodemography in continuous time series spanning millennia. Assessing the impact of investigation bias on radiocarbon-based reconstructions of paleodemography requires comparison with other paleodemographic proxies.

The American Southwest provides one the best available case studies for assessing the role of investigation bias in radiocarbon-based reconstructions of paleodemography. It also has one of the most well-documented cases of population growth and decline in the world {[}\protect\hyperlink{ref-Bocinsky2016}{23}--\protect\hyperlink{ref-Freeman2018PNAS}{27}{]}. For almost a century, researchers have collected multiple proxies, such as dendrochronological records, settlement size records, and ceramic seriation records to address the magnitude of this population growth and decline. However, a critical knowledge gap still exists for the exact scale of population decline across and beyond the Southwest {[}\protect\hyperlink{ref-Cameron2010}{28},\protect\hyperlink{ref-Boyer2010}{29}{]}. This gap is because the less intensively researched regions lack multiple proxies for robust reconstructions of paleodemography. In these regions, radiocarbon data are the only available proxy time series for reconstructing paleodemography. Over the past six years, we have collected all available radiocarbon data across and beyond the Southwest {[}\protect\hyperlink{ref-Robinson2019}{26}{]}. Initial analyses of this new radiocarbon dataset have revealed extensive population decline from AD 1300--1600 across and beyond the Southwest {[}\protect\hyperlink{ref-Robinson2019}{26},\protect\hyperlink{ref-Freeman2018PNAS}{27}{]}. Therefore, the geographic extent of Late Prehistoric population decline could be much broader than previously expected.

The potential importance of this widespread population decline for research on Late Prehistoric archaeology of the American West requires comparison of the sensitivity of lower-resolution radiocarbon data against other higher-resolution palaeodemographic proxies. In the most well-documented regions (such as the northern American Southwest) we expect radiocarbon records to reflect the highest amount of investigation bias, as researchers have the option of selecting from a suite of other higher-resolution proxies, including dendrochronology and ceramic seriation. Investigation bias can create an artificial `edge effect' spuriously appearing as a decrease in population. Here we develop a new method for arraying dendrochronological records as time series for comparison against radiocarbon time series. We focus on the broader region of the Upland US Southwest (UUSW) for which there is a robust dendrochronological record {[}\protect\hyperlink{ref-Bocinsky2016}{23}{]}.

We also focus on the more concentrated and intensively investigated Central Mesa Verde (CMV) and Northern Rio Grande (NRG) regions of the UUSW (Figure \ref{fig:Figure1}), for which multiproxy palaeodemographic reconstructions were produced by the Village Ecodynamics Project (VEP) {[}\protect\hyperlink{ref-Ortman2016}{30},\protect\hyperlink{ref-Schwindt2016}{31}{]}. The VEP is an interdisciplinary, multi-method research initiative aimed at better understanding demography of ancestral Pueblo farmers and human-environment interaction across the UUSW. As part of the project, VEP researchers cataloged data on all known archaeological habitation sites within a 4,600-km2 area in the CMV {[}\protect\hyperlink{ref-Schwindt2016}{31}{]} and a 6,900-km2 area in the NRG {[}\protect\hyperlink{ref-Ortman2016}{30}{]}, to which most of the CMV population migrated during the thirteenth century AD {[}\protect\hyperlink{ref-Ortman2016}{30}{]}. They integrated data including pitstructure and room count estimates, the presence of diagnostic architectural features, surface ceramic tallies, surveyor assessments, and limited dendrochronological data in an empirical Bayesian framework to develop population estimates for each region resolved to periods ranging from 20 to 125 years in length {[}\protect\hyperlink{ref-Ortman2016}{30},\protect\hyperlink{ref-Schwindt2016}{31}{]}. The VEP estimates are the highest quality regional palaeodemographic reconstructions currently available in the UUSW, and are completely independent of the radiocarbon records for the region. They are therefore useful in exploring the impact of investigation bias in the radiocarbon record.

\begin{figure}

{\centering \includegraphics[width=0.5\linewidth]{/Users/bocinsky/git/publications/robinson2020/analysis/figures/Figure1-1}

}

\caption{The UUSW as defined in this study (region with black border). The white dashed area represents the UUSW as defined in {[}\protect\hyperlink{ref-Bocinsky2016}{23}{]}. VEP study areas are white boxes. CMV: Central Mesa Verde; NRG: Northern Rio Grande.}\label{fig:Figure1}
\end{figure}

We expect that substantial investigation bias is present in radiocarbon records if those records substantially diverge from the dendrochronological and multiproxy reconstructions. If the lower-resolution radiocarbon records reveal similar trends to the two other records, we expect investigation bias does not substantially impact the ability of radiocarbon records to be used as relative measures of the directionality of population change through time. This would, in turn, lend more confidence to radiocarbon records across the American West that reveal extensive post AD 1300 population declines but lack other palaeodemographic proxies.

\hypertarget{methods}{%
\section*{Methods}\label{methods}}
\addcontentsline{toc}{section}{Methods}

Here we compare two large collections of radiocarbon and tree-ring dates drawn from the UUSW. After harmonizing those datasets with one another and calibrating the radiocarbon dates, we calculate binned summed probability distributions across the UUSW, CMV, and NRG. The radiocarbon dates we analyse here are drawn from the National Science Foundation supported project, ``Populating a Radiocarbon Database for North America,'' which have been uploaded to the Canadian Archaeological Radiocarbon Database (CARD; \url{https://www.canadianarchaeology.ca}). Tree-ring data are those presented in {[}\protect\hyperlink{ref-Bocinsky2016}{23}{]}. Both datasets are available on the Digital Archaeological Record (see Data Accessibility).

\hypertarget{study-area}{%
\subsection*{Study Area}\label{study-area}}
\addcontentsline{toc}{subsection}{Study Area}

The UUSW was geographically defined by {[}\protect\hyperlink{ref-Bocinsky2016}{23}{]} as the region between 105--113°W and 32--38°N. Here, we refine that definition in two ways. First, because location data for our radiocarbon database are only resolved to the county level for most sites {[}\protect\hyperlink{ref-Robinson2019}{26}{]}, we define our study region to include all counties whose geographic centroids fall within the region from {[}\protect\hyperlink{ref-Bocinsky2016}{23}{]}. Second, because of the generally poor tree-ring date record in the Sonoran Desert, we exclude the Phoenix and Tucson basins by dropping Maricopa, Pinal, and Pima counties in Arizona. Figure \ref{fig:Figure1} shows the original UUSW from {[}\protect\hyperlink{ref-Bocinsky2016}{23}{]} and revised study area used in this analysis, as well as the two VEP study areas.

\hypertarget{period-of-interest}{%
\subsection*{Period of Interest}\label{period-of-interest}}
\addcontentsline{toc}{subsection}{Period of Interest}

Here, we focus our analyses on the period for which data are most abundant: 1750--150 BP, or AD 200--1800. This period includes the periods of the VEP population estimates for the CMV (AD 600--1280) and NRG (AD 900--1760), and extends 400 years earlier than the CMV estimates to allow us to assess trends in the radiocarbon record leading up to the VEP reconstructions, and to control for edge effects. While the VEP demographic estimates are for ancestral Pueblo populations only, it is likely that the tree-ring records include samples from Spanish colonial contexts after AD 1600 or so.

\hypertarget{datasets}{%
\subsection*{Datasets}\label{datasets}}
\addcontentsline{toc}{subsection}{Datasets}

\begin{verbatim}
#> [1] "Calibrating radiocarbon ages..."
#> 
#> [1] "Done."
\end{verbatim}

We cleaned the radiocarbon database by first removing all non-archaeological dates. To ensure our data are reasonably precise, we removed all radiocarbon dates with 1σ errors greater than 300 years and more than 25\% of the uncalibrated age. We removed dates with duplicate lab numbers whose radiocarbon ages, 1σ errors, or county information did not match. We then trimmed these data to counties within the UUSW. We retained radiocarbon dates that likely derive from our period of interest by first calibrating all of the dates, then retaining dates whose 95\% calibrated probability masses overlapped 1750--150 BP.

We draw on the tree-ring date database presented in {[}\protect\hyperlink{ref-Bocinsky2016}{23}{]}. These data were compiled by Tim Kohler and Rebecca Higgins from many contributors, and the vast majority of dates were produced by the Laboratory of Tree Ring Research at the University of Arizona. The original dataset contained 32,863 cutting, near-cutting, and non-cutting dates from across the US Southwest. Here, we only use the dates that are most likely to be accurate --- cutting dates (`B', `G', `L', `c', or `r' dates, with or without a `+') for which we know the year the tree died, and near-cutting dates (`v' or `v+' dates) which are likely within 0--3 years of the true date of the outermost ring. The {[}\protect\hyperlink{ref-Bocinsky2016}{23}{]} data contain site locations; for this analysis, we first spatially cropped the database to our study area, and then translated those locations to the county level. We converted the original date determinations to BP to match calibrated radiocarbon year conventions, and retained dates within 1750--150 BP.

Finally, due to much of the radiocarbon data only being resolved to the county level, we had to approximate inclusion in the VEP, CMV and NRG study areas. The CMV study area conforms well to Montezuma county, Colorado; the NRG overlaps with Los Alamos, Rio Arriba, Sandoval, Santa Fe, and Taos counties in New Mexico. In the analyses below, dates were included as ``within'' each study area if they derive from those respective counties.

Our final dataset includes 1,531 radiocarbon dates from 482 unique sites, dating from 2260--90 BP (uncalibrated; 309 BC--1860 AD) and 15,077 tree-ring dates from 770 unique sites, dating from 1749 -- 152 BP (201--1798 AD). Table \ref{tab:Table1} presents the counts of dates and sites of each date type in the CMV and NRG study areas.

\begin{table}

\caption{\label{tab:Table1}Counts of radiocarbon and tree-ring dates and sites in this study.}
\centering
\begin{tabular}[t]{lrrrr}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{2}{c}{Radiocarbon} & \multicolumn{2}{c}{Tree-ring} \\
\cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5}
Study Area & Dates & Sites & Dates & Sites\\
\midrule
UUSW Total & 1,531 & 482 & 15,077 & 770\\
CMV & 36 & 18 & 4,435 & 219\\
NRG & 83 & 40 & 2,306 & 146\\
\bottomrule
\end{tabular}
\end{table}

Date densities across the study area were very uneven, with high counts of radiocarbon dates deriving from Navajo County, Arizona, and almost a third of the tree-ring dates in our study being from Montezuma County, Colorado. Below, we overcome some of these sampling issues by binning dates by site and occupation phase. Still, these disparities likely demonstrate more fundamental differences in research design across archaeological projects. There are very few radiocarbon dates from the two VEP study areas, indicating a clear bias towards using tree-ring dating where (and when) possible. Figure \ref{fig:Figure2} presents the number of radiocarbon and tree-ring dates by county within our study area. Figure \ref{fig:Figure3} presents the number of unique sites with radiocarbon and tree-ring dates by county within our study area. There are similar ratios of radiocarbon dates to sites with radiocarbon dates for the CMV and NRG (roughly 2:1), but sites in the CMV tend to have a third more tree-ring dates than those in the NRG (20:1 versus 15:1, respectively).

\begin{figure}

{\centering \includegraphics[width=0.45\linewidth]{/Users/bocinsky/git/publications/robinson2020/analysis/figures/Figure2-1} \includegraphics[width=0.45\linewidth]{/Users/bocinsky/git/publications/robinson2020/analysis/figures/Figure2-2}

}

\caption{The number of radiocarbon and tree-ring dates by county in this study.}\label{fig:Figure2}
\end{figure}

\begin{figure}

{\centering \includegraphics[width=0.45\linewidth]{/Users/bocinsky/git/publications/robinson2020/analysis/figures/Figure3-1} \includegraphics[width=0.45\linewidth]{/Users/bocinsky/git/publications/robinson2020/analysis/figures/Figure3-2}

}

\caption{The number of sites with radiocarbon and tree-ring dates by county in this study.}\label{fig:Figure3}
\end{figure}

\hypertarget{preparing-the-spds}{%
\subsection*{Preparing the SPDs}\label{preparing-the-spds}}
\addcontentsline{toc}{subsection}{Preparing the SPDs}

To prepare the radiocarbon dataset, we used functions provided by the rcarbon package for R {[}\protect\hyperlink{ref-Bevan2019}{1}{]}. We calibrated the radiocarbon dates using the methods of {[}\protect\hyperlink{ref-Bronk2008}{32}{]}, using the `intcal13' calibration curve {[}\protect\hyperlink{ref-Reimer2013}{33}{]}. To control for the inconsistent number of dates among the archaeological sites represented in the database (a form of investigation bias), we aggregated radiocarbon dates known to be from the same phase of an archaeological site using the rcarbon::binPrep function using the single linkage agglomeration method with an h-value of 100 (see {[}Supplementary Material{]} for a discussion of various h-values). This creates clusters of dates from the same site whose mean values (uncalibrated) are within 100 years of each other, and splits site occupations into distinct phases separated by 100 years or more. We constructed summed probability distributions (SPDs) from the calibrated dates using the \texttt{rcarbon::spd} function. We averaged probability distributions of binned dates, and then constructed SPDs from the resulting site-phase probability distributions.

We treated the tree-ring dates in as similar a way possible as the radiocarbon dates. We modeled faux-calibrated tree-ring date probabilities as discrete degenerate distributions (\(\Pr(X=k_0)=1\)) with \(k_0\) equal to the tree-ring date --- in other words, a probability of \(1\) at the date value, and \(0\) elsewhere. This is akin to modeling the tree ring date as a normal distribution with a mean of the date value, and a standard deviation of zero. Thinking of tree ring dates this way allows us to use the same binning and SPD algorithms as for radiocarbon dates.\footnote{One might consider alternative probability distributions for non-cutting or near-cutting dates. The probability mass function in such a case would be zero for years prior to the date of the outer ring of the sample, and non-zero for subsequent years, presumably decaying quickly. For instance, we could imagine a study that derives empirical probability distributions for non-cutting dates based on cutting-dates from the same contexts.} Here, we binned the tree-ring dates into site-phases using the rcarbon::binPrep function and constructing SPDs using \texttt{rcarbon::spd}, both with the same parameter selections as used for the radiocarbon samples. As above, we averaged probability distributions of binned dates, and then constructed SPDs from the site-phase distributions.

\hypertarget{results}{%
\section*{Results}\label{results}}
\addcontentsline{toc}{section}{Results}

Figure \ref{fig:Figure4} presents radiocarbon and tree-ring SPDs for the UUSW, CMV, and NRG regions, as well as VEP population reconstructions for the CMV and NRG regions. Across the UUSW, we see striking agreement in directionality and timing between the radiocarbon and tree-ring series (Figure \ref{fig:Figure4}A); both series trend upwards until the late thirteenth century, and then decline rapidly. The tree-ring series --- the light orange line is the raw SPD, and the dark line is smoothed --- shows more variation on annual to decadal timescales, and yet matches the broad trends of the radiocarbon SPD. Roughly 150-year periodic fluctuations in the tree-ring SPD, described as periods of ``exploration'' and ``exploitation'' by {[}\protect\hyperlink{ref-Bocinsky2016}{23}{]}, are not present in the radiocarbon SPD, the result of a much smaller sample size of site-phases and lower temporal resolution of radiocarbon dating relative to tree-ring dating.

\begin{figure}

{\centering \includegraphics[width=0.75\linewidth]{/Users/bocinsky/git/publications/robinson2020/analysis/figures/Figure4-1}

}

\caption{Radiocarbon and tree-ring density distributions in the UUSW (A), CMV (B), and NRG (C) regions, and VEP population estimates for the CMV and NRG. The dark orange tree-ring density curve smooths the raw tree-ring density (light orange) using a 21-year center-aligned Gaussian kernel with a 5-year standard deviation, following {[}\protect\hyperlink{ref-Bocinsky2016}{23}{]}. In Panel A, the radiocarbon density (y-axis) is scaled by a factor of 3 to facilitate comparison with the tree-ring density over the target time period. The radiocarbon and tree-ring densities (y-axis) in panels B and C are scaled such that the area under their densities is equal to the area under the VEP population estimates (i.e., the cumulative population).}\label{fig:Figure4}
\end{figure}

Comparisons between the CMV and NRG regional SPDs and the VEP population estimates are also revealing. In Figures \ref{fig:Figure4}B and \ref{fig:Figure4}C, the regional SPDs have been scaled such that the areas under their densities (the orange and green lines) are equal to the area under the VEP population estimates (the black lines). For both regions, the radiocarbon and tree-ring SPDs capture the broad directionality of the VEP population estimates. In the CMV region (Figure \ref{fig:Figure4}B), the radiocarbon SPD has substantial probability mass prior to the VEP reconstruction, which is likely due to sampling bias in the radiocarbon record that is skewed in this region towards older (pre-Pueblo) sites. The radiocarbon SPD also does not capture the Pueblo I (AD 750--900) population peak (which may be too short-lived to be captured by such a small sample of radiocarbon dates), and seems to begin declining by the early thirteenth century AD. The tree-ring SPDs better capture the variability in population, though include a decline in the mid AD 1100s that is not in the VEP population estimates. In the NRG region (Figure \ref{fig:Figure4}C), both SPDs rise until about AD 1250, then decline thereafter. The high peak in the tree-ring record at AD 1250, far in excess of the VEP population estimate, is particularly interesting --- it likely signals the net impact of the immigration of many thousands of people to the NRG region, and the resulting (and simultaneous) burst in construction. Such a sudden and synchronous demographic and architectural event is hard to capture in the relatively lower resolution and sparsely sampled radiocarbon record, which tends to be smeared across sub-centennial scales courtesy of the calibration process {[}\protect\hyperlink{ref-Baillie1991}{34}{]}.

These results yield key insights on the relative sensitivity of the radiocarbon record to capture long-term, coarse-grained demographic change across the Southwest. As stated in the introduction, our aim is to understand how well the radiocarbon record captures population declines across the Southwest, as we need to develop a measure of confidence in radiocarbon records from regions across and beyond the Southwest that lack other demographic proxies. Both the radiocarbon and tree-ring SPDs show declines post-AD 1300 across all regions. In both the CMV and NRG the decline in the radiocarbon SPD begins earlier than in the VEP population series. This may indicate that radiocarbon time series make the decline seem to begin earlier than it actually does, which could be caused by a `cliff' in the radiocarbon calibration curve from AD 1200--1300 that is surrounded by two plateaus at AD 1100--1200 and AD 1400 {[}\protect\hyperlink{ref-Reimer2013}{33}{]}.

Comparisons of the radiocarbon and dendrochronological record with the architecture/ceramic record show similarities and differences that reinforce the need for multiple-proxy studies of palaeodemography when moving between broader and finer spatial scales. While the multi-proxy population reconstruction for the CMV region was similar to the radiocarbon and dendrochronological records, it showed a different trend for the NRG region (lower panel, Figure \ref{fig:Figure4}). There, the population reconstruction peaks from AD 1300--1400, whereas the radiocarbon record declines, at first quickly and then slowly. These different trends between the VEP population reconstructions and dendrochronological and radiocarbon proxies reflect some investigation bias in both the radiocarbon and dendrochronological records at smaller, intra-regional scales, which calls for caution in the use of radiocarbon records to assess possible paleo-migration scenarios at finer spatial scales. This result reinforces other work that has called for the use of multiple proxies as cross-checks on radiocarbon time series {[}\protect\hyperlink{ref-Williams2012}{9}{]}.

The central aim in comparing radiocarbon to dendrochronological records here was to assess the potential impacts of investigation bias in radiocarbon records, caused by researchers not taking radiocarbon samples in favour of other types of chronometric techniques (e.g., diagnostic artefact counts, tree-ring dates). Specifically, we are interested in the potential impact of this kind of bias in the post AD 1300 decline of radiocarbon SPDs across and beyond the Southwest. Similarities in the decline of both radiocarbon and dendrochronological records across the UUSW suggest that, at this particular regional scale, radiocarbon records reflect long-term, coarse-grained trends in paleodemography. Indeed, not only do we see corresponding declines in the radiocarbon and dendrochronological records at this large scale, but these patterns are also consistent with previous work on demographic and social change in the American Southwest: the Coalescent Communities Database, a Southwest-wide database of room counts and occupation spans of habitation sites with 12 rooms or more, supports a ``population decline that began between AD 1300 and 1350 and continued to European contact (Hill et al.~2004:Figures 2,4)'' {[}\protect\hyperlink{ref-Hill2004}{24},\protect\hyperlink{ref-Phillips2018}{35}{]}.

\hypertarget{conclusion}{%
\section*{Conclusion}\label{conclusion}}
\addcontentsline{toc}{section}{Conclusion}

Understanding the geographic scale of post-AD 1300 human population declines in the American Southwest requires a reliance on radiocarbon-based reconstructions of palaeodemography, as radiocarbon is often the only proxy record available in many regions for developing continuous time series. However, radiocarbon records may include investigation bias caused by researchers not selecting radiocarbon samples in favour of Late Prehistoric diagnostic artefacts or other dating methods. In this paper we develop a new method for systematically comparing radiocarbon time series against multiple palaeodemographic proxies. We selected three regions---the Upland US Southwest, Central Mesa Verde, and the Northern Rio Grande---where we expected the most investigation bias to occur, because in these regions multiple higher resolution dating techniques are available. Results indicated that all three palaeodemographic proxies reveal post-AD 1300 population declines. While there is some variability at finer spatial and temporal scales that warrant the collection of other data that can be used to cross-check and validate the resolution of radiocarbon records, at coarse-grained, large spatial and temporal scales radiocarbon records provide a robust relative measure of long-term palaeodemographic trends. These results lend confidence to single proxy radiocarbon-based reconstructions of palaeodemography outside the Southwest that suggest post-AD 1300 population declines in many parts of North America.

Investigation bias is an inevitable issue that all researchers using radiocarbon records have to address. The results of this paper advance our confidence in radiocarbon records as a coarse-grained measure of the directionality of human paleodemographic trends. The methods developed can be used for other regions of the world that have dendrochronological records, such as Europe {[}\protect\hyperlink{ref-Kuniholm2001}{36}--\protect\hyperlink{ref-Hillam1990}{38}{]}, where comparisons with radiocarbon records could help to advance knowledge of Neolithic `booms and busts' of populations {[}\protect\hyperlink{ref-Shennan2013}{7},\protect\hyperlink{ref-Whitehouse2014}{39}{]} and to overcome limitations of the Bronze Age radiocarbon calibration plateau {[}\protect\hyperlink{ref-Armit2013}{18}{]}.

\hypertarget{acknowledgements}{%
\paragraph*{Acknowledgements}\label{acknowledgements}}
\addcontentsline{toc}{paragraph}{Acknowledgements}

We thank the guest editorial board for inviting us to contribute to this volume. We thank two anonymous reviewers for their helpful suggestions for improvement. We thank our colleagues in the PAGES People3000 working group, and PAGES for funding this working group. We thank the National Foundation for supporting this research. Lastly, we thank the generations of different stakeholders working across the Southwest for their collection, storage and publication of the data presented here.

\hypertarget{data-accessibility}{%
\paragraph*{Data Accessibility}\label{data-accessibility}}
\addcontentsline{toc}{paragraph}{Data Accessibility}

A compendium of data and code used in this analysis is available on Github at \url{https://github.com/people3k/robinson2020} and is archived with Zenodo at \url{https://doi.org/10.5281/zenodo.3647599}. Radiocarbon data used in this analysis are archived with the Digital Archaeological Record (tDAR) and are available at \url{https://doi.org/10.6067/XCV8455460}. Tree-ring data used here are archived with tDAR at \url{https://doi.org/10.6067/XCV82J6D7B}. Archaeological site locations are protected by US federal and state statutes. In order to protect these sensitive data, we pre-processed both the radiocarbon and tree-ring data; data included in the Github repository do not include site locations, and researchers will need to gain permission from the State Historic Preservation Office (SHPO) of each state in order to access the site location data.

\hypertarget{ethics}{%
\paragraph*{Ethics}\label{ethics}}
\addcontentsline{toc}{paragraph}{Ethics}

No ethical issues are associated with this paper.

\hypertarget{authors-contributions}{%
\paragraph*{Authors' contributions}\label{authors-contributions}}
\addcontentsline{toc}{paragraph}{Authors' contributions}

E.R. and R.K.B designed the study, collected and analysed the data, and wrote the first draft. J.F. and D.B collected and organized data. R.L.K collected data. All authors revised/improved the manuscript.

\hypertarget{competing-interests}{%
\paragraph*{Competing interests}\label{competing-interests}}
\addcontentsline{toc}{paragraph}{Competing interests}

We have no competing interests.

\hypertarget{funding}{%
\paragraph*{Funding}\label{funding}}
\addcontentsline{toc}{paragraph}{Funding}

Funding for this research was provided by NSF grants BCS-1418858, BCS-1624061, BCS-1822033 and PAGES (Past Global Changes) funding to the Working Group \href{http://www.pastglobalchanges.org/ini/wg/arctic2k/170-initiatives/working-group/people-3000/1732-people-3000}{Paleoclimate and the Peopling of the Earth (PEOPLE 3000)}.

\newpage

\hypertarget{references}{%
\section*{References}\label{references}}
\addcontentsline{toc}{section}{References}

\hypertarget{refs}{}
\begin{cslreferences}
\leavevmode\hypertarget{ref-Bevan2019}{}%
1. Bevan A, Palmisano A, Woodbridge J, Fyfe R, Roberts CN, Shennan S. 2019 The changing face of the Mediterranean---Land cover, demography and environmental change: Introduction and overview. \emph{The Holocene} \textbf{29}, 703--707. (doi:\href{https://doi.org/10.1177/0959683619826688}{10.1177/0959683619826688})

\leavevmode\hypertarget{ref-Crema2016}{}%
2. Crema ER, Habu J, Kobayashi K, Madella M. 2016 Summed probability distribution of \textsuperscript{14}C dates suggests regional divergences in the population dynamics of the Jomon period in eastern Japan. \emph{PLoS One} \textbf{11}, e0154809. (doi:\href{https://doi.org/10.1371/journal.pone.0154809}{10.1371/journal.pone.0154809})

\leavevmode\hypertarget{ref-Goldberg2016}{}%
3. Goldberg A, Mychajliw AM, Hadly EA. 2016 Post-invasion demography of prehistoric humans in South America. \emph{Nature} \textbf{532}, 232--235. (doi:\href{https://doi.org/10.1038/nature17176}{10.1038/nature17176})

\leavevmode\hypertarget{ref-Palmisano2019}{}%
4. Palmisano A \emph{et al.} 2019 Holocene landscape dynamics and long-term population trends in the Levant. \emph{The Holocene} \textbf{29}, 708--727. (doi:\href{https://doi.org/10.1177/0959683619826642}{10.1177/0959683619826642})

\leavevmode\hypertarget{ref-Peros2010}{}%
5. Peros MC, Munoz SE, Gajewski K, Viau AE. 2010 Prehistoric demography of North America inferred from radiocarbon data. \emph{Journal of Archaeological Science} \textbf{37}, 656--664. (doi:\href{https://doi.org/10.1016/j.jas.2009.10.029}{10.1016/j.jas.2009.10.029})

\leavevmode\hypertarget{ref-Russell2009}{}%
6. Russell T, Steele J. 2009 A geo-referenced radiocarbon database for Early Iron Age sites in sub-Saharan Africa: Initial analysis. \emph{Southern African Humanities} \textbf{21}, 327--344.

\leavevmode\hypertarget{ref-Shennan2013}{}%
7. Shennan S, Downey SS, Timpson A, Edinborough K, Colledge S, Kerig T, Manning K, Thomas MG. 2013 Regional population collapse followed initial agriculture booms in mid-holocene europe. \emph{Nature communications} \textbf{4}, 1--8.

\leavevmode\hypertarget{ref-Wang2014}{}%
8. Wang C, Lu H, Zhang J, Gu Z, He K. 2014 Prehistoric demographic fluctuations in china inferred from radiocarbon data and their linkage with climate change over the past 50,000 years. \emph{Quaternary Science Reviews} \textbf{98}, 45--59.

\leavevmode\hypertarget{ref-Williams2012}{}%
9. Williams AN. 2012 The use of summed radiocarbon probability distributions in archaeology: A review of methods. \emph{Journal of Archaeological Science} \textbf{39}, 578--589.

\leavevmode\hypertarget{ref-Berry1982}{}%
10. Berry MS. 1982 \emph{Time, space and transition in anasazi prehistory}. Salt Lake City, Utah: University of Utah Press.

\leavevmode\hypertarget{ref-Rick1987}{}%
11. Rick JW. 1987 Dates as data: An examination of the Peruvian preceramic radiocarbon record. \emph{American Antiquity} \textbf{52}, 55--73. (doi:\href{https://doi.org/10.2307/281060}{10.2307/281060})

\leavevmode\hypertarget{ref-Crombe2014}{}%
12. Crombé P, Robinson E. 2014 \textsuperscript{14}C dates as demographic proxies in Neolithisation models of northwestern Europe: A critical assessment using Belgium and northeast France as a case-study. \emph{Journal of Archaeological Science} \textbf{52}, 558--566. (doi:\href{https://doi.org/10.1016/j.jas.2014.02.001}{10.1016/j.jas.2014.02.001})

\leavevmode\hypertarget{ref-Freeman2018}{}%
13. Freeman J, Byers DA, Robinson E, Kelly RL. 2018 Culture process and the interpretation of radiocarbon data. \emph{Radiocarbon} \textbf{60}, 453--467. (doi:\href{https://doi.org/10.1017/RDC.2017.124}{10.1017/RDC.2017.124})

\leavevmode\hypertarget{ref-Naudinot2014}{}%
14. Naudinot N, Tomasso A, Tozzi C, Peresani M. 2014 Changes in mobility patterns as a factor of \textsuperscript{14}C date density variation in the Late Epigravettian of Northern Italy and Southeastern France. \emph{Journal of Archaeological Science} \textbf{52}, 578--590. (doi:\href{https://doi.org/10.1016/j.jas.2014.05.021}{10.1016/j.jas.2014.05.021})

\leavevmode\hypertarget{ref-Bluhm2019}{}%
15. Bluhm LE, Surovell TA. 2019 Validation of a global model of taphonomic bias using geologic radiocarbon ages. \emph{Quaternary Research} \textbf{91}, 325--328. (doi:\href{https://doi.org/10.1017/qua.2018.78}{10.1017/qua.2018.78})

\leavevmode\hypertarget{ref-Surovell2007}{}%
16. Surovell TA, Brantingham PJ. 2007 A note on the use of temporal frequency distributions in studies of prehistoric demography. \emph{Journal of Archaeological Science} \textbf{34}, 1868--1877.

\leavevmode\hypertarget{ref-Surovell2009}{}%
17. Surovell TA, Finley JB, Smith GM, Brantingham PJ, Kelly R. 2009 Correcting temporal frequency distributions for taphonomic bias. \emph{Journal of archaeological Science} \textbf{36}, 1715--1724.

\leavevmode\hypertarget{ref-Armit2013}{}%
18. Armit I, Swindles GT, Becker K. 2013 From dates to demography in later prehistoric Ireland? Experimental approaches to the meta-analysis of large \textsuperscript{14}C data-sets. \emph{Journal of archaeological science} \textbf{40}, 433--438. (doi:\href{https://doi.org/10.1016/j.jas.2012.08.039}{10.1016/j.jas.2012.08.039})

\leavevmode\hypertarget{ref-Bamforth2012}{}%
19. Bamforth DB, Grund B. 2012 Radiocarbon calibration curves, summed probability distributions, and early Paleoindian population trends in North America. \emph{Journal of Archaeological Science} \textbf{39}, 1768--1774. (doi:\href{https://doi.org/10.1016/j.jas.2012.01.017}{10.1016/j.jas.2012.01.017})

\leavevmode\hypertarget{ref-Contreras2014}{}%
20. Contreras DA, Meadows J. 2014 Summed radiocarbon calibrations as a population proxy: A critical evaluation using a realistic simulation approach. \emph{Journal of Archaeological Science} \textbf{52}, 591--608. (doi:\href{https://doi.org/10.1016/j.jas.2014.05.030}{10.1016/j.jas.2014.05.030})

\leavevmode\hypertarget{ref-Michczynski2006}{}%
21. Michczynski A, Michczynska DJ. 2006 The effect of PDF peaks' height increase during calibration of radiocarbon date sets. \emph{Geochronometria} \textbf{25}, 1--4.

\leavevmode\hypertarget{ref-Timpson2014}{}%
22. Timpson A, Colledge S, Crema E, Edinborough K, Kerig T, Manning K, Thomas MG, Shennan S. 2014 Reconstructing regional population fluctuations in the european neolithic using radiocarbon dates: A new case-study using an improved method. \emph{Journal of Archaeological Science} \textbf{52}, 549--557.

\leavevmode\hypertarget{ref-Bocinsky2016}{}%
23. Bocinsky RK, Rush J, Kintigh KW, Kohler TA. 2016 Exploration and exploitation in the macrohistory of the pre-Hispanic Pueblo Southwest. \emph{Science Advances} \textbf{2}, e1501532. (doi:\href{https://doi.org/10.1126/sciadv.1501532}{10.1126/sciadv.1501532})

\leavevmode\hypertarget{ref-Hill2004}{}%
24. Hill JB, Clark JJ, Doelle WH, Lyons PD. 2004 Prehistoric demography in the Southwest: Migration, coalescence, and Hohokam population decline. \emph{American Antiquity} \textbf{69}, 689--716. (doi:\href{https://doi.org/10.2307/4128444}{10.2307/4128444})

\leavevmode\hypertarget{ref-Kohler2010b}{}%
25. Kohler TA, Varien MD, Wright AM, editors. 2010 \emph{Leaving Mesa Verde: Peril and Change in the Thirteenth-Century Southwest}. Tucson, Arizona: The Amerind Foundation; University of Arizona Press.

\leavevmode\hypertarget{ref-Robinson2019}{}%
26. Robinson E, Nicholson C, Kelly RL. 2019 The importance of spatial data to open-access national archaeological databases and the development of paleodemography research. \emph{Advances in Archaeological Practice} \textbf{7}, 395--408. (doi:\href{https://doi.org/10.1017/aap.2019.29}{10.1017/aap.2019.29})

\leavevmode\hypertarget{ref-Freeman2018PNAS}{}%
27. Freeman J, Baggio JA, Robinson E, Byers DA, Gayo E, Finley JB, Meyer JA, Kelly RL, Anderies JM. 2018 Synchronization of energy consumption by human societies throughout the Holocene. \emph{Proceedings of the National Academy of Sciences} \textbf{115}, 9962--9967. (doi:\href{https://doi.org/10.1073/pnas.1802859115}{10.1073/pnas.1802859115})

\leavevmode\hypertarget{ref-Cameron2010}{}%
28. Cameron CM. 2010 Advances in understanding the thirteenth-century depopulation of the northern Southwest. In \emph{Leaving mesa verde: Peril and change in the thirteenth century southwest} (eds T Kohler, M Varien, A Wright), pp. 346--363. Tucson, Arizona: The Amerind Foundation; University of Arizona Press.

\leavevmode\hypertarget{ref-Boyer2010}{}%
29. Boyer JL, Moore JL, Lakatos SA, Akins NJ, Wilson CD, Blinman E. 2010 Remodeling immigration: A Northern Rio Grande perspective on depopulation, migration, and donation-side models. In \emph{Leaving mesa verde: Peril and change in the thirteenth century southwest} (eds T Kohler, M Varien, A Wright), pp. 285--323. Tucson, Arizona: The Amerind Foundation; University of Arizona Press.

\leavevmode\hypertarget{ref-Ortman2016}{}%
30. Ortman SG. 2016 Uniform probability density analysis and population history in the northern Rio Grande. \emph{Journal of Archaeological Method and Theory} \textbf{23}, 95--126. (doi:\href{https://doi.org/10.1007/s10816-014-9227-6}{10.1007/s10816-014-9227-6})

\leavevmode\hypertarget{ref-Schwindt2016}{}%
31. Schwindt DM, Bocinsky RK, Ortman SG, Glowacki DM, Varien MD, Kohler TA. 2016 The social consequences of climate change in the central Mesa Verde region. \emph{American Antiquity} \textbf{81}, 74--96.

\leavevmode\hypertarget{ref-Bronk2008}{}%
32. Bronk Ramsey C. 2008 Radiocarbon dating: Revolutions in understanding. \emph{Archaeometry} \textbf{50}, 249--275. (doi:\href{https://doi.org/10.1111/j.1475-4754.2008.00394.x}{10.1111/j.1475-4754.2008.00394.x})

\leavevmode\hypertarget{ref-Reimer2013}{}%
33. Reimer PJ \emph{et al.} 2013 IntCal13 and Marine13 radiocarbon age calibration curves 0--50,000 years cal BP. \emph{Radiocarbon} \textbf{55}, 1869--1887. (doi:\href{https://doi.org/10.2458/azu_js_rc.55.16947}{10.2458/azu\_js\_rc.55.16947})

\leavevmode\hypertarget{ref-Baillie1991}{}%
34. Baillie MG. 1991 Suck-in and smear: Two related chronological problems for the 90s. \emph{Journal of theoretical archaeology} \textbf{2}, 12--16.

\leavevmode\hypertarget{ref-Phillips2018}{}%
35. Phillips DA, Wearing HJ, Clark JJ. 2018 Village growth, emerging infectious disease, and the end of the Neolithic Demographic Transition in the Southwest United States and Northwest Mexico. \emph{American Antiquity} \textbf{83}, 263--280. (doi:\href{https://doi.org/10.1017/aaq.2018.3}{10.1017/aaq.2018.3})

\leavevmode\hypertarget{ref-Kuniholm2001}{}%
36. Kuniholm PI. 2001 Dendrochronology and other applications of tree-ring studies in archaeology. In \emph{The handbook of archaeological sciences} (eds DR Brothwell, AM Pollard), London: John Wiley \& Sons, Ltd.

\leavevmode\hypertarget{ref-Haneca2009}{}%
37. Haneca K, Čufar K, Beeckman H. 2009 Oaks, tree-rings and wooden cultural heritage: A review of the main characteristics and applications of oak dendrochronology in Europe. \emph{Journal of Archaeological Science} \textbf{36}, 1--11. (doi:\href{https://doi.org/10.1016/j.jas.2008.07.005}{10.1016/j.jas.2008.07.005})

\leavevmode\hypertarget{ref-Hillam1990}{}%
38. Hillam J, Groves CM, Brown DM, Baillie MG, Coles JM, Coles BJ. 1990 Dendrochronology of the English Neolithic. \emph{Antiquity} \textbf{64}, 210--220. (doi:\href{https://doi.org/10.1017/S0003598X00077826}{10.1017/S0003598X00077826})

\leavevmode\hypertarget{ref-Whitehouse2014}{}%
39. Whitehouse NJ \emph{et al.} 2014 Neolithic agriculture on the European western frontier: The boom and bust of early farming in Ireland. \emph{Journal of Archaeological Science} \textbf{51}, 181--205.

\leavevmode\hypertarget{ref-rcarbon}{}%
40. Bevan A, Crema ER. 2020 \emph{rcarbon: Methods for calibrating and analysing radiocarbon dates}. See \url{https://CRAN.R-project.org/package=rcarbon}.

\leavevmode\hypertarget{ref-Marwick2017}{}%
41. Marwick B. 2017 Computational reproducibility in archaeological research: Basic principles and a case study of their implementation. \emph{Journal of Archaeological Method and Theory} \textbf{24}, 424--450. (doi:\href{https://doi.org/10.1007/s10816-015-9272-9}{10.1007/s10816-015-9272-9})

\leavevmode\hypertarget{ref-rrtools}{}%
42. Marwick B. 2019 \emph{rrtools: Creates a Reproducible Research Compendium}. See \url{https://github.com/benmarwick/rrtools}.
\end{cslreferences}

\newpage

\hypertarget{supplementary-material-cut-width-sensitivity-for-temporal-binning}{%
\section*{Supplementary Material: Cut-width sensitivity for temporal binning}\label{supplementary-material-cut-width-sensitivity-for-temporal-binning}}
\addcontentsline{toc}{section}{Supplementary Material: Cut-width sensitivity for temporal binning}

Uneven sampling across archaeological contexts is a persistant problem when constructing summed probability distributions from radiocarbon dates, and perhaps even a larger issue for the more ubiquitous tree-ring dates in regions like the Upland US Southwest. In SPDs of radiocarbon dates, it is conventional to combine dates from single sites prior to creating the SPDs. Dates from separate components or phases of multi-component sites should be counted separately, however.

To control for inconsistent sample size across the archaeological sites represented in the database, we aggregated dates known to be from the same phase of same archaeological site using the \texttt{rcarbon::binPrep} function using the ``single linkage'' agglomeration method with a cut width (or \(h\)-value) of 100 {[}\protect\hyperlink{ref-rcarbon}{40}{]}. This clusters dates from the same site that are within 100 years of each other, and splits site occupations into distinct phases separated by 100 years or more. Binned calibrated dates are then summed, and the resulting probability mass function is divided by the number of dates in the site-phase.

Because the wider archaeological SPD research community still lacks a standard protocol for determining \(h\), we perform a sensitivity analysis to demonstrate that the resulting SPDs for radiocarbon and tree-ring dates stabilize at a cut width of 100 or so. We calculated SPDs at 25-year intervals from 0 (only duplicate dates are binned) to 150 years (effectively binning by site).

As can be seen in Figure \ref{fig:FigureS1}, binning in general has little impact on the resulting SPDs, but cut widths above \(h=100\) years are virtually indistinguishable. The shape of the radiocarbon SPD (panel A) stabilizes above \(h=0\) and its values stabilize by \(h=100\). The tree-ring SPD arguably stabilizes in both shape and value above \(h=75\) or so, but in the analysis we use \(h=100\) to keep our methods consistent between the radiocarbon and tree-ring SPDs.

\begin{figure}

{\centering \includegraphics[width=0.75\linewidth]{/Users/bocinsky/git/publications/robinson2020/analysis/figures/FigureS1-1}

}

\caption{Radiocarbon (A) and tree-ring (B) density distributions in the UUSW, given varying cut widths for binning dates within sites. The density distributions for the UUSW are normalized. Years are given in AD.}\label{fig:FigureS1}
\end{figure}

\newpage

\hypertarget{colophon}{%
\section*{Colophon}\label{colophon}}
\addcontentsline{toc}{section}{Colophon}

This report is part of a research compendium developed using the \texttt{rrtools} package for reproducible research {[}\protect\hyperlink{ref-Marwick2017}{41}{]}. This report was generated on 2020-04-16 22:50:13 using the following computational environment and dependencies:

\begin{verbatim}
#> - Session info --------------------------------------------------------------------------------------------------------------------------------
#>  setting  value                       
#>  version  R version 3.6.3 (2020-02-29)
#>  os       macOS Catalina 10.15.4      
#>  system   x86_64, darwin15.6.0        
#>  ui       RStudio                     
#>  language (EN)                        
#>  collate  en_US.UTF-8                 
#>  ctype    en_US.UTF-8                 
#>  tz       America/Denver              
#>  date     2020-04-16                  
#> 
#> - Packages ------------------------------------------------------------------------------------------------------------------------------------
#>  package        * version    date       lib source        
#>  abind            1.4-5      2016-07-21 [1] CRAN (R 3.6.0)
#>  assertthat       0.2.1      2019-03-21 [1] CRAN (R 3.6.3)
#>  backports        1.1.6      2020-04-05 [1] CRAN (R 3.6.2)
#>  bookdown         0.18       2020-03-05 [1] CRAN (R 3.6.0)
#>  callr            3.4.3      2020-03-28 [1] CRAN (R 3.6.2)
#>  class            7.3-15     2019-01-01 [1] CRAN (R 3.6.3)
#>  classInt         0.4-3      2020-04-07 [1] CRAN (R 3.6.2)
#>  cli              2.0.2      2020-02-28 [1] CRAN (R 3.6.3)
#>  codetools        0.2-16     2018-12-24 [1] CRAN (R 3.6.3)
#>  colorspace       1.4-1      2019-03-18 [1] CRAN (R 3.6.0)
#>  crayon           1.3.4      2017-09-16 [1] CRAN (R 3.6.3)
#>  crul             0.9.0      2019-11-06 [1] CRAN (R 3.6.0)
#>  curl             4.3        2019-12-02 [1] CRAN (R 3.6.3)
#>  DBI              1.1.0      2019-12-15 [1] CRAN (R 3.6.0)
#>  deldir           0.1-25     2020-02-03 [1] CRAN (R 3.6.0)
#>  desc             1.2.0      2018-05-01 [1] CRAN (R 3.6.3)
#>  devtools         2.3.0      2020-04-10 [1] CRAN (R 3.6.3)
#>  digest           0.6.25     2020-02-23 [1] CRAN (R 3.6.3)
#>  doParallel       1.0.15     2019-08-02 [1] CRAN (R 3.6.0)
#>  dplyr            0.8.5      2020-03-07 [1] CRAN (R 3.6.0)
#>  e1071            1.7-3      2019-11-26 [1] CRAN (R 3.6.0)
#>  egg              0.4.5      2019-07-13 [1] CRAN (R 3.6.0)
#>  ellipsis         0.3.0      2019-09-20 [1] CRAN (R 3.6.3)
#>  evaluate         0.14       2019-05-28 [1] CRAN (R 3.6.3)
#>  fansi            0.4.1      2020-01-08 [1] CRAN (R 3.6.3)
#>  farver           2.0.3      2020-01-16 [1] CRAN (R 3.6.0)
#>  foreach          1.5.0      2020-03-30 [1] CRAN (R 3.6.3)
#>  foreign          0.8-75     2020-01-20 [1] CRAN (R 3.6.3)
#>  fs               1.4.1      2020-04-04 [1] CRAN (R 3.6.2)
#>  generics         0.0.2      2018-11-29 [1] CRAN (R 3.6.0)
#>  geojson          0.3.2      2019-01-31 [1] CRAN (R 3.6.0)
#>  geojsonio        0.9.2      2020-04-07 [1] CRAN (R 3.6.3)
#>  geojsonlint      0.4.0      2020-02-13 [1] CRAN (R 3.6.0)
#>  ggplot2        * 3.3.0      2020-03-05 [1] CRAN (R 3.6.0)
#>  git2r            0.26.1     2019-06-29 [1] CRAN (R 3.6.3)
#>  glue             1.4.0      2020-04-03 [1] CRAN (R 3.6.2)
#>  goftest          1.2-2      2019-12-02 [1] CRAN (R 3.6.0)
#>  gridExtra        2.3        2017-09-09 [1] CRAN (R 3.6.0)
#>  gtable           0.3.0      2019-03-25 [1] CRAN (R 3.6.0)
#>  here             0.1        2017-05-28 [1] CRAN (R 3.6.0)
#>  hms              0.5.3      2020-01-08 [1] CRAN (R 3.6.0)
#>  htmltools        0.4.0      2019-10-04 [1] CRAN (R 3.6.3)
#>  httpcode         0.3.0      2020-04-10 [1] CRAN (R 3.6.2)
#>  httr             1.4.1      2019-08-05 [1] CRAN (R 3.6.3)
#>  iterators        1.0.12     2019-07-26 [1] CRAN (R 3.6.0)
#>  jqr              1.1.0      2018-10-22 [1] CRAN (R 3.6.0)
#>  jsonlite         1.6.1      2020-02-02 [1] CRAN (R 3.6.3)
#>  jsonvalidate     1.1.0      2019-06-25 [1] CRAN (R 3.6.0)
#>  kableExtra       1.1.0      2019-03-16 [1] CRAN (R 3.6.0)
#>  KernSmooth       2.23-16    2019-10-15 [1] CRAN (R 3.6.3)
#>  knitr            1.28       2020-02-06 [1] CRAN (R 3.6.3)
#>  labeling         0.3        2014-08-23 [1] CRAN (R 3.6.0)
#>  lattice          0.20-38    2018-11-04 [1] CRAN (R 3.6.3)
#>  lazyeval         0.2.2      2019-03-15 [1] CRAN (R 3.6.3)
#>  lifecycle        0.2.0      2020-03-06 [1] CRAN (R 3.6.0)
#>  lubridate        1.7.8      2020-04-06 [1] CRAN (R 3.6.2)
#>  magrittr       * 1.5        2014-11-22 [1] CRAN (R 3.6.3)
#>  maptools         0.9-9      2019-12-01 [1] CRAN (R 3.6.0)
#>  Matrix           1.2-18     2019-11-27 [1] CRAN (R 3.6.3)
#>  memoise          1.1.0      2017-04-21 [1] CRAN (R 3.6.3)
#>  mgcv             1.8-31     2019-11-09 [1] CRAN (R 3.6.3)
#>  munsell          0.5.0      2018-06-12 [1] CRAN (R 3.6.0)
#>  nlme             3.1-144    2020-02-06 [1] CRAN (R 3.6.3)
#>  pillar           1.4.3      2019-12-20 [1] CRAN (R 3.6.0)
#>  pkgbuild         1.0.6      2019-10-09 [1] CRAN (R 3.6.3)
#>  pkgconfig        2.0.3      2019-09-22 [1] CRAN (R 3.6.0)
#>  pkgload          1.0.2      2018-10-29 [1] CRAN (R 3.6.3)
#>  polyclip         1.10-0     2019-03-14 [1] CRAN (R 3.6.0)
#>  prettyunits      1.1.1      2020-01-24 [1] CRAN (R 3.6.3)
#>  processx         3.4.2      2020-02-09 [1] CRAN (R 3.6.3)
#>  ps               1.3.2      2020-02-13 [1] CRAN (R 3.6.3)
#>  purrr            0.3.3      2019-10-18 [1] CRAN (R 3.6.3)
#>  R6               2.4.1      2019-11-12 [1] CRAN (R 3.6.3)
#>  rcarbon          1.3.1      2020-03-18 [1] CRAN (R 3.6.0)
#>  RColorBrewer     1.1-2      2014-12-07 [1] CRAN (R 3.6.0)
#>  Rcpp             1.0.4      2020-03-17 [1] CRAN (R 3.6.0)
#>  readr            1.3.1      2018-12-21 [1] CRAN (R 3.6.0)
#>  remotes          2.1.1      2020-02-15 [1] CRAN (R 3.6.3)
#>  rgeos            0.5-2      2019-10-03 [1] CRAN (R 3.6.0)
#>  rlang            0.4.5      2020-03-01 [1] CRAN (R 3.6.3)
#>  rmapshaper       0.4.4      2020-04-01 [1] CRAN (R 3.6.2)
#>  rmarkdown        2.1        2020-01-20 [1] CRAN (R 3.6.0)
#>  robinson2020   * 1.0.0.9000 2020-04-17 [1] local         
#>  rpart            4.1-15     2019-04-12 [1] CRAN (R 3.6.3)
#>  rprojroot        1.3-2      2018-01-03 [1] CRAN (R 3.6.3)
#>  rstudioapi       0.11       2020-02-07 [1] CRAN (R 3.6.3)
#>  rvest            0.3.5      2019-11-08 [1] CRAN (R 3.6.0)
#>  scales           1.1.0      2019-11-18 [1] CRAN (R 3.6.0)
#>  sessioninfo      1.1.1      2018-11-05 [1] CRAN (R 3.6.3)
#>  sf               0.9-2      2020-04-14 [1] CRAN (R 3.6.2)
#>  sp               1.4-1      2020-02-28 [1] CRAN (R 3.6.0)
#>  spatstat         1.63-3     2020-03-01 [1] CRAN (R 3.6.0)
#>  spatstat.data    1.4-3      2020-01-26 [1] CRAN (R 3.6.0)
#>  spatstat.utils   1.17-0     2020-02-07 [1] CRAN (R 3.6.0)
#>  stringi          1.4.6      2020-02-17 [1] CRAN (R 3.6.3)
#>  stringr          1.4.0      2019-02-10 [1] CRAN (R 3.6.3)
#>  tensor           1.5        2012-05-05 [1] CRAN (R 3.6.0)
#>  testthat         2.3.2      2020-03-02 [1] CRAN (R 3.6.3)
#>  tibble           3.0.0      2020-03-30 [1] CRAN (R 3.6.2)
#>  tidyr            1.0.2      2020-01-24 [1] CRAN (R 3.6.0)
#>  tidyselect       1.0.0      2020-01-27 [1] CRAN (R 3.6.0)
#>  tinytex          0.21       2020-03-31 [1] CRAN (R 3.6.2)
#>  units            0.6-6      2020-03-16 [1] CRAN (R 3.6.0)
#>  usethis          1.6.0      2020-04-09 [1] CRAN (R 3.6.3)
#>  V8               3.0.2      2020-03-14 [1] CRAN (R 3.6.0)
#>  vctrs            0.2.4      2020-03-10 [1] CRAN (R 3.6.0)
#>  viridisLite      0.3.0      2018-02-01 [1] CRAN (R 3.6.0)
#>  webshot          0.5.2      2019-11-22 [1] CRAN (R 3.6.0)
#>  withr            2.1.2      2018-03-15 [1] CRAN (R 3.6.3)
#>  xfun             0.13       2020-04-13 [1] CRAN (R 3.6.2)
#>  xml2             1.3.1      2020-04-09 [1] CRAN (R 3.6.2)
#>  yaml             2.2.1      2020-02-01 [1] CRAN (R 3.6.3)
#> 
#> [1] /Library/Frameworks/R.framework/Versions/3.6/Resources/library
\end{verbatim}

The current Git commit details are:

\begin{verbatim}
#> Local:    master /Users/bocinsky/git/publications/robinson2020
#> Remote:   master @ origin (https://github.com/people3k/robinson2020)
#> Head:     [802ee87] 2020-03-24: working towards resubmission
\end{verbatim}

\end{document}

### 2020-04-17 analysis/robinson2020.Rmd
added kableExtra

@@
 knitr::opts_chunk$set(
-  cache = TRUE,
+  cache = FALSE,
   echo = FALSE,
@@
         x = Date_BP,
-        errors = SD
+        errors = SD,
+        verbose = FALSE
       ) %>%

### 2020-04-17 analysis/robinson2020.Rmd
using pander for colophon

@@
     short: Dendrochronological and radiocarbon dates in the American Southwest
-date: "`r format(lubridate::as_date('2020-04-16'), '%d %B, %Y')`"
+date: "`r format(Sys.Date(), '%B %d, %Y')`"
 author:
@@
 bibliography: robinson2020_references.bib
-csl: "philosophical-transactions-of-the-royal-society-b.csl"
+# csl: "philosophical-transactions-of-the-royal-society-b.csl"
+csl: "journal-of-archaeological-science.csl"
 link-citations: true
@@
 knitr::opts_chunk$set(
-  cache = FALSE,
+  cache = TRUE,
   echo = FALSE,
@@
 Radiocarbon time series were used as a complement to earlier periods that lacked robust dendrochronological records.
-Berry [@Berry1982: 120] noted that they were “not even remotely related to population size”, but rather the “relative probability of occupation through time” and “the major trends in these charts (i.e., peaks and troughs) should provide a reliable indication of the direction of change”.
-Later, Rick [@Rick1987: 56] more formally outlined the use of radiocarbon time series as a method for paleodemography by developing an ‘inference chain’ that detailed the various biases involved in their use and interpretation.
-Rick identified three biases [@Rick1987]: creation bias, preservation bias, and investigation bias.
+Berry [-@Berry1982: 120] noted that they were “not even remotely related to population size”, but rather the “relative probability of occupation through time” and “the major trends in these charts (i.e., peaks and troughs) should provide a reliable indication of the direction of change”.
+Later, Rick [-@Rick1987: 56] more formally outlined the use of radiocarbon time series as a method for paleodemography by developing an ‘inference chain’ that detailed the various biases involved in their use and interpretation.
+Rick identified three biases: creation bias, preservation bias, and investigation bias [@Rick1987].
 Over the past decade, as archaeologists have assembled and analysed large radiocarbon datasets from various regions of the world, there has been considerable focus on these biases, notably creation [@Crombe2014; @Freeman2018; @Naudinot2014] and preservation bias [@Peros2010; @Williams2012; @Bluhm2019; @Surovell2007; @Surovell2009], as well as bias caused by the radiocarbon calibration process [@Williams2012; @Armit2013; @Bamforth2012; @Contreras2014; @Michczynski2006; @Timpson2014].
@@
 (ref:fig1) The UUSW as defined in this study (region with black border).
-The white dashed area represents the UUSW as defined in [@Bocinsky2016].
+The white dashed area represents the UUSW as defined by @Bocinsky2016.
 VEP study areas are white boxes.
@@
 The radiocarbon dates we analyse here are drawn from the National Science Foundation supported project, “Populating a Radiocarbon Database for North America,” which have been uploaded to the Canadian Archaeological Radiocarbon Database (CARD; https://www.canadianarchaeology.ca).
-Tree-ring data are those presented in [@Bocinsky2016].
-Both datasets are available on the Digital Archaeological Record (see Data Accessibility).
+Tree-ring data are those presented by @Bocinsky2016.
+Both datasets are available on the Digital Archaeological Record (see [Data Accessibility]).
 
 ## Study Area {-}
-The UUSW was geographically defined by [@Bocinsky2016] as the region between 105--113°W and 32--38°N.
+The UUSW was geographically defined by @Bocinsky2016 as the region between 105--113°W and 32--38°N.
 Here, we refine that definition in two ways.
-First, because location data for our radiocarbon database are only resolved to the county level for most sites [@Robinson2019], we define our study region to include all counties whose geographic centroids fall within the region from [@Bocinsky2016].
+First, because location data for our radiocarbon database are only resolved to the county level for most sites [@Robinson2019], we define our study region to include all counties whose geographic centroids fall within the region from @Bocinsky2016.
 Second, because of the generally poor tree-ring date record in the Sonoran Desert, we exclude the Phoenix and Tucson basins by dropping Maricopa, Pinal, and Pima counties in Arizona.
-Figure \@ref(fig:Figure1) shows the original UUSW from [@Bocinsky2016] and revised study area used in this analysis, as well as the two VEP study areas.
+Figure \@ref(fig:Figure1) shows the original UUSW from @Bocinsky2016 and revised study area used in this analysis, as well as the two VEP study areas.
 
@@
 To prepare the radiocarbon dataset, we used functions provided by the rcarbon package for R [@Bevan2019].
-We calibrated the radiocarbon dates using the methods of [@Bronk2008], using the ‘intcal13’ calibration curve [@Reimer2013].
+We calibrated the radiocarbon dates using the methods of @Bronk2008, using the ‘intcal13’ calibration curve [@Reimer2013].
 To control for the inconsistent number of dates among  the archaeological sites represented in the database (a form of investigation bias), we aggregated radiocarbon dates known to be from the same phase of an archaeological site using the rcarbon::binPrep function using the single linkage agglomeration method with an h-value of 100 (see [Supplementary Material] for a discussion of various h-values).
@@
 We constructed summed probability distributions (SPDs) from the calibrated dates using the `rcarbon::spd` function.
-We averaged probability distributions of binned dates, and then constructed SPDs from the resulting site-phase probability distributions.
+We averaged probability distributions of binned dates, and then constructed SPDs from the resulting site-phase probability distributions.^[Using this method, a site-phase with a single date and a site-phase with 10 dates would each contribute equally to the resulting SPD.
+This method of developing the SPD is therefore a model of the relative abundance of site occupation events through time rather than of dates through time.
+This method does not account for investigation biases deriving from the definition of “site” or the decision to use a particular dating technique at a site in the first place.]
 
@@
 The tree-ring series — the light orange line is the raw SPD, and the dark line is smoothed — shows more variation on annual to decadal timescales, and yet matches the broad trends of the radiocarbon SPD.
-Roughly 150-year periodic fluctuations in the tree-ring SPD, described as periods of “exploration” and “exploitation” by [@Bocinsky2016], are not present in the radiocarbon SPD, the result of a much smaller sample size of site-phases and lower temporal resolution of radiocarbon dating relative to tree-ring dating.
+Roughly 150-year periodic fluctuations in the tree-ring SPD, described as periods of “exploration” and “exploitation” by @Bocinsky2016, are not present in the radiocarbon SPD, the result of a much smaller sample size of site-phases and lower temporal resolution of radiocarbon dating relative to tree-ring dating.
 
 (ref:fig4) Radiocarbon and tree-ring density distributions in the UUSW (A), CMV (B), and NRG (C) regions, and VEP population estimates for the CMV and NRG.
-The dark orange tree-ring density curve smooths the raw tree-ring density (light orange) using a 21-year center-aligned Gaussian kernel with a 5-year standard deviation, following [@Bocinsky2016].
+The dark orange tree-ring density curve smooths the raw tree-ring density (light orange) using a 21-year center-aligned Gaussian kernel with a 5-year standard deviation, following @Bocinsky2016.
 In Panel A, the radiocarbon density (y-axis) is scaled by a factor of 3 to facilitate comparison with the tree-ring density over the target time period.
@@
 Similarities in the decline of both radiocarbon and dendrochronological records across the UUSW suggest that, at this particular regional scale, radiocarbon records reflect long-term, coarse-grained trends in paleodemography.
-Indeed, not only do we see corresponding declines in the radiocarbon and dendrochronological records at this large scale, but these patterns are also consistent with previous work on demographic and social change in the American Southwest: the Coalescent Communities Database, a Southwest-wide database of room counts and occupation spans of habitation sites with 12 rooms or more, supports a “population decline that began between AD 1300 and 1350 and continued to European contact (Hill et al.
-2004:Figures 2,4)” [@Phillips2018; @Hill2004].
+Indeed, not only do we see corresponding declines in the radiocarbon and dendrochronological records at this large scale, but these patterns are also consistent with previous work on demographic and social change in the American Southwest: the Coalescent Communities Database, a Southwest-wide database of room counts and occupation spans of habitation sites with 12 rooms or more, supports a “population decline that began between AD 1300 and 1350 and continued to European contact” [@Phillips2018, following @Hill2004: Figures 2,4].
 
@@
 
-This report is part of a research compendium developed using the `rrtools` package for reproducible research [@Marwick2017, @rrtools].
+This report is part of a research compendium developed using the `rrtools` package for reproducible research [@Marwick2017; @rrtools].
 This report was generated on `r Sys.time()` using the following computational environment and dependencies:
 
-```{r colophon, cache = FALSE}
+```{r colophon-r, cache = FALSE}
 # which R packages and versions?
-if ("devtools" %in% installed.packages()) devtools::session_info()
+if (requireNamespace("pander", quietly = TRUE)) pander::pander(sessionInfo())
 ```
-
-The current Git commit details are:
-
-```{r}
-# what commit is this file at?
-if ("git2r" %in% installed.packages() &
-  git2r::in_repository(path = ".")) {
-  git2r::repository(here::here())
-}
-```

### 2020-04-17 analysis/robinson2020.tex
using pander for colophon

@@
 ]{sa}
-
-% \usepackage{graphicx}
-
-% \usepackage{fontspec,xunicode}
-
-% \usepackage{polyglossia}
-% \setdefaultlanguage[variant=usmax]{english}
 
@@
 \newenvironment{cslreferences}%
-  {}%
+  {\setlength{\parindent}{0pt}%
+  \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces}%
   {\par}
@@
 The spatial and temporal ubiquity of archaeological radiocarbon dates makes them an unprecedented proxy for human paleodemography.
-The past decade has witnessed a rapid increase in the use of archaeological radiocarbon time series to reconstruct prehistoric population growth and decline in various regions of the world {[}\protect\hyperlink{ref-Bevan2019}{1}--\protect\hyperlink{ref-Williams2012}{9}{]}.
-One of the first aggregations of radiocarbon dates for palaeodemographic analysis was by Berry {[}\protect\hyperlink{ref-Berry1982}{10}{]} in his study of population growth, decline and migration across the American Southwest.
+The past decade has witnessed a rapid increase in the use of archaeological radiocarbon time series to reconstruct prehistoric population growth and decline in various regions of the world (Bevan et al., \protect\hyperlink{ref-Bevan2019}{2019}; Crema et al., \protect\hyperlink{ref-Crema2016}{2016}; Goldberg et al., \protect\hyperlink{ref-Goldberg2016}{2016}; Palmisano et al., \protect\hyperlink{ref-Palmisano2019}{2019}; Peros et al., \protect\hyperlink{ref-Peros2010}{2010}; Russell and Steele, \protect\hyperlink{ref-Russell2009}{2009}; Shennan et al., \protect\hyperlink{ref-Shennan2013}{2013}; Wang et al., \protect\hyperlink{ref-Wang2014}{2014}; Williams, \protect\hyperlink{ref-Williams2012}{2012}).
+One of the first aggregations of radiocarbon dates for palaeodemographic analysis was by Berry (Berry, \protect\hyperlink{ref-Berry1982}{1982}) in his study of population growth, decline and migration across the American Southwest.
 Radiocarbon time series were used as a complement to earlier periods that lacked robust dendrochronological records.
-Berry {[}\protect\hyperlink{ref-Berry1982}{10}{]} noted that they were ``not even remotely related to population size'', but rather the ``relative probability of occupation through time'' and ``the major trends in these charts (i.e., peaks and troughs) should provide a reliable indication of the direction of change''.
-Later, Rick {[}\protect\hyperlink{ref-Rick1987}{11}{]} more formally outlined the use of radiocarbon time series as a method for paleodemography by developing an `inference chain' that detailed the various biases involved in their use and interpretation.
-Rick identified three biases {[}\protect\hyperlink{ref-Rick1987}{11}{]}: creation bias, preservation bias, and investigation bias.
-Over the past decade, as archaeologists have assembled and analysed large radiocarbon datasets from various regions of the world, there has been considerable focus on these biases, notably creation {[}\protect\hyperlink{ref-Crombe2014}{12}--\protect\hyperlink{ref-Naudinot2014}{14}{]} and preservation bias {[}\protect\hyperlink{ref-Peros2010}{5},\protect\hyperlink{ref-Williams2012}{9},\protect\hyperlink{ref-Bluhm2019}{15}--\protect\hyperlink{ref-Surovell2009}{17}{]}, as well as bias caused by the radiocarbon calibration process {[}\protect\hyperlink{ref-Williams2012}{9},\protect\hyperlink{ref-Armit2013}{18}--\protect\hyperlink{ref-Timpson2014}{22}{]}.
+Berry (\protect\hyperlink{ref-Berry1982}{1982}: 120) noted that they were ``not even remotely related to population size'', but rather the ``relative probability of occupation through time'' and ``the major trends in these charts (i.e., peaks and troughs) should provide a reliable indication of the direction of change''.
+Later, Rick (\protect\hyperlink{ref-Rick1987}{1987}: 56) more formally outlined the use of radiocarbon time series as a method for paleodemography by developing an `inference chain' that detailed the various biases involved in their use and interpretation.
+Rick identified three biases: creation bias, preservation bias, and investigation bias (Rick, \protect\hyperlink{ref-Rick1987}{1987}).
+Over the past decade, as archaeologists have assembled and analysed large radiocarbon datasets from various regions of the world, there has been considerable focus on these biases, notably creation (Crombé and Robinson, \protect\hyperlink{ref-Crombe2014}{2014}; Freeman, Byers, et al., \protect\hyperlink{ref-Freeman2018}{2018}; Naudinot et al., \protect\hyperlink{ref-Naudinot2014}{2014}) and preservation bias (Bluhm and Surovell, \protect\hyperlink{ref-Bluhm2019}{2019}; Peros et al., \protect\hyperlink{ref-Peros2010}{2010}; Surovell and Brantingham, \protect\hyperlink{ref-Surovell2007}{2007}; Surovell et al., \protect\hyperlink{ref-Surovell2009}{2009}; Williams, \protect\hyperlink{ref-Williams2012}{2012}), as well as bias caused by the radiocarbon calibration process (Armit et al., \protect\hyperlink{ref-Armit2013}{2013}; Bamforth and Grund, \protect\hyperlink{ref-Bamforth2012}{2012}; Contreras and Meadows, \protect\hyperlink{ref-Contreras2014}{2014}; Michczynski and Michczynska, \protect\hyperlink{ref-Michczynski2006}{2006}; Timpson et al., \protect\hyperlink{ref-Timpson2014}{2014}; Williams, \protect\hyperlink{ref-Williams2012}{2012}).
 Of the three biases Rick identified, investigation bias has been the most understudied thus far.
@@
 The American Southwest provides one the best available case studies for assessing the role of investigation bias in radiocarbon-based reconstructions of paleodemography.
-It also has one of the most well-documented cases of population growth and decline in the world {[}\protect\hyperlink{ref-Bocinsky2016}{23}--\protect\hyperlink{ref-Freeman2018PNAS}{27}{]}.
+It also has one of the most well-documented cases of population growth and decline in the world (Bocinsky et al., \protect\hyperlink{ref-Bocinsky2016}{2016}; Freeman, Baggio, et al., \protect\hyperlink{ref-Freeman2018PNAS}{2018}; Hill et al., \protect\hyperlink{ref-Hill2004}{2004}; Kohler et al., \protect\hyperlink{ref-Kohler2010b}{2010}; Robinson et al., \protect\hyperlink{ref-Robinson2019}{2019}).
 For almost a century, researchers have collected multiple proxies, such as dendrochronological records, settlement size records, and ceramic seriation records to address the magnitude of this population growth and decline.
-However, a critical knowledge gap still exists for the exact scale of population decline across and beyond the Southwest {[}\protect\hyperlink{ref-Cameron2010}{28},\protect\hyperlink{ref-Boyer2010}{29}{]}.
+However, a critical knowledge gap still exists for the exact scale of population decline across and beyond the Southwest (Boyer et al., \protect\hyperlink{ref-Boyer2010}{2010}; Cameron, \protect\hyperlink{ref-Cameron2010}{2010}).
 This gap is because the less intensively researched regions lack multiple proxies for robust reconstructions of paleodemography.
 In these regions, radiocarbon data are the only available proxy time series for reconstructing paleodemography.
-Over the past six years, we have collected all available radiocarbon data across and beyond the Southwest {[}\protect\hyperlink{ref-Robinson2019}{26}{]}.
-Initial analyses of this new radiocarbon dataset have revealed extensive population decline from AD 1300--1600 across and beyond the Southwest {[}\protect\hyperlink{ref-Robinson2019}{26},\protect\hyperlink{ref-Freeman2018PNAS}{27}{]}.
+Over the past six years, we have collected all available radiocarbon data across and beyond the Southwest (Robinson et al., \protect\hyperlink{ref-Robinson2019}{2019}).
+Initial analyses of this new radiocarbon dataset have revealed extensive population decline from AD 1300--1600 across and beyond the Southwest (Freeman, Baggio, et al., \protect\hyperlink{ref-Freeman2018PNAS}{2018}; Robinson et al., \protect\hyperlink{ref-Robinson2019}{2019}).
 Therefore, the geographic extent of Late Prehistoric population decline could be much broader than previously expected.
@@
 Here we develop a new method for arraying dendrochronological records as time series for comparison against radiocarbon time series.
-We focus on the broader region of the Upland US Southwest (UUSW) for which there is a robust dendrochronological record {[}\protect\hyperlink{ref-Bocinsky2016}{23}{]}.
+We focus on the broader region of the Upland US Southwest (UUSW) for which there is a robust dendrochronological record (Bocinsky et al., \protect\hyperlink{ref-Bocinsky2016}{2016}).
 
-We also focus on the more concentrated and intensively investigated Central Mesa Verde (CMV) and Northern Rio Grande (NRG) regions of the UUSW (Figure \ref{fig:Figure1}), for which multiproxy palaeodemographic reconstructions were produced by the Village Ecodynamics Project (VEP) {[}\protect\hyperlink{ref-Ortman2016}{30},\protect\hyperlink{ref-Schwindt2016}{31}{]}.
+We also focus on the more concentrated and intensively investigated Central Mesa Verde (CMV) and Northern Rio Grande (NRG) regions of the UUSW (Figure \ref{fig:Figure1}), for which multiproxy palaeodemographic reconstructions were produced by the Village Ecodynamics Project (VEP) (Ortman, \protect\hyperlink{ref-Ortman2016}{2016}; Schwindt et al., \protect\hyperlink{ref-Schwindt2016}{2016}).
 The VEP is an interdisciplinary, multi-method research initiative aimed at better understanding demography of ancestral Pueblo farmers and human-environment interaction across the UUSW.
-As part of the project, VEP researchers cataloged data on all known archaeological habitation sites within a 4,600-km2 area in the CMV {[}\protect\hyperlink{ref-Schwindt2016}{31}{]} and a 6,900-km2 area in the NRG {[}\protect\hyperlink{ref-Ortman2016}{30}{]}, to which most of the CMV population migrated during the thirteenth century AD {[}\protect\hyperlink{ref-Ortman2016}{30}{]}.
-They integrated data including pitstructure and room count estimates, the presence of diagnostic architectural features, surface ceramic tallies, surveyor assessments, and limited dendrochronological data in an empirical Bayesian framework to develop population estimates for each region resolved to periods ranging from 20 to 125 years in length {[}\protect\hyperlink{ref-Ortman2016}{30},\protect\hyperlink{ref-Schwindt2016}{31}{]}.
+As part of the project, VEP researchers cataloged data on all known archaeological habitation sites within a 4,600-km2 area in the CMV (Schwindt et al., \protect\hyperlink{ref-Schwindt2016}{2016}) and a 6,900-km2 area in the NRG (Ortman, \protect\hyperlink{ref-Ortman2016}{2016}), to which most of the CMV population migrated during the thirteenth century AD (Ortman, \protect\hyperlink{ref-Ortman2016}{2016}).
+They integrated data including pitstructure and room count estimates, the presence of diagnostic architectural features, surface ceramic tallies, surveyor assessments, and limited dendrochronological data in an empirical Bayesian framework to develop population estimates for each region resolved to periods ranging from 20 to 125 years in length (Ortman, \protect\hyperlink{ref-Ortman2016}{2016}; Schwindt et al., \protect\hyperlink{ref-Schwindt2016}{2016}).
 The VEP estimates are the highest quality regional palaeodemographic reconstructions currently available in the UUSW, and are completely independent of the radiocarbon records for the region.
@@
 \caption{The UUSW as defined in this study (region with black border).
-The white dashed area represents the UUSW as defined in {[}\protect\hyperlink{ref-Bocinsky2016}{23}{]}.
+The white dashed area represents the UUSW as defined by Bocinsky et al.
+(\protect\hyperlink{ref-Bocinsky2016}{2016}).
 VEP study areas are white boxes.
@@
 The radiocarbon dates we analyse here are drawn from the National Science Foundation supported project, ``Populating a Radiocarbon Database for North America,'' which have been uploaded to the Canadian Archaeological Radiocarbon Database (CARD; \url{https://www.canadianarchaeology.ca}).
-Tree-ring data are those presented in {[}\protect\hyperlink{ref-Bocinsky2016}{23}{]}.
-Both datasets are available on the Digital Archaeological Record (see Data Accessibility).
+Tree-ring data are those presented by Bocinsky et al.
+(\protect\hyperlink{ref-Bocinsky2016}{2016}).
+Both datasets are available on the Digital Archaeological Record (see \protect\hyperlink{data-accessibility}{Data Accessibility}).
 
@@
 
-The UUSW was geographically defined by {[}\protect\hyperlink{ref-Bocinsky2016}{23}{]} as the region between 105--113°W and 32--38°N.
+The UUSW was geographically defined by Bocinsky et al.
+(\protect\hyperlink{ref-Bocinsky2016}{2016}) as the region between 105--113°W and 32--38°N.
 Here, we refine that definition in two ways.
-First, because location data for our radiocarbon database are only resolved to the county level for most sites {[}\protect\hyperlink{ref-Robinson2019}{26}{]}, we define our study region to include all counties whose geographic centroids fall within the region from {[}\protect\hyperlink{ref-Bocinsky2016}{23}{]}.
+First, because location data for our radiocarbon database are only resolved to the county level for most sites (Robinson et al., \protect\hyperlink{ref-Robinson2019}{2019}), we define our study region to include all counties whose geographic centroids fall within the region from Bocinsky et al.
+(\protect\hyperlink{ref-Bocinsky2016}{2016}).
 Second, because of the generally poor tree-ring date record in the Sonoran Desert, we exclude the Phoenix and Tucson basins by dropping Maricopa, Pinal, and Pima counties in Arizona.
-Figure \ref{fig:Figure1} shows the original UUSW from {[}\protect\hyperlink{ref-Bocinsky2016}{23}{]} and revised study area used in this analysis, as well as the two VEP study areas.
+Figure \ref{fig:Figure1} shows the original UUSW from Bocinsky et al.
+(\protect\hyperlink{ref-Bocinsky2016}{2016}) and revised study area used in this analysis, as well as the two VEP study areas.
 
@@
 
-\begin{verbatim}
-#> [1] "Calibrating radiocarbon ages..."
-#> 
-#> [1] "Done."
-\end{verbatim}
-
 We cleaned the radiocarbon database by first removing all non-archaeological dates.
@@
 
-We draw on the tree-ring date database presented in {[}\protect\hyperlink{ref-Bocinsky2016}{23}{]}.
+We draw on the tree-ring date database presented in (Bocinsky et al., \protect\hyperlink{ref-Bocinsky2016}{2016}).
 These data were compiled by Tim Kohler and Rebecca Higgins from many contributors, and the vast majority of dates were produced by the Laboratory of Tree Ring Research at the University of Arizona.
@@
 Here, we only use the dates that are most likely to be accurate --- cutting dates (`B', `G', `L', `c', or `r' dates, with or without a `+') for which we know the year the tree died, and near-cutting dates (`v' or `v+' dates) which are likely within 0--3 years of the true date of the outermost ring.
-The {[}\protect\hyperlink{ref-Bocinsky2016}{23}{]} data contain site locations; for this analysis, we first spatially cropped the database to our study area, and then translated those locations to the county level.
+The (Bocinsky et al., \protect\hyperlink{ref-Bocinsky2016}{2016}) data contain site locations; for this analysis, we first spatially cropped the database to our study area, and then translated those locations to the county level.
 We converted the original date determinations to BP to match calibrated radiocarbon year conventions, and retained dates within 1750--150 BP.
@@
 
-To prepare the radiocarbon dataset, we used functions provided by the rcarbon package for R {[}\protect\hyperlink{ref-Bevan2019}{1}{]}.
-We calibrated the radiocarbon dates using the methods of {[}\protect\hyperlink{ref-Bronk2008}{32}{]}, using the `intcal13' calibration curve {[}\protect\hyperlink{ref-Reimer2013}{33}{]}.
+To prepare the radiocarbon dataset, we used functions provided by the rcarbon package for R (Bevan et al., \protect\hyperlink{ref-Bevan2019}{2019}).
+We calibrated the radiocarbon dates using the methods of Bronk Ramsey (\protect\hyperlink{ref-Bronk2008}{2008}), using the `intcal13' calibration curve (Reimer et al., \protect\hyperlink{ref-Reimer2013}{2013}).
 To control for the inconsistent number of dates among the archaeological sites represented in the database (a form of investigation bias), we aggregated radiocarbon dates known to be from the same phase of an archaeological site using the rcarbon::binPrep function using the single linkage agglomeration method with an h-value of 100 (see {[}Supplementary Material{]} for a discussion of various h-values).
@@
 We constructed summed probability distributions (SPDs) from the calibrated dates using the \texttt{rcarbon::spd} function.
-We averaged probability distributions of binned dates, and then constructed SPDs from the resulting site-phase probability distributions.
+We averaged probability distributions of binned dates, and then constructed SPDs from the resulting site-phase probability distributions.\footnote{Using this method, a site-phase with a single date and a site-phase with 10 dates would each contribute equally to the resulting SPD.
+This method of developing the SPD is therefore a model of the relative abundance of site occupation events through time rather than of dates through time.
+This method does not account for investigation biases deriving from the definition of ``site'' or the decision to use a particular dating technique at a site in the first place.}
 
@@
 The tree-ring series --- the light orange line is the raw SPD, and the dark line is smoothed --- shows more variation on annual to decadal timescales, and yet matches the broad trends of the radiocarbon SPD.
-Roughly 150-year periodic fluctuations in the tree-ring SPD, described as periods of ``exploration'' and ``exploitation'' by {[}\protect\hyperlink{ref-Bocinsky2016}{23}{]}, are not present in the radiocarbon SPD, the result of a much smaller sample size of site-phases and lower temporal resolution of radiocarbon dating relative to tree-ring dating.
+Roughly 150-year periodic fluctuations in the tree-ring SPD, described as periods of ``exploration'' and ``exploitation'' by Bocinsky et al.
+(\protect\hyperlink{ref-Bocinsky2016}{2016}), are not present in the radiocarbon SPD, the result of a much smaller sample size of site-phases and lower temporal resolution of radiocarbon dating relative to tree-ring dating.
 
@@
 \caption{Radiocarbon and tree-ring density distributions in the UUSW (A), CMV (B), and NRG (C) regions, and VEP population estimates for the CMV and NRG.
-The dark orange tree-ring density curve smooths the raw tree-ring density (light orange) using a 21-year center-aligned Gaussian kernel with a 5-year standard deviation, following {[}\protect\hyperlink{ref-Bocinsky2016}{23}{]}.
+The dark orange tree-ring density curve smooths the raw tree-ring density (light orange) using a 21-year center-aligned Gaussian kernel with a 5-year standard deviation, following Bocinsky et al.
+(\protect\hyperlink{ref-Bocinsky2016}{2016}).
 In Panel A, the radiocarbon density (y-axis) is scaled by a factor of 3 to facilitate comparison with the tree-ring density over the target time period.
@@
 The high peak in the tree-ring record at AD 1250, far in excess of the VEP population estimate, is particularly interesting --- it likely signals the net impact of the immigration of many thousands of people to the NRG region, and the resulting (and simultaneous) burst in construction.
-Such a sudden and synchronous demographic and architectural event is hard to capture in the relatively lower resolution and sparsely sampled radiocarbon record, which tends to be smeared across sub-centennial scales courtesy of the calibration process {[}\protect\hyperlink{ref-Baillie1991}{34}{]}.
+Such a sudden and synchronous demographic and architectural event is hard to capture in the relatively lower resolution and sparsely sampled radiocarbon record, which tends to be smeared across sub-centennial scales courtesy of the calibration process (Baillie, \protect\hyperlink{ref-Baillie1991}{1991}).
 
@@
 In both the CMV and NRG the decline in the radiocarbon SPD begins earlier than in the VEP population series.
-This may indicate that radiocarbon time series make the decline seem to begin earlier than it actually does, which could be caused by a `cliff' in the radiocarbon calibration curve from AD 1200--1300 that is surrounded by two plateaus at AD 1100--1200 and AD 1400 {[}\protect\hyperlink{ref-Reimer2013}{33}{]}.
+This may indicate that radiocarbon time series make the decline seem to begin earlier than it actually does, which could be caused by a `cliff' in the radiocarbon calibration curve from AD 1200--1300 that is surrounded by two plateaus at AD 1100--1200 and AD 1400 (Reimer et al., \protect\hyperlink{ref-Reimer2013}{2013}).
 
@@
 These different trends between the VEP population reconstructions and dendrochronological and radiocarbon proxies reflect some investigation bias in both the radiocarbon and dendrochronological records at smaller, intra-regional scales, which calls for caution in the use of radiocarbon records to assess possible paleo-migration scenarios at finer spatial scales.
-This result reinforces other work that has called for the use of multiple proxies as cross-checks on radiocarbon time series {[}\protect\hyperlink{ref-Williams2012}{9}{]}.
+This result reinforces other work that has called for the use of multiple proxies as cross-checks on radiocarbon time series (Williams, \protect\hyperlink{ref-Williams2012}{2012}, p.
+@Crombe2014).
 
@@
 Similarities in the decline of both radiocarbon and dendrochronological records across the UUSW suggest that, at this particular regional scale, radiocarbon records reflect long-term, coarse-grained trends in paleodemography.
-Indeed, not only do we see corresponding declines in the radiocarbon and dendrochronological records at this large scale, but these patterns are also consistent with previous work on demographic and social change in the American Southwest: the Coalescent Communities Database, a Southwest-wide database of room counts and occupation spans of habitation sites with 12 rooms or more, supports a ``population decline that began between AD 1300 and 1350 and continued to European contact (Hill et al.~2004:Figures 2,4)'' {[}\protect\hyperlink{ref-Hill2004}{24},\protect\hyperlink{ref-Phillips2018}{35}{]}.
+Indeed, not only do we see corresponding declines in the radiocarbon and dendrochronological records at this large scale, but these patterns are also consistent with previous work on demographic and social change in the American Southwest: the Coalescent Communities Database, a Southwest-wide database of room counts and occupation spans of habitation sites with 12 rooms or more, supports a ``population decline that began between AD 1300 and 1350 and continued to European contact'' (Phillips et al., \protect\hyperlink{ref-Phillips2018}{2018}, following @Hill2004: Figures 2,4).
 
@@
 The results of this paper advance our confidence in radiocarbon records as a coarse-grained measure of the directionality of human paleodemographic trends.
-The methods developed can be used for other regions of the world that have dendrochronological records, such as Europe {[}\protect\hyperlink{ref-Kuniholm2001}{36}--\protect\hyperlink{ref-Hillam1990}{38}{]}, where comparisons with radiocarbon records could help to advance knowledge of Neolithic `booms and busts' of populations {[}\protect\hyperlink{ref-Shennan2013}{7},\protect\hyperlink{ref-Whitehouse2014}{39}{]} and to overcome limitations of the Bronze Age radiocarbon calibration plateau {[}\protect\hyperlink{ref-Armit2013}{18}{]}.
+The methods developed can be used for other regions of the world that have dendrochronological records, such as Europe (Haneca et al., \protect\hyperlink{ref-Haneca2009}{2009}; Hillam et al., \protect\hyperlink{ref-Hillam1990}{1990}; Kuniholm, \protect\hyperlink{ref-Kuniholm2001}{2001}), where comparisons with radiocarbon records could help to advance knowledge of Neolithic `booms and busts' of populations (Shennan et al., \protect\hyperlink{ref-Shennan2013}{2013}; Whitehouse et al., \protect\hyperlink{ref-Whitehouse2014}{2014}) and to overcome limitations of the Bronze Age radiocarbon calibration plateau (Armit et al., \protect\hyperlink{ref-Armit2013}{2013}).
 
@@
 \begin{cslreferences}
+\leavevmode\hypertarget{ref-Armit2013}{}%
+Armit, I., Swindles, G.T., Becker, K., 2013.
+From dates to demography in later prehistoric Ireland?
+Experimental approaches to the meta-analysis of large \textsuperscript{14}C data-sets.
+Journal of archaeological science 40, 433--438.
+\url{https://doi.org/10.1016/j.jas.2012.08.039}
+
+\leavevmode\hypertarget{ref-Baillie1991}{}%
+Baillie, M.G., 1991. Suck-in and smear: Two related chronological problems for the 90s. Journal of theoretical archaeology 2, 12--16.
+
+\leavevmode\hypertarget{ref-Bamforth2012}{}%
+Bamforth, D.B., Grund, B., 2012.
+Radiocarbon calibration curves, summed probability distributions, and early Paleoindian population trends in North America.
+Journal of Archaeological Science 39, 1768--1774.
+\url{https://doi.org/10.1016/j.jas.2012.01.017}
+
+\leavevmode\hypertarget{ref-Berry1982}{}%
+Berry, M.S., 1982. Time, space and transition in anasazi prehistory. University of Utah Press, Salt Lake City, Utah.
+
+\leavevmode\hypertarget{ref-rcarbon}{}%
+Bevan, A., Crema, E.R., 2020. rcarbon: Methods for calibrating and analysing radiocarbon dates.
+
 \leavevmode\hypertarget{ref-Bevan2019}{}%
-1.
-Bevan A, Palmisano A, Woodbridge J, Fyfe R, Roberts CN, Shennan S.
-2019 The changing face of the Mediterranean---Land cover, demography and environmental change: Introduction and overview.
-\emph{The Holocene} \textbf{29}, 703--707.
-(doi:\href{https://doi.org/10.1177/0959683619826688}{10.1177/0959683619826688})
+Bevan, A., Palmisano, A., Woodbridge, J., Fyfe, R., Roberts, C.N., Shennan, S., 2019.
+The changing face of the Mediterranean---Land cover, demography and environmental change: Introduction and overview.
+The Holocene 29, 703--707.
+\url{https://doi.org/10.1177/0959683619826688}
+
+\leavevmode\hypertarget{ref-Bluhm2019}{}%
+Bluhm, L.E., Surovell, T.A., 2019. Validation of a global model of taphonomic bias using geologic radiocarbon ages. Quaternary Research 91, 325--328. \url{https://doi.org/10.1017/qua.2018.78}
+
+\leavevmode\hypertarget{ref-Bocinsky2016}{}%
+Bocinsky, R.K., Rush, J., Kintigh, K.W., Kohler, T.A., 2016.
+Exploration and exploitation in the macrohistory of the pre-Hispanic Pueblo Southwest.
+Science Advances 2, e1501532.
+\url{https://doi.org/10.1126/sciadv.1501532}
+
+\leavevmode\hypertarget{ref-Boyer2010}{}%
+Boyer, J.L., Moore, J.L., Lakatos, S.A., Akins, N.J., Wilson, C.D., Blinman, E., 2010.
+Remodeling immigration: A Northern Rio Grande perspective on depopulation, migration, and donation-side models, in: Kohler, T., Varien, M., Wright, A.
+(Eds.), Leaving Mesa Verde: Peril and Change in the Thirteenth Century Southwest.
+The Amerind Foundation; University of Arizona Press, Tucson, Arizona, pp.
+285--323.
+
+\leavevmode\hypertarget{ref-Bronk2008}{}%
+Bronk Ramsey, C., 2008. Radiocarbon dating: Revolutions in understanding. Archaeometry 50, 249--275. \url{https://doi.org/10.1111/j.1475-4754.2008.00394.x}
+
+\leavevmode\hypertarget{ref-Cameron2010}{}%
+Cameron, C.M., 2010.
+Advances in understanding the thirteenth-century depopulation of the northern Southwest, in: Kohler, T., Varien, M., Wright, A.
+(Eds.), Leaving Mesa Verde: Peril and Change in the Thirteenth Century Southwest.
+The Amerind Foundation; University of Arizona Press, Tucson, Arizona, pp.
+346--363.
+
+\leavevmode\hypertarget{ref-Contreras2014}{}%
+Contreras, D.A., Meadows, J., 2014.
+Summed radiocarbon calibrations as a population proxy: A critical evaluation using a realistic simulation approach.
+Journal of Archaeological Science 52, 591--608.
+\url{https://doi.org/10.1016/j.jas.2014.05.030}
 
 \leavevmode\hypertarget{ref-Crema2016}{}%
-2.
-Crema ER, Habu J, Kobayashi K, Madella M.
-2016 Summed probability distribution of \textsuperscript{14}C dates suggests regional divergences in the population dynamics of the Jomon period in eastern Japan.
-\emph{PLoS One} \textbf{11}, e0154809.
-(doi:\href{https://doi.org/10.1371/journal.pone.0154809}{10.1371/journal.pone.0154809})
+Crema, E.R., Habu, J., Kobayashi, K., Madella, M., 2016.
+Summed probability distribution of \textsuperscript{14}C dates suggests regional divergences in the population dynamics of the Jomon period in eastern Japan.
+PLoS One 11, e0154809.
+\url{https://doi.org/10.1371/journal.pone.0154809}
+
+\leavevmode\hypertarget{ref-Crombe2014}{}%
+Crombé, P., Robinson, E., 2014.
+\textsuperscript{14}C dates as demographic proxies in Neolithisation models of northwestern Europe: A critical assessment using Belgium and northeast France as a case-study.
+Journal of Archaeological Science 52, 558--566.
+\url{https://doi.org/10.1016/j.jas.2014.02.001}
+
+\leavevmode\hypertarget{ref-Freeman2018PNAS}{}%
+Freeman, J., Baggio, J.A., Robinson, E., Byers, D.A., Gayo, E., Finley, J.B., Meyer, J.A., Kelly, R.L., Anderies, J.M., 2018.
+Synchronization of energy consumption by human societies throughout the Holocene.
+Proceedings of the National Academy of Sciences 115, 9962--9967.
+\url{https://doi.org/10.1073/pnas.1802859115}
+
+\leavevmode\hypertarget{ref-Freeman2018}{}%
+Freeman, J., Byers, D.A., Robinson, E., Kelly, R.L., 2018. Culture process and the interpretation of radiocarbon data. Radiocarbon 60, 453--467. \url{https://doi.org/10.1017/RDC.2017.124}
 
 \leavevmode\hypertarget{ref-Goldberg2016}{}%
-3.
-Goldberg A, Mychajliw AM, Hadly EA.
-2016 Post-invasion demography of prehistoric humans in South America.
-\emph{Nature} \textbf{532}, 232--235.
-(doi:\href{https://doi.org/10.1038/nature17176}{10.1038/nature17176})
+Goldberg, A., Mychajliw, A.M., Hadly, E.A., 2016. Post-invasion demography of prehistoric humans in South America. Nature 532, 232--235. \url{https://doi.org/10.1038/nature17176}
+
+\leavevmode\hypertarget{ref-Haneca2009}{}%
+Haneca, K., Čufar, K., Beeckman, H., 2009.
+Oaks, tree-rings and wooden cultural heritage: A review of the main characteristics and applications of oak dendrochronology in Europe.
+Journal of Archaeological Science 36, 1--11.
+\url{https://doi.org/10.1016/j.jas.2008.07.005}
+
+\leavevmode\hypertarget{ref-Hill2004}{}%
+Hill, J.B., Clark, J.J., Doelle, W.H., Lyons, P.D., 2004.
+Prehistoric demography in the Southwest: Migration, coalescence, and Hohokam population decline.
+American Antiquity 69, 689--716.
+\url{https://doi.org/10.2307/4128444}
+
+\leavevmode\hypertarget{ref-Hillam1990}{}%
+Hillam, J., Groves, C.M., Brown, D.M., Baillie, M.G., Coles, J.M., Coles, B.J., 1990. Dendrochronology of the English Neolithic. Antiquity 64, 210--220. \url{https://doi.org/10.1017/S0003598X00077826}
+
+\leavevmode\hypertarget{ref-Kohler2010b}{}%
+Kohler, T.A., Varien, M.D., Wright, A.M. (Eds.), 2010. Leaving Mesa Verde: Peril and Change in the Thirteenth-Century Southwest. The Amerind Foundation; University of Arizona Press, Tucson, Arizona.
+
+\leavevmode\hypertarget{ref-Kuniholm2001}{}%
+Kuniholm, P.I., 2001.
+Dendrochronology and other applications of tree-ring studies in archaeology, in: Brothwell, D.R., Pollard, A.M.
+(Eds.), The Handbook of Archaeological Sciences.
+John Wiley \& Sons, Ltd., London.
+
+\leavevmode\hypertarget{ref-rrtools}{}%
+Marwick, B., 2019. rrtools: Creates a Reproducible Research Compendium.
+
+\leavevmode\hypertarget{ref-Marwick2017}{}%
+Marwick, B., 2017.
+Computational reproducibility in archaeological research: Basic principles and a case study of their implementation.
+Journal of Archaeological Method and Theory 24, 424--450.
+\url{https://doi.org/10.1007/s10816-015-9272-9}
+
+\leavevmode\hypertarget{ref-Michczynski2006}{}%
+Michczynski, A., Michczynska, D.J., 2006. The effect of PDF peaks' height increase during calibration of radiocarbon date sets. Geochronometria 25, 1--4.
+
+\leavevmode\hypertarget{ref-Naudinot2014}{}%
+Naudinot, N., Tomasso, A., Tozzi, C., Peresani, M., 2014.
+Changes in mobility patterns as a factor of \textsuperscript{14}C date density variation in the Late Epigravettian of Northern Italy and Southeastern France.
+Journal of Archaeological Science 52, 578--590.
+\url{https://doi.org/10.1016/j.jas.2014.05.021}
+
+\leavevmode\hypertarget{ref-Ortman2016}{}%
+Ortman, S.G., 2016.
+Uniform probability density analysis and population history in the northern Rio Grande.
+Journal of Archaeological Method and Theory 23, 95--126.
+\url{https://doi.org/10.1007/s10816-014-9227-6}
 
 \leavevmode\hypertarget{ref-Palmisano2019}{}%
-4.
-Palmisano A \emph{et al.} 2019 Holocene landscape dynamics and long-term population trends in the Levant.
-\emph{The Holocene} \textbf{29}, 708--727.
-(doi:\href{https://doi.org/10.1177/0959683619826642}{10.1177/0959683619826642})
+Palmisano, A., Woodbridge, J., Roberts, C.N., Bevan, A., Fyfe, R., Shennan, S., Cheddadi, R., Greenberg, R., Kaniewski, D., Langgut, D., others, 2019.
+Holocene landscape dynamics and long-term population trends in the Levant.
+The Holocene 29, 708--727.
+\url{https://doi.org/10.1177/0959683619826642}
 
 \leavevmode\hypertarget{ref-Peros2010}{}%
-5.
-Peros MC, Munoz SE, Gajewski K, Viau AE.
-2010 Prehistoric demography of North America inferred from radiocarbon data.
-\emph{Journal of Archaeological Science} \textbf{37}, 656--664.
-(doi:\href{https://doi.org/10.1016/j.jas.2009.10.029}{10.1016/j.jas.2009.10.029})
+Peros, M.C., Munoz, S.E., Gajewski, K., Viau, A.E., 2010.
+Prehistoric demography of North America inferred from radiocarbon data.
+Journal of Archaeological Science 37, 656--664.
+\url{https://doi.org/10.1016/j.jas.2009.10.029}
+
+\leavevmode\hypertarget{ref-Phillips2018}{}%
+Phillips, D.A., Wearing, H.J., Clark, J.J., 2018.
+Village growth, emerging infectious disease, and the end of the Neolithic Demographic Transition in the Southwest United States and Northwest Mexico.
+American Antiquity 83, 263--280.
+\url{https://doi.org/10.1017/aaq.2018.3}
+
+\leavevmode\hypertarget{ref-Reimer2013}{}%
+Reimer, P.J., Bard, E., Bayliss, A., Beck, J.W., Blackwell, P.G., Ramsey, C.B., Buck, C.E., Cheng, H., Edwards, R.L., Friedrich, M., others, 2013.
+IntCal13 and Marine13 radiocarbon age calibration curves 0--50,000 years cal BP.
+Radiocarbon 55, 1869--1887.
+\url{https://doi.org/10.2458/azu_js_rc.55.16947}
+
+\leavevmode\hypertarget{ref-Rick1987}{}%
+Rick, J.W., 1987. Dates as data: An examination of the Peruvian preceramic radiocarbon record. American Antiquity 52, 55--73. \url{https://doi.org/10.2307/281060}
+
+\leavevmode\hypertarget{ref-Robinson2019}{}%
+Robinson, E., Nicholson, C., Kelly, R.L., 2019.
+The importance of spatial data to open-access national archaeological databases and the development of paleodemography research.
+Advances in Archaeological Practice 7, 395--408.
+\url{https://doi.org/10.1017/aap.2019.29}
 
 \leavevmode\hypertarget{ref-Russell2009}{}%
-6. Russell T, Steele J. 2009 A geo-referenced radiocarbon database for Early Iron Age sites in sub-Saharan Africa: Initial analysis. \emph{Southern African Humanities} \textbf{21}, 327--344.
+Russell, T., Steele, J., 2009. A geo-referenced radiocarbon database for Early Iron Age sites in sub-Saharan Africa: Initial analysis. Southern African Humanities 21, 327--344.
+
+\leavevmode\hypertarget{ref-Schwindt2016}{}%
+Schwindt, D.M., Bocinsky, R.K., Ortman, S.G., Glowacki, D.M., Varien, M.D., Kohler, T.A., 2016.
+The social consequences of climate change in the central Mesa Verde region.
+American Antiquity 81, 74--96.
 
 \leavevmode\hypertarget{ref-Shennan2013}{}%
-7.
-Shennan S, Downey SS, Timpson A, Edinborough K, Colledge S, Kerig T, Manning K, Thomas MG.
-2013 Regional population collapse followed initial agriculture booms in mid-holocene europe.
-\emph{Nature communications} \textbf{4}, 1--8.
+Shennan, S., Downey, S.S., Timpson, A., Edinborough, K., Colledge, S., Kerig, T., Manning, K., Thomas, M.G., 2013.
+Regional population collapse followed initial agriculture booms in mid-holocene europe.
+Nature communications 4, 1--8.
+
+\leavevmode\hypertarget{ref-Surovell2007}{}%
+Surovell, T.A., Brantingham, P.J., 2007. A note on the use of temporal frequency distributions in studies of prehistoric demography. Journal of Archaeological Science 34, 1868--1877.
+
+\leavevmode\hypertarget{ref-Surovell2009}{}%
+Surovell, T.A., Finley, J.B., Smith, G.M., Brantingham, P.J., Kelly, R., 2009. Correcting temporal frequency distributions for taphonomic bias. Journal of archaeological Science 36, 1715--1724.
+
+\leavevmode\hypertarget{ref-Timpson2014}{}%
+Timpson, A., Colledge, S., Crema, E., Edinborough, K., Kerig, T., Manning, K., Thomas, M.G., Shennan, S., 2014.
+Reconstructing regional population fluctuations in the european neolithic using radiocarbon dates: A new case-study using an improved method.
+Journal of Archaeological Science 52, 549--557.
 
 \leavevmode\hypertarget{ref-Wang2014}{}%
-8.
-Wang C, Lu H, Zhang J, Gu Z, He K.
-2014 Prehistoric demographic fluctuations in china inferred from radiocarbon data and their linkage with climate change over the past 50,000 years.
-\emph{Quaternary Science Reviews} \textbf{98}, 45--59.
+Wang, C., Lu, H., Zhang, J., Gu, Z., He, K., 2014.
+Prehistoric demographic fluctuations in china inferred from radiocarbon data and their linkage with climate change over the past 50,000 years.
+Quaternary Science Reviews 98, 45--59.
+
+\leavevmode\hypertarget{ref-Whitehouse2014}{}%
+Whitehouse, N.J., Schulting, R.J., McClatchie, M., Barratt, P., McLaughlin, T.R., Bogaard, A., Colledge, S., Marchant, R., Gaffrey, J., Bunting, M.J., 2014.
+Neolithic agriculture on the European western frontier: The boom and bust of early farming in Ireland.
+Journal of Archaeological Science 51, 181--205.
 
 \leavevmode\hypertarget{ref-Williams2012}{}%
-9. Williams AN. 2012 The use of summed radiocarbon probability distributions in archaeology: A review of methods. \emph{Journal of Archaeological Science} \textbf{39}, 578--589.
-
-\leavevmode\hypertarget{ref-Berry1982}{}%
-10. Berry MS. 1982 \emph{Time, space and transition in anasazi prehistory}. Salt Lake City, Utah: University of Utah Press.
-
-\leavevmode\hypertarget{ref-Rick1987}{}%
-11.
-Rick JW.
-1987 Dates as data: An examination of the Peruvian preceramic radiocarbon record.
-\emph{American Antiquity} \textbf{52}, 55--73.
-(doi:\href{https://doi.org/10.2307/281060}{10.2307/281060})
-
-\leavevmode\hypertarget{ref-Crombe2014}{}%
-12.
-Crombé P, Robinson E.
-2014 \textsuperscript{14}C dates as demographic proxies in Neolithisation models of northwestern Europe: A critical assessment using Belgium and northeast France as a case-study.
-\emph{Journal of Archaeological Science} \textbf{52}, 558--566.
-(doi:\href{https://doi.org/10.1016/j.jas.2014.02.001}{10.1016/j.jas.2014.02.001})
-
-\leavevmode\hypertarget{ref-Freeman2018}{}%
-13.
-Freeman J, Byers DA, Robinson E, Kelly RL.
-2018 Culture process and the interpretation of radiocarbon data.
-\emph{Radiocarbon} \textbf{60}, 453--467.
-(doi:\href{https://doi.org/10.1017/RDC.2017.124}{10.1017/RDC.2017.124})
-
-\leavevmode\hypertarget{ref-Naudinot2014}{}%
-14.
-Naudinot N, Tomasso A, Tozzi C, Peresani M.
-2014 Changes in mobility patterns as a factor of \textsuperscript{14}C date density variation in the Late Epigravettian of Northern Italy and Southeastern France.
-\emph{Journal of Archaeological Science} \textbf{52}, 578--590.
-(doi:\href{https://doi.org/10.1016/j.jas.2014.05.021}{10.1016/j.jas.2014.05.021})
-
-\leavevmode\hypertarget{ref-Bluhm2019}{}%
-15.
-Bluhm LE, Surovell TA.
-2019 Validation of a global model of taphonomic bias using geologic radiocarbon ages.
-\emph{Quaternary Research} \textbf{91}, 325--328.
-(doi:\href{https://doi.org/10.1017/qua.2018.78}{10.1017/qua.2018.78})
-
-\leavevmode\hypertarget{ref-Surovell2007}{}%
-16. Surovell TA, Brantingham PJ. 2007 A note on the use of temporal frequency distributions in studies of prehistoric demography. \emph{Journal of Archaeological Science} \textbf{34}, 1868--1877.
-
-\leavevmode\hypertarget{ref-Surovell2009}{}%
-17. Surovell TA, Finley JB, Smith GM, Brantingham PJ, Kelly R. 2009 Correcting temporal frequency distributions for taphonomic bias. \emph{Journal of archaeological Science} \textbf{36}, 1715--1724.
-
-\leavevmode\hypertarget{ref-Armit2013}{}%
-18.
-Armit I, Swindles GT, Becker K.
-2013 From dates to demography in later prehistoric Ireland?
-Experimental approaches to the meta-analysis of large \textsuperscript{14}C data-sets.
-\emph{Journal of archaeological science} \textbf{40}, 433--438.
-(doi:\href{https://doi.org/10.1016/j.jas.2012.08.039}{10.1016/j.jas.2012.08.039})
-
-\leavevmode\hypertarget{ref-Bamforth2012}{}%
-19.
-Bamforth DB, Grund B.
-2012 Radiocarbon calibration curves, summed probability distributions, and early Paleoindian population trends in North America.
-\emph{Journal of Archaeological Science} \textbf{39}, 1768--1774.
-(doi:\href{https://doi.org/10.1016/j.jas.2012.01.017}{10.1016/j.jas.2012.01.017})
-
-\leavevmode\hypertarget{ref-Contreras2014}{}%
-20.
-Contreras DA, Meadows J.
-2014 Summed radiocarbon calibrations as a population proxy: A critical evaluation using a realistic simulation approach.
-\emph{Journal of Archaeological Science} \textbf{52}, 591--608.
-(doi:\href{https://doi.org/10.1016/j.jas.2014.05.030}{10.1016/j.jas.2014.05.030})
-
-\leavevmode\hypertarget{ref-Michczynski2006}{}%
-21. Michczynski A, Michczynska DJ. 2006 The effect of PDF peaks' height increase during calibration of radiocarbon date sets. \emph{Geochronometria} \textbf{25}, 1--4.
-
-\leavevmode\hypertarget{ref-Timpson2014}{}%
-22.
-Timpson A, Colledge S, Crema E, Edinborough K, Kerig T, Manning K, Thomas MG, Shennan S.
-2014 Reconstructing regional population fluctuations in the european neolithic using radiocarbon dates: A new case-study using an improved method.
-\emph{Journal of Archaeological Science} \textbf{52}, 549--557.
-
-\leavevmode\hypertarget{ref-Bocinsky2016}{}%
-23.
-Bocinsky RK, Rush J, Kintigh KW, Kohler TA.
-2016 Exploration and exploitation in the macrohistory of the pre-Hispanic Pueblo Southwest.
-\emph{Science Advances} \textbf{2}, e1501532.
-(doi:\href{https://doi.org/10.1126/sciadv.1501532}{10.1126/sciadv.1501532})
-
-\leavevmode\hypertarget{ref-Hill2004}{}%
-24.
-Hill JB, Clark JJ, Doelle WH, Lyons PD.
-2004 Prehistoric demography in the Southwest: Migration, coalescence, and Hohokam population decline.
-\emph{American Antiquity} \textbf{69}, 689--716.
-(doi:\href{https://doi.org/10.2307/4128444}{10.2307/4128444})
-
-\leavevmode\hypertarget{ref-Kohler2010b}{}%
-25.
-Kohler TA, Varien MD, Wright AM, editors.
-2010 \emph{Leaving Mesa Verde: Peril and Change in the Thirteenth-Century Southwest}.
-Tucson, Arizona: The Amerind Foundation; University of Arizona Press.
-
-\leavevmode\hypertarget{ref-Robinson2019}{}%
-26.
-Robinson E, Nicholson C, Kelly RL.
-2019 The importance of spatial data to open-access national archaeological databases and the development of paleodemography research.
-\emph{Advances in Archaeological Practice} \textbf{7}, 395--408.
-(doi:\href{https://doi.org/10.1017/aap.2019.29}{10.1017/aap.2019.29})
-
-\leavevmode\hypertarget{ref-Freeman2018PNAS}{}%
-27.
-Freeman J, Baggio JA, Robinson E, Byers DA, Gayo E, Finley JB, Meyer JA, Kelly RL, Anderies JM.
-2018 Synchronization of energy consumption by human societies throughout the Holocene.
-\emph{Proceedings of the National Academy of Sciences} \textbf{115}, 9962--9967.
-(doi:\href{https://doi.org/10.1073/pnas.1802859115}{10.1073/pnas.1802859115})
-
-\leavevmode\hypertarget{ref-Cameron2010}{}%
-28.
-Cameron CM.
-2010 Advances in understanding the thirteenth-century depopulation of the northern Southwest.
-In \emph{Leaving mesa verde: Peril and change in the thirteenth century southwest} (eds T Kohler, M Varien, A Wright), pp.
-346--363.
-Tucson, Arizona: The Amerind Foundation; University of Arizona Press.
-
-\leavevmode\hypertarget{ref-Boyer2010}{}%
-29.
-Boyer JL, Moore JL, Lakatos SA, Akins NJ, Wilson CD, Blinman E.
-2010 Remodeling immigration: A Northern Rio Grande perspective on depopulation, migration, and donation-side models.
-In \emph{Leaving mesa verde: Peril and change in the thirteenth century southwest} (eds T Kohler, M Varien, A Wright), pp.
-285--323.
-Tucson, Arizona: The Amerind Foundation; University of Arizona Press.
-
-\leavevmode\hypertarget{ref-Ortman2016}{}%
-30.
-Ortman SG.
-2016 Uniform probability density analysis and population history in the northern Rio Grande.
-\emph{Journal of Archaeological Method and Theory} \textbf{23}, 95--126.
-(doi:\href{https://doi.org/10.1007/s10816-014-9227-6}{10.1007/s10816-014-9227-6})
-
-\leavevmode\hypertarget{ref-Schwindt2016}{}%
-31.
-Schwindt DM, Bocinsky RK, Ortman SG, Glowacki DM, Varien MD, Kohler TA.
-2016 The social consequences of climate change in the central Mesa Verde region.
-\emph{American Antiquity} \textbf{81}, 74--96.
-
-\leavevmode\hypertarget{ref-Bronk2008}{}%
-32.
-Bronk Ramsey C.
-2008 Radiocarbon dating: Revolutions in understanding.
-\emph{Archaeometry} \textbf{50}, 249--275.
-(doi:\href{https://doi.org/10.1111/j.1475-4754.2008.00394.x}{10.1111/j.1475-4754.2008.00394.x})
-
-\leavevmode\hypertarget{ref-Reimer2013}{}%
-33.
-Reimer PJ \emph{et al.} 2013 IntCal13 and Marine13 radiocarbon age calibration curves 0--50,000 years cal BP.
-\emph{Radiocarbon} \textbf{55}, 1869--1887.
-(doi:\href{https://doi.org/10.2458/azu_js_rc.55.16947}{10.2458/azu\_js\_rc.55.16947})
-
-\leavevmode\hypertarget{ref-Baillie1991}{}%
-34. Baillie MG. 1991 Suck-in and smear: Two related chronological problems for the 90s. \emph{Journal of theoretical archaeology} \textbf{2}, 12--16.
-
-\leavevmode\hypertarget{ref-Phillips2018}{}%
-35.
-Phillips DA, Wearing HJ, Clark JJ.
-2018 Village growth, emerging infectious disease, and the end of the Neolithic Demographic Transition in the Southwest United States and Northwest Mexico.
-\emph{American Antiquity} \textbf{83}, 263--280.
-(doi:\href{https://doi.org/10.1017/aaq.2018.3}{10.1017/aaq.2018.3})
-
-\leavevmode\hypertarget{ref-Kuniholm2001}{}%
-36.
-Kuniholm PI.
-2001 Dendrochronology and other applications of tree-ring studies in archaeology.
-In \emph{The handbook of archaeological sciences} (eds DR Brothwell, AM Pollard), London: John Wiley \& Sons, Ltd.
-
-\leavevmode\hypertarget{ref-Haneca2009}{}%
-37.
-Haneca K, Čufar K, Beeckman H.
-2009 Oaks, tree-rings and wooden cultural heritage: A review of the main characteristics and applications of oak dendrochronology in Europe.
-\emph{Journal of Archaeological Science} \textbf{36}, 1--11.
-(doi:\href{https://doi.org/10.1016/j.jas.2008.07.005}{10.1016/j.jas.2008.07.005})
-
-\leavevmode\hypertarget{ref-Hillam1990}{}%
-38.
-Hillam J, Groves CM, Brown DM, Baillie MG, Coles JM, Coles BJ.
-1990 Dendrochronology of the English Neolithic.
-\emph{Antiquity} \textbf{64}, 210--220.
-(doi:\href{https://doi.org/10.1017/S0003598X00077826}{10.1017/S0003598X00077826})
-
-\leavevmode\hypertarget{ref-Whitehouse2014}{}%
-39.
-Whitehouse NJ \emph{et al.} 2014 Neolithic agriculture on the European western frontier: The boom and bust of early farming in Ireland.
-\emph{Journal of Archaeological Science} \textbf{51}, 181--205.
-
-\leavevmode\hypertarget{ref-rcarbon}{}%
-40. Bevan A, Crema ER. 2020 \emph{rcarbon: Methods for calibrating and analysing radiocarbon dates}. See \url{https://CRAN.R-project.org/package=rcarbon}.
-
-\leavevmode\hypertarget{ref-Marwick2017}{}%
-41.
-Marwick B.
-2017 Computational reproducibility in archaeological research: Basic principles and a case study of their implementation.
-\emph{Journal of Archaeological Method and Theory} \textbf{24}, 424--450.
-(doi:\href{https://doi.org/10.1007/s10816-015-9272-9}{10.1007/s10816-015-9272-9})
-
-\leavevmode\hypertarget{ref-rrtools}{}%
-42. Marwick B. 2019 \emph{rrtools: Creates a Reproducible Research Compendium}. See \url{https://github.com/benmarwick/rrtools}.
+Williams, A.N., 2012. The use of summed radiocarbon probability distributions in archaeology: A review of methods. Journal of Archaeological Science 39, 578--589.
 \end{cslreferences}
@@
 
-To control for inconsistent sample size across the archaeological sites represented in the database, we aggregated dates known to be from the same phase of same archaeological site using the \texttt{rcarbon::binPrep} function using the ``single linkage'' agglomeration method with a cut width (or \(h\)-value) of 100 {[}\protect\hyperlink{ref-rcarbon}{40}{]}.
+To control for inconsistent sample size across the archaeological sites represented in the database, we aggregated dates known to be from the same phase of same archaeological site using the \texttt{rcarbon::binPrep} function using the ``single linkage'' agglomeration method with a cut width (or \(h\)-value) of 100 (Bevan and Crema, \protect\hyperlink{ref-rcarbon}{2020}).
 This clusters dates from the same site that are within 100 years of each other, and splits site occupations into distinct phases separated by 100 years or more.
@@
 
-This report is part of a research compendium developed using the \texttt{rrtools} package for reproducible research {[}\protect\hyperlink{ref-Marwick2017}{41}{]}.
-This report was generated on 2020-04-16 22:50:13 using the following computational environment and dependencies:
+This report is part of a research compendium developed using the \texttt{rrtools} package for reproducible research (Marwick, \protect\hyperlink{ref-rrtools}{2019}, \protect\hyperlink{ref-Marwick2017}{2017}).
+This report was generated on 2020-04-17 07:45:54 using the following computational environment and dependencies:
 
-\begin{verbatim}
-#> - Session info --------------------------------------------------------------------------------------------------------------------------------
-#>  setting  value                       
-#>  version  R version 3.6.3 (2020-02-29)
-#>  os       macOS Catalina 10.15.4      
-#>  system   x86_64, darwin15.6.0        
-#>  ui       RStudio                     
-#>  language (EN)                        
-#>  collate  en_US.UTF-8                 
-#>  ctype    en_US.UTF-8                 
-#>  tz       America/Denver              
-#>  date     2020-04-16                  
-#> 
-#> - Packages ------------------------------------------------------------------------------------------------------------------------------------
-#>  package        * version    date       lib source        
-#>  abind            1.4-5      2016-07-21 [1] CRAN (R 3.6.0)
-#>  assertthat       0.2.1      2019-03-21 [1] CRAN (R 3.6.3)
-#>  backports        1.1.6      2020-04-05 [1] CRAN (R 3.6.2)
-#>  bookdown         0.18       2020-03-05 [1] CRAN (R 3.6.0)
-#>  callr            3.4.3      2020-03-28 [1] CRAN (R 3.6.2)
-#>  class            7.3-15     2019-01-01 [1] CRAN (R 3.6.3)
-#>  classInt         0.4-3      2020-04-07 [1] CRAN (R 3.6.2)
-#>  cli              2.0.2      2020-02-28 [1] CRAN (R 3.6.3)
-#>  codetools        0.2-16     2018-12-24 [1] CRAN (R 3.6.3)
-#>  colorspace       1.4-1      2019-03-18 [1] CRAN (R 3.6.0)
-#>  crayon           1.3.4      2017-09-16 [1] CRAN (R 3.6.3)
-#>  crul             0.9.0      2019-11-06 [1] CRAN (R 3.6.0)
-#>  curl             4.3        2019-12-02 [1] CRAN (R 3.6.3)
-#>  DBI              1.1.0      2019-12-15 [1] CRAN (R 3.6.0)
-#>  deldir           0.1-25     2020-02-03 [1] CRAN (R 3.6.0)
-#>  desc             1.2.0      2018-05-01 [1] CRAN (R 3.6.3)
-#>  devtools         2.3.0      2020-04-10 [1] CRAN (R 3.6.3)
-#>  digest           0.6.25     2020-02-23 [1] CRAN (R 3.6.3)
-#>  doParallel       1.0.15     2019-08-02 [1] CRAN (R 3.6.0)
-#>  dplyr            0.8.5      2020-03-07 [1] CRAN (R 3.6.0)
-#>  e1071            1.7-3      2019-11-26 [1] CRAN (R 3.6.0)
-#>  egg              0.4.5      2019-07-13 [1] CRAN (R 3.6.0)
-#>  ellipsis         0.3.0      2019-09-20 [1] CRAN (R 3.6.3)
-#>  evaluate         0.14       2019-05-28 [1] CRAN (R 3.6.3)
-#>  fansi            0.4.1      2020-01-08 [1] CRAN (R 3.6.3)
-#>  farver           2.0.3      2020-01-16 [1] CRAN (R 3.6.0)
-#>  foreach          1.5.0      2020-03-30 [1] CRAN (R 3.6.3)
-#>  foreign          0.8-75     2020-01-20 [1] CRAN (R 3.6.3)
-#>  fs               1.4.1      2020-04-04 [1] CRAN (R 3.6.2)
-#>  generics         0.0.2      2018-11-29 [1] CRAN (R 3.6.0)
-#>  geojson          0.3.2      2019-01-31 [1] CRAN (R 3.6.0)
-#>  geojsonio        0.9.2      2020-04-07 [1] CRAN (R 3.6.3)
-#>  geojsonlint      0.4.0      2020-02-13 [1] CRAN (R 3.6.0)
-#>  ggplot2        * 3.3.0      2020-03-05 [1] CRAN (R 3.6.0)
-#>  git2r            0.26.1     2019-06-29 [1] CRAN (R 3.6.3)
-#>  glue             1.4.0      2020-04-03 [1] CRAN (R 3.6.2)
-#>  goftest          1.2-2      2019-12-02 [1] CRAN (R 3.6.0)
-#>  gridExtra        2.3        2017-09-09 [1] CRAN (R 3.6.0)
-#>  gtable           0.3.0      2019-03-25 [1] CRAN (R 3.6.0)
-#>  here             0.1        2017-05-28 [1] CRAN (R 3.6.0)
-#>  hms              0.5.3      2020-01-08 [1] CRAN (R 3.6.0)
-#>  htmltools        0.4.0      2019-10-04 [1] CRAN (R 3.6.3)
-#>  httpcode         0.3.0      2020-04-10 [1] CRAN (R 3.6.2)
-#>  httr             1.4.1      2019-08-05 [1] CRAN (R 3.6.3)
-#>  iterators        1.0.12     2019-07-26 [1] CRAN (R 3.6.0)
-#>  jqr              1.1.0      2018-10-22 [1] CRAN (R 3.6.0)
-#>  jsonlite         1.6.1      2020-02-02 [1] CRAN (R 3.6.3)
-#>  jsonvalidate     1.1.0      2019-06-25 [1] CRAN (R 3.6.0)
-#>  kableExtra       1.1.0      2019-03-16 [1] CRAN (R 3.6.0)
-#>  KernSmooth       2.23-16    2019-10-15 [1] CRAN (R 3.6.3)
-#>  knitr            1.28       2020-02-06 [1] CRAN (R 3.6.3)
-#>  labeling         0.3        2014-08-23 [1] CRAN (R 3.6.0)
-#>  lattice          0.20-38    2018-11-04 [1] CRAN (R 3.6.3)
-#>  lazyeval         0.2.2      2019-03-15 [1] CRAN (R 3.6.3)
-#>  lifecycle        0.2.0      2020-03-06 [1] CRAN (R 3.6.0)
-#>  lubridate        1.7.8      2020-04-06 [1] CRAN (R 3.6.2)
-#>  magrittr       * 1.5        2014-11-22 [1] CRAN (R 3.6.3)
-#>  maptools         0.9-9      2019-12-01 [1] CRAN (R 3.6.0)
-#>  Matrix           1.2-18     2019-11-27 [1] CRAN (R 3.6.3)
-#>  memoise          1.1.0      2017-04-21 [1] CRAN (R 3.6.3)
-#>  mgcv             1.8-31     2019-11-09 [1] CRAN (R 3.6.3)
-#>  munsell          0.5.0      2018-06-12 [1] CRAN (R 3.6.0)
-#>  nlme             3.1-144    2020-02-06 [1] CRAN (R 3.6.3)
-#>  pillar           1.4.3      2019-12-20 [1] CRAN (R 3.6.0)
-#>  pkgbuild         1.0.6      2019-10-09 [1] CRAN (R 3.6.3)
-#>  pkgconfig        2.0.3      2019-09-22 [1] CRAN (R 3.6.0)
-#>  pkgload          1.0.2      2018-10-29 [1] CRAN (R 3.6.3)
-#>  polyclip         1.10-0     2019-03-14 [1] CRAN (R 3.6.0)
-#>  prettyunits      1.1.1      2020-01-24 [1] CRAN (R 3.6.3)
-#>  processx         3.4.2      2020-02-09 [1] CRAN (R 3.6.3)
-#>  ps               1.3.2      2020-02-13 [1] CRAN (R 3.6.3)
-#>  purrr            0.3.3      2019-10-18 [1] CRAN (R 3.6.3)
-#>  R6               2.4.1      2019-11-12 [1] CRAN (R 3.6.3)
-#>  rcarbon          1.3.1      2020-03-18 [1] CRAN (R 3.6.0)
-#>  RColorBrewer     1.1-2      2014-12-07 [1] CRAN (R 3.6.0)
-#>  Rcpp             1.0.4      2020-03-17 [1] CRAN (R 3.6.0)
-#>  readr            1.3.1      2018-12-21 [1] CRAN (R 3.6.0)
-#>  remotes          2.1.1      2020-02-15 [1] CRAN (R 3.6.3)
-#>  rgeos            0.5-2      2019-10-03 [1] CRAN (R 3.6.0)
-#>  rlang            0.4.5      2020-03-01 [1] CRAN (R 3.6.3)
-#>  rmapshaper       0.4.4      2020-04-01 [1] CRAN (R 3.6.2)
-#>  rmarkdown        2.1        2020-01-20 [1] CRAN (R 3.6.0)
-#>  robinson2020   * 1.0.0.9000 2020-04-17 [1] local         
-#>  rpart            4.1-15     2019-04-12 [1] CRAN (R 3.6.3)
-#>  rprojroot        1.3-2      2018-01-03 [1] CRAN (R 3.6.3)
-#>  rstudioapi       0.11       2020-02-07 [1] CRAN (R 3.6.3)
-#>  rvest            0.3.5      2019-11-08 [1] CRAN (R 3.6.0)
-#>  scales           1.1.0      2019-11-18 [1] CRAN (R 3.6.0)
-#>  sessioninfo      1.1.1      2018-11-05 [1] CRAN (R 3.6.3)
-#>  sf               0.9-2      2020-04-14 [1] CRAN (R 3.6.2)
-#>  sp               1.4-1      2020-02-28 [1] CRAN (R 3.6.0)
-#>  spatstat         1.63-3     2020-03-01 [1] CRAN (R 3.6.0)
-#>  spatstat.data    1.4-3      2020-01-26 [1] CRAN (R 3.6.0)
-#>  spatstat.utils   1.17-0     2020-02-07 [1] CRAN (R 3.6.0)
-#>  stringi          1.4.6      2020-02-17 [1] CRAN (R 3.6.3)
-#>  stringr          1.4.0      2019-02-10 [1] CRAN (R 3.6.3)
-#>  tensor           1.5        2012-05-05 [1] CRAN (R 3.6.0)
-#>  testthat         2.3.2      2020-03-02 [1] CRAN (R 3.6.3)
-#>  tibble           3.0.0      2020-03-30 [1] CRAN (R 3.6.2)
-#>  tidyr            1.0.2      2020-01-24 [1] CRAN (R 3.6.0)
-#>  tidyselect       1.0.0      2020-01-27 [1] CRAN (R 3.6.0)
-#>  tinytex          0.21       2020-03-31 [1] CRAN (R 3.6.2)
-#>  units            0.6-6      2020-03-16 [1] CRAN (R 3.6.0)
-#>  usethis          1.6.0      2020-04-09 [1] CRAN (R 3.6.3)
-#>  V8               3.0.2      2020-03-14 [1] CRAN (R 3.6.0)
-#>  vctrs            0.2.4      2020-03-10 [1] CRAN (R 3.6.0)
-#>  viridisLite      0.3.0      2018-02-01 [1] CRAN (R 3.6.0)
-#>  webshot          0.5.2      2019-11-22 [1] CRAN (R 3.6.0)
-#>  withr            2.1.2      2018-03-15 [1] CRAN (R 3.6.3)
-#>  xfun             0.13       2020-04-13 [1] CRAN (R 3.6.2)
-#>  xml2             1.3.1      2020-04-09 [1] CRAN (R 3.6.2)
-#>  yaml             2.2.1      2020-02-01 [1] CRAN (R 3.6.3)
-#> 
-#> [1] /Library/Frameworks/R.framework/Versions/3.6/Resources/library
-\end{verbatim}
+\textbf{R version 3.6.3 (2020-02-29)}
 
-The current Git commit details are:
+\textbf{Platform:} x86\_64-apple-darwin15.6.0 (64-bit)
 
-\begin{verbatim}
-#> Local:    master /Users/bocinsky/git/publications/robinson2020
-#> Remote:   master @ origin (https://github.com/people3k/robinson2020)
-#> Head:     [802ee87] 2020-03-24: working towards resubmission
-\end{verbatim}
+\textbf{locale:}
+en\_US.UTF-8\textbar\textbar en\_US.UTF-8\textbar\textbar en\_US.UTF-8\textbar\textbar C\textbar\textbar en\_US.UTF-8\textbar\textbar en\_US.UTF-8
+
+\textbf{attached base packages:}
+\emph{stats}, \emph{graphics}, \emph{grDevices}, \emph{utils}, \emph{datasets}, \emph{methods} and \emph{base}
+
+\textbf{other attached packages:}
+\emph{ggplot2(v.3.3.0)}, \emph{robinson2020(v.1.0.0.9000)} and \emph{magrittr(v.1.5)}
+
+\textbf{loaded via a namespace (and not attached):}
+\emph{Rcpp(v.1.0.4)}, \emph{compiler(v.3.6.3)}, \emph{pillar(v.1.4.3)}, \emph{tools(v.3.6.3)}, \emph{digest(v.0.6.25)}, \emph{evaluate(v.0.14)}, \emph{lifecycle(v.0.2.0)}, \emph{tibble(v.3.0.0)}, \emph{gtable(v.0.3.0)}, \emph{pkgconfig(v.2.0.3)}, \emph{rlang(v.0.4.5)}, \emph{cli(v.2.0.2)}, \emph{yaml(v.2.2.1)}, \emph{xfun(v.0.13)}, \emph{withr(v.2.1.2)}, \emph{stringr(v.1.4.0)}, \emph{dplyr(v.0.8.5)}, \emph{knitr(v.1.28)}, \emph{vctrs(v.0.2.4)}, \emph{rprojroot(v.1.3-2)}, \emph{grid(v.3.6.3)}, \emph{tidyselect(v.1.0.0)}, \emph{glue(v.1.4.0)}, \emph{here(v.0.1)}, \emph{R6(v.2.4.1)}, \emph{fansi(v.0.4.1)}, \emph{rmarkdown(v.2.1)}, \emph{bookdown(v.0.18)}, \emph{pander(v.0.6.3)}, \emph{purrr(v.0.3.3)}, \emph{backports(v.1.1.6)}, \emph{scales(v.1.1.0)}, \emph{ellipsis(v.0.3.0)}, \emph{htmltools(v.0.4.0)}, \emph{assertthat(v.0.2.1)}, \emph{colorspace(v.1.4-1)}, \emph{stringi(v.1.4.6)}, \emph{munsell(v.0.5.0)} and \emph{crayon(v.1.3.4)}
 

### 2020-04-17 analysis/robinson2020.Rmd
bumping to r 3.6.3 in dockerfile

@@
 
-Our final dataset includes `r nrow(all_data$Radiocarbon) %>% format(big.mark = ',')` radiocarbon dates from `r all_data$Radiocarbon$SiteID %>% unique() %>% length %>% format(big.mark = ',')` unique sites, dating from `r max(all_data$Radiocarbon$Date_BP)`--`r min(all_data$Radiocarbon$Date_BP)` BP (uncalibrated; `r -1 * (1950 - max(all_data$Radiocarbon$Date_BP)) - 1` BC--`r 1950 - min(all_data$Radiocarbon$Date_BP)` AD) and `r nrow(all_data[['Tree-ring']]) %>% format(big.mark = ',')` tree-ring dates from `r all_data[["Tree-ring"]]$SiteID %>% unique() %>% length %>% format(big.mark = ',')` unique sites, dating from `r max(all_data[['Tree-ring']]$Date_BP)` -- `r min(all_data[['Tree-ring']]$Date_BP)` BP (`r 1950 - max(all_data[['Tree-ring']]$Date_BP)`--`r 1950 - min(all_data[['Tree-ring']]$Date_BP)` AD).
+Our final dataset includes `r nrow(all_data$Radiocarbon) %>% format(big.mark = ',')` radiocarbon dates from `r all_data$Radiocarbon$SiteID %>% unique() %>% length %>% format(big.mark = ',')` unique sites, dating from `r max(all_data$Radiocarbon$Date_BP)`--`r min(all_data$Radiocarbon$Date_BP)` BP (uncalibrated; `r -1 * (1950 - max(all_data$Radiocarbon$Date_BP)) - 1` BC--`r 1950 - min(all_data$Radiocarbon$Date_BP)` AD) and `r nrow(all_data[['Tree-ring']]) %>% format(big.mark = ',')` tree-ring dates from `r all_data[["Tree-ring"]]$SiteID %>% unique() %>% length %>% format(big.mark = ',')` unique sites, dating from `r max(all_data[['Tree-ring']]$Date_BP)`--`r min(all_data[['Tree-ring']]$Date_BP)` BP (`r 1950 - max(all_data[['Tree-ring']]$Date_BP)`--`r 1950 - min(all_data[['Tree-ring']]$Date_BP)` AD).
 Table \@ref(tab:Table1) presents the counts of dates and sites of each date type in the CMV and NRG study areas.
@@
 
+rm(radiocarbon_spd, dendro_spd)
+gc()
+gc()
+
 vep_demography <-

### 2020-04-17 analysis/robinson2020.tex
bumping to r 3.6.3 in dockerfile

@@
 
-Our final dataset includes 1,531 radiocarbon dates from 482 unique sites, dating from 2260--90 BP (uncalibrated; 309 BC--1860 AD) and 15,077 tree-ring dates from 770 unique sites, dating from 1749 -- 152 BP (201--1798 AD).
+Our final dataset includes 1,531 radiocarbon dates from 482 unique sites, dating from 2260--90 BP (uncalibrated; 309 BC--1860 AD) and 15,077 tree-ring dates from 770 unique sites, dating from 1749--152 BP (201--1798 AD).
 Table \ref{tab:Table1} presents the counts of dates and sites of each date type in the CMV and NRG study areas.
@@
 This report is part of a research compendium developed using the \texttt{rrtools} package for reproducible research (Marwick, \protect\hyperlink{ref-rrtools}{2019}, \protect\hyperlink{ref-Marwick2017}{2017}).
-This report was generated on 2020-04-17 07:45:54 using the following computational environment and dependencies:
+This report was generated on 2020-04-17 08:14:58 using the following computational environment and dependencies:
 

### 2020-04-17 analysis/robinson2020.Rmd
bumping to r 3.6.3 in dockerfile

@@
       fig_caption: yes
+      extra_dependencies:
+        caption: ["labelfont={bf}"]
       pandoc_args:
@@
 
-We draw on the tree-ring date database presented in [@Bocinsky2016].
+We draw on the tree-ring date database presented by @Bocinsky2016.
 These data were compiled by Tim Kohler and Rebecca Higgins from many contributors, and the vast majority of dates were produced by the Laboratory of Tree Ring Research at the University of Arizona.
@@
 Here, we only use the dates that are most likely to be accurate — cutting dates (‘B’, ‘G’, ‘L’, ‘c’, or ‘r’ dates, with or without a ‘+’) for which we know the year the tree died, and near-cutting dates (‘v’ or ‘v+’ dates) which are likely within 0--3 years of the true date of the outermost ring.
-The [@Bocinsky2016] data contain site locations; for this analysis, we first spatially cropped the database to our study area, and then translated those locations to the county level.
+The @Bocinsky2016 data contain site locations; for this analysis, we first spatially cropped the database to our study area, and then translated those locations to the county level.
 We converted the original date determinations to BP to match calibrated radiocarbon year conventions, and retained dates within 1750--150 BP.
@@
 rm(radiocarbon_spd, dendro_spd)
-gc()
-gc()
+gc(verbose = FALSE)
+gc(verbose = FALSE)
 
@@
 Similarities in the decline of both radiocarbon and dendrochronological records across the UUSW suggest that, at this particular regional scale, radiocarbon records reflect long-term, coarse-grained trends in paleodemography.
-Indeed, not only do we see corresponding declines in the radiocarbon and dendrochronological records at this large scale, but these patterns are also consistent with previous work on demographic and social change in the American Southwest: the Coalescent Communities Database, a Southwest-wide database of room counts and occupation spans of habitation sites with 12 rooms or more, supports a “population decline that began between AD 1300 and 1350 and continued to European contact” [@Phillips2018, following @Hill2004: Figures 2,4].
+Indeed, not only do we see corresponding declines in the radiocarbon and dendrochronological records at this large scale, but these patterns are also consistent with previous work on demographic and social change in the American Southwest: the Coalescent Communities Database, a Southwest-wide database of room counts and occupation spans of habitation sites with 12 rooms or more, supports a “population decline that began between AD 1300 and 1350 and continued to European contact” [@Phillips2018; following @Hill2004: Figures 2,4].
 

### 2020-04-17 analysis/robinson2020.tex
bumping to r 3.6.3 in dockerfile

@@
 % Pandoc header
+\usepackage[labelfont={bf}]{caption}
 
@@
 
-We draw on the tree-ring date database presented in (Bocinsky et al., \protect\hyperlink{ref-Bocinsky2016}{2016}).
+We draw on the tree-ring date database presented by Bocinsky et al.
+(\protect\hyperlink{ref-Bocinsky2016}{2016}).
 These data were compiled by Tim Kohler and Rebecca Higgins from many contributors, and the vast majority of dates were produced by the Laboratory of Tree Ring Research at the University of Arizona.
@@
 Here, we only use the dates that are most likely to be accurate --- cutting dates (`B', `G', `L', `c', or `r' dates, with or without a `+') for which we know the year the tree died, and near-cutting dates (`v' or `v+' dates) which are likely within 0--3 years of the true date of the outermost ring.
-The (Bocinsky et al., \protect\hyperlink{ref-Bocinsky2016}{2016}) data contain site locations; for this analysis, we first spatially cropped the database to our study area, and then translated those locations to the county level.
+The Bocinsky et al.
+(\protect\hyperlink{ref-Bocinsky2016}{2016}) data contain site locations; for this analysis, we first spatially cropped the database to our study area, and then translated those locations to the county level.
 We converted the original date determinations to BP to match calibrated radiocarbon year conventions, and retained dates within 1750--150 BP.
@@
 
+\begin{verbatim}
+#>           used (Mb) gc trigger   (Mb) limit (Mb)  max used   (Mb)
+#> Ncells  716943 38.3    1232191   65.9         NA   1232191   65.9
+#> Vcells 1639461 12.6  508769831 3881.7     716800 507690663 3873.4
+#>           used (Mb) gc trigger   (Mb) limit (Mb)  max used   (Mb)
+#> Ncells  717083 38.3    1232191   65.9         NA   1232191   65.9
+#> Vcells 1639786 12.6  407015865 3105.3     716800 507690663 3873.4
+\end{verbatim}
+
 \begin{figure}
@@
 Similarities in the decline of both radiocarbon and dendrochronological records across the UUSW suggest that, at this particular regional scale, radiocarbon records reflect long-term, coarse-grained trends in paleodemography.
-Indeed, not only do we see corresponding declines in the radiocarbon and dendrochronological records at this large scale, but these patterns are also consistent with previous work on demographic and social change in the American Southwest: the Coalescent Communities Database, a Southwest-wide database of room counts and occupation spans of habitation sites with 12 rooms or more, supports a ``population decline that began between AD 1300 and 1350 and continued to European contact'' (Phillips et al., \protect\hyperlink{ref-Phillips2018}{2018}, following @Hill2004: Figures 2,4).
+Indeed, not only do we see corresponding declines in the radiocarbon and dendrochronological records at this large scale, but these patterns are also consistent with previous work on demographic and social change in the American Southwest: the Coalescent Communities Database, a Southwest-wide database of room counts and occupation spans of habitation sites with 12 rooms or more, supports a ``population decline that began between AD 1300 and 1350 and continued to European contact'' (Phillips et al., \protect\hyperlink{ref-Phillips2018}{2018}; following Hill et al., \protect\hyperlink{ref-Hill2004}{2004}: Figures 2,4).
 
@@
 This report is part of a research compendium developed using the \texttt{rrtools} package for reproducible research (Marwick, \protect\hyperlink{ref-rrtools}{2019}, \protect\hyperlink{ref-Marwick2017}{2017}).
-This report was generated on 2020-04-17 08:14:58 using the following computational environment and dependencies:
+This report was generated on 2020-04-17 08:31:33 using the following computational environment and dependencies:
 

### 2020-04-17 analysis/robinson2020.Rmd
bumping to r 3.6.3 in dockerfile

@@
 journal: Preprint
-documentclass: sa
+# documentclass: sa
 bibliography: robinson2020_references.bib
@@
     bookdown::pdf_document2:
-      template: templates/template.tex
+      # template: templates/template.tex
       latex_engine: xelatex
@@
       fig_caption: yes
+      toc: false
       extra_dependencies:
@@
       - --lua-filter=templates/author-info-blocks.lua
-      - --lua-filter=templates/pagebreak.lua
+      # - --lua-filter=templates/pagebreak.lua
 editor_options: 
@@
 
-rm(radiocarbon_spd, dendro_spd)
-gc(verbose = FALSE)
-gc(verbose = FALSE)
-
 vep_demography <-

### 2020-04-17 analysis/robinson2020.tex
bumping to r 3.6.3 in dockerfile

@@
+% Options for packages loaded elsewhere
+\PassOptionsToPackage{unicode}{hyperref}
+\PassOptionsToPackage{hyphens}{url}
+%
 \documentclass[
-]{sa}
-
+]{article}
+\usepackage{lmodern}
+\usepackage{amssymb,amsmath}
+\usepackage{ifxetex,ifluatex}
+\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
+  \usepackage[T1]{fontenc}
+  \usepackage[utf8]{inputenc}
+  \usepackage{textcomp} % provide euro and other symbols
+\else % if luatex or xetex
+  \usepackage{unicode-math}
+  \defaultfontfeatures{Scale=MatchLowercase}
+  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
+\fi
+% Use upquote if available, for straight quotes in verbatim environments
+\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
+\IfFileExists{microtype.sty}{% use microtype if available
+  \usepackage[]{microtype}
+  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
+}{}
+\makeatletter
+\@ifundefined{KOMAClassName}{% if non-KOMA class
+  \IfFileExists{parskip.sty}{%
+    \usepackage{parskip}
+  }{% else
+    \setlength{\parindent}{0pt}
+    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
+}{% if KOMA class
+  \KOMAoptions{parskip=half}}
+\makeatother
+\usepackage{xcolor}
+\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
+\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
+\hypersetup{
+  pdfauthor={Erick Robinson1,; Kyle Bocinsky2,3; Darcy Bird4; Jacob Freeman1; Robert L. Kelly5},
+  pdfkeywords={true},
+  hidelinks,
+  pdfcreator={LaTeX via pandoc}}
+\urlstyle{same} % disable monospaced font for URLs
+\usepackage[margin=1in]{geometry}
+\usepackage{longtable,booktabs}
+% Correct order of tables after \paragraph or \subparagraph
+\usepackage{etoolbox}
+\makeatletter
+\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
+\makeatother
+% Allow footnotes in longtable head/foot
+\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
+\makesavenoteenv{longtable}
+\usepackage{graphicx}
+\makeatletter
+\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
+\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
+\makeatother
+% Scale images if necessary, so that they will not overflow the page
+% margins by default, and it is still possible to overwrite the defaults
+% using explicit options in \includegraphics[width, height, ...]{}
+\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
+% Set default figure placement to htbp
+\makeatletter
+\def\fps@figure{htbp}
+\makeatother
+\setlength{\emergencystretch}{3em} % prevent overfull lines
 \providecommand{\tightlist}{%
   \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
+\setcounter{secnumdepth}{5}
+\usepackage[labelfont={bf}]{caption}
+\newlength{\cslhangindent}
+\setlength{\cslhangindent}{1.5em}
+\newenvironment{cslreferences}%
+  {\setlength{\parindent}{0pt}%
+  \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces}%
+  {\par}
 
-\author{
-Erick Robinson\textsuperscript{1,*}
+\title{true}
+\author{Erick Robinson\textsuperscript{1,*} \and Kyle Bocinsky\textsuperscript{2,3} \and Darcy Bird\textsuperscript{4} \and Jacob Freeman\textsuperscript{1} \and Robert L. Kelly\textsuperscript{5}}
+\date{April 17, 2020}
 
-\And
-Kyle Bocinsky\textsuperscript{2,3}
-
-\And
-Darcy Bird\textsuperscript{4}
-
-\And
-Jacob Freeman\textsuperscript{1}
-
-\And
-Robert L. Kelly\textsuperscript{5}
-
-}
-
-\title{Dendrochronological dates confirm a Late Prehistoric population decline in the American Southwest derived from radiocarbon dates}
-
-\Plainauthor{Erick Robinson\textsuperscript{1,*}, Kyle Bocinsky\textsuperscript{2,3}, Darcy Bird\textsuperscript{4}, Jacob Freeman\textsuperscript{1}, Robert L. Kelly\textsuperscript{5}}
-
-\Shorttitle{Dendrochronological and radiocarbon dates in the American Southwest}
-
-\Abstract{
+\begin{document}
+\maketitle
+\begin{abstract}
 The northern American Southwest provides one of the most well-documented cases of human population growth and decline in the world.
@@
 These results lend confidence to single proxy radiocarbon-based reconstructions of paleodemography outside the Southwest that suggest post-AD 1300 population declines in many parts of North America.
-}
-
-\Keywords{dates-as-data, radiocarbon summed probability distributions, dendrochronology, palaeodemography, southwestern United States}
-
-%% publication information
-%% \Volume{50}
-%% \Issue{9}
-%% \Month{June}
-%% \Year{2012}
-%% \Submitdate{}
-%% \Acceptdate{2012-06-04}
-
-\Address{
-          }
-
-%% Publication information
-%% Edit as needed depending on status of manuscript.
-%% Replace Preprint with journal short name of accepted
-%% Pubs if copyright is open compliant. Add punctuation
-%% for formatting as desired.
-\Journal{Preprint}
-% \Volume{}
-% \Issue{} %Enter as "(Issue)" for APA
-% \Pages{} %Enter as "pp. m---n" for APA
-% \Year{}
-\Submitdate{\today} %Enter date you submitted preprint for review
-\Acceptdate{\today} %Enter date your paper was accepted for publication
-\DOI{
-DOI HERE
-} %Enter final DOI here
-
-\newlength{\cslhangindent}
-\setlength{\cslhangindent}{1.5em}
-\newenvironment{cslreferences}%
-  {\setlength{\parindent}{0pt}%
-  \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces}%
-  {\par}
-
-% Pandoc header
-\usepackage[labelfont={bf}]{caption}
-
-\begin{document}
+\end{abstract}
 
@@
 (\protect\hyperlink{ref-Bocinsky2016}{2016}), are not present in the radiocarbon SPD, the result of a much smaller sample size of site-phases and lower temporal resolution of radiocarbon dating relative to tree-ring dating.
-
-\begin{verbatim}
-#>           used (Mb) gc trigger   (Mb) limit (Mb)  max used   (Mb)
-#> Ncells  716943 38.3    1232191   65.9         NA   1232191   65.9
-#> Vcells 1639461 12.6  508769831 3881.7     716800 507690663 3873.4
-#>           used (Mb) gc trigger   (Mb) limit (Mb)  max used   (Mb)
-#> Ncells  717083 38.3    1232191   65.9         NA   1232191   65.9
-#> Vcells 1639786 12.6  407015865 3105.3     716800 507690663 3873.4
-\end{verbatim}
 
@@
 This report is part of a research compendium developed using the \texttt{rrtools} package for reproducible research (Marwick, \protect\hyperlink{ref-rrtools}{2019}, \protect\hyperlink{ref-Marwick2017}{2017}).
-This report was generated on 2020-04-17 08:31:33 using the following computational environment and dependencies:
+This report was generated on 2020-04-17 11:04:15 using the following computational environment and dependencies:
 

### 2020-04-17 analysis/robinson2020.Rmd
bumping to r 3.6.3 in dockerfile

@@
       # template: templates/template.tex
-      latex_engine: xelatex
-      keep_tex: true
+      latex_engine: lualatex
+      # keep_tex: true
       fig_caption: yes

### 2020-04-17 analysis/robinson2020.Rmd
bumping to r 3.6.3 in dockerfile

@@
       latex_engine: lualatex
-      # keep_tex: true
       fig_caption: yes
@@
       - --lua-filter=templates/author-info-blocks.lua
-      # - --lua-filter=templates/pagebreak.lua
+      - --lua-filter=templates/pagebreak.lua
 editor_options: 
@@
 if (requireNamespace("pander", quietly = TRUE)) pander::pander(sessionInfo())
+gc()
+gc()
 ```

### 2020-04-20 analysis/robinson2020.Rmd
debugging latex

@@
       correspondence: true
-  - Kyle Bocinsky:
+  - R. Kyle Bocinsky:
       email: kbocinsky@crowcanyon.org
@@
 journal: Preprint
-# documentclass: sa
+documentclass: sa
 bibliography: robinson2020_references.bib
 # csl: "philosophical-transactions-of-the-royal-society-b.csl"
-csl: "journal-of-archaeological-science.csl"
+# csl: "journal-of-archaeological-science.csl"
+csl: "apa.csl"
 link-citations: true
@@
     bookdown::pdf_document2:
-      # template: templates/template.tex
+      template: templates/template.tex
       latex_engine: lualatex
@@
   chunk_output_type: console
+header-includes:
+  \newcommand{\beginsupplement}{\setcounter{table}{0}  \renewcommand{\thetable}{S\arabic{table}} \setcounter{figure}{0}       \renewcommand{\thefigure}{S\arabic{figure}}}
 ---
@@
     caption = "Counts of radiocarbon and tree-ring dates and sites in this study.",
+    format = "latex",
     booktabs = TRUE,
@@
 \newpage
-
 # References {-}
@@
 \newpage
-
+\beginsupplement
 # Supplementary Material: Cut-width sensitivity for temporal binning {-}
@@
 \newpage
-
 # Colophon {-}
-
 This report is part of a research compendium developed using the `rrtools` package for reproducible research [@Marwick2017; @rrtools].
@@
 # which R packages and versions?
-if (requireNamespace("pander", quietly = TRUE)) pander::pander(sessionInfo())
-gc()
-gc()
+sess <- sessionInfo()
+sess$basePkgs %<>% sort()
+sess$loadedOnly %<>% magrittr::extract(sort(names(sess$loadedOnly)))
+if (requireNamespace("pander", quietly = TRUE)) pander::pander(sess)
 ```

### 2020-04-20 analysis/test.Rmd
debugging latex

--- 2020-04-20 analysis/robinson2020.Rmd
@@
       toc: false
+      # tables: true
       extra_dependencies:
         caption: ["labelfont={bf}"]
+        # colortbl: null
       pandoc_args:
@@
   geom_sf(data = four_corners_states) +
-  geom_raster(
-    data = robinson2020::swus_hillshade_500m,
-    mapping = aes(
-      x = x,
-      y = y,
-      alpha = ID
-    ),
-    na.rm = TRUE
-  ) +
+  # geom_raster(
+  #   data = robinson2020::swus_hillshade_500m,
+  #   mapping = aes(
+  #     x = x,
+  #     y = y,
+  #     alpha = ID
+  #   ),
+  #   na.rm = TRUE
+  # ) +
   scale_alpha(
@@
 ```
-
-Date densities across the study area were very uneven, with high counts of radiocarbon dates deriving from Navajo County, Arizona, and almost a third of the tree-ring dates in our study being from Montezuma County, Colorado.
-Below, we overcome some of these sampling issues by binning dates by site and occupation phase.
-Still, these disparities likely demonstrate more fundamental differences in research design across archaeological projects.
-There are very few radiocarbon dates from the two VEP study areas, indicating a clear bias towards using tree-ring dating where (and when) possible.
-Figure \@ref(fig:Figure2) presents the number of radiocarbon and tree-ring dates by county within our study area.
-Figure \@ref(fig:Figure3) presents the number of unique sites with radiocarbon and tree-ring dates by county within our study area.
-There are similar ratios of radiocarbon dates to sites with radiocarbon dates for the CMV and NRG (roughly 2:1), but sites in the CMV tend to have a third more tree-ring dates than those in the NRG (20:1 versus 15:1, respectively).
-
-```{r Figure2, fig.height=9, fig.width=6.5, fig.cap='The number of radiocarbon and tree-ring dates by county in this study.', fig.show = 'hold', out.width = '45%'}
-county_counts <-
-  all_data %>%
-  dplyr::bind_rows() %>%
-  dplyr::group_by(Type, County) %>%
-  dplyr::summarise(Dates = dplyr::n()) %>%
-  dplyr::left_join(uusw_counties,
-    by = "County"
-  ) %>%
-  sf::st_as_sf()
-
-theme_map <- function(base_size = 6.5,
-                      base_family = "") {
-  theme_bw(
-    base_size = base_size,
-    base_family = base_family
-  ) %+replace%
-    theme(
-      axis.line = element_blank(),
-      axis.text = element_blank(),
-      axis.ticks = element_blank(),
-      axis.title = element_blank(),
-      panel.background = element_blank(),
-      panel.border = element_blank(),
-      panel.grid = element_blank(),
-      panel.grid.major = element_line(colour = "transparent"),
-      panel.grid.minor = element_line(colour = "transparent"),
-      legend.key = element_blank(),
-      legend.position = "bottom",
-      legend.background = element_blank(),
-      legend.key.width = unit(0.15, "npc"),
-      legend.text = element_text(size = rel(1)),
-      plot.background = element_blank(),
-      plot.margin = margin(t = 0, r = 0, b = 0, l = 0, unit = "npc")
-    )
-}
-
-county_counts %>%
-  dplyr::filter(Type == "Radiocarbon") %>%
-  ggplot() +
-  geom_sf(
-    data = four_corners_states,
-    fill = NA
-  ) +
-  geom_sf(aes(fill = Dates),
-    color = "black"
-  ) +
-  scale_fill_distiller(
-    name = "Number of Radiocarbon Dates",
-    palette = "Greys",
-    direction = 1,
-    limits = c(0, 200),
-    minor_breaks = seq(25, 175, 50),
-    breaks = seq(0, 200, 50),
-    expand = expansion(mult = 0, add = 0),
-    guide = guide_colorbar(title.position = "top")
-  ) +
-  geom_sf(
-    data = uusw_counties,
-    fill = NA,
-    color = "black"
-  ) +
-  theme_map(16)
-
-county_counts %>%
-  dplyr::filter(Type == "Tree-ring") %>%
-  ggplot() +
-  geom_sf(
-    data = four_corners_states,
-    fill = NA
-  ) +
-  geom_sf(aes(fill = Dates),
-    color = "black"
-  ) +
-  scale_fill_distiller(
-    name = "Number of Tree-ring Dates",
-    palette = "Greys",
-    direction = 1,
-    limits = c(0, 5000),
-    minor_breaks = seq(500, 4500, 1000),
-    breaks = seq(0, 5000, 1000),
-    expand = expansion(mult = 0, add = 0),
-    guide = guide_colorbar(title.position = "top")
-  ) +
-  geom_sf(
-    data = uusw_counties,
-    fill = NA,
-    color = "black"
-  ) +
-  theme_map(16)
-```
-
-```{r Figure3, fig.height=9, fig.width=6.5, fig.cap='The number of sites with radiocarbon and tree-ring dates by county in this study.', fig.show = 'hold', out.width = '45%'}
-county_counts <-
-  all_data %>%
-  dplyr::bind_rows() %>%
-  dplyr::select(Type, SiteID, County) %>%
-  dplyr::distinct() %>%
-  dplyr::group_by(Type, County) %>%
-  dplyr::summarise(Dates = dplyr::n()) %>%
-  dplyr::left_join(uusw_counties,
-    by = "County"
-  ) %>%
-  sf::st_as_sf()
-
-theme_map <- function(base_size = 6.5,
-                      base_family = "") {
-  theme_bw(
-    base_size = base_size,
-    base_family = base_family
-  ) %+replace%
-    theme(
-      axis.line = element_blank(),
-      axis.text = element_blank(),
-      axis.ticks = element_blank(),
-      axis.title = element_blank(),
-      panel.background = element_blank(),
-      panel.border = element_blank(),
-      panel.grid = element_blank(),
-      panel.grid.major = element_line(colour = "transparent"),
-      panel.grid.minor = element_line(colour = "transparent"),
-      legend.key = element_blank(),
-      legend.position = "bottom",
-      legend.background = element_blank(),
-      legend.key.width = unit(0.15, "npc"),
-      legend.text = element_text(size = rel(1)),
-      plot.background = element_blank(),
-      plot.margin = margin(t = 0, r = 0, b = 0, l = 0, unit = "npc")
-    )
-}
-
-county_counts %>%
-  dplyr::filter(Type == "Radiocarbon") %>%
-  ggplot() +
-  geom_sf(
-    data = four_corners_states,
-    fill = NA
-  ) +
-  geom_sf(aes(fill = Dates),
-    color = "black"
-  ) +
-  scale_fill_distiller(
-    name = "Number of Sites with Radiocarbon Dates",
-    palette = "Greys",
-    direction = 1,
-    limits = c(0, 80),
-    minor_breaks = seq(10, 70, 20),
-    breaks = seq(0, 80, 20),
-    expand = expansion(mult = 0, add = 0),
-    guide = guide_colorbar(title.position = "top")
-  ) +
-  geom_sf(
-    data = uusw_counties,
-    fill = NA,
-    color = "black"
-  ) +
-  theme_map(16)
-
-county_counts %>%
-  dplyr::filter(Type == "Tree-ring") %>%
-  ggplot() +
-  geom_sf(
-    data = four_corners_states,
-    fill = NA
-  ) +
-  geom_sf(aes(fill = Dates),
-    color = "black"
-  ) +
-  scale_fill_distiller(
-    name = "Number of Sites with Tree-ring Dates",
-    palette = "Greys",
-    direction = 1,
-    limits = c(0, 250),
-    minor_breaks = seq(25, 225, 50),
-    breaks = seq(0, 250, 50),
-    expand = expansion(mult = 0, add = 0),
-    guide = guide_colorbar(title.position = "top")
-  ) +
-  geom_sf(
-    data = uusw_counties,
-    fill = NA,
-    color = "black"
-  ) +
-  theme_map(16)
-```
-
-## Preparing the SPDs {-}
-To prepare the radiocarbon dataset, we used functions provided by the rcarbon package for R [@Bevan2019].
-We calibrated the radiocarbon dates using the methods of @Bronk2008, using the ‘intcal13’ calibration curve [@Reimer2013].
-To control for the inconsistent number of dates among  the archaeological sites represented in the database (a form of investigation bias), we aggregated radiocarbon dates known to be from the same phase of an archaeological site using the rcarbon::binPrep function using the single linkage agglomeration method with an h-value of 100 (see [Supplementary Material] for a discussion of various h-values).
-This creates clusters of dates from the same site whose mean values (uncalibrated) are within 100 years of each other, and splits site occupations into distinct phases separated by 100 years or more.
-We constructed summed probability distributions (SPDs) from the calibrated dates using the `rcarbon::spd` function.
-We averaged probability distributions of binned dates, and then constructed SPDs from the resulting site-phase probability distributions.^[Using this method, a site-phase with a single date and a site-phase with 10 dates would each contribute equally to the resulting SPD.
-This method of developing the SPD is therefore a model of the relative abundance of site occupation events through time rather than of dates through time.
-This method does not account for investigation biases deriving from the definition of “site” or the decision to use a particular dating technique at a site in the first place.]
-
-```{r radiocarbon-spds}
-timerange <- c(10000, 0)
-
-radiocarbon_spd <-
-  all_data$Radiocarbon %>%
-  split(., .$`Study Area`) %>%
-  tibble::tibble(
-    `Study Area` = names(.),
-    Dates = .
-  ) %>%
-  dplyr::bind_rows(tibble::tibble(
-    `Study Area` = "UUSW",
-    Dates = list(all_data$Radiocarbon)
-  )) %>%
-  dplyr::rowwise() %>%
-  dplyr::mutate(
-    Calibrated = list(Dates %$%
-      rcarbon::calibrate(
-        x = Date_BP,
-        errors = SD,
-        timeRange = timerange,
-        calMatrix = TRUE,
-        verbose = FALSE
-      )),
-    Bins = list(rcarbon::binPrep(
-      sites = Dates$SiteID,
-      ages = Dates$Date_BP,
-      h = 100,
-      method = "single"
-    )),
-    SPD = list(rcarbon::spd(
-      x = Calibrated,
-      bins = Bins,
-      timeRange = timerange,
-      spdnormalised = TRUE,
-      verbose = FALSE
-    ))
-  ) %>%
-  dplyr::ungroup() %>%
-  dplyr::mutate(Density = SPD %>%
-    purrr::map(function(x) {
-      x$grid %>%
-        tibble::as_tibble() %$%
-        tibble::tibble(
-          Type = "Radiocarbon Density",
-          Date_BP = calBP,
-          Density = PrDens
-        )
-    })) %T>%
-  readr::write_rds(
-    here::here("analysis/data/derived_data/radiocarbon_spd.rds"),
-    compress = "gz"
-  )
-```
-
-We treated the tree-ring dates in as similar a way possible as the radiocarbon dates.
-We modeled faux-calibrated tree-ring date probabilities as discrete degenerate distributions ($\Pr(X=k_0)=1$) with $k_0$ equal to the tree-ring date — in other words, a probability of $1$ at the date value, and $0$ elsewhere.
-This is akin to modeling the tree ring date as a normal distribution with a mean of the date value, and a standard deviation of zero.
-Thinking of tree ring dates this way allows us to use the same binning and SPD algorithms as for radiocarbon dates.^[One might consider alternative probability distributions for non-cutting or near-cutting dates.
-The probability mass function in such a case would be zero for years prior to the date of the outer ring of the sample, and non-zero for subsequent years, presumably decaying quickly.
-For instance, we could imagine a study that derives empirical probability distributions for non-cutting dates based on cutting-dates from the same contexts.] Here, we binned the tree-ring dates into site-phases using the rcarbon::binPrep function and constructing SPDs using `rcarbon::spd`, both with the same parameter selections as used for the radiocarbon samples.
-As above, we averaged probability distributions of binned dates, and then constructed SPDs from the site-phase distributions.
-
-```{r treering-spds}
-calibrate_tree_ring_dates <- function(x, timeRange = c(50000, 0)) {
-  out <- list(
-    metadata = tibble::tibble(
-      DateID = 1:length(x),
-      CRA = x,
-      Error = 0,
-      Details = NA,
-      CalCurve = "intcal13",
-      ResOffsets = 0,
-      ResErrors = 0,
-      StartBP = timeRange[1],
-      EndBP = timeRange[2],
-      Normalised = TRUE,
-      F14C = FALSE,
-      CalEPS = 1e-05
-    ),
-    grids = NA,
-    calmatrix =
-      tibble::tibble(
-        index = 1:length(x),
-        year_bp = x,
-        value = TRUE
-      ) %>%
-        tidyr::pivot_wider(
-          names_from = index,
-          values_from = value
-        ) %>%
-        dplyr::bind_rows(tibble::tibble(year_bp = timeRange[1]:timeRange[2])) %>%
-        dplyr::filter(!duplicated(year_bp)) %>%
-        dplyr::arrange(-year_bp) %>%
-        dplyr::filter(year_bp >= 0) %>%
-        tibble::column_to_rownames("year_bp") %>%
-        as.matrix() %>%
-        tidyr::replace_na(0) %>%
-        magrittr::set_colnames(NULL)
-  )
-
-  attr(out, "class") <- c("CalDates", "list")
-
-  return(out)
-}
-
-dendro_spd <-
-  all_data$`Tree-ring` %>%
-  split(., .$`Study Area`) %>%
-  tibble::tibble(
-    `Study Area` = names(.),
-    Dates = .
-  ) %>%
-  dplyr::bind_rows(tibble::tibble(
-    `Study Area` = "UUSW",
-    Dates = list(all_data$`Tree-ring`)
-  )) %>%
-  dplyr::rowwise() %>%
-  dplyr::mutate(
-    Calibrated = list(Dates %$%
-      calibrate_tree_ring_dates(
-        x = Date_BP,
-        timeRange = timerange
-      )),
-    Bins = list(rcarbon::binPrep(
-      sites = Dates$SiteID,
-      ages = Dates$Date_BP,
-      h = 100,
-      method = "single"
-    )),
-    SPD = list(rcarbon::spd(
-      x = Calibrated,
-      bins = Bins,
-      timeRange = timerange,
-      spdnormalised = TRUE,
-      verbose = FALSE
-    ))
-  ) %>%
-  dplyr::ungroup() %>%
-  dplyr::mutate(Density = SPD %>%
-    purrr::map(function(x) {
-      x$grid %>%
-        tibble::as_tibble() %$%
-        tibble::tibble(
-          Type = "Tree-ring Density",
-          Date_BP = calBP,
-          Density = PrDens
-        )
-    })) %T>%
-  readr::write_rds(
-    here::here("analysis/data/derived_data/dendro_spd.rds"),
-    compress = "gz"
-  )
-```
-
-# Results {-}
-Figure \@ref(fig:Figure4) presents radiocarbon and tree-ring SPDs for the UUSW, CMV, and NRG regions, as well as VEP population reconstructions for the CMV and NRG regions.
-Across the UUSW, we see striking agreement in directionality and timing between the radiocarbon and tree-ring series (Figure \@ref(fig:Figure4)A); both series trend upwards until the late thirteenth century, and then decline rapidly.
-The tree-ring series — the light orange line is the raw SPD, and the dark line is smoothed — shows more variation on annual to decadal timescales, and yet matches the broad trends of the radiocarbon SPD.
-Roughly 150-year periodic fluctuations in the tree-ring SPD, described as periods of “exploration” and “exploitation” by @Bocinsky2016, are not present in the radiocarbon SPD, the result of a much smaller sample size of site-phases and lower temporal resolution of radiocarbon dating relative to tree-ring dating.
-
-(ref:fig4) Radiocarbon and tree-ring density distributions in the UUSW (A), CMV (B), and NRG (C) regions, and VEP population estimates for the CMV and NRG.
-The dark orange tree-ring density curve smooths the raw tree-ring density (light orange) using a 21-year center-aligned Gaussian kernel with a 5-year standard deviation, following @Bocinsky2016.
-In Panel A, the radiocarbon density (y-axis) is scaled by a factor of 3 to facilitate comparison with the tree-ring density over the target time period.
-The radiocarbon and tree-ring densities (y-axis) in panels B and C are scaled such that the area under their densities is equal to the area under the VEP population estimates (i.e., the cumulative population).
-
-```{r Figure4, fig.height=6.5, fig.width=6.5, fig.cap='(ref:fig4)', fig.show = 'hold', out.width = '75%'}
-
-# This is a 21-year wide Gaussian distribution with a
-# mean of 0 and standard deviation of 5 years.
-# This is the smoother used in Bocinsky et al. 2016.
-smoother <- dnorm(seq(-10, 10, 1), sd = 5)
-
-all_density <-
-  dendro_spd %>%
-  dplyr::select(`Study Area`, Density) %>%
-  tidyr::unnest(cols = c(Density)) %>%
-  dplyr::bind_rows(radiocarbon_spd %>%
-    dplyr::select(`Study Area`, Density) %>%
-    tidyr::unnest(cols = c(Density))) %>%
-  dplyr::mutate(`Year AD` = 1950 - Date_BP) %>%
-  dplyr::select(
-    `Study Area`,
-    Type,
-    `Year AD`,
-    Density
-  )
-
-vep_demography <-
-  robinson2020::vep_demography %>%
-  dplyr::mutate(End = ifelse(`Study Area` == "CMV" & Period == 20, 1300, End)) %>%
-  dplyr::rowwise() %>%
-  dplyr::mutate(`Year AD` = list(Start:(End - 1))) %>%
-  dplyr::select(
-    `Study Area`,
-    `Year AD`,
-    `Period`,
-    Population
-  ) %>%
-  tidyr::unnest(`Year AD`)
-
-dplyr::left_join(all_density,
-  vep_demography,
-  by = c(
-    "Study Area",
-    "Year AD"
-  )
-) %>%
-  dplyr::group_by(
-    `Study Area`,
-    Type
-  ) %>%
-  dplyr::mutate(`Scaled Density` = Density * (sum(Population, na.rm = T) / sum(Density * !is.na(Population), na.rm = TRUE))) %>%
-  dplyr::group_by(
-    `Study Area`,
-    Type,
-    Period
-  ) %>%
-  dplyr::group_by(
-    `Study Area`,
-    Type
-  ) %>%
-  dplyr::mutate(
-    `Raw Density` = ifelse(`Study Area` == "UUSW", Density, `Scaled Density`),
-    `Smoothed Density` = ifelse(Type == "Tree-ring Density", stats::filter(`Raw Density`, filter = smoother), `Raw Density`),
-    `Raw Density` = ifelse(`Study Area` == "UUSW", `Raw Density`, NA),
-    `Raw Density` = ifelse(Type == "Tree-ring Density", `Raw Density`, NA)
-  ) %>%
-  dplyr::select(
-    `Study Area`,
-    Type,
-    `Year AD`,
-    Population,
-    `Raw Density`,
-    `Smoothed Density`,
-    # `Binned Density`
-  ) %>%
-  dplyr::ungroup() %>%
-  dplyr::mutate(
-    Type = factor(Type,
-      levels = c(
-        "Tree-ring Density",
-        "Radiocarbon Density",
-        "VEP Population"
-      ),
-      ordered = TRUE
-    ),
-    `Study Area` = factor(`Study Area`,
-      levels = c(
-        "UUSW",
-        "CMV",
-        "NRG"
-      ),
-      labels = c(
-        "Upland US Southwest Density",
-        "Central Mesa Verde Population",
-        "Northern Rio Grande Population"
-      ),
-      ordered = TRUE
-    )
-  ) %>%
-  dplyr::mutate(`Smoothed Density` = ifelse(`Study Area` == "Upland US Southwest Density" & Type == "Radiocarbon Density",
-    (`Smoothed Density` * 3),
-    `Smoothed Density`
-  )) %>%
-  {
-    ggplot(
-      .,
-      aes(x = `Year AD`)
-    ) +
-      geom_line(aes(
-        y = `Raw Density`,
-        color = Type
-      ),
-      size = 0.5,
-      alpha = 0.5
-      ) +
-      geom_line(aes(
-        y = `Population`
-      ),
-      size = 1.25
-      ) +
-      geom_line(aes(
-        y = `Smoothed Density`,
-        color = Type
-      ),
-      size = 0.75
-      ) +
-      scale_color_manual(
-        values = c(
-          "Tree-ring Density" = "#d95f02",
-          "Radiocarbon Density" = "#1b9e77",
-          "VEP Population" = "black"
-        ),
-        limits = c(
-          "Tree-ring Density",
-          "Radiocarbon Density",
-          "VEP Population"
-        )
-      ) +
-      xlab("Year AD") +
-      scale_x_continuous(
-        limits = c(200, 1800),
-        breaks = seq(200, 1800, 200),
-        expand = expansion(0, 0)
-      ) +
-      theme_minimal(18) +
-      theme(
-        legend.title = element_blank(),
-        legend.justification = c(0, 1),
-        legend.position = "bottom",
-        legend.text = element_text(size = 12),
-        axis.title.x = element_text(size = 12),
-        axis.text.x = element_text(size = 10),
-        axis.title.y = element_blank(),
-        axis.text.y = element_text(size = 10),
-        # panel.grid.major.y = element_blank(),
-        # panel.grid.minor.y = element_blank(),
-        plot.margin = unit(c(0.25, 0.5, 0.25, 0.25), "in")
-      ) +
-      facet_wrap(
-        # rows =
-        vars(`Study Area`),
-        ncol = 1,
-        scales = "free_y"
-      )
-  } %>%
-  egg::tag_facet(
-    open = "",
-    close = "",
-    tag_pool = paste0(LETTERS, ". ", c(
-      "Upland US Southwest Density",
-      "Central Mesa Verde Population",
-      "Northern Rio Grande Population"
-    )),
-    hjust = 0,
-    vjust = 1.5,
-    size = 4
-  )
-```
-
-Comparisons between the CMV and NRG regional SPDs and the VEP population estimates are also revealing.
-In Figures \@ref(fig:Figure4)B and \@ref(fig:Figure4)C, the regional SPDs have been scaled such that the areas under their densities (the orange and green lines) are equal to the area under the VEP population estimates (the black lines).
-For both regions, the radiocarbon and tree-ring SPDs capture the broad directionality of the VEP population estimates.
-In the CMV region (Figure \@ref(fig:Figure4)B), the radiocarbon SPD has substantial probability mass prior to the VEP reconstruction, which is likely due to sampling bias in the radiocarbon record that is skewed in this region towards older (pre-Pueblo) sites.
-The radiocarbon SPD also does not capture the Pueblo I (AD 750--900) population peak (which may be too short-lived to be captured by such a small sample of radiocarbon dates), and seems to begin declining by the early thirteenth century AD.
-The tree-ring SPDs better capture the variability in population, though include a decline in the mid AD 1100s that is not in the VEP population estimates.
-In the NRG region (Figure \@ref(fig:Figure4)C), both SPDs rise until about AD 1250, then decline thereafter.
-The high peak in the tree-ring record at AD 1250, far in excess of the VEP population estimate, is particularly interesting — it likely signals the net impact of the immigration of many thousands of people to the NRG region, and the resulting (and simultaneous) burst in construction.
-Such a sudden and synchronous demographic and architectural event is hard to capture in the relatively lower resolution and sparsely sampled radiocarbon record, which tends to be smeared across sub-centennial scales courtesy of the calibration process [@Baillie1991].
-
-These results yield key insights on the relative sensitivity of the radiocarbon record to capture long-term, coarse-grained demographic change across the Southwest.
-As stated in the introduction, our aim is to understand how well the radiocarbon record captures population declines across the Southwest, as we need to develop a measure of confidence in radiocarbon records from regions across and beyond the Southwest that lack other demographic proxies.
-Both the radiocarbon and tree-ring SPDs show declines post-AD 1300 across all regions.
-In both the CMV and NRG the decline in the radiocarbon SPD begins earlier than in the VEP population series.
-This may indicate that radiocarbon time series make the decline seem to begin earlier than it actually does, which could be caused by a ‘cliff’ in the radiocarbon calibration curve from AD 1200--1300 that is surrounded by two plateaus at AD 1100--1200 and AD 1400 [@Reimer2013].
-
-Comparisons of the radiocarbon and dendrochronological record with the architecture/ceramic record show similarities and differences that reinforce the need for multiple-proxy studies of palaeodemography when moving between broader and finer spatial scales.
-While the multi-proxy population reconstruction for the CMV region was similar to the radiocarbon and dendrochronological records, it showed a different trend for the NRG region (lower panel, Figure \@ref(fig:Figure4)).
-There, the population reconstruction peaks from AD 1300--1400, whereas the radiocarbon record declines, at first quickly and then slowly.
-These different trends between the VEP population reconstructions and dendrochronological and radiocarbon proxies reflect some investigation bias in both the radiocarbon and dendrochronological records at smaller, intra-regional scales, which calls for caution in the use of radiocarbon records to assess possible paleo-migration scenarios at finer spatial scales.
-This result reinforces other work that has called for the use of multiple proxies as cross-checks on radiocarbon time series [@Williams2012, @Crombe2014].
-
-The central aim in comparing radiocarbon to dendrochronological records here was to assess the potential impacts of investigation bias in radiocarbon records, caused by researchers not taking radiocarbon samples in favour of other types of chronometric techniques (e.g., diagnostic artefact counts, tree-ring dates).
-Specifically, we are interested in the potential impact of this kind of bias in the post AD 1300 decline of radiocarbon SPDs across and beyond the Southwest.
-Similarities in the decline of both radiocarbon and dendrochronological records across the UUSW suggest that, at this particular regional scale, radiocarbon records reflect long-term, coarse-grained trends in paleodemography.
-Indeed, not only do we see corresponding declines in the radiocarbon and dendrochronological records at this large scale, but these patterns are also consistent with previous work on demographic and social change in the American Southwest: the Coalescent Communities Database, a Southwest-wide database of room counts and occupation spans of habitation sites with 12 rooms or more, supports a “population decline that began between AD 1300 and 1350 and continued to European contact” [@Phillips2018; following @Hill2004: Figures 2,4].
-
-# Conclusion {-}
-Understanding the geographic scale of post-AD 1300 human population declines in the American Southwest requires a reliance on radiocarbon-based reconstructions of palaeodemography, as radiocarbon is often the only proxy record available in many regions for developing continuous time series.
-However, radiocarbon records may include investigation bias caused by researchers not selecting radiocarbon samples in favour of Late Prehistoric diagnostic artefacts or other dating methods.
-In this paper we develop a new method for systematically comparing radiocarbon time series against multiple palaeodemographic proxies.
-We selected three regions—the Upland US Southwest, Central Mesa Verde, and the Northern Rio Grande—where we expected the most investigation bias to occur, because in these regions multiple higher resolution dating techniques are available.
-Results indicated that all three palaeodemographic proxies reveal post-AD 1300 population declines.
-While there is some variability at finer spatial and temporal scales that warrant the collection of other data that can be used to cross-check and validate the resolution of radiocarbon records, at coarse-grained, large spatial and temporal scales radiocarbon records provide a robust relative measure of long-term palaeodemographic trends.
-These results lend confidence to single proxy radiocarbon-based reconstructions of palaeodemography outside the Southwest that suggest post-AD 1300 population declines in many parts of North America.
-
-Investigation bias is an inevitable issue that all researchers using radiocarbon records have to address.
-The results of this paper advance our confidence in radiocarbon records as a coarse-grained measure of the directionality of human paleodemographic trends.
-The methods developed can be used for other regions of the world that have dendrochronological records, such as Europe [@Kuniholm2001; @Haneca2009; @Hillam1990], where comparisons with radiocarbon records could help to advance knowledge of Neolithic ‘booms and busts’ of populations [@Shennan2013; @Whitehouse2014] and to overcome limitations of the Bronze Age radiocarbon calibration plateau [@Armit2013].
-
-#### Acknowledgements {-}
-We thank the guest editorial board for inviting us to contribute to this volume.
-We thank two anonymous reviewers for their helpful suggestions for improvement.
-We thank our colleagues in the PAGES People3000 working group, and PAGES for funding this working group.
-We thank the National Foundation for supporting this research.
-Lastly, we thank the generations of different stakeholders working across the Southwest for their collection, storage and publication of the data presented here.
-
-#### Data Accessibility {-}
-A compendium of data and code used in this analysis is available on Github at https://github.com/people3k/robinson2020 and is archived with Zenodo at https://doi.org/10.5281/zenodo.3647599.
-Radiocarbon data used in this analysis are archived with the Digital Archaeological Record (tDAR) and are available at https://doi.org/10.6067/XCV8455460.
-Tree-ring data used here are archived with tDAR at https://doi.org/10.6067/XCV82J6D7B.
-Archaeological site locations are protected by US federal and state statutes.
-In order to protect these sensitive data, we pre-processed both the radiocarbon and tree-ring data; data included in the Github repository do not include site locations, and researchers will need to gain permission from the State Historic Preservation Office (SHPO) of each state in order to access the site location data.
-
-#### Ethics {-}
-No ethical issues are associated with this paper.
-
-#### Authors’ contributions {-}
-E.R.
-and R.K.B designed the study, collected and analysed the data, and wrote the first draft.
-J.F.
-and D.B collected and organized data.
-R.L.K collected data.
-All authors revised/improved the manuscript.
-
-#### Competing interests {-}
-We have no competing interests.
-
-#### Funding {-}
-Funding for this research was provided by NSF grants BCS-1418858, BCS-1624061, BCS-1822033 and PAGES (Past Global Changes) funding to the Working Group [Paleoclimate and the Peopling of the Earth (PEOPLE 3000)](http://www.pastglobalchanges.org/ini/wg/arctic2k/170-initiatives/working-group/people-3000/1732-people-3000).
-
-<!-- The following line inserts a page break  -->
-\newpage
-# References {-}
-<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->
-<div id="refs"></div>
-
-<!-- The following line inserts a page break  -->
-\newpage
-\beginsupplement
-# Supplementary Material: Cut-width sensitivity for temporal binning {-}
-Uneven sampling across archaeological contexts is a persistant problem when constructing summed probability distributions from radiocarbon dates, and perhaps even a larger issue for the more ubiquitous tree-ring dates in regions like the Upland US Southwest.
-In SPDs of radiocarbon dates, it is conventional to combine dates from single sites prior to creating the SPDs.
-Dates from separate components or phases of multi-component sites should be counted separately, however.
-
-To control for inconsistent sample size across the archaeological sites represented in the database, we aggregated dates known to be from the same phase of same archaeological site using the `rcarbon::binPrep` function using the “single linkage” agglomeration method with a cut width (or $h$-value) of 100 [@rcarbon].
-This clusters dates from the same site that are within 100 years of each other, and splits site occupations into distinct phases separated by 100 years or more.
-Binned calibrated dates are then summed, and the resulting probability mass function is divided by the number of dates in the site-phase.
-
-Because the wider archaeological SPD research community still lacks a standard protocol for determining $h$, we perform a sensitivity analysis to demonstrate that the resulting SPDs for radiocarbon and tree-ring dates stabilize at a cut width of 100 or so.
-We calculated SPDs at 25-year intervals from 0 (only duplicate dates are binned) to 150 years (effectively binning by site).
-
-As can be seen in Figure \@ref(fig:FigureS1), binning in general has little impact on the resulting SPDs, but cut widths above $h=100$ years are virtually indistinguishable.
-The shape of the radiocarbon SPD (panel A) stabilizes above $h=0$ and its values stabilize by $h=100$.
-The tree-ring SPD arguably stabilizes in both shape and value above $h=75$ or so, but in the analysis we use $h=100$ to keep our methods consistent between the radiocarbon and tree-ring SPDs.
-
-```{r FigureS1, fig.height=6.5, fig.width=6.5, fig.cap='Radiocarbon (A) and tree-ring (B) density distributions in the UUSW, given varying cut widths for binning dates within sites.
-The density distributions for the UUSW are normalized.
-Years are given in AD.', fig.show = 'hold', out.width = '75%'}
-cut_widths <- seq(0, 150, 25)
-
-radiocarbon_spd <-
-  here::here("analysis/data/derived_data/radiocarbon_spd.rds") %>%
-  readr::read_rds()
-
-dendro_spd <-
-  here::here("analysis/data/derived_data/dendro_spd.rds") %>%
-  readr::read_rds()
-
-# This is a 21-year wide Gaussian distribution with a
-# mean of 0 and standard deviation of 5 years.
-# This is the smoother used in Bocinsky et al. 2016.
-smoother <- dnorm(seq(-10, 10, 1), sd = 5)
-
-timerange <- c(10000, 0)
-
-spds <-
-  radiocarbon_spd %>%
-  dplyr::filter(`Study Area` == "UUSW") %>%
-  dplyr::bind_rows(dendro_spd %>%
-    dplyr::filter(`Study Area` == "UUSW")) %>%
-  dplyr::mutate(Type = c(
-    "Radiocarbon",
-    "Tree-ring"
-  )) %>%
-  dplyr::select(Type, Dates, Calibrated) %>%
-  dplyr::rowwise() %>%
-  dplyr::mutate(
-    Densities =
-      list(
-        cut_widths %>%
-          magrittr::set_names(., .) %>%
-          purrr::map_dfr(~ rcarbon::spd(
-            x = Calibrated,
-            timeRange = timerange,
-            bins =
-              rcarbon::binPrep(
-                sites = Dates$SiteID,
-                ages = Dates$Date_BP,
-                h = .x,
-                method = "single"
-              ),
-            spdnormalised = TRUE,
-            verbose = FALSE
-          ) %$%
-            grid %>%
-            tibble::as_tibble() %$%
-            tibble::tibble(
-              Date_AD = 1950 - calBP,
-              Density = PrDens
-            ),
-          .id = "h"
-          ) %>%
-          dplyr::mutate(
-            `Cut Width (h)` =
-              factor(h,
-                levels = cut_widths,
-                ordered = TRUE
-              )
-          ) %>%
-          dplyr::filter(Date_AD > 0) %>%
-          dplyr::select(`Cut Width (h)`,
-            `Year AD` = Date_AD,
-            Density
-          )
-      )
-  ) %>%
-  dplyr::select(Type, Densities) %>%
-  tidyr::unnest(cols = c(Densities))
-
-spds %>%
-  dplyr::group_by(Type, `Cut Width (h)`) %>%
-  dplyr::mutate(
-    Density =
-      ifelse(Type == "Tree-ring",
-        stats::filter(Density,
-          filter = smoother
-        ),
-        Density
-      )
-  ) %>%
-  dplyr::ungroup() %>%
-  {
-    ggplot2::ggplot(
-      .,
-      ggplot2::aes(
-        x = `Year AD`,
-        y = Density,
-        color = `Cut Width (h)`
-      )
-    ) +
-      ggplot2::geom_line() +
-      ggplot2::facet_wrap("Type",
-        nrow = 2,
-        scales = "free_y"
-      ) +
-      xlab("Year AD") +
-      scale_x_continuous(
-        limits = c(200, 1800),
-        breaks = seq(200, 1800, 200),
-        expand = expansion(0, 0)
-      ) +
-      scale_y_continuous(
-        limits = c(0, NA),
-        expand = expansion(0, 0)
-      ) +
-      theme_minimal(18) +
-      theme(
-        legend.title = element_text(size = 12),
-        legend.position = "bottom",
-        legend.text = element_text(size = 12),
-        axis.title.x = element_text(size = 12),
-        axis.text.x = element_text(size = 10),
-        axis.title.y = element_blank(),
-        axis.text.y = element_text(size = 10),
-        plot.margin = unit(c(0.25, 0.5, 0.25, 0.25), "in")
-      ) +
-      guides(color = guide_legend(nrow = 1))
-  } %>%
-  egg::tag_facet(
-    open = "",
-    close = "",
-    tag_pool = paste0(LETTERS, ". ", c(
-      "Radiocarbon Dates",
-      "Tree-ring Dates"
-    )),
-    hjust = 0,
-    vjust = 1.5,
-    size = 4
-  )
-```
-
-\newpage
-# Colophon {-}
-This report is part of a research compendium developed using the `rrtools` package for reproducible research [@Marwick2017; @rrtools].
-This report was generated on `r Sys.time()` using the following computational environment and dependencies:
-
-```{r colophon-r, cache = FALSE}
-# which R packages and versions?
-sess <- sessionInfo()
-sess$basePkgs %<>% sort()
-sess$loadedOnly %<>% magrittr::extract(sort(names(sess$loadedOnly)))
-if (requireNamespace("pander", quietly = TRUE)) pander::pander(sess)
-```

### 2020-04-20 analysis/robinson2020.Rmd
still trying to get travis working

@@
 # csl: "philosophical-transactions-of-the-royal-society-b.csl"
-# csl: "journal-of-archaeological-science.csl"
 csl: "apa.csl"
@@
 These different trends between the VEP population reconstructions and dendrochronological and radiocarbon proxies reflect some investigation bias in both the radiocarbon and dendrochronological records at smaller, intra-regional scales, which calls for caution in the use of radiocarbon records to assess possible paleo-migration scenarios at finer spatial scales.
-This result reinforces other work that has called for the use of multiple proxies as cross-checks on radiocarbon time series [@Williams2012, @Crombe2014].
+This result reinforces other work that has called for the use of multiple proxies as cross-checks on radiocarbon time series [@Williams2012; @Crombe2014].
 

### 2020-04-20 analysis/robinson2020.Rmd
still trying to get travis working

@@
       template: templates/template.tex
-      latex_engine: lualatex
+      latex_engine: xelatex
       fig_caption: yes

### 2020-04-20 analysis/robinson2020.Rmd
still trying to get travis working

@@
 library(ggplot2)
+
+# Fix for deprecated ggplot2::expand_scale function
+if (utils::packageVersion("ggplot2") >= "3.3.0") {
+  expand_scale <- expansion
+}
 ```
@@
     breaks = seq(0, 200, 50),
-    expand = expansion(mult = 0, add = 0),
+    expand = expand_scale(mult = 0, add = 0),
     guide = guide_colorbar(title.position = "top")
@@
     breaks = seq(0, 5000, 1000),
-    expand = expansion(mult = 0, add = 0),
+    expand = expand_scale(mult = 0, add = 0),
     guide = guide_colorbar(title.position = "top")
@@
     breaks = seq(0, 80, 20),
-    expand = expansion(mult = 0, add = 0),
+    expand = expand_scale(mult = 0, add = 0),
     guide = guide_colorbar(title.position = "top")
@@
     breaks = seq(0, 250, 50),
-    expand = expansion(mult = 0, add = 0),
+    expand = expand_scale(mult = 0, add = 0),
     guide = guide_colorbar(title.position = "top")
@@
         breaks = seq(200, 1800, 200),
-        expand = expansion(0, 0)
+        expand = expand_scale(0, 0)
       ) +
@@
         limits = c(0, NA),
-        expand = expansion(0, 0)
+        expand = expand_scale(0, 0)
       ) +

### 2020-04-21 analysis/robinson2020.Rmd
still trying to get travis working

@@
       - southwestern United States
-doi: [DOI HERE]
+doi: [10.31235/osf.io/hbecz]
 journal: Preprint
@@
       template: templates/template.tex
-      latex_engine: xelatex
+      latex_engine: lualatex
       fig_caption: yes
@@
   chunk_output_type: console
-header-includes:
-  \newcommand{\beginsupplement}{\setcounter{table}{0}  \renewcommand{\thetable}{S\arabic{table}} \setcounter{figure}{0}       \renewcommand{\thefigure}{S\arabic{figure}}}
+header-includes: >
+  \newcommand{\beginsupplement}{\setcounter{table}{0}  
+  \renewcommand{\thetable}{S\arabic{table}} \setcounter{figure}{0}       
+  \renewcommand{\thefigure}{S\arabic{figure}}} 
+#  \setmainfont[Ligatures=TeX]{Nimbus Roman No9 L}
 ---
@@
 We calibrated the radiocarbon dates using the methods of @Bronk2008, using the ‘intcal13’ calibration curve [@Reimer2013].
-To control for the inconsistent number of dates among  the archaeological sites represented in the database (a form of investigation bias), we aggregated radiocarbon dates known to be from the same phase of an archaeological site using the rcarbon::binPrep function using the single linkage agglomeration method with an h-value of 100 (see [Supplementary Material] for a discussion of various h-values).
+To control for the inconsistent number of dates among  the archaeological sites represented in the database (a form of investigation bias), we aggregated radiocarbon dates known to be from the same phase of an archaeological site using the rcarbon::binPrep function using the single linkage agglomeration method with an h-value of 100 (see [Supplementary Material](#supp) for a discussion of various h-values).
 This creates clusters of dates from the same site whose mean values (uncalibrated) are within 100 years of each other, and splits site occupations into distinct phases separated by 100 years or more.
@@
 \beginsupplement
-# Supplementary Material: Cut-width sensitivity for temporal binning {-}
+# Supplementary Material: Cut-width sensitivity for temporal binning {- #supp}
 Uneven sampling across archaeological contexts is a persistant problem when constructing summed probability distributions from radiocarbon dates, and perhaps even a larger issue for the more ubiquitous tree-ring dates in regions like the Upland US Southwest.

### 2020-04-26 analysis/robinson2020.Rmd
getting travis working

@@
 link-citations: true
+params:
+  cache: TRUE
 output: 
@@
 knitr::opts_chunk$set(
-  cache = TRUE,
+  cache = params$cache,
   echo = FALSE,

### 2020-04-26 analysis/robinson2020.Rmd
getting travis working

@@
 if (requireNamespace("pander", quietly = TRUE)) pander::pander(sess)
+
+gc()
+gc()
 ```

### 2020-04-26 README.Rmd
getting travis working

@@
 Robert L. Kelly[![](https://orcid.org/sites/default/files/images/orcid_16x16.png)](https://orcid.org/0000-0001-9737-0152), 
-(2020). _`r Title`_. *Philosophical Transactions B*. In Press.
+(In Press). `r Title`. *Philosophical Transactions of the Royal Society B*.
 
@@
 
-> Bocinsky, R. Kyle, Darcy Bird, and Erick Robinson, (`r format(Sys.Date(), "%Y")`). _Compendium of R code and data for `r Title`_. Accessed `r format(Sys.Date(), "%d %b %Y")`.
+> Bocinsky, R. Kyle, Darcy Bird, and Erick Robinson, (2020). _Compendium of R code and data for `r Title`_. Accessed \[TODAY'S DATE\].
 

### 2020-04-27 analysis/robinson2020.Rmd
getting travis working

--- 2020-04-21 analysis/robinson2020.Rmd
@@
 link-citations: true
+params:
+  clean: FALSE
 output: 
@@
   \renewcommand{\thefigure}{S\arabic{figure}}} 
-#  \setmainfont[Ligatures=TeX]{Nimbus Roman No9 L}
+  \setmainfont[Ligatures=TeX]{Nimbus Roman No9 L}
 ---
@@
   expand_scale <- expansion
+}
+
+if (params$clean) {
+  unlink(here::here("analysis/robinson2020.pdf"))
+  unlink(here::here("analysis/robinson2020_cache"), recursive = TRUE)
+  unlink(here::here("analysis/figures/."), recursive = TRUE)
+  unlink(here::here("analysis/data/derived_data/."), recursive = TRUE)
 }
@@
 sess <- sessionInfo()
+sess$locale <- Sys.getlocale("LC_COLLATE")
 sess$basePkgs %<>% sort()
@@
 ```
+
+```{r gc, echo = FALSE, message = FALSE, results='hide'}
+gc()
+gc()
+```

### 2020-04-27 analysis/robinson2020.Rmd
getting travis working

@@
   unlink(here::here("analysis/robinson2020_cache"), recursive = TRUE)
-  unlink(here::here("analysis/figures/."), recursive = TRUE)
-  unlink(here::here("analysis/data/derived_data/."), recursive = TRUE)
+  unlink(here::here("analysis/figures/."))
+  unlink(here::here("analysis/data/derived_data/."))
 }
@@
 )) {
+  download.file("https://core.tdar.org/filestore/download/455460/567756",
+    destfile = here::here("analysis/data/raw_data/robinson_et_al_2020_radiocarbon.csv")
+  )
+
   readr::read_csv(

### 2020-04-27 analysis/robinson2020.Rmd
getting travis working

@@
   unlink(here::here("analysis/figures/."))
-  unlink(here::here("analysis/data/derived_data/."))
+  # unlink(here::here("analysis/data/derived_data/."))
 }

### 2020-04-27 analysis/robinson2020.Rmd
getting travis working

--- 2020-04-27 analysis/robinson2020.Rmd
@@
   unlink(here::here("analysis/figures/."), recursive = TRUE)
-  unlink(here::here("analysis/data/derived_data/."), recursive = TRUE)
+  # unlink(here::here("analysis/data/derived_data/."), recursive = TRUE)
 }
@@
 ```{r treering-spds}
-calibrate_tree_ring_dates <- function(x, timeRange = c(50000, 0)) {
+calibrate_tree_ring_dates <-
+  function(x, timeRange = c(50000, 0)) {
     out <- list(

### 2020-04-27 analysis/robinson2020.Rmd
getting travis working

@@
 
-```{r treering-spds}
+```{r treering-spds, results = 'hide'}
+
+gc()
+gc()
+
 calibrate_tree_ring_dates <-

### 2020-04-27 analysis/robinson2020.Rmd
getting travis working
added nass back in

@@
   unlink(here::here("analysis/figures/."), recursive = TRUE)
-  # unlink(here::here("analysis/data/derived_data/."), recursive = TRUE)
 }
@@
 ```{r treering-spds, results = 'hide'}
-
 gc()
@@
 The tree-ring SPD arguably stabilizes in both shape and value above $h=75$ or so, but in the analysis we use $h=100$ to keep our methods consistent between the radiocarbon and tree-ring SPDs.
+
+```{r gc2, echo = FALSE, message = FALSE, results='hide'}
+gc()
+gc()
+```
 

### 2020-04-27 analysis/robinson2020.Rmd
getting travis working

@@
 
+```{r gc3, echo = FALSE, message = FALSE, results='hide'}
+# This chunk is here to deal with memory limits of Travis CI
+rm(list = ls())
+gc()
+gc()
+```
+
 Comparisons between the CMV and NRG regional SPDs and the VEP population estimates are also revealing.
@@
 The tree-ring SPD arguably stabilizes in both shape and value above $h=75$ or so, but in the analysis we use $h=100$ to keep our methods consistent between the radiocarbon and tree-ring SPDs.
-
-```{r gc2, echo = FALSE, message = FALSE, results='hide'}
-gc()
-gc()
-```
 
@@
 ```{r gc, echo = FALSE, message = FALSE, results='hide'}
+# This chunk is here to deal with memory limits of Travis CI
 gc()

### 2020-12-01 README.Rmd
publication version

--- 2020-04-17 README.Rmd
@@
 Robert L. Kelly[![](https://orcid.org/sites/default/files/images/orcid_16x16.png)](https://orcid.org/0000-0001-9737-0152), 
-(2020). _`r Title`_. *Philosophical Transactions B*. In Press.
+(2020). `r Title`. *Philosophical Transactions of the Royal Society B* 376:20190718. https://doi.org/10.1098/rstb.2019.0718
 
@@
 
-- [:file_folder: paper](/analysis/paper): R Markdown source document for the manuscript.
+- [:page_facing_up: robinson2020](/analysis/robinson2020): R Markdown source document for the manuscript.
 Includes code to reproduce the figures and tables generated by the analysis.
-It also has a rendered version, `paper.html`, suitable for reading (the code is replaced by figures and tables in this file).
+It also has a rendered version, `robinson2020.pdf`, suitable for reading (the code is replaced by figures and tables in this file).
 - [:file_folder: data](/analysis/data): Data used in the analysis.  

### 2020-12-02 README.Rmd
publication version

@@
 [![Travis build status](https://travis-ci.org/people3k/robinson2020.svg?branch=master)](https://travis-ci.org/people3k/robinson2020)
-[![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh/people3k/robinson2020/master?urlpath=rstudio)
+[![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh/people3k/robinson2020/test?urlpath=rstudio)
 
@@
 Jacob Freeman [![](https://orcid.org/sites/default/files/images/orcid_16x16.png)](https://orcid.org/0000-0001-7402-8450), 
-Robert L. Kelly[![](https://orcid.org/sites/default/files/images/orcid_16x16.png)](https://orcid.org/0000-0001-9737-0152), 
-(2020). `r Title`. *Philosophical Transactions of the Royal Society B* 376:20190718. https://doi.org/10.1098/rstb.2019.0718
+Robert L.
+Kelly[![](https://orcid.org/sites/default/files/images/orcid_16x16.png)](https://orcid.org/0000-0001-9737-0152), (2020).
+`r Title`.
+*Philosophical Transactions of the Royal Society B* 376:20190718.
+https://doi.org/10.1098/rstb.2019.0718
 
